### 2014-10-08 abxD01_analysis.Rmd
Fig1&2 added

---
title: "AbxD01 Analysis"
author: "Alyx Schubert"
date: "September 29, 2014"
output: word_document
---

# Introduction

This is a digital notebook to accompany the paper titled, “_” that will be published in _. It was written in R markdown and converted to html using the R knitr package. This enables us to embed the results of our analyses directly into the text to allow for a completely reproducible data analysis pipeline. A github repository is available where you can pull down your own version of the notebook to modify our analysis or adapt it to your analysis.

This document was generated using mothur v.1.33 and R.

# Prcessing 16S gene sequence data in mothur

## First, get the required files.
Sequence files can be found here _.

The following files can be found through the mothur wiki: 
silva.bacteria.fasta
trainset9_032012.pds.tax
trainset9_032012.pds.fasta

The following files can be found on the github repository for this notebook:
abxD01.files

## Now for the curation pipeline.
These sequences were generated using the Illumina MiSeq Platform. For data processing, we followed the MiSeq SOP outlined in the  [Kozich et al. manuscript](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3753973/). These commands come from the outline described in the [MiSeq SOP](http://www.mothur.org/wiki/MiSeq_SOP).

```
make.contigs(file=abxD01.files, processors=14)
summary.seqs(fasta=abxD01.trim.contigs.fasta, processors=14)
screen.seqs(fasta=abxD01.trim.contigs.fasta, group=abxD01.contigs.groups, summary=abxD01.trim.contigs.summary, maxambig=0, maxlength=275, processors=14)
unique.seqs(fasta=abxD01.trim.contigs.good.fasta)
count.seqs(name=abxD01.trim.contigs.good.names, group=abxD01.contigs.good.groups)
summary.seqs(count=abxD01.trim.contigs.good.count_table)
#Make sure silva.bacteria.fasta file in folder
pcr.seqs(fasta=silva.bacteria.fasta, start=11894, end=25319, keepdots=F, processors=14)
system(mv silva.bacteria.pcr.fasta silva.v4.fasta)
summary.seqs(fasta=silva.v4.fasta)
align.seqs(fasta=abxD01.trim.contigs.good.unique.fasta, reference=silva.v4.fasta, processors=14)
summary.seqs(fasta=abxD01.trim.contigs.good.unique.align, count=abxD01.trim.contigs.good.count_table)
screen.seqs(fasta=abxD01.trim.contigs.good.unique.align, count=abxD01.trim.contigs.good.count_table, summary=abxD01.trim.contigs.good.unique.summary, start=1968, end=11550, maxhomop=8)
summary.seqs(fasta=current, count=current)
filter.seqs(fasta=abxD01.trim.contigs.good.unique.good.align, vertical=T, trump=.)
unique.seqs(fasta=abxD01.trim.contigs.good.unique.good.filter.fasta, count=abxD01.trim.contigs.good.good.count_table)
pre.cluster(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.fasta, count=abxD01.trim.contigs.good.unique.good.filter.count_table, diffs=2)
chimera.uchime(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.fasta, count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.count_table, dereplicate=t, processors=14)
remove.seqs(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.fasta, accnos=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.accnos)
summary.seqs(fasta=current, count=current)
#Make sure the reference and training set files are in the analysis folder
classify.seqs(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.fasta, count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.count_table, reference=trainset9_032012.pds.fasta, taxonomy=trainset9_032012.pds.tax, cutoff=80)
remove.lineage(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.fasta, count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.count_table, taxonomy=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.taxonomy, taxon=Chloroplast-Mitochondria-unknown-Archaea-Eukaryota)
remove.groups(count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.pick.count_table, fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta, taxonomy=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.pick.taxonomy, groups=mock1-mock2-mock3-mock4-mock5-mock6-mock7-mock8)

#renamed fasta, count_table, taxonomy to have shorter names
system(cp abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.pick.pick.taxonomy abxD01.final.taxonomy)
system(cp abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.pick.pick.count_table abxD01.final.count_table)
system(cp abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.fasta abxD01.final.fasta)

set.current(fasta=abxD01.final.fasta, count=abxD01.final.count_table, taxonomy=abxD01.final.taxonomy)

```

## Creating phylotypes

Continuing the pipeline outlined on the mothur wiki's MiSeq SOP.

```
#PHYLOTYPE analysis:
#Note: label=1 is the genus level, which is the finest you can do, then label=2 is family and so forth
phylotype(taxonomy=abxD01.final.taxonomy)
make.shared(list=abxD01.final.tx.list, count=abxD01.final.count_table)
count.groups()
#The following size 1625 was chosen based on the results of count.groups(). I tried to minimize the number of samples at Day 0 that were dropped at that minimum cutoff.
sub.sample(shared=abxD01.final.tx.shared, size=1625)
classify.otu(list=abxD01.final.tx.list, count=abxD01.final.count_table, taxonomy=abxD01.final.taxonomy)
#tx.2=family, tx.3=order, etc
get.relabund(shared=abxD01.final.tx.2.subsample.shared) 
get.relabund(shared=abxD01.final.tx.3.subsample.shared)
get.relabund(shared=abxD01.final.tx.4.subsample.shared) 
get.relabund(shared=abxD01.final.tx.5.subsample.shared) 
get.oturep(column=abxD01.final.dist, list=abxD01.final.tx.list, label=2-3-4-5, fasta=abxD01.final.fasta, count=abxD01.final.count_table)

```

## Creating OTUs

It is worth noting that while phylotype analysis followed the listed protocol exactly, we deviated from the listed SOP for the OTU analysis.  This was due to the extensive size of this dataset and limitations in our computing abilities. Thus, in a **separate** folder we copied over the following files from earlier in the protocol:
abxD01.trim.contigs.good.unique.good.filter.unique.fasta
abxD01.trim.contigs.good.unique.good.filter.count_table
trainset9_032012.pds.fasta
trainset9_032012.pds.tax

**If this isn't done in a new folder, you will write over all your previous files because they have the same name.**

```
#Changed diffs from 2 to 3
pre.cluster(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.fasta, count=abxD01.trim.contigs.good.unique.good.filter.count_table, diffs=3)
chimera.uchime(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.fasta, count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.count_table, dereplicate=t, processors=14)
remove.seqs(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.fasta, accnos=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.accnos)
summary.seqs(fasta=current, count=current)
#Make sure the reference and training set files are in the analysis folder
classify.seqs(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.fasta, count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.count_table, reference=trainset9_032012.pds.fasta, taxonomy=trainset9_032012.pds.tax, cutoff=80)
remove.lineage(fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.fasta, count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.count_table, taxonomy=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.taxonomy, taxon=Chloroplast-Mitochondria-unknown-Archaea-Eukaryota)
remove.groups(count=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.pick.count_table, fasta=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.fasta, taxonomy=abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.pick.taxonomy, groups=mock1-mock2-mock3-mock4-mock5-mock6-mock7-mock8)

#renamed fasta, count_table, taxonomy to have shorter names
system(cp abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pds.wang.pick.pick.taxonomy abxD01.final.taxonomy)
system(cp abxD01.trim.contigs.good.unique.good.filter.unique.precluster.uchime.pick.pick.pick.count_table abxD01.final.count_table)
system(cp abxD01.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.fasta abxD01.final.fasta)

set.current(fasta=abxD01.final.fasta, count=abxD01.final.count_table, taxonomy=abxD01.final.taxonomy)

#OTU ANALYSIS:
#changed SOP taxlevel from 4 to 5
cluster.split(fasta=abxD01.final.fasta, count=abxD01.final.count_table, taxonomy=abxD01.final.taxonomy, splitmethod=classify, taxlevel=5, cutoff=0.20)
make.shared(list=abxD01.final.an.unique_list.list, count=abxD01.final.count_table, label=0.03)
count.groups()

#Note: the size varies based on what you see on count.groups, need to check the distribution of day 0 groups and clindamycin group
sub.sample(shared=abxD01.final.an.unique_list.shared, size=1625)
classify.otu(list=abxD01.final.an.unique_list.list, count=abxD01.final.count_table, taxonomy=abxD01.final.taxonomy, label=0.03, cutoff=80)
get.relabund(shared=abxD01.final.an.unique_list.0.03.subsample.shared)
get.oturep(column=abxD01.final.dist, list=abxD01.final.an.unique_list.list, label=0.03, fasta=abxD01.final.fasta, count=abxD01.final.count_table)
collect.single(shared=abxD01.final.an.unique_list.0.03.subsample.shared, calc=chao-invsimpson-sobs-simpsoneven-shannon-jclass, freq=100)
rarefaction.single(freq=100)
#Note summary.single--use the original shared file, and put subsample as same as before, will subsample 1000 times and get avgs
summary.single(calc=nseqs-coverage-chao-sobs-invsimpson-simpsoneven-shannoneven-shannon, shared=abxD01.final.an.unique_list.shared, subsample=1625)
```

# Data Analysis and Paper Figures

## Figure 1

This figure was built using the following R code.  The input files are listed in Github.

```{r, eval=FALSE}
#################################################
#Purpose: Convert csv matrix of columns you want converted into barcharts
#input: csv file following this heading format format:  sampleID, expgroup, nextDayCFU, Otu001, Otu002, Otu003, etc
#ex:file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.5.subsample.relabund.topdose.forlogscale2.csv
#Note: It is assumed that anything after the expgroup column are vectors containing data to be split by the groups in 'expgroup' 
#      and averaged with SD measured. 
#Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
#################################################
# Parameters to change:
# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.1.subsample.1.pick.relabund.topdose2.forlogscale.csv", header=T)
fileIDS<-read.csv("~/Desktop/mothur/abxD01/barcharts/topdose_tx1_barchart_ids.csv", header=T)
# Y Labels for each graph: 
abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")

# If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
# The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
sortbyphyl<-TRUE

# Highlight all and run!
#################################################

file<-file[,-1] #delete the first col of group names
avgs=NULL
avgs <-data.frame(levels(file$expgroup))
colnames(avgs) <- c("expgroup")
stds=NULL
stds <-data.frame(levels(file$expgroup))
colnames(stds) <- c("expgroup")

#Calculate average cdiff
avgCD=NULL
sdCD=NULL
avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
avgCD<-format(avgCD, scientific=TRUE, digits=2)
sdCD<-tapply(file$nextDayCFU, file$expgroup, sd)
file<-file[,-2] #delete the nexDayCFU column

for(i in 1:ncol(file)){

  if(i==1){
    ids<-names(file)
  }
  else{
    columni <- data.frame(file[,i])
    colnames(columni) <- c("columndata")

    #calculates mean for each column in file by expgroup and stores it in avgs data frame
    poop<- aggregate(columni$columndata~file$expgroup, FUN=mean) #poop stores the aggregated means for each group for this specific column
    colnames(poop) <- c(ids[1], ids[i])

    avgs<- merge(avgs, poop, by ="expgroup")

    #calculates st dev for each column in file by expgroup and stores it in stds data frame
    poop2<- aggregate(columni$columndata~file$expgroup, FUN=sd) #poop stores the aggregated means for each group for this specific column
    colnames(poop2) <- c(ids[1], ids[i])

    stds<- merge(stds, poop2, by ="expgroup")
  }
}

ordIDS <- fileIDS[order(fileIDS[,2]), ] #sort by the phyla
totalOTU <- dim(ordIDS)[1]

########for doing sort by phylum then within phylum (by relabund)
if(sortbyphyl == TRUE){
  ordIDS<-cbind(ordIDS, 0)

  for (i in 1:totalOTU){ #loop to fill out the final column for the phyla groups, while phyla is sorted!!
    if(i==1){
      ordIDS[i, 4] <- 1
    }
    else{
      if(ordIDS[i, 2]==ordIDS[i-1, 2]){
        ordIDS[i, 4] <- ordIDS[i-1, 4]
      }
      else{
        ordIDS[i, 4] <- ordIDS[i-1, 4] + 1
      }

    }
  }

  names(ordIDS)[4]<-paste("phynum")

  ##add row for phylum number to the avgs?
  #avgs<-rbind(avgs, 0)[c(9, 1, 2, 3, 4, 5, 6, 7, 8),] #this is hard coded for the top doses... also dont rerun this multiple times!!! might need to change
  avgs<-rbind(0,avgs) #this is hard coded for the top doses... also dont rerun this multiple times!!! might need to change, will give error

  numphyla<-max(ordIDS[,4])

  #test_avg <- avgs[, rev(order(avgs[1,2:length(avgs)-1]))] #sort all but other column (last)

  physums <- data.frame(c(1:numphyla))
  physums <- cbind(physums, 0)
  colnames(physums) <- c("Num", "Sum")

  leng<-length(avgs)

  #jtime<-NULL
  #itime<-NULL
  #inside<-FALSE
  #count<-0
  j<-1
  ##This loop is to fill the physums data frame with the sum of the relabund for each phylum represented
  for(i in 2:(leng-1)){ #searching through avgs, so leng is number of otus, -1 because i dont want to deal with "Other" column
    found<-FALSE
    #itime<-rbind(itime, i)
    j<-1
    while(found==FALSE){ #searching through ordIDS
      test<-ordIDS[j, 1] == names(avgs)[i]
      if(test) { 
        pnum<-ordIDS[j, 4]
        found<-TRUE
        inside<-TRUE
        avgs[1,i]<-pnum
      }
      else{
        j <- j + 1
        #count<-count+1
        #jtime<-rbind(jtime, j)

      }
    } #WARNING: could cause an error if dont find the ID, theres no check for the end of the list

    physums[pnum, 2] <- physums[pnum, 2] + avgs[2, i] ##add the relabund of the otu   
  }

  y<-rev(order(physums$Sum))
  x<-avgs[1, 2:(leng-1)]

  ord_physums<-physums[rev(order(physums[,2])),]
  sort_avgs<-data.frame(avgs$expgroup)
  names(sort_avgs)[1]<-c("expgroup")

  for(i in 1:numphyla){ #loop is to sort by decreasing phyla and perform internal sort within phylum, return the new sorted avgs to go into mavgs
    val<-NULL
    phy<-which(x %in% c(ord_physums[i,1])) #phy returns the indices for where that phyNum is
    for(j in 1:length(phy)){ #using locations of phylum i to find in avgs
      val[j]<-avgs[2,phy[j]+1]
    }
    phy<-phy[rev(order(val))]

    for(j in 1:length(phy)){ #now loop through the sorted sub cols and add them to the sort_avgs
      sort_avgs<-cbind(sort_avgs, avgs[,phy[j]+1])
      names(sort_avgs)[length(sort_avgs)]<-names(avgs)[phy[j]+1]
    }
  }

  attach(avgs)
  sort_avgs<-cbind(sort_avgs, Other)
  detach(avgs)

  sort_avgs <- sort_avgs[-1,] #get rid of phylum first row from earlier sorting

  #loop for getting the right ids in the right order as sort
  ids<-data.frame(1:(totalOTU))

  row.names(sort_avgs)<-sort_avgs$expgroup
  sort_avgs<-sort_avgs[,-1]

  for (i in 1:(length(sort_avgs)-1)){
    found<-FALSE
    #itime<-rbind(itime, i)
    j<-1
    while(found==FALSE){ #searching through ordIDS
      test<-ordIDS[j, 1] == names(sort_avgs)[i]
      if(test) { 
        ids[i,2]<-ordIDS[j, 3]
        found<-TRUE
      }
      else{
        j <- j + 1
        #count<-count+1
        #jtime<-rbind(jtime, j)
      }
    }

  }

  names(ids)<-c("x", "name")
  ids<-rbind(ids, data.frame(x=totalOTU+1, name= "Other"))

  mavgs<-as.matrix(sort_avgs)

  leng<-dim(mavgs)[2]

  numgr <- nlevels(file$expgroup)
} ##end if(sortbyphyl == TRUE){

####order the results by the first group alpha numerically
if(sortbyphyl == FALSE){

  row.names(avgs)<-avgs$expgroup
  avgs<-avgs[,-1]
  ordered_avgs<- avgs[, rev(order(avgs[1,1:length(avgs)-1]))] #sort all but other column (last)
  attach(avgs)
  ordered_avgs <- cbind(ordered_avgs, Other) #put other back on
  detach(avgs)

  mavgs<-as.matrix(ordered_avgs)
  leng<-dim(mavgs)[2]

  numgr <- nlevels(file$expgroup)

  ids<-data.frame(1:(totalOTU))

  for (i in 1:(length(ordered_avgs)-1)){
    found<-FALSE
    #itime<-rbind(itime, i)
    j<-1
    while(found==FALSE){ #searching through ordIDS
      test<-ordIDS[j, 1] == names(ordered_avgs)[i]
      if(test) { 
        ids[i,2]<-ordIDS[j, 3]
        found<-TRUE
      }
      else{
        j <- j + 1
        #count<-count+1
        #jtime<-rbind(jtime, j)
      }
    }

  }

  names(ids)<-c("x", "name")
  ids<-rbind(ids, data.frame(x=totalOTU+1, name= "Other"))

}  #end if(sortbyphyl == FALSE){

###started to do the stds table... didnt complete
#stds<-stds[,-1]
#mstds<-as.matrix(stds)

###PLOT PARAMETERS
par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)

for(j in 1:numgr){

  if(j != numgr){
    barplot(mavgs[j, 1:leng], ylab=NULL, col="black", yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
    #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
    # k <- k+1
    axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
    mtext(abx[j], side=2, line=6, cex=.8)
    mtext(avgCD[j], side=2, line=4.5, cex=.8)
    abline(h=c(0.0001, 1), lwd=3)
    abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
    abline(h=c(.25,.5, .75), col="black", lty="dashed")
  }

  else{ #the last graph needs different margins    
    label<-barplot(mavgs[j, 1:leng], col="black",ylab=NULL,  yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
    #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
    axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
    mtext(abx[j], side=2, line=6, cex=.8)
    mtext(avgCD[j], side=2, line=4.5, cex=.8)
    axis(1, at=(label[,1]), labels=FALSE)
    text(label[,1]+.13, .00005, label=ids[,2], xpd=NA, pos=2, srt=45, cex=1.2)
    #text(-3.9,.0001, label=expression(paste(italic("C.d."), " CFU/g Feces:")), xpd=NA, pos=2, srt=90, cex=1.1)
    abline(h=c(0.0001,1), lwd=3)
    abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
    abline(h=c(.25,.5, .75), col="black", lty="dashed")

  }

}
par(mfrow=c(1, 1))

##############END CODE###########################
#################################################
```

The graph can be sorted 2 ways either by the untreated's decreasing relabund, or by the most abundant within each phylum. For each treatment, the resulting *C. difficile* CFU/g Feces is also labeled on the Y axis below the treatment name. File saved as "topdose2_tx1_sorted_10x20.pdf"

### TO DO STILL
- change the "0" to be min o 0.001 instead
- add the 1/2 dashed lines
- also wanted to try to make this table in the other direction: Phyla/OTUs as each graph with different shadings or colors for each treatment... will be good for titration graphs

### 2014-11-11 abxD01_analysis.Rmd
updated markdown file

@@
 
+This figure shows the differences in the communities' structures following different antibiotic treatments.
+The average *C.
+difficile* CFU/g Feces is also calculated in this code for each treatment group.
+
 This figure was built using the following R code.  The input files are listed in Github.
@@
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.1.subsample.1.pick.relabund.topdose2.forlogscale.csv", header=T)
+file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
 fileIDS<-read.csv("~/Desktop/mothur/abxD01/barcharts/topdose_tx1_barchart_ids.csv", header=T)
@@
 - also wanted to try to make this table in the other direction: Phyla/OTUs as each graph with different shadings or colors for each treatment... will be good for titration graphs
+
+## Figure 2
+
+This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1.
+
+### Figure 2 Correlation Calculations
+
+Before we get to how this graph was made, we first used an R script to calculate the spearman correlation of OTUs' relative abundance.
+
+The required file is located on Github.
+
+```{r, eval=FALSE}
+
+meta <- read.table('~/Desktop/mothur/abxD01/correlation/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
+#meta<-meta[1:96] #change based on number of OTUs above .05%, then add 2 for first two cols
+c<-1
+otu <- c()
+cor.spear <- c()
+pval.spear <- c()
+cor.ken = c()
+pval.ken = c()
+for(i in 3:length(meta)){
+  otu[c] <- colnames(meta[i])
+  cor.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$estimate
+  pval.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$p.value
+  cor.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$estimate #good to see because kendall handles ties
+  pval.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
+  c <- c+1
+}
+
+pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
+pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons
+
+results = NULL
+results <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
+colnames(results) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
+results <- results[order(results[,3]),]  #order by pvalue column=3
+write.table(results[1:dim(results)[1],], file="~/Desktop/mothur/abxD01/correlation/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
+
+```
+
+### Figure 2 Graph
+
+Necessary files are on github, and it was made using the following code:
+
+```{r, eval=FALSE}
+
+##Paper Figure: correlations for topdose 
+c<-read.csv("~/Desktop/mothur/abxD01/correlation/corr_allSig_topdose2_stripchart.csv", header=T)
+labels=c("Lachnospiraceae", "Ruminococcaceae", "Clostridia", "Lactobacillales", "Firmicutes", "Bacillales", "Porphyromonadaceae", "Bacteroidales", "Bacteroidetes", "Actinobacteria", "Proteobacteria", "Anaeroplasma", "Deinococcus", "Unclassified")
+ns=c("n=18", "n=9", "n=8", "n=4", "n=2", "n=1", "n=15", "n=4", "n=1", "n=4", "n=3", "n=1", "n=1", "n=3")
+
+par(mar=c(12, 7, 2, 2) +0.1, mgp=c(5, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
+stripchart(c$cor~c$order, vertical=TRUE, ylab="Correlation", ylim=c(.85, -.85), cex.axis=1.1, cex=2, pch=21, lwd=3, col="black", xaxt="n", method="jitter",  cex.lab=1.7, yaxt="n")
+axis(1, cex.axis=1, at=(1:nlevels(c$name)), labels=FALSE)
+text(1:nlevels(c$name)+.1, par("usr")[3]-0.16, label=labels, xpd=NA, pos=2, srt=45, cex=1.2)
+text(1:nlevels(c$name), par("usr")[3]-0.01, label=ns, xpd=NA, pos=1, cex=.9)
+axis(2, cex.axis=1.1, at=seq(-0.85,+0.85, by=.1), las=1)
+abline(v=c(1:nlevels(c$name)), col="dark gray")
+#abline(h=c(.3, -.3), col="dark gray", lty="dashed", lwd=4) #change to something else
+abline(h=0, col="black", lwd=2)
+
+```
+
+Figure saved as "corr_allSig_topdose2_10x8.pdf".
+
+### TO DO STILL
+- Do I want to change the groupings?
+
+## Figure 3
+
+This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.
+
+### TO DO STILL
+- I want to change the graph so that the OTUs are their own graphs and the titrations are side by side
+
+## Figure 4
+
+This figure is a heatmap showing the correlation analysis results for the original set of antibiotic treatments and doses with the titration data sets.
+
+### TO DO STILL
+- Add "ns" to the graph where not significant
+
+## Figure 5
+
+This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole.
+
+### TO DO STILL
+- Do for ampicillin too
+- Pick the same groups showed in Figure 2.

### 2014-11-13 abxD01_analysis.Rmd
backup

@@
 date: "September 29, 2014"
-output: word_document
+output: pdf_document
 ---
@@
 - Pick the same groups showed in Figure 2.
+
+## Figure 6
+
+This figure shows the results of our model predicting *C. difficile* colonization given relative abundances of a subset of bacteria.
+
+### TO DO STILL
+- Make figure!

### 2014-11-13 abxD01.r

@@ -1,0 +1,1922 @@
+#function to plot the error bars, either SEM or SD
+##################################################
+error.bar <- function(x, y, upper, lower=upper, length=0.1,...){
+  if(length(x) != length(y) | length(y) !=length(lower) | length(lower) != length(upper))
+    stop("vectors must be same length")
+  arrows(x,y+upper, x, y-lower, angle=90, code=3, length=length, ...)
+}
+
+###Deconstructed Stacked bar charts
+###################################
+###Deconstructed Stacked bar charts
+
+c<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.2.subsample.2.pick.relabund.D0.csv", header=T)
+b<-read.csv("~/Desktop/mothur/abxD01/correlation/abxD01.final.tx.2.subsample.2.pick.relabund.D01.csv", header=T)
+
+avgCD<-tapply(b$nextDayCFU, b$exp, mean)
+sdCD<-tapply(b$nextDayCFU, b$exp, sd)
+
+avg1<-tapply(c$Otu001, c$exp, mean)
+sd1<-tapply(c$Otu001, c$exp, sd)
+
+avg14<-tapply(c$Otu014, c$exp, mean)
+sd14<-tapply(c$Otu014, c$exp, sd)
+
+avg27<-tapply(c$Otu027, c$exp, mean)
+sd27<-tapply(c$Otu027, c$exp, sd)
+
+avg7<-tapply(c$Otu007, c$exp, mean)
+sd7<-tapply(c$Otu007, c$exp, sd)
+
+par(mfrow=c(5, 1)) 
+
+plotCD<-barplot(avgCD, col="black", cex.names=.6) 
+error.bar(plotCD, avgCD, sdCD) #plot SD
+
+leg<-c("cdiff CFU", "porphyromonadaceae -0.65", "alistipes -0.68", "clostridia -0.58", "enterobacteriaceae +0.55")
+col<-c("black", "blue","darkgreen", "darkorange", "red")
+#pch<-c(16, 16, 16)
+legend("topright", legend=leg, pch=16, col=col,  cex=1, bty="n")
+
+plot1<-barplot(avg1, col="blue", ylim=c(0, .8), cex.names=.6) #how can i change this to log scale
+error.bar(plot1, avg1, sd1) #plot SD
+
+plot14<-barplot(avg14, col="darkgreen", ylim=c(0, .8), cex.names=.6)
+error.bar(plot14, avg14, sd14) #plot SD
+
[…]

### 2014-11-20 abxD01_analysis.Rmd
major update
examples of all figures put up, with code currently working with for
each, to do list for each figure largely up-to-date

@@
 
+The code for this is still in the works. But is uploaded to github under Figure 3 folder.
+
+Input files on github:
+1. abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv
+2. abxD01/barcharts/allvanctitr_tx2_barchart_ids.csv
+
 ### TO DO STILL
 - I want to change the graph so that the OTUs are their own graphs and the titrations are side by side
+-once code is working then put up the other titration data files
 
@@
 
+2 input files found in Figure 4 github folder:
+1. correl_heatmap_0.01rel_topdose2_newtitr.csv
+2. correl_heatmapSIDE_0.01rel_topdose2_newtitr.csv
+
+```{r, eval=FALSE}
+library(gplots)
+correl<- read.csv("~/Desktop/mothur/abxD01/correlation/correl_heatmap_0.01rel_topdose2_newtitr.csv", header=T)
+
+row.names(correl)<-correl$OTU
+correl_matrix<-data.matrix(correl)
+correl_matrix<-correl_matrix[,-1]
+correl_matrix<-data.matrix(correl_matrix)
+my_palette <- colorRampPalette(c("blue", "white", "red"))(n = 299)
+breaks = c(seq(-1,-.33,length=100),seq(-.33,.33,length=100),seq(.33,1,length=100))#changed to equal numbers
+
+#par(mar=c(4, 6, 3, 4.5) +0.1) #default is 5.1 4.1 4.1 2.1
+
+side<-scan("~/Desktop/mothur/abxD01/correlation/correl_heatmapSIDE_0.01rel_topdose2_newtitr.csv", what=",", sep=",")
+side2<-read.csv("~/Desktop/mothur/abxD01/correlation/correl_heatmapSIDE_0.01rel_topdose2_newtitr.csv", header=T)
+side3<-as.matrix(side2)
+#lmat=rbind(c(4,3), c(2,1))
+#lmat=rbind(c(4,3), c(1,2))
+
+#lmat order: 1)row dendrogram, 2) heatmap, 3)col?, 4)key?--key disappeared
+lmat = rbind(3:4, 1:2)
+lmat = rbind(c(2,4), c(1,3))
+lmat = rbind(c(0,3), c(2,1))
+lmat=NULL
+lwid=NULL
+lhei=NULL
+lwid = c(.5, 2, 4)
+lhei = c(.7, 4)
+
+correl_heatmap<-heatmap.2(correl_matrix, scale="none",  col=my_palette, breaks=breaks, density.info="none", lhei=lhei,lwid=lwid, RowSideColors=side3[,2], labRow=side3[,1], trace="none",margins=c(8, 25), dendrogram="none",  Rowv=FALSE, Colv=FALSE,  na.color="black", cexCol=1.5,  key=TRUE, keysize=1, cexRow=1) #then plot was 7x8in, portrait
+
+```
+
+Each row in this heatmap represents a single OTU.
+The side colors show broad taxonomic groups of each OTU.
+The subsequent columns show the correlation value for each OTU (Day 0) with Day 1 *C.
+difficile* colonization levels by plating.
+The first column represents the correlations calculations over all the original antibiotics used in the study.
+The last 3 columns represent the correlation calculations over the titration experiments for the given antibiotic.
+An example pdf of this figure is on github.
+
 ### TO DO STILL
 - Add "ns" to the graph where not significant
+- Is this the order of drugs I want-keep consistent
+- for all the significant values for each pair of original + abx rows that are both significant, quantify that significance for each column (abx treatment)... see personal notes for more
+- change xlabel names
+- change code to get the key back
 
@@
 
-This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole.
+This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole. The figure is shown as a pdf on github.
+
+Input files found on github:
+1. abxD01.final.tx.2.subsample.2.pick.metro.relabund.d0s.txt
+
+```{r, eval=FALSE}
+#First calculating:
+b<-read.delim("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.2.subsample.2.pick.metro.relabund.d0s.txt", header=T)
+
+avg1<-tapply(b$Otu001, b$order, mean)
+sd1<-tapply(b$Otu001, b$order, sd)
+
+avg11<-tapply(b$Otu011, b$order, mean)
+sd11<-tapply(b$Otu011, b$order, sd)
+
+avg02<-tapply(b$Otu002, b$order, mean)
+sd02<-tapply(b$Otu002, b$order, sd)
+
+avg03<-tapply(b$Otu003, b$order, mean)
+sd03<-tapply(b$Otu003, b$order, sd)
+
+avg7<-tapply(b$Otu007, b$order, mean)
+sd7<-tapply(b$Otu007, b$order, sd)
+
+avg05<-tapply(b$Otu005, b$order, mean)
+sd05<-tapply(b$Otu005, b$order, sd)
+
+avg16<-tapply(b$Otu016, b$order, mean)
+sd16<-tapply(b$Otu016, b$order, sd)
+
+avg06<-tapply(b$Otu006, b$order, mean)
+sd06<-tapply(b$Otu006, b$order, sd)
+
+avg04<-tapply(b$Otu004, b$order, mean)
+sd04<-tapply(b$Otu004, b$order, sd)
+
+#Now graphing:
+par(mfrow=c(4, 3))
+
+par(mar=c(2, 4, 2, 3) +0.1)
+plot1<-barplot(avg1, col="darkblue", cex.axis=1.3, xaxt="n") 
+error.bar(plot1, avg1, sd1) #plot SD
+
+par(mar=c(2, 3, 2, 3) +0.1)
+plot03<-barplot(avg03, col="steelblue", cex.axis=1.3, xaxt="n")
+error.bar(plot03, avg03, sd03) #plot SD
+
+plot11<-barplot(avg11, col="darkgreen", cex.axis=1.3, xaxt="n")
+error.bar(plot11, avg11, sd11) #plot SD
+
+par(mar=c(2, 4, 2, 3) +0.1)
+plot02<-barplot(avg02, col="darkred", cex.axis=1.3, xaxt="n")
+error.bar(plot02, avg02, sd02) #plot SD
+
+par(mar=c(2, 3, 2, 3) +0.1)
+plot05<-barplot(avg05, col="orange2", cex.axis=1.3, xaxt="n")
+error.bar(plot05, avg05, sd05) #plot SD
+
+plot04<-barplot(avg04, col="gold1", cex.axis=1.3, xaxt="n")
+error.bar(plot04, avg04, sd04) #plot SD
+
+par(mar=c(2, 4, 2, 3) +0.1)
+plot16<-barplot(avg16, col="deeppink3", cex.axis=1.3, xaxt="n")
+error.bar(plot16, avg16, sd16) #plot SD
+axis(1, at=plot16, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+par(mar=c(2, 3, 2, 3) +0.1)
+plot7<-barplot(avg7, col="blueviolet", cex.axis=1.3, xaxt="n")
+error.bar(plot7, avg7, sd7) #plot SD
+axis(1, at=plot7, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+plot06<-barplot(avg06, col="black", cex.axis=1.3, xaxt="n")
+error.bar(plot06, avg06, sd06) #plot SD
+axis(1, at=plot06, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+#Now figure legend:
+plot(1, type="n", axes=F, xlab="", ylab="")
+
+leg<-c("Porphyromonadaceae", "Bacteroides","Erysipelotrichaceae" )
+col<-c("darkblue",  "steelblue", "darkgreen" )
+#pch<-c(16, 16, 16)
+legend("left", legend=leg, pch=15, col=col,  cex=1.7, bty="n", ncol=1)
+
+plot(1, type="n", axes=F, xlab="", ylab="")
+
+leg<-c("Lachnospiraceae",  "Lactobacillus", "Ruminococcaceae" )
+col<-c("darkred", "orange2", "gold1" )
+#pch<-c(16, 16, 16)
+legend("left", legend=leg, pch=15, col=col,  cex=1.7, bty="n", ncol=1)
+
+plot(1, type="n", axes=F, xlab="", ylab="")
+
+leg<-c("Bifidobacterium",  "Enterobacteriaceae","Akkermansia" )
+col<-c("deeppink3", "blueviolet", "black" )
+#pch<-c(16, 16, 16)
+legend("left", legend=leg, pch=15, col=col,  cex=1.7, bty="n", ncol=1)
+
+par(mfrow=c(1, 1))
+
+```
 
@@
 - Do for ampicillin too
-- Pick the same groups showed in Figure 2.
+- Pick the same groups showed in Figure 2. AND as side heatmap.
 
@@
 
+### Figure 6: Model Building
+
+In order to determine the best possible set of models, I first narrowed down the list of candidate OTUs to be considered in the model.
+This candidate list was created using several criteria, based on the data from the original set of antibiotic experiments alone:
+
+1. Using top results of random forest's feature selection. 
+2. Incorporating both strongly positively correlated OTUs and strongly negatively correlated OTUs. We used the spearman correlation analysis for this, see Figure 2.
+3.
+Using sparCC in mothur to determine any strong correlations (>90%) among the candidate list OTUs.
+We found the highest correlation between OTUs to be ___% This was not high enough to consider eliminating from the list.
+4. Limiting the candidate list to 15 OTUs.
+
+For the model, I limited models to include anywhere from 1 to 10 parameters.
+I log transformed the *C.
+difficile* CFU/g feces on Day 1 for inclusion in a linear model.
+The OTU data was in the form of a shared file.
+
+The script for running this code is called "abxD01.lmModel.compare.R". The input files for this script are:
+1. abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.15otus.rfnegpos.csv
+
+I realize I could've done the modelling comparison script in a loop and shortened the code length significantly, but deal with it.
+
 ### TO DO STILL
-- Make figure!
+- Make figure! for modelling results
+- The criteria for picking the otus, these results can be listed in supplementary tables possibly. 
+- Consider rerunning the model building code using a bigger candidate list, or using higher level taxonomy!!!
+- Calculate the BIC for each--change code for this

### 2014-11-20 Figure 3/abxD01.barcharts.xOTU.test.r

@@ -1,0 +1,355 @@
+#################################################
+#Purpose: Convert csv matrix of columns you want converted into barcharts
+#input: csv file following this heading format format:  sampleID, expgroup, (nextDayCFU,) Otu001, Otu002, Otu003, etc
+#ex:file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.5.subsample.relabund.topdose.forlogscale2.csv
+#Note: It is assumed that anything after the expgroup column are vectors containing data to be split by the groups in 'expgroup' 
+#      and averaged with SD measured. 
+#Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
+#################################################
+# Parameters to change:
+# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
+file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Desktop/mothur/abxD01/barcharts/allvanctitr_tx2_barchart_ids.csv", header=T)
+# Y Labels for each graph: 
+abx<-c("Untreated", "Vanc 0.625", "Vanc 0.3", "Vanc 0.1")
+
+# If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
+# The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
+sortbyphyl<-TRUE
+
+# If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
+# If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
+graphbyphyl<-TRUE
+
+
+# Highlight all and run!
+#################################################
+
[…]

### 2014-11-20 Figure 6/abxD01.lmModel.compare.R

@@ -1,0 +1,1089 @@
+require(foreign)
+require(ggplot2)
+require(MASS)
+
+
+## linear model
+#######################################
+## linear model
+## File headings: SampleName nextDayCFU OTU1 OTU2 ... total OTU candidates being tested, candidate number should be at least 10
+## nextDayCFU should be log transformed
+## IN file type is *.csv
+
+
+
+
+topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.15otus.rfnegpos.csv", header=T)
+td<-topdose[,-1]
+ids<-names(td)
+ids = ids[-1]
+
+numOTU=length(td)-1
+
+results = data.frame(matrix(0, ncol=numOTU+5)) #table as follows: model# AIC numVar intercept OTU1B OTU2B ... etc all included
+colnames(results)[1:5] = c("Modelnum", "AIC", "rsq", "vars", "intCoeff")
+colnames(results)[6:(numOTU+5)] = ids
+#make results headings
+
+#results = NULL
+#results = data.frame(NULL)
+
+
+#all combos for a model with 1 variable
+for(i in 1:numOTU){ 
+  m1 = NULL
+  err = NULL
+  err<-tryCatch(
+{
+  m1<-lm(td[,1] ~ td[,(i+1)])
+},
+error = function(cond) {
+  message(cond)
+  return(NULL)     
+},
+finally = {}
+  )
+
+if(is.null(err)) {
+  results[i, 1] = i
+  results[i, 2] = c("NA")
+  results[i, 3] = c("NA")
+  results[i, 4] = c(1)
+  results[i, 5] = names(td)[(i+1)] #add more names as in formula
+}
+else{
+  results[i, 1] = i #model number
+  results[i, 2] = AIC(m1) #AIC 
+  coeff=m1$coefficients
+  names(coeff)[2] = names(td)[i+1]
+  vars = length(coeff)-1
+  results[i, 3] = summary(m1)$r.squared
+  results[i, 4] = vars #number of variables in model
+  results[i, 5] = coeff[1] #intercept coefficient
+  
+  for(j in 6:(numOTU+5)){  #fill results columns for each OTU whether NA or a coefficient value
+    for(k in 2:(vars+1)){
+      if(names(coeff)[k]==names(results)[j]){
+        results[i, j] =   coeff[k]  #print coefficient value
+      }
+      else{
+        if(k>vars){
[…]

### 2014-11-20 abxD01_analysis.Rmd
updated r markdown files
finished code for fig 3, update fig 1

@@
 date: "September 29, 2014"
-output: pdf_document
+output: word_document
 ---
@@
 ### TO DO STILL
-- change the "0" to be min o 0.001 instead
+- change the "0" to be min of 0.001 instead of 0.0001 relabund
+    - **completed 11/20/14, just make sure future files are changed for other graphs using the "abxD01.barcharts.xOTU.test.r"**
 - add the 1/2 dashed lines
+    - **completed 11/20/14**
 - also wanted to try to make this table in the other direction: Phyla/OTUs as each graph with different shadings or colors for each treatment... will be good for titration graphs
@@
 
+This figure shows correlation values for OTUs with next day *C. difficile* CFU levels calculated across the original set of antibiotic experiments.
+
 Necessary files are on github, and it was made using the following code:
@@
 - I want to change the graph so that the OTUs are their own graphs and the titrations are side by side
--once code is working then put up the other titration data files
+- once code is working then put up the other titration data files
+- change min relabund from 0.0001 to 0.001 for cef and strep
+    - **11/20/14, changed the vanc titrations min**
 

### 2014-11-20 Figure 3/abxD01.barcharts.xOTU.test.r

@@ -8,7 +8,7 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
 fileIDS<-read.csv("~/Desktop/mothur/abxD01/barcharts/allvanctitr_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
 abx<-c("Untreated", "Vanc 0.625", "Vanc 0.3", "Vanc 0.1")
@@ -274,29 +274,29 @@ if(graphbyphyl==FALSE){
   for(j in 1:numgr){
     
     if(j != numgr){
-      barplot(mavgs[j, 1:leng], ylab=NULL, col="black", yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
+      barplot(mavgs[j, 1:leng], ylab=NULL, col="black", yaxt="n", xaxt="n", ylim=c(0.001, 1), log="y", cex.names=3)
       #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
       # k <- k+1
-      axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
+      axis(2, las=1, at=c(.001, .01, .1, 1), labels=c(0, .01, .1, 1), cex.axis=1.1)
       mtext(abx[j], side=2, line=6, cex=.8)
       mtext(avgCD[j], side=2, line=4.5, cex=.8)
-      abline(h=c(0.0001, 1), lwd=3)
-      abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
-      abline(h=c(.25,.5, .75), col="black", lty="dashed")
+      abline(h=c(0.001, 1), lwd=3) #min/max
[…]

### 2014-11-20 abxD01_analysis.Rmd
updated figure 3 files

@@
 - I want to change the graph so that the OTUs are their own graphs and the titrations are side by side
+    -**11/20/14, completed most of the graphical parameters.*
 - once code is working then put up the other titration data files
@@
     - **11/20/14, changed the vanc titrations min**
+- *Possibly work on merging singles*
+- *Possibly work on getting plots on top of grid lines*
 

### 2014-11-20 Figure 3/abxD01.barcharts.xOTU.test.r

@@ -311,7 +311,7 @@ if(graphbyphyl==FALSE){
 if(graphbyphyl==TRUE){  
   par(mfrow=c(numphyla+1, 1)) #+2 to give extra labeling space--numphyl doesn't include the Other group
 #  par(mfrow=c(3, 1)) #temporarily for testing
-  par(mar=c(2, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is  par(mar=c(5, 4, 4, 2 ) +0.1, mgp=c(3, 1, 0)) bot/left/top/right, also default mgp is c(3,1,0)
+  par(mar=c(2.5, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is  par(mar=c(5, 4, 4, 2 ) +0.1, mgp=c(3, 1, 0)) bot/left/top/right, also default mgp is c(3,1,0)
   colors= gray.colors(numgr, start=0, end=1, alpha=NULL)
   currphy=ids[1, 4]
   k <- 1
@@ -330,13 +330,14 @@ if(graphbyphyl==TRUE){
           break}
       }
      
-    label<-barplot(mavgs[,j:(leng+j-1)], beside=TRUE, ylab=ids[j,3], col=colors, yaxt="n", xaxt="n", ylim=c(0.001, 1), log="y", cex.names=5)
+    label<-barplot(mavgs[,j:(leng+j-1)],, beside=TRUE, ylab=ids[j,3], col=colors, yaxt="n", xaxt="n", ylim=c(0.001, 1), log="y", cex.names=5)
+    
     axis(2, las=1, at=c(.001, .01, .1, 1), labels=c(0, .01, .1, 1), cex.axis=1.1)
    # mtext("Relative Abundance", side=2, line=6, cex=.8)
     #mtext(avgCD[j], side=2, line=4.5, cex=.8)
     abline(h=c(0.001, 1), lwd=3) #min/max
-    abline(h=c(0.01, 0.1), col="black", lty="longdash", lwd=2)
-    abline(h=c(0.005, 0.05, 0.25, 0.5, 0.75), col="black", lty="dashed")
+    abline(h=c(0.01, 0.1), col="gray", lty="longdash", lwd=2)
+    abline(h=c(0.005, 0.05, 0.25, 0.5, 0.75), col="gray", lty="dashed")
 
     labelAVG=apply(label, 2, mean)
     axis(1, at=(labelAVG), labels=FALSE)

### 2014-11-21 abxD01_analysis.Rmd
removed the to do lists from the rmkdwn file

--- 2014-11-20 abxD01_analysis.Rmd
@@
 
-### TO DO STILL
-- change the "0" to be min of 0.001 instead of 0.0001 relabund
-    - **completed 11/20/14, just make sure future files are changed for other graphs using the "abxD01.barcharts.xOTU.test.r"**
-- add the 1/2 dashed lines
-    - **completed 11/20/14**
-- also wanted to try to make this table in the other direction: Phyla/OTUs as each graph with different shadings or colors for each treatment... will be good for titration graphs
-
 ## Figure 2
@@
 
-### TO DO STILL
-- Do I want to change the groupings?
-
 ## Figure 3
@@
 2. abxD01/barcharts/allvanctitr_tx2_barchart_ids.csv
-
-### TO DO STILL
-- I want to change the graph so that the OTUs are their own graphs and the titrations are side by side
-- once code is working then put up the other titration data files
-- change min relabund from 0.0001 to 0.001 for cef and strep
-    - **11/20/14, changed the vanc titrations min**
 
@@
 An example pdf of this figure is on github.
-
-### TO DO STILL
-- Add "ns" to the graph where not significant
-- Is this the order of drugs I want-keep consistent
-- for all the significant values for each pair of original + abx rows that are both significant, quantify that significance for each column (abx treatment)... see personal notes for more
-- change xlabel names
-- change code to get the key back
 
@@
 
-### TO DO STILL
-- Do for ampicillin too
-- Pick the same groups showed in Figure 2. AND as side heatmap.
-
 ## Figure 6
@@
 I realize I could've done the modelling comparison script in a loop and shortened the code length significantly, but deal with it.
-
-### TO DO STILL
-- Make figure! for modelling results
-- The criteria for picking the otus, these results can be listed in supplementary tables possibly. 
-- Consider rerunning the model building code using a bigger candidate list, or using higher level taxonomy!!!
-- Calculate the BIC for each--change code for this

### 2014-12-02 abxD01_analysis.Rmd
Rmd includes RF analysis, added BIC to lm script

@@
 
-1. Using top results of random forest's feature selection. 
-2. Incorporating both strongly positively correlated OTUs and strongly negatively correlated OTUs. We used the spearman correlation analysis for this, see Figure 2.
+1.
+Using top 15 results from random forest's feature selection.
+These OTUs had the highest % increase in MSE in the random forest regression model.
+These 15 were the top OTUs that were also not "unclassified".
+2.
+Incorporating both the top 5 most positively correlated OTUs and 10 most negatively correlated OTUs.
+We used the spearman correlation analysis for this, see Figure 2.
+The number of significantly negatively correlated OTUs far outnumber the number of signifacntly positive correlated OTUs.
+Due to this difference we chose to incorporate more negative OTUs over positive.
 3.
 Using sparCC in mothur to determine any strong correlations (>90%) among the candidate list OTUs.
-We found the highest correlation between OTUs to be ___% This was not high enough to consider eliminating from the list.
-4. Limiting the candidate list to 15 OTUs.
+We found the highest correlation between OTUs to be 84.0% between OTUs 6 & 7, which subsequently dropped off to 70.1%.
+This was not high enough to consider eliminating from the list.
 
@@
 I realize I could've done the modelling comparison script in a loop and shortened the code length significantly, but deal with it.
+
+### Figure 6: Model Building - Random Forest's Feature Selection
+
+The input file includes the Day 0 shared values for OTUs (whose relative abundance was equal to or greater than 1%) and the log transformed next day CFU values (log(CFU)).
+
+Then in r: 
+```{r, eval=TRUE}
+topdose_logtrans<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.topdose2.regression.logtrans.filter16mintot.csv", header=T)
+training_logtrans<-topdose_logtrans[,-1]
+newtit_logtrans<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.newtitration.regression.logtrans.filter16mintot.csv", header=T)
+testing_logtrans<-newtit_logtrans[,-1] #new titration data only
+delay_logtrans<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.delays.regression.logtrans.csv", header=T)
+testing2_logtrans<-delay_logtrans[,-1]
+
+library(randomForest)
+
+#fit the randomforest model
+model_logtrans <- randomForest(nextDayCFU~., 
+                      data = training_logtrans,  outscale=TRUE,
+                      importance=TRUE, proximity=TRUE,
+                      keep.forest=TRUE, ntree=5000
+)
+plot(model_logtrans)
+
+print(model_logtrans)
+
+#what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
+varImpPlot(model_logtrans, type=1)
+imp<-importance(model_logtrans)
+most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
+
+#predict the outcome of the testing data
+predicted_logtrans <- predict(model_logtrans, newdata=testing_logtrans[ ,-1])
+predicted2_logtrans <- predict(model_logtrans, newdata=testing2_logtrans[ ,-1])
+
+# what is the proportion variation explained in the outcome of the testing data?
+# i.e., what is 1-(SSerror/SStotal)
+actual_logtrans <- testing_logtrans$nextDayCFU
+actual2_logtrans <- testing2_logtrans$nextDayCFU
+
+ybar = mean(actual_logtrans)
+SSres = sum((actual_logtrans-predicted_logtrans)^2)
+SStot = sum((actual_logtrans-ybar)^2)
+rsq = 1-(SSres/SStot)
+
+#rsq_logtrans <- 1-sum((actual_logtrans-predicted_logtrans)^2)/sum((actual_logtrans-mean(actual_logtrans))^2)
+print(rsq)
+
+ybar2 = mean(actual2_logtrans)
+SSres2 = sum((actual2_logtrans-predicted2_logtrans)^2)
+SStot2 = sum((actual2_logtrans-ybar2)^2)
+rsq2 = 1-(SSres2/SStot2)
+
+print(rsq2)
+
+write.table(cbind(actual_logtrans, predicted_logtrans), file="~/Desktop/mothur/abxD01/rf/RF.regression.topTitr.txt", sep="\t", row.names=FALSE)
+
+```
+
+The "plot(model_logtrans)" shows that setting ntree=5000 is enough to get a consistently low error.
+
+We can also see from printing the results that the percent variation explained is 88.4%.

### 2014-12-02 Figure 6/abxD01.lmModel.compare.R

@@ -13,22 +13,22 @@ require(MASS)
 
 
 
-topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.15otus.rfnegpos.csv", header=T)
+topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.19otus.rfnegpos.csv", header=T)
 td<-topdose[,-1]
 ids<-names(td)
 ids = ids[-1]
 
 numOTU=length(td)-1
 
-results = data.frame(matrix(0, ncol=numOTU+5)) #table as follows: model# AIC numVar intercept OTU1B OTU2B ... etc all included
-colnames(results)[1:5] = c("Modelnum", "AIC", "rsq", "vars", "intCoeff")
-colnames(results)[6:(numOTU+5)] = ids
+results = data.frame(matrix(0, ncol=numOTU+6)) #table as follows: model# AIC numVar intercept OTU1B OTU2B ... etc all included
+colnames(results)[1:6] = c("Modelnum", "AIC", "BIC", "rsq", "vars", "intCoeff")
+colnames(results)[7:(numOTU+6)] = ids
 #make results headings
 
 #results = NULL
 #results = data.frame(NULL)
 
-
+######################################################
 #all combos for a model with 1 variable
 for(i in 1:numOTU){ 
   m1 = NULL
@@ -48,20 +48,22 @@ if(is.null(err)) {
   results[i, 1] = i
   results[i, 2] = c("NA")
   results[i, 3] = c("NA")
-  results[i, 4] = c(1)
-  results[i, 5] = names(td)[(i+1)] #add more names as in formula
+  results[i, 4] = c("NA")
+  results[i, 5] = c(1)
+  results[i, 6] = names(td)[(i+1)] #add more names as in formula
 }
 else{
   results[i, 1] = i #model number
   results[i, 2] = AIC(m1) #AIC 
+  results[i, 3] = BIC(m1) #BIC 
   coeff=m1$coefficients
   names(coeff)[2] = names(td)[i+1]
   vars = length(coeff)-1
-  results[i, 3] = summary(m1)$r.squared
-  results[i, 4] = vars #number of variables in model
-  results[i, 5] = coeff[1] #intercept coefficient
+  results[i, 4] = summary(m1)$r.squared
[…]

### 2014-12-02 abxD01.r

@@ -1600,6 +1600,8 @@ print(model_logtrans)
 
 #what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
 varImpPlot(model_logtrans, type=1)
+imp<-importance(model_logtrans)
+most.imp<-rownames(imp[order(imp[, "MeanDecreaseAccuracy"], decreasing=T)[1:5],])
 
 #predict the outcome of the testing data
 predicted_logtrans <- predict(model_logtrans, newdata=testing_logtrans[ ,-1])

### 2014-12-03 abxD01_analysis.Rmd
used leaps r pkg to determine top linear models

@@
 
-The script for running this code is called "abxD01.lmModel.compare.R". The input files for this script are:
+The script for running this code is called "abxD01.lmModel.compare2.R". The input files for this script are:
 1. abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.15otus.rfnegpos.csv
 
-I realize I could've done the modelling comparison script in a loop and shortened the code length significantly, but deal with it.
+I found a program "leaps" in r which will calculate all subsets of a model for a given range of parameters in the model.
+That is what is used in the above .R file.
+The methods I used were an exhaustive search and sequential replacement, which is a combination of both forward and backward stepwise directions for model building.
+Both of these methods produced similar results.
+Interestingly, the top models in both methods included 5 common OTUs: 6, 7, 20, 39, and 283 corresponding to a porphyromonadaceae, porphyromonadaceae, alistipes, lachnospiraceae, and blautia OTU respectively.
+The blautia OTU is the only one in the group positively (though low and not significantly) correlated with *C.
+difficile* CFU.
+There are also several OTUs from the candidate list that did not appear in the top models.
 

### 2014-12-03 Figure 6/abxD01.lmModel.compare2.R

@@ -1,0 +1,28 @@
+library(leaps)
+
+topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.19otus.rfnegpos.csv", header=T)
+td<-topdose[,-1]
+attach(td)
+ids<-names(td)
+ids = ids[-1]
+
+#this will make 50 models--the 5 best for each model with 1 to 10 variables
+leaps<-regsubsets(nextDayCFU ~ Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00283 + Otu00431, data=td, nbest=3, nvmax=10)
+#shows the best models and which variables to include
+summary(leaps)
+#shows the value for each parameter in the model along with intercept
+vcov(leaps, 50) ##the second parameter corresponds to model number, so 50 is the max/last model
+#coef
+coef(leaps, 1:2)
+#scales=bic, Cp, and adjr2, r2
+plot(leaps,scale="r2")
+plot(leapsf,scale="adjr2")
+plot(leaps,scale="bic")
+
+leapsf<-regsubsets(nextDayCFU ~ Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00283 + Otu00431, data=td, nbest=3, nvmax=10, method="forward")
+leapsb<-regsubsets(nextDayCFU ~ Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00283 + Otu00431, data=td, nbest=3, nvmax=10, method="backward")
[…]

### 2014-12-08 abxD01_analysis.Rmd
fixed heat map figure, added ns

@@
 
-The following files can be found through the mothur wiki: 
+The following files can be found through the [mothur wiki](http://www.mothur.org/wiki/MiSeq_SOP): 
 silva.bacteria.fasta
@@
 
-The following files can be found on the github repository for this notebook:
+The following files can be found on the [github repository for this notebook](https://github.com/SchlossLab/abxD01):
 abxD01.files
@@
 
-#par(mar=c(4, 6, 3, 4.5) +0.1) #default is 5.1 4.1 4.1 2.1
-
 side<-scan("~/Desktop/mothur/abxD01/correlation/correl_heatmapSIDE_0.01rel_topdose2_newtitr.csv", what=",", sep=",")
@@
 side3<-as.matrix(side2)
-#lmat=rbind(c(4,3), c(2,1))
-#lmat=rbind(c(4,3), c(1,2))
 
-#lmat order: 1)row dendrogram, 2) heatmap, 3)col?, 4)key?--key disappeared
-lmat = rbind(3:4, 1:2)
-lmat = rbind(c(2,4), c(1,3))
-lmat = rbind(c(0,3), c(2,1))
-lmat=NULL
-lwid=NULL
-lhei=NULL
-lwid = c(.5, 2, 4)
-lhei = c(.7, 4)
+sig <- read.csv("~/Desktop/mothur/abxD01/correlation/correl_heatmap_0.01rel_topdose2_newtitr_significance.csv",header=T,stringsAsFactors=F)
+sig<-sig[,-1]
 
-correl_heatmap<-heatmap.2(correl_matrix, scale="none",  col=my_palette, breaks=breaks, density.info="none", lhei=lhei,lwid=lwid, RowSideColors=side3[,2], labRow=side3[,1], trace="none",margins=c(8, 25), dendrogram="none",  Rowv=FALSE, Colv=FALSE,  na.color="black", cexCol=1.5,  key=TRUE, keysize=1, cexRow=1) #then plot was 7x8in, portrait
+par(mar=c(5, 4, 4, 2) +0.1) #default is 5.1 4.1 4.1 2.1
+correl_heatmap<-heatmap.2(correl_matrix, scale="none",  col=my_palette, breaks=breaks, density.info="none",  RowSideColors=side3[,2], labRow=side3[,1], trace="none",margins=c(8, 25), dendrogram="none",  Rowv=FALSE, Colv=FALSE,  na.color="black", cexCol=.9,  key=TRUE, keysize=1, cexRow=1, cellnote=as.matrix(sig), notecol="dark gray", notecex = 0.5) #then plot was 7x17in, portrait
 

### 2014-12-11 abxD01.barcharts.xOTU.test.r
minor changes

@@ -8,7 +8,7 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
 fileIDS<-read.csv("~/Desktop/mothur/abxD01/barcharts/allvanctitr_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
 abx<-c("Untreated", "Vanc 0.625", "Vanc 0.3", "Vanc 0.1")

### 2014-12-18 Figure 6/abxD01.optimizingModel.Rmd
worked on model optimization
pats new formula for r2, new model optimization markdown file under
figure 6 including html output

---
title: "Model Optimization"
output: html_document
---

# Researching the best OTUs for improved predictions
First I was curious what a random forest model would look like trained on the delay data. Mostly I was interested in the OTUs determined by randomforest's feature selection algorithm that overlapped with the current working 5 OTU model.

``` {r, eval=TRUE}
#################################
###rf of the delay data
delay<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.delay.regression.logtrans.filter16mintot.csv", header=T)
delay<-delay[,-1]

library(randomForest)

#fit the randomforest model
delay_model <- randomForest(nextDayCFU~., 
                               data = delay,  outscale=TRUE,
                               importance=TRUE, proximity=TRUE,
                               keep.forest=TRUE, ntree=5000
)
plot(delay_model)

print(delay_model)

#what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
varImpPlot(delay_model, type=1)
imp<-importance(delay_model)
most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
```

5000 trees looks apropriate (flatlined) when looking at the plot(delay_model) graph. The % Var explained was 74.31. Not too shabby! (Just as a note, I'm reporting the original results I got, so the numbers will be slightly different, as well as some differences in the importance plot.)

The OTUs that overlapped included 6 and 39, which were in the 5 OTU model. OTU11 has most influence, followed by 1, 5, 23, 21, 3, 6, 39 before dropping off in %IncMSE.  I'm curious to see what inclusion of OTU11 (Ecoli) will have on the effect of the model. I like the idea of including that because it's the most positively correlated OTU with *C. difficile* CFU the next day.

The next thing I was interested in was seeing which OTUs rf identified when building a model based on the titration and original data sets:

``` {r, eval=TRUE}
#################################
###rf of titration + topdose data
toptit<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.toptit.regression.logtrans.filter16mintot.csv", header=T)
toptit<-toptit[,-1]

library(randomForest)

#fit the randomforest model
toptit_model <- randomForest(nextDayCFU~., 
                            data = toptit,  outscale=TRUE,
                            importance=TRUE, proximity=TRUE,
                            keep.forest=TRUE, ntree=5000
)
plot(toptit_model)

print(toptit_model)

#what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
varImpPlot(toptit_model, type=1)
imp<-importance(toptit_model)
most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])

```

% Var explained goes up to 90%! The OTUs, or really the ONE OTU that far and away was important was OTU3, followed by 13, and 27.

# Revising and Expanding on the 5 OTU model

Testing the inclusion of either OTU 11 or OTU 3 or both and how it affects the other 5 OTUs.
```{r, eval=TRUE}
###Reperform the model building forcing certain OTUs to be present, such as OTU11 and OTU3.

library(leaps)

topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.19otus.rfnegpos.csv", header=T)
td<-topdose[,-1]
attach(td)
ids<-names(td)
ids = ids[-1]

#this will make 30 models--the 3 best for each model with 1 to 10 variables
leap11<-regsubsets(nextDayCFU ~ Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00283 + Otu00431, data=td, nbest=3, nvmax=10, force.in=c(5)) #the force.in parameter takes 2=OTU3, 5=OTU11
#shows the best models and which variables to include
summary(leap11)
#shows the value for each parameter in the model along with intercept
vcov(leap11, 50) ##the second parameter corresponds to model number, so 50 is the max/last model
#coef
coef(leap11, 1:2)
#scales=bic, Cp, and adjr2, r2
plot(leap11,scale="adjr2")
plot(leap11,scale="bic")
```

After looking at the results I don't think it affects the other 5 OTUs in the model at all. Though I haven't tested for improved predictions yet!

I next want to read in a topdose/titration combo file and see how the top linear models change or not for the combined data. Do the 5 OTUs stay the same in terms of importance?

```{r, eval=TRUE}
###read in 
toptit<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.toptit.logtrans.19otus.rfnegpos.csv", header=T)
tt<-toptit[,-1]
attach(tt)
ids<-names(tt)
ids = ids[-1]

#this will make 30 models--the 3 best for each model with 1 to 10 variables
leap11<-regsubsets(nextDayCFU ~ Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00283 + Otu00431, data=tt, nbest=3, nvmax=10, force.in=NULL) #the force.in parameter takes 2=OTU3, 5=OTU11
#shows the best models and which variables to include
summary(leap11)
#shows the value for each parameter in the model along with intercept
vcov(leap11, 50) ##the second parameter corresponds to model number, so 50 is the max/last model
#coef
coef(leap11, 1:2)
#scales=bic, Cp, and adjr2, r2
plot(leap11,scale="adjr2")
plot(leap11,scale="bic")
```

### 2014-12-18 Figure 6/abxD01.lmModel.compare2.R

@@ -120,11 +120,31 @@ SSres = sum((actual.2-predictlm5.2)^2)
 rsq = 1-(SSres/SStot)
 rsq
 
+
+#use this! ALYXXXXXX
+#ybar <- apply(res, 2, mean)
+#num<-sum((res$actual-ybar["actual"])*(res$predict-ybar["predict"]))
+#denA <- sum((res$actual-ybar["actual"])^2)
+#denB <- sum((res$predict-ybar["predict"])^2)
+
+
+# ybar = colMeans(actual)[1]
+# SStot = sum((actual-ybar)^2)
+# SSres = sum((actual-predictlm3delay)^2)
+# rsq = 1-(SSres/SStot)
+# rsq
+
+
 #adjusted r^2
 numer <- ((1-rsq)*(n-1))
 denom <- (n-p-1)
 adjr2 <- (1 - numer/denom)
 adjr2
+
+
+
+
+
 
 res<-cbind(actual.2, predictlm5.2)
 names(res) <- c( "actual", "predict")

### 2015-01-22 abxD01_analysis.Rmd
minor changes
worked on model optimization
pats new formula for r2, new model optimization markdown file under
figure 6 including html output
Figure 1 associated stats

@@
 
-## Figure 1
+## Figure 1: Bargraph
 
@@
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
-fileIDS<-read.csv("~/Desktop/mothur/abxD01/barcharts/topdose_tx1_barchart_ids.csv", header=T)
+file<-read.csv("/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
+fileIDS<-read.csv("/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
@@
 File saved as "topdose2_tx1_sorted_10x20.pdf"
+
+### Figure 1: Statistics
+
+To compare CFU levels: 
+```{r, eval=TRUE}
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
+
+data <- file[ which(file$expgroup != '1untr'), ] #choose experimental groups
+data$nextDayCFU<-log10(data$nextDayCFU+1) #log transform CFUs
+anova(lm(data$nextDayCFU ~ data$expgroup)) #perform anova across experimental groups for CFU
+
+#Which groups are different? Ciprofloxacin was clearly lower
+data <- data[ which(data$expgroup != '2cipro'), ] #remove ciprofloxacin to see now if there's a difference in remaining groups
+anova(lm(data$nextDayCFU ~ data$expgroup))
+
+```
+Alternatively:
+```{r, eval=TRUE}
+data<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
+
+data <- file[ which(file$expgroup != '1untr'), ] #choose experimental groups
+data$nextDayCFU<-log10(data$nextDayCFU+1) #log transform CFUs
+CFU.anova <- aov(nextDayCFU ~ expgroup, data = data)
+CFU.tukey <- TukeyHSD(CFU.anova)
+
+library(multcomp)
+library(multcompView)
+multcompLetters4(CFU.anova, CFU.tukey)
+```
+The CFU levels were significantly different between the ciprofloxacin treated communities and all other treatments.
+
+Amova calculation in mothur to determine the uniqueness of each community structure:
+```
+dist.shared(shared=abxD01.final.an.unique_list.0.03.topdose2.shared, calc=thetayc-jclass, subsample=1625)
+amova(phylip=abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.dist, design=topdose_design_byABX.txt, iters=1000000)
+
+```
+This results in the file called "abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.amova" located in the github folder. The results of these are summarized in "amova8expgroups.xlsx".
 

### 2015-01-22 abxD01.barcharts.xOTU.test.r

@@ -8,10 +8,10 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 3/allstreptitr_tx2_barchart_ids.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
-abx<-c("Untreated", "Cef 0.5", "Cef 0.3", "Cef 0.1")
+abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
 
 # If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
 # The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
@@ -19,7 +19,7 @@ sortbyphyl<-TRUE
 
 # If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
 # If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
-graphbyphyl<-TRUE
+graphbyphyl<-FALSE
 
 
 # Highlight all and run!

### 2015-01-22 abxD01_analysis.Rmd
fixed min a bun in fig 1 graph, added alternative

@@
 
-This figure was built using the following R code.  The input files are listed in Github.
-
-```{r, eval=FALSE}
-#################################################
-#Purpose: Convert csv matrix of columns you want converted into barcharts
-#input: csv file following this heading format format:  sampleID, expgroup, nextDayCFU, Otu001, Otu002, Otu003, etc
-#ex:file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.5.subsample.relabund.topdose.forlogscale2.csv
-#Note: It is assumed that anything after the expgroup column are vectors containing data to be split by the groups in 'expgroup' 
-#      and averaged with SD measured. 
-#Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
-#################################################
-# Parameters to change:
-# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
-fileIDS<-read.csv("/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", header=T)
-# Y Labels for each graph: 
-abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
-
-# If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
-# The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
-sortbyphyl<-TRUE
-
-# Highlight all and run!
-#################################################
-
-file<-file[,-1] #delete the first col of group names
-avgs=NULL
-avgs <-data.frame(levels(file$expgroup))
-colnames(avgs) <- c("expgroup")
-stds=NULL
-stds <-data.frame(levels(file$expgroup))
-colnames(stds) <- c("expgroup")
-
-#Calculate average cdiff
-avgCD=NULL
-sdCD=NULL
-avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
-avgCD<-format(avgCD, scientific=TRUE, digits=2)
-sdCD<-tapply(file$nextDayCFU, file$expgroup, sd)
-file<-file[,-2] #delete the nexDayCFU column
-
-for(i in 1:ncol(file)){
-
-  if(i==1){
-    ids<-names(file)
-  }
-  else{
-    columni <- data.frame(file[,i])
-    colnames(columni) <- c("columndata")
-
-    #calculates mean for each column in file by expgroup and stores it in avgs data frame
-    poop<- aggregate(columni$columndata~file$expgroup, FUN=mean) #poop stores the aggregated means for each group for this specific column
-    colnames(poop) <- c(ids[1], ids[i])
-
-    avgs<- merge(avgs, poop, by ="expgroup")
-
-    #calculates st dev for each column in file by expgroup and stores it in stds data frame
-    poop2<- aggregate(columni$columndata~file$expgroup, FUN=sd) #poop stores the aggregated means for each group for this specific column
-    colnames(poop2) <- c(ids[1], ids[i])
-
-    stds<- merge(stds, poop2, by ="expgroup")
-  }
-}
-
-ordIDS <- fileIDS[order(fileIDS[,2]), ] #sort by the phyla
-totalOTU <- dim(ordIDS)[1]
-
-########for doing sort by phylum then within phylum (by relabund)
-if(sortbyphyl == TRUE){
-  ordIDS<-cbind(ordIDS, 0)
-
-  for (i in 1:totalOTU){ #loop to fill out the final column for the phyla groups, while phyla is sorted!!
-    if(i==1){
-      ordIDS[i, 4] <- 1
-    }
-    else{
-      if(ordIDS[i, 2]==ordIDS[i-1, 2]){
-        ordIDS[i, 4] <- ordIDS[i-1, 4]
-      }
-      else{
-        ordIDS[i, 4] <- ordIDS[i-1, 4] + 1
-      }
-
-    }
-  }
-
-  names(ordIDS)[4]<-paste("phynum")
-
-  ##add row for phylum number to the avgs?
-  #avgs<-rbind(avgs, 0)[c(9, 1, 2, 3, 4, 5, 6, 7, 8),] #this is hard coded for the top doses... also dont rerun this multiple times!!! might need to change
-  avgs<-rbind(0,avgs) #this is hard coded for the top doses... also dont rerun this multiple times!!! might need to change, will give error
-
-  numphyla<-max(ordIDS[,4])
-
-  #test_avg <- avgs[, rev(order(avgs[1,2:length(avgs)-1]))] #sort all but other column (last)
-
-  physums <- data.frame(c(1:numphyla))
-  physums <- cbind(physums, 0)
-  colnames(physums) <- c("Num", "Sum")
-
-  leng<-length(avgs)
-
-  #jtime<-NULL
-  #itime<-NULL
-  #inside<-FALSE
-  #count<-0
-  j<-1
-  ##This loop is to fill the physums data frame with the sum of the relabund for each phylum represented
-  for(i in 2:(leng-1)){ #searching through avgs, so leng is number of otus, -1 because i dont want to deal with "Other" column
-    found<-FALSE
-    #itime<-rbind(itime, i)
-    j<-1
-    while(found==FALSE){ #searching through ordIDS
-      test<-ordIDS[j, 1] == names(avgs)[i]
-      if(test) { 
-        pnum<-ordIDS[j, 4]
-        found<-TRUE
-        inside<-TRUE
-        avgs[1,i]<-pnum
-      }
-      else{
-        j <- j + 1
-        #count<-count+1
-        #jtime<-rbind(jtime, j)
-
-      }
-    } #WARNING: could cause an error if dont find the ID, theres no check for the end of the list
-
-    physums[pnum, 2] <- physums[pnum, 2] + avgs[2, i] ##add the relabund of the otu   
-  }
-
-  y<-rev(order(physums$Sum))
-  x<-avgs[1, 2:(leng-1)]
-
-  ord_physums<-physums[rev(order(physums[,2])),]
-  sort_avgs<-data.frame(avgs$expgroup)
-  names(sort_avgs)[1]<-c("expgroup")
-
-  for(i in 1:numphyla){ #loop is to sort by decreasing phyla and perform internal sort within phylum, return the new sorted avgs to go into mavgs
-    val<-NULL
-    phy<-which(x %in% c(ord_physums[i,1])) #phy returns the indices for where that phyNum is
-    for(j in 1:length(phy)){ #using locations of phylum i to find in avgs
-      val[j]<-avgs[2,phy[j]+1]
-    }
-    phy<-phy[rev(order(val))]
-
-    for(j in 1:length(phy)){ #now loop through the sorted sub cols and add them to the sort_avgs
-      sort_avgs<-cbind(sort_avgs, avgs[,phy[j]+1])
-      names(sort_avgs)[length(sort_avgs)]<-names(avgs)[phy[j]+1]
-    }
-  }
-
-  attach(avgs)
-  sort_avgs<-cbind(sort_avgs, Other)
-  detach(avgs)
-
-  sort_avgs <- sort_avgs[-1,] #get rid of phylum first row from earlier sorting
-
-  #loop for getting the right ids in the right order as sort
-  ids<-data.frame(1:(totalOTU))
-
-  row.names(sort_avgs)<-sort_avgs$expgroup
-  sort_avgs<-sort_avgs[,-1]
-
-  for (i in 1:(length(sort_avgs)-1)){
-    found<-FALSE
-    #itime<-rbind(itime, i)
-    j<-1
-    while(found==FALSE){ #searching through ordIDS
-      test<-ordIDS[j, 1] == names(sort_avgs)[i]
-      if(test) { 
-        ids[i,2]<-ordIDS[j, 3]
-        found<-TRUE
-      }
-      else{
-        j <- j + 1
-        #count<-count+1
-        #jtime<-rbind(jtime, j)
-      }
-    }
-
-  }
-
-  names(ids)<-c("x", "name")
-  ids<-rbind(ids, data.frame(x=totalOTU+1, name= "Other"))
-
-  mavgs<-as.matrix(sort_avgs)
-
-  leng<-dim(mavgs)[2]
-
-  numgr <- nlevels(file$expgroup)
-} ##end if(sortbyphyl == TRUE){
-
-####order the results by the first group alpha numerically
-if(sortbyphyl == FALSE){
-
-  row.names(avgs)<-avgs$expgroup
-  avgs<-avgs[,-1]
-  ordered_avgs<- avgs[, rev(order(avgs[1,1:length(avgs)-1]))] #sort all but other column (last)
-  attach(avgs)
-  ordered_avgs <- cbind(ordered_avgs, Other) #put other back on
-  detach(avgs)
-
-  mavgs<-as.matrix(ordered_avgs)
-  leng<-dim(mavgs)[2]
-
-  numgr <- nlevels(file$expgroup)
-
-  ids<-data.frame(1:(totalOTU))
-
-  for (i in 1:(length(ordered_avgs)-1)){
-    found<-FALSE
-    #itime<-rbind(itime, i)
-    j<-1
-    while(found==FALSE){ #searching through ordIDS
-      test<-ordIDS[j, 1] == names(ordered_avgs)[i]
-      if(test) { 
-        ids[i,2]<-ordIDS[j, 3]
-        found<-TRUE
-      }
-      else{
-        j <- j + 1
-        #count<-count+1
-        #jtime<-rbind(jtime, j)
-      }
-    }
-
-  }
-
-  names(ids)<-c("x", "name")
-  ids<-rbind(ids, data.frame(x=totalOTU+1, name= "Other"))
-
-}  #end if(sortbyphyl == FALSE){
-
-###started to do the stds table... didnt complete
-#stds<-stds[,-1]
-#mstds<-as.matrix(stds)
-
-###PLOT PARAMETERS
-par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
-par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
-
-for(j in 1:numgr){
-
-  if(j != numgr){
-    barplot(mavgs[j, 1:leng], ylab=NULL, col="black", yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
-    #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
-    # k <- k+1
-    axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
-    mtext(abx[j], side=2, line=6, cex=.8)
-    mtext(avgCD[j], side=2, line=4.5, cex=.8)
-    abline(h=c(0.0001, 1), lwd=3)
-    abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
-    abline(h=c(.25,.5, .75), col="black", lty="dashed")
-  }
-
-  else{ #the last graph needs different margins    
-    label<-barplot(mavgs[j, 1:leng], col="black",ylab=NULL,  yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
-    #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
-    axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
-    mtext(abx[j], side=2, line=6, cex=.8)
-    mtext(avgCD[j], side=2, line=4.5, cex=.8)
-    axis(1, at=(label[,1]), labels=FALSE)
-    text(label[,1]+.13, .00005, label=ids[,2], xpd=NA, pos=2, srt=45, cex=1.2)
-    #text(-3.9,.0001, label=expression(paste(italic("C.d."), " CFU/g Feces:")), xpd=NA, pos=2, srt=90, cex=1.1)
-    abline(h=c(0.0001,1), lwd=3)
-    abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
-    abline(h=c(.25,.5, .75), col="black", lty="dashed")
-
-  }
-
-}
-par(mfrow=c(1, 1))
-
-##############END CODE###########################
-#################################################
-```
-
+This figure was built using the R script called "abxD01.barcharts.xOTU.test.r" in the abxD01 folder.The input files are listed in Github.
 The graph can be sorted 2 ways either by the untreated's decreasing relabund, or by the most abundant within each phylum.
@@
 difficile* CFU/g Feces is also labeled on the Y axis below the treatment name.
-File saved as "topdose2_tx1_sorted_10x20.pdf"
+File saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf"."
 
@@
 
-### Figure 2 Correlation Calculations
+### Figure 2: Correlation Calculations
 
@@
 
-### Figure 2 Graph
+### Figure 2: Stipchart
 

### 2015-01-22 abxD01.barcharts.xOTU.test.r

@@ -8,8 +8,8 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
 abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
 
@@ -265,7 +265,7 @@ if(sortbyphyl == FALSE){
 
 
 
-
+#######################################
 ###PLOT PARAMETERS
 if(graphbyphyl==FALSE){
   par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
@@ -295,7 +295,7 @@ if(graphbyphyl==FALSE){
       text(label[,1]+.13, .0005, label=ids[,2], xpd=NA, pos=2, srt=45, cex=1.2)
       #text(-3.9,.0001, label=expression(paste(italic("C.d."), " CFU/g Feces:")), xpd=NA, pos=2, srt=90, cex=1.1)
       abline(h=c(0.001,1), lwd=3) #min/max
-      abline(h=c(0.01, 0.01), col="black", lty="longdash", lwd=2)
+      abline(h=c(0.01, 0.1), col="black", lty="longdash", lwd=2)
       abline(h=c(0.005, 0.05, 0.25, 0.5, 0.75), col="black", lty="dashed") 
     }  
     

### 2015-01-25 abxD01_analysis.Rmd
fixed figure 2

@@
 difficile* CFU/g Feces is also labeled on the Y axis below the treatment name.
+The parameters I changed in the r script were:
+sortbyphyl<-TRUE
+graphbyphyl<-FALSE
+abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
+
 File saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf"."
@@
 
-meta <- read.table('~/Desktop/mothur/abxD01/correlation/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
+meta <- read.table('~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
 #meta<-meta[1:96] #change based on number of OTUs above .05%, then add 2 for first two cols
@@
 results <- results[order(results[,3]),]  #order by pvalue column=3
-write.table(results[1:dim(results)[1],], file="~/Desktop/mothur/abxD01/correlation/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
+write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
 
@@
 ##Paper Figure: correlations for topdose 
-c<-read.csv("~/Desktop/mothur/abxD01/correlation/corr_allSig_topdose2_stripchart.csv", header=T)
-labels=c("Lachnospiraceae", "Ruminococcaceae", "Clostridia", "Lactobacillales", "Firmicutes", "Bacillales", "Porphyromonadaceae", "Bacteroidales", "Bacteroidetes", "Actinobacteria", "Proteobacteria", "Anaeroplasma", "Deinococcus", "Unclassified")
-ns=c("n=18", "n=9", "n=8", "n=4", "n=2", "n=1", "n=15", "n=4", "n=1", "n=4", "n=3", "n=1", "n=1", "n=3")
+otuCorrs<-read.csv("~/Documents/Github/abxD01/Figure 2/corr_avgAbove0.001_topdose2_stripchart.csv", header=T)
+#labels=c("Lachnospiraceae", "Ruminococcaceae", "Clostridia", "Lactobacillales", "Firmicutes", "Bacillales", "Porphyromonadaceae", "Bacteroidales", "Bacteroidetes", "Actinobacteria", "Proteobacteria", "Anaeroplasma", "Deinococcus", "Unclassified")
+#ns=c("n=18", "n=9", "n=8", "n=4", "n=2", "n=1", "n=15", "n=4", "n=1", "n=4", "n=3", "n=1", "n=1", "n=3")
 
-par(mar=c(12, 7, 2, 2) +0.1, mgp=c(5, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
-stripchart(c$cor~c$order, vertical=TRUE, ylab="Correlation", ylim=c(.85, -.85), cex.axis=1.1, cex=2, pch=21, lwd=3, col="black", xaxt="n", method="jitter",  cex.lab=1.7, yaxt="n")
-axis(1, cex.axis=1, at=(1:nlevels(c$name)), labels=FALSE)
-text(1:nlevels(c$name)+.1, par("usr")[3]-0.16, label=labels, xpd=NA, pos=2, srt=45, cex=1.2)
-text(1:nlevels(c$name), par("usr")[3]-0.01, label=ns, xpd=NA, pos=1, cex=.9)
-axis(2, cex.axis=1.1, at=seq(-0.85,+0.85, by=.1), las=1)
-abline(v=c(1:nlevels(c$name)), col="dark gray")
+#test<-otuCorrs[otuCorrs$phyl=="Bacteroidetes",]
+#length(unique(test$family))
+
+pch <- c(1,1,1,1, 6,6, 6, 6, 6, 6, 6, 6,0, 0, 9, 2, 2, 4)
+par(mar=c(12, 7, 2, 2) +0.1, mgp=c(4, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
+stripchart(otuCorrs$corr ~ otuCorrs$graphOrder, vertical=TRUE, ylab="Spearman Correlation", ylim=c(-1, 1), cex.axis=1.1, cex=2, pch=pch, lwd=2, col="black",  method="jitter",  jitter=0.25, cex.lab=1.3, yaxt="n", xaxt="n")
+axis(1, cex.axis=1, at=(1:nlevels(otuCorrs$family)), labels=FALSE)
+text(1:nlevels(otuCorrs$family)+.1, par("usr")[3]-0.05, label=unique(otuCorrs$family), xpd=NA, pos=2, srt=45, cex=1.2)
+#text(1:nlevels(otuCorrs$family), par("usr")[3]-0.01, label=ns, xpd=NA, pos=1, cex=.9)
+axis(2, cex.axis=1.1, at=seq(-1,+1, by=.1), las=1)
+abline(v=c(1:nlevels(otuCorrs$family)), col="dark gray")
 #abline(h=c(.3, -.3), col="dark gray", lty="dashed", lwd=4) #change to something else
 abline(h=0, col="black", lwd=2)
+abline(h=c(0.25, 0.5, 0.75, -0.25, -0.5, -0.75), col="light gray", lwd=2, lty="dashed")
+legend("top", legend=unique(otuCorrs$phyl), pch=c(1, 6, 0, 9, 2, 4), cex=0.7, bty="o", horiz=TRUE)
 
@@
 
-Figure saved as "corr_allSig_topdose2_10x8.pdf".
+Figure saved as "corr_avgAbove0.001_topdose2_10x8.pdf".
 
@@
 
-The code for this is still in the works. But is uploaded to github under Figure 3 folder.
+The code for this is the script "abxD01.barcharts.xOTU.test.r" in the paper's home Github directory. Opening this file, the parameters I changed in the top were:
+sortbyphyl<-TRUE
+graphbyphyl<-TRUE
+abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
+file
+fileIDS
 
+The file/fileIDS should be matched by the titration.
 Input files on github:
-1. abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv
-2. abxD01/barcharts/allvanctitr_tx2_barchart_ids.csv
+1. abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv
+2. abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv
+3. abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv
+4. allceftitr_tx2_barchart_ids.csv
+5. allstreptitr_tx2_barchart_ids.csv
+6. allvanctitr_tx2_barchart_ids.csv
+
+For this graph I decided to take out the untreated communities to highlight the differences among titration groups. So in r I changed the file a bit:
+
+```{r, eval=FALSE}
+file<-file[file$expgroup!="1untrStrep",]
+#alternatively for the other titrated antibiotics it's "1untrCef", or "1untrVanc"
+```
 

### 2015-01-25 abxD01.barcharts.xOTU.test.r

@@ -8,10 +8,10 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 3/allstreptitr_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
-abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
+abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
 
 # If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
 # The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
@@ -19,7 +19,7 @@ sortbyphyl<-TRUE
 
 # If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
 # If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
-graphbyphyl<-FALSE
+graphbyphyl<-TRUE
 
 
 # Highlight all and run!
@@ -209,7 +209,7 @@ if(sortbyphyl == TRUE){
   
   leng<-dim(mavgs)[2]
   
-  numgr <- nlevels(file$expgroup)
+  numgr <- length(unique(file$expgroup))
 } ##end if(sortbyphyl == TRUE){
 
 
@@ -226,7 +226,7 @@ if(sortbyphyl == FALSE){
   mavgs<-as.matrix(ordered_avgs)
   leng<-dim(mavgs)[2]
   
-  numgr <- nlevels(file$expgroup)
+  numgr <- length(unique(file$expgroup))
   
   
   ids<-data.frame(1:(totalOTU))

### 2015-01-26 abxD01_analysis.Rmd
figure 5 updates

@@
 
+### Figure 3: Barchart
+
 This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.
@@
 #alternatively for the other titrated antibiotics it's "1untrCef", or "1untrVanc"
+```
+
+### Figure 3: Statistics
+
+```{r, eval=FALSE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrStrep",]
+kruskal.test(file$Otu001 ~ file$expgroup)
+kruskal.test(file$Otu003 ~ file$expgroup)
+kruskal.test(file$Otu002 ~ file$expgroup)
+kruskal.test(file$Otu005 ~ file$expgroup)
+kruskal.test(file$Otu004 ~ file$expgroup)
+kruskal.test(file$Otu014 ~ file$expgroup)
+kruskal.test(file$Otu006 ~ file$expgroup)
+kruskal.test(file$Otu007 ~ file$expgroup)
+kruskal.test(file$Otu009 ~ file$expgroup)
+kruskal.test(file$Otu015 ~ file$expgroup)
+kruskal.test(file$Otu017 ~ file$expgroup)
+kruskal.test(file$Other ~ file$expgroup)
+
+test <- pairwise.wilcox.test(file$Otu001, file$expgroup, p.adj="BH")
+comparisons <- length(test$p.value[]) - sum(is.na(test$p.value))
+
+pairwise.wilcox.test(file$Otu003, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu005, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu002, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu004, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu014, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu006, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu007, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu009, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu015, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu017, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Other, file$expgroup, p.adj="BH")
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrVanc",]
+kruskal.test(file$Otu001 ~ file$expgroup)
+kruskal.test(file$Otu002 ~ file$expgroup)
+kruskal.test(file$Otu003 ~ file$expgroup)
+kruskal.test(file$Otu005 ~ file$expgroup)
+kruskal.test(file$Otu004 ~ file$expgroup)
+kruskal.test(file$Otu014 ~ file$expgroup)
+kruskal.test(file$Otu006 ~ file$expgroup)
+kruskal.test(file$Otu007 ~ file$expgroup)
+kruskal.test(file$Otu008 ~ file$expgroup)
+kruskal.test(file$Otu019 ~ file$expgroup)
+kruskal.test(file$Otu018 ~ file$expgroup)
+kruskal.test(file$Otu017 ~ file$expgroup)
+kruskal.test(file$Other ~ file$expgroup)
+
+pairwise.wilcox.test(file$Otu001, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu003, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu005, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu002, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu004, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu014, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu006, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu007, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu008, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu019, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu018, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu017, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Other, file$expgroup, p.adj="BH")
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrCef",]
+kruskal.test(file$Otu001 ~ file$expgroup)
+kruskal.test(file$Otu002 ~ file$expgroup)
+kruskal.test(file$Otu003 ~ file$expgroup)
+kruskal.test(file$Otu005 ~ file$expgroup)
+kruskal.test(file$Otu004 ~ file$expgroup)
+kruskal.test(file$Otu014 ~ file$expgroup)
+kruskal.test(file$Otu006 ~ file$expgroup)
+kruskal.test(file$Otu007 ~ file$expgroup)
+kruskal.test(file$Otu008 ~ file$expgroup)
+kruskal.test(file$Otu009 ~ file$expgroup)
+kruskal.test(file$Otu019 ~ file$expgroup)
+kruskal.test(file$Otu018 ~ file$expgroup)
+kruskal.test(file$Otu016 ~ file$expgroup)
+kruskal.test(file$Otu024 ~ file$expgroup)
+kruskal.test(file$Otu011 ~ file$expgroup)
+kruskal.test(file$Other ~ file$expgroup)
+
+pairwise.wilcox.test(file$Otu001, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu003, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu005, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu002, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu004, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu014, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu006, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu007, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu008, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu009, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu011, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu019, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu018, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu016, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Otu024, file$expgroup, p.adj="BH")
+pairwise.wilcox.test(file$Other, file$expgroup, p.adj="BH")
+
 ```
@@
 
+### Figure 5: Random Forest Analysis
+
+Performed RF analysis on family level phylotypes:
+```{r, eval=FALSE}
+topdose_tx2<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.2.shared.rf.topdose.regression.logtrans.csv", header=T)
+topdose_tx2<-topdose_tx2[,-1]
+
+library(randomForest)
+
+#fit the randomforest model
+model_tx2 <- randomForest(nextDayCFU~., 
+                      data = topdose_tx2,  outscale=TRUE,
+                      importance=TRUE, proximity=TRUE,
+                      keep.forest=TRUE, ntree=5000
+)
+plot(model_tx2)
+
+print(model_tx2)
+
+#what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
+par(mfrow=c(1, 1)) 
+varImpPlot(model_tx2, type=1)
+imp<-importance(model_tx2)
+most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
+```
+
+### Figure 5: Barchart
+
 This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole. The figure is shown as a pdf on github.
@@
 Input files found on github:
-1. abxD01.final.tx.2.subsample.2.pick.metro.relabund.d0s.txt
+1. abxD01.final.tx.2.subsample.2.metro.relabund.D0s.csv
 
@@
 #First calculating:
-b<-read.delim("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.2.subsample.2.pick.metro.relabund.d0s.txt", header=T)
+metroD0<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.2.metro.relabund.D0s.csv", header=T)
 
-avg1<-tapply(b$Otu001, b$order, mean)
-sd1<-tapply(b$Otu001, b$order, sd)
+avg01<-tapply(metroD0$Otu001, metroD0$order, mean)
+sd01<-tapply(metroD0$Otu001, metroD0$order, sd)
 
-avg11<-tapply(b$Otu011, b$order, mean)
-sd11<-tapply(b$Otu011, b$order, sd)
+avg14<-tapply(metroD0$Otu014, metroD0$order, mean)
+sd14<-tapply(metroD0$Otu014, metroD0$order, sd)
 
-avg02<-tapply(b$Otu002, b$order, mean)
-sd02<-tapply(b$Otu002, b$order, sd)
+avg7<-tapply(metroD0$Otu007, metroD0$order, mean)
+sd7<-tapply(metroD0$Otu007, metroD0$order, sd)
 
-avg03<-tapply(b$Otu003, b$order, mean)
-sd03<-tapply(b$Otu003, b$order, sd)
+avg09<-tapply(metroD0$Otu009, metroD0$order, mean)
+sd09<-tapply(metroD0$Otu009, metroD0$order, sd)
 
-avg7<-tapply(b$Otu007, b$order, mean)
-sd7<-tapply(b$Otu007, b$order, sd)
+avg02<-tapply(metroD0$Otu002, metroD0$order, mean)
+sd02<-tapply(metroD0$Otu002, metroD0$order, sd)
 
-avg05<-tapply(b$Otu005, b$order, mean)
-sd05<-tapply(b$Otu005, b$order, sd)
+avg13<-tapply(metroD0$Otu013, metroD0$order, mean)
+sd13<-tapply(metroD0$Otu013, metroD0$order, sd)
 
-avg16<-tapply(b$Otu016, b$order, mean)
-sd16<-tapply(b$Otu016, b$order, sd)
+avg03<-tapply(metroD0$Otu003, metroD0$order, mean)
+sd03<-tapply(metroD0$Otu003, metroD0$order, sd)
 
-avg06<-tapply(b$Otu006, b$order, mean)
-sd06<-tapply(b$Otu006, b$order, sd)
+avg05<-tapply(metroD0$Otu005, metroD0$order, mean)
+sd05<-tapply(metroD0$Otu005, metroD0$order, sd)
 
-avg04<-tapply(b$Otu004, b$order, mean)
-sd04<-tapply(b$Otu004, b$order, sd)
+avg08<-tapply(metroD0$Otu008, metroD0$order, mean)
+sd08<-tapply(metroD0$Otu008, metroD0$order, sd)
+
+avg06<-tapply(metroD0$Otu006, metroD0$order, mean)
+sd06<-tapply(metroD0$Otu006, metroD0$order, sd)
+
+avg04<-tapply(metroD0$Otu004, metroD0$order, mean)
+sd04<-tapply(metroD0$Otu004, metroD0$order, sd)
+
+avg12<-tapply(metroD0$Otu012, metroD0$order, mean)
+sd12<-tapply(metroD0$Otu012, metroD0$order, sd)
+
+# avg11<-tapply(metroD0$Otu011, metroD0$order, mean)
+# sd11<-tapply(metroD0$Otu011, metroD0$order, sd)
+# 
+# avg05<-tapply(metroD0$Otu005, metroD0$order, mean)
+# sd05<-tapply(metroD0$Otu005, metroD0$order, sd)
+# 
+# avg16<-tapply(metroD0$Otu016, metroD0$order, mean)
+# sd16<-tapply(metroD0$Otu016, metroD0$order, sd)
 
@@
 par(mar=c(2, 4, 2, 3) +0.1)
-plot1<-barplot(avg1, col="darkblue", cex.axis=1.3, xaxt="n") 
-error.bar(plot1, avg1, sd1) #plot SD
+plot01<-barplot(avg01, col="darkblue", cex.axis=1.3, xaxt="n") 
+error.bar(plot01, avg01, sd01) #plot SD
 

### 2015-01-26 Figure 6/abxD01.lmModel.compare.R

@@ -1,1116 +1,0 @@
-require(foreign)
-require(ggplot2)
-require(MASS)
-
-
-## linear model
-#######################################
-## linear model
-## File headings: SampleName nextDayCFU OTU1 OTU2 ... total OTU candidates being tested, candidate number should be at least 10
-## nextDayCFU should be log transformed
-## IN file type is *.csv
-
-
-
-
-topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.19otus.rfnegpos.csv", header=T)
-td<-topdose[,-1]
-ids<-names(td)
-ids = ids[-1]
-
-numOTU=length(td)-1
-
-results = data.frame(matrix(0, ncol=numOTU+6)) #table as follows: model# AIC numVar intercept OTU1B OTU2B ... etc all included
-colnames(results)[1:6] = c("Modelnum", "AIC", "BIC", "rsq", "vars", "intCoeff")
-colnames(results)[7:(numOTU+6)] = ids
-#make results headings
-
-#results = NULL
-#results = data.frame(NULL)
-
-######################################################
-#all combos for a model with 1 variable
-for(i in 1:numOTU){ 
-  m1 = NULL
-  err = NULL
-  err<-tryCatch(
-{
-  m1<-lm(td[,1] ~ td[,(i+1)])
-},
-error = function(cond) {
-  message(cond)
-  return(NULL)     
-},
-finally = {}
-  )
-
-if(is.null(err)) {
-  results[i, 1] = i
-  results[i, 2] = c("NA")
-  results[i, 3] = c("NA")
-  results[i, 4] = c("NA")
-  results[i, 5] = c(1)
-  results[i, 6] = names(td)[(i+1)] #add more names as in formula
-}
-else{
-  results[i, 1] = i #model number
-  results[i, 2] = AIC(m1) #AIC 
-  results[i, 3] = BIC(m1) #BIC 
-  coeff=m1$coefficients
-  names(coeff)[2] = names(td)[i+1]
-  vars = length(coeff)-1
-  results[i, 4] = summary(m1)$r.squared
-  results[i, 5] = vars #number of variables in model
-  results[i, 6] = coeff[1] #intercept coefficient
-  
-  for(j in 7:(numOTU+6)){  #fill results columns for each OTU whether NA or a coefficient value
-    for(k in 2:(vars+1)){
-      if(names(coeff)[k]==names(results)[j]){
[…]

### 2015-01-26 Figure 6/abxD01.lmModel.compare2.R

@@ -122,10 +122,33 @@ rsq
 
 
 #use this! ALYXXXXXX
-#ybar <- apply(res, 2, mean)
-#num<-sum((res$actual-ybar["actual"])*(res$predict-ybar["predict"]))
-#denA <- sum((res$actual-ybar["actual"])^2)
-#denB <- sum((res$predict-ybar["predict"])^2)
+ybar <- apply(res, 2, mean)
+num<-sum((res$actual-ybar["actual"])*(res$predict-ybar["predict"]))
+denA <- sum((res$actual-ybar["actual"])^2)
+denB <- sum((res$predict-ybar["predict"])^2)
+
+rsq <- 1-(num/(denA*denB))
+
+# 
+# 
+# #predict the outcome of the testing data
+# predicted_logtrans <- predict(model_logtrans, newdata=testing_logtrans[ ,-1])
+# predicted2_logtrans <- predict(model_logtrans, newdata=testing2_logtrans[ ,-1])
+# 
+# # what is the proportion variation explained in the outcome of the testing data?
+# # i.e., what is 1-(SSerror/SStotal)
+# actual_logtrans <- testing_logtrans$nextDayCFU
+# actual2_logtrans <- testing2_logtrans$nextDayCFU
+# 
+# ybar <- apply(res, 2, mean)
+# num<-sum((res$actual-ybar["actual"])*(res$predict-ybar["predict"]))
+# denA <- sum((res$actual-ybar["actual"])^2)
+# denB <- sum((res$predict-ybar["predict"])^2)
+# 
+# rsq <- 1-(num/(denA*denB))
+# 
+# print(rsq)
+# 
 
 
 # ybar = colMeans(actual)[1]
@@ -134,14 +157,14 @@ rsq
 # rsq = 1-(SSres/SStot)
 # rsq
 
-
-#adjusted r^2
-numer <- ((1-rsq)*(n-1))
-denom <- (n-p-1)
-adjr2 <- (1 - numer/denom)
-adjr2
-
-
+# 
+# #adjusted r^2
+# numer <- ((1-rsq)*(n-1))
+# denom <- (n-p-1)
+# adjr2 <- (1 - numer/denom)
+# adjr2
+# 
+# 
 
 
 

### 2015-01-26 abxD01.barcharts.xOTU.test.r

@@ -20,6 +20,7 @@ sortbyphyl<-TRUE
 # If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
 # If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
 graphbyphyl<-TRUE
+file<-file[file$expgroup!="1untrStrep",]
 
 
 # Highlight all and run!
@@ -270,6 +271,7 @@ if(sortbyphyl == FALSE){
 if(graphbyphyl==FALSE){
   par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
   par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+  color_transparent <- adjustcolor("black", alpha.f = 0.1) 
   
   for(j in 1:numgr){
     
@@ -281,8 +283,8 @@ if(graphbyphyl==FALSE){
       mtext(abx[j], side=2, line=6, cex=.8)
       mtext(avgCD[j], side=2, line=4.5, cex=.8)
       abline(h=c(0.001, 1), lwd=3) #min/max
-      abline(h=c(0.01, 0.1), col="black", lty="longdash", lwd=2)
-      abline(h=c(0.005, 0.05, 0.25, 0.5, 0.75), col="black", lty="dashed")
+      abline(h=c(0.01, 0.1), col=color_transparent, lty="longdash", lwd=2)
+      abline(h=c(0.005, 0.05, 0.25, 0.5, 0.75), col=color_transparent, lty="dashed")
     }
     
     else{ #the last graph needs different margins    
@@ -295,8 +297,8 @@ if(graphbyphyl==FALSE){
       text(label[,1]+.13, .0005, label=ids[,2], xpd=NA, pos=2, srt=45, cex=1.2)
       #text(-3.9,.0001, label=expression(paste(italic("C.d."), " CFU/g Feces:")), xpd=NA, pos=2, srt=90, cex=1.1)
       abline(h=c(0.001,1), lwd=3) #min/max
-      abline(h=c(0.01, 0.1), col="black", lty="longdash", lwd=2)
-      abline(h=c(0.005, 0.05, 0.25, 0.5, 0.75), col="black", lty="dashed") 
+      abline(h=c(0.01, 0.1), col=color_transparent, lty="longdash", lwd=2)
[…]

### 2015-01-26 abxD01_analysis.Rmd
figure update

@@
 
-avg14<-tapply(metroD0$Otu014, metroD0$order, mean)
-sd14<-tapply(metroD0$Otu014, metroD0$order, sd)
+# avg14<-tapply(metroD0$Otu014, metroD0$order, mean)
+# sd14<-tapply(metroD0$Otu014, metroD0$order, sd)
 
-avg7<-tapply(metroD0$Otu007, metroD0$order, mean)
-sd7<-tapply(metroD0$Otu007, metroD0$order, sd)
+avg07<-tapply(metroD0$Otu007, metroD0$order, mean)
+sd07<-tapply(metroD0$Otu007, metroD0$order, sd)
 
-avg09<-tapply(metroD0$Otu009, metroD0$order, mean)
-sd09<-tapply(metroD0$Otu009, metroD0$order, sd)
+# avg09<-tapply(metroD0$Otu009, metroD0$order, mean)
+# sd09<-tapply(metroD0$Otu009, metroD0$order, sd)
 
@@
 
-avg12<-tapply(metroD0$Otu012, metroD0$order, mean)
-sd12<-tapply(metroD0$Otu012, metroD0$order, sd)
+# avg12<-tapply(metroD0$Otu012, metroD0$order, mean)
+# sd12<-tapply(metroD0$Otu012, metroD0$order, sd)
 
@@
 #Now graphing:
-par(mfrow=c(4, 3))
+par(mfrow=c(3, 3))
 
 par(mar=c(2, 4, 2, 3) +0.1)
-plot01<-barplot(avg01, col="darkblue", cex.axis=1.3, xaxt="n") 
-error.bar(plot01, avg01, sd01) #plot SD
+plot01<-barplot(avg01, col="black", cex.axis=1.3, xaxt="n") 
+#error.bar(plot01, avg01, sd01) #plot SD
+axis(1, at=plot01, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+# par(mar=c(2, 4, 2, 3) +0.1)
+# plot14<-barplot(avg14, col="black", cex.axis=1.3, xaxt="n") 
+# #error.bar(plot14, avg14, sd14) #plot SD
+# axis(1, at=plot14, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 
 par(mar=c(2, 3, 2, 3) +0.1)
-plot03<-barplot(avg03, col="steelblue", cex.axis=1.3, xaxt="n")
-error.bar(plot03, avg03, sd03) #plot SD
+plot07<-barplot(avg07, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot07, avg07, sd07) #plot SD
+axis(1, at=plot07, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 
-plot11<-barplot(avg11, col="darkgreen", cex.axis=1.3, xaxt="n")
-error.bar(plot11, avg11, sd11) #plot SD
+# par(mar=c(2, 3, 2, 3) +0.1)
+# plot09<-barplot(avg09, col="black", cex.axis=1.3, xaxt="n")
+# #error.bar(plot09, avg09, sd09) #plot SD
+# axis(1, at=plot09, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 
 par(mar=c(2, 4, 2, 3) +0.1)
-plot02<-barplot(avg02, col="darkred", cex.axis=1.3, xaxt="n")
-error.bar(plot02, avg02, sd02) #plot SD
+plot02<-barplot(avg02, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot02, avg02, sd02) #plot SD
+axis(1, at=plot02, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+par(mar=c(2, 4, 2, 3) +0.1)
+plot13<-barplot(avg13, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot13, avg13, sd13) #plot SD
+axis(1, at=plot13, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 
 par(mar=c(2, 3, 2, 3) +0.1)
-plot05<-barplot(avg05, col="orange2", cex.axis=1.3, xaxt="n")
-error.bar(plot05, avg05, sd05) #plot SD
-
-plot04<-barplot(avg04, col="gold1", cex.axis=1.3, xaxt="n")
-error.bar(plot04, avg04, sd04) #plot SD
-
-par(mar=c(2, 4, 2, 3) +0.1)
-plot16<-barplot(avg16, col="deeppink3", cex.axis=1.3, xaxt="n")
-error.bar(plot16, avg16, sd16) #plot SD
-axis(1, at=plot16, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+plot03<-barplot(avg03, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot03, avg03, sd03) #plot SD
+axis(1, at=plot03, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 
 par(mar=c(2, 3, 2, 3) +0.1)
-plot7<-barplot(avg7, col="blueviolet", cex.axis=1.3, xaxt="n")
-error.bar(plot7, avg7, sd7) #plot SD
-axis(1, at=plot7, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+plot05<-barplot(avg05, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot05, avg05, sd05) #plot SD
+axis(1, at=plot05, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 
+par(mar=c(2, 3, 2, 3) +0.1)
+plot08<-barplot(avg08, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot08, avg08, sd08) #plot SD
+axis(1, at=plot08, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+par(mar=c(2, 3, 2, 3) +0.1)
 plot06<-barplot(avg06, col="black", cex.axis=1.3, xaxt="n")
-error.bar(plot06, avg06, sd06) #plot SD
+#error.bar(plot06, avg06, sd06) #plot SD
 axis(1, at=plot06, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+par(mar=c(2, 3, 2, 3) +0.1)
+plot04<-barplot(avg04, col="black", cex.axis=1.3, xaxt="n")
+#error.bar(plot04, avg04, sd04) #plot SD
+axis(1, at=plot04, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+
+# par(mar=c(2, 3, 2, 3) +0.1)
+# plot12<-barplot(avg12, col="black", cex.axis=1.3, xaxt="n")
+# #error.bar(plot12, avg12, sd12) #plot SD
+# axis(1, at=plot12, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+#
+
+# par(mar=c(2, 4, 2, 3) +0.1)
+# plot11<-barplot(avg11, col="black", cex.axis=1.3, xaxt="n")
+# error.bar(plot11, avg11, sd11) #plot SD
+# axis(1, at=plot11, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
+# 
+# par(mar=c(2, 4, 2, 3) +0.1)
+# plot16<-barplot(avg16, col="balck", cex.axis=1.3, xaxt="n")
+# error.bar(plot16, avg16, sd16) #plot SD
+# axis(1, at=plot16, labels=c("D0", "D0 Recovered"), cex.axis=1.7)
 

### 2015-01-29 Figure 6/abxD01.optimizingModel.Rmd
worked on modeling

@@
 difficile* CFU the next day.
+
+I wanted to use the information in the titration data set to improve the model.
+The original model was really good at predicting high levels of *C.
+difficile*, but not very good at lower levels.
+The titration data set includes groups with lower levels of *C.
+difficile*.
+So I'm running random forest with the titration data alone:
+
+``` {r, eval=TRUE}
+#################################
+###rf of the titration data
+rf.titr<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.newtitration.regression.logtrans.filter16mintot.noUntr.csv", header=T)
+rf.titr<-rf.titr[,-1]
+
+library(randomForest)
+
+#fit the randomforest model
+rf.titr.model <- randomForest(nextDayCFU~., 
+                               data = rf.titr,  outscale=TRUE,
+                               importance=TRUE, proximity=TRUE,
+                               keep.forest=TRUE, ntree=5000
+)
+plot(rf.titr.model)
+print(rf.titr.model)
+
+#what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
+varImpPlot(rf.titr.model, type=1)
+imp<-importance(rf.titr.model)
+most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
+
+```
+
+The % variance explained was 90.95. Otu13, followed by 3 and 1 were the most obvious influences on the model.
+
+I've also performed this same script to a file without the untreated titration groups.
+This yields 82.69% variance explained.
+Otu3 is the most important OTU in this dataset, followed by 13, 1, 20, 42.
+In fact all the top OTUs are mostly porphyromonadaceae, a lachnospiraceae, and alistipes.
+It almost seems from this data like resistance is more about the loss of these protective bugs as opposed to growth of *C.
+difficile* promoting bacteria.
 

### 2015-01-29 Figure 6/abxD01.lmModel.compare2.R

@@ -80,8 +80,9 @@ summary(lm5)
 p <- 5  #number of parameters
 
 titr<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.lm.newtitration.logtrans.filter16mintotal.19otus.noUntr.csv", header=T)
+actual<-as.data.frame(titr[,2])
+row.names(actual) <- titr[,1]
 titr<-titr[,-1]
-actual<-as.data.frame(titr[,1])
 titr<-titr[,-1]
 n <- dim(titr)[1] #number of samples
 
@@ -101,22 +102,28 @@ pred.w.clim <-predict(lm5, newtit, interval = "confidence")
 matplot(actual, cbind(pred.w.clim, pred.w.plim[, -1]), lty = c(1, 2, 2, 3, 3), type="l", ylab = "predicted y")
 
 
-ybar = colMeans(actual)[1]
-SStot = sum((actual-ybar)^2)
-SSres = sum((actual-predictlm5)^2)
-rsq = 1-(SSres/SStot)
-rsq
+results<-cbind(actual, predictlm5$fit)
+names(results) <- c( "actual", "predict")
+plot(results$actual, results$predict)
 
-#adjusted r^2
-numer <- ((1-rsq)*(n-1))
-denom <- (n-p-1)
-adjr2 <- (1 - numer/denom)
-adjr2
+ybar <- apply(results, 2, mean)
+num<-sum((results$actual-ybar["actual"])*(results$predict-ybar["predict"]))
+denA <- sum((results$actual-ybar["actual"])^2)
+denB <- sum((results$predict-ybar["predict"])^2)
 
-res<-cbind(actual, predictlm5)
-names(res) <- c( "actual", "predict")
-plot(res$actual, res$predict)
+rsq <- (num^2)/(denA*denB) #calculated from the square of the sample correlation coefficient, little r squared as opposed to big R squared
+print(rsq)
 
+results[(results$actual == 0 & results$predict < 4 & results$predict > 1.7),]
+resid <- (results$actual - results$predict)
+results <- cbind(results, resid)
+absResid <- abs(results$actual - results$predict)
+results <- cbind(results, absResid)
+ord_residuals <- order(-results$absResid)
+ord_residuals <- results[ord_residuals,]
+
+plot(results$predict, results$resid)
[…]

### 2015-01-30 abxD01_analysis.Rmd
finished a lot of model analysis, wrote functions

@@
 ```{r, eval=TRUE}
-topdose_logtrans<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.topdose2.regression.logtrans.filter16mintot.csv", header=T)
+topdose_logtrans<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.shared.rf.topdose2.regression.logtrans.filter16mintot.filter16mintot.csv", header=T)
 training_logtrans<-topdose_logtrans[,-1]
@@
 imp<-importance(model_logtrans)
-most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
+most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:20],])
+
+write.table(imp, file="~/Desktop/mothur/abxD01/rf/RF.regression.topdose2.filter.filter.txt", sep="\t", row.names=TRUE)
 

### 2015-01-30 Figure 6/abxD01.lmModel.compare2.R

@@ -1,83 +1,186 @@
 library(leaps)
 
 topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.19otus.rfnegpos.csv", header=T)
-td<-topdose[,-1]
+actual<-as.data.frame(topdose[,2]) #save the actual results in new df
+row.names(actual) <- topdose[,1] #save the group names
+td<-topdose[,-1] 
 attach(td)
 ids<-names(td)
-ids = ids[-1]
+ids = ids[-1] #ids of OTUs in topdose
 
+
+
+
+
+#Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00053 + Otu00431
 #this will make 30 models--the 3 best for each model with 1 to 10 variables
-leaps<-regsubsets(nextDayCFU ~ Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00283 + Otu00431, data=td, nbest=3, nvmax=10)
+which(ids == "Otu00003")
+inGroup <- c(which(ids == "Otu00003"), which(ids == "Otu00007"), which(ids == "Otu000020"), which(ids == "Otu00039"), which(ids == "Otu00013"))
+outGroup <- c(which(ids == "Otu00006"), which(ids == "Otu00015"))
+leaps.3_7_20_39_13.no6_15<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, force.in=inGroup, force.out=outGroup)
+plot(leaps.3_7_20_39_13.no6_15, scale="adjr2", main="leaps.3_7_20_39_13.no6_15")
+plot(leaps.3_7_20_39_13.no6_15, scale="bic", main="leaps.3_7_20_39_13.no6_15")
+#leaps.13, force in otu13
+#leaps.3 force in otu3
+#leaps.3.no6
+#leaps.3.no7
+#leaps.no6
+#leaps.no7
[…]

### 2015-01-30 abxD01_analysis.Rmd
going through results to tweak things

@@
 
+For comparison of the CFU's among titration groups: 
+```{r, eval=FALSE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrStrep",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+strep_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p < 0.0001
+#strep_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# all significantly different
+strep.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(strep.aov)
+#TukeyHSD(strep.aov)
+strep_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
+# all significantly different
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrCef",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+cef_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p = 0.003
+#cef_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# mid and low dose not different, but high different than mid and low
+#cef.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(cef.aov)
+#TukeyHSD(cef.aov)
+cef_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
+# all significantly different
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrVanc",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+vanc_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p = 0.0016
+#vanc_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# mid and low dose not different, but high different than mid and low
+vanc.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(vanc.aov)
+#TukeyHSD(vanc.aov)
+vanc_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
+# mid and low dose not different, but high different than mid and low
+
+```
+
+For comparisons between family-level phylotypes:
+Performed inside abxD01.barcharts.xOTU.test.r
 ```{r, eval=FALSE}

### 2015-01-30 Figure 6/abxD01.lmModel.compare2.R

@@ -18,8 +18,8 @@ which(ids == "Otu00003")
 inGroup <- c(which(ids == "Otu00003"), which(ids == "Otu00007"), which(ids == "Otu000020"), which(ids == "Otu00039"), which(ids == "Otu00013"))
 outGroup <- c(which(ids == "Otu00006"), which(ids == "Otu00015"))
 leaps.3_7_20_39_13.no6_15<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, force.in=inGroup, force.out=outGroup)
-plot(leaps.3_7_20_39_13.no6_15, scale="adjr2", main="leaps.3_7_20_39_13.no6_15")
-plot(leaps.3_7_20_39_13.no6_15, scale="bic", main="leaps.3_7_20_39_13.no6_15")
+plot(leaps.3_7_20_39_15.no6_13, scale="adjr2", main="leaps.3_7_20_39_15.no6_13")
+plot(leaps.3_7_20_39_15.no6_13, scale="bic", main="leaps.3_7_20_39_15.no6_13")
 #leaps.13, force in otu13
 #leaps.3 force in otu3
 #leaps.3.no6

### 2015-01-30 abxD01.barcharts.xOTU.test.r

@@ -8,8 +8,8 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.allmetro.forlogscale.fig5.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 5/allmetro_tx2_barchart_ids.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 3/allvanctitr_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
 abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
 
@@ -20,7 +20,7 @@ sortbyphyl<-TRUE
 # If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
 # If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
 graphbyphyl<-TRUE
-#file<-file[file$expgroup!="1untrStrep",]
+file<-file[file$expgroup!="1untrVanc",]
 
 
 # Highlight all and run!
@@ -343,7 +343,8 @@ if(graphbyphyl==TRUE){
     otus <- dimnames(mavgs)[2][[1]][j:(leng+j-1)]
     m <- j
     for (i in 1:leng){      
-      results.wilcox <- pairwise.wilcox.test(file[,which(names(file)==otus[i])], file$expgroup, p.adj="BH")
+      #results.wilcox <- pairwise.wilcox.test(file[,which(names(file)==otus[i])], file$expgroup, p.adj="BH")
+      results.wilcox <- pairwise.t.test(file[,which(names(file)==otus[i])], file$expgroup, p.adj="BH")
       
       #double check that all values were calculated, if not put as n.s.
       if(results.wilcox$p.value[1] == "NaN"){

### 2015-01-30 Figure 6/abxD01.optimizingModel.Rmd
fix graphs, model

@@
 ###rf of the delay data
-delay<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.delay.regression.logtrans.filter16mintot.csv", header=T)
+delay<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.delay.regression.logtrans.filter16mintot.noUntr.csv", header=T)
 delay<-delay[,-1]

### 2015-01-30 abxD01_analysis.Rmd
fix graphs, model

@@
 
+Statistical comparison of delayed and on time CFU results:
+```{r, eval=FALSE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.allmetro.forlogscale.fig5.csv", header=T)
+file<-file[file$expgroup!="1untr",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+metro_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p < 0.0001
+#strep_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# all significantly different
+metro.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(metr.aov)
+#TukeyHSD(strep.aov)
+metro_ttest <- t.test(file$nextDayCFU~file$expgroup)
+# both significantly different
+```
+
 ## Figure 6

### 2015-01-30 Figure 6/abxD01.lmModel.compare2.R

@@ -14,12 +14,12 @@ ids = ids[-1] #ids of OTUs in topdose
 
 #Otu00002 + Otu00003 + Otu00006 + Otu00007 + Otu00011 + Otu00013 + Otu00015 + Otu00019 + Otu00020 + Otu00023 + Otu00027 + Otu00029 + Otu00039 + Otu00044 + Otu00065 + Otu00078 + Otu00120 + Otu00053 + Otu00431
 #this will make 30 models--the 3 best for each model with 1 to 10 variables
-which(ids == "Otu00003")
-inGroup <- c(which(ids == "Otu00003"), which(ids == "Otu00007"), which(ids == "Otu000020"), which(ids == "Otu00039"), which(ids == "Otu00013"))
-outGroup <- c(which(ids == "Otu00006"), which(ids == "Otu00015"))
-leaps.3_7_20_39_13.no6_15<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, force.in=inGroup, force.out=outGroup)
-plot(leaps.3_7_20_39_15.no6_13, scale="adjr2", main="leaps.3_7_20_39_15.no6_13")
-plot(leaps.3_7_20_39_15.no6_13, scale="bic", main="leaps.3_7_20_39_15.no6_13")
+#which(ids == "Otu00003")
+inGroup <- c(which(ids == "Otu00007"), which(ids == "Otu000020"), which(ids == "Otu00039"), which(ids == "Otu00013"), which(ids == "Otu00003"), which(ids == "Otu00015"), which(ids == "Otu00053"))
+outGroup <- c(which(ids == "Otu00006"))
+leaps.3_7_20_39_13_15_53.no6<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, force.in=inGroup, force.out=outGroup)
+plot(leaps.3_7_20_39_13_15_53.no6, scale="adjr2", main="leaps.3_7_20_39_13_15_53.no6")
+plot(leaps.3_7_20_39_13_15_53.no6, scale="bic", main="leaps.3_7_20_39_13_15_53.no6")
 #leaps.13, force in otu13
 #leaps.3 force in otu3
 #leaps.3.no6
[…]

### 2015-01-30 abxD01.barcharts.xOTU.test.r

@@ -8,8 +8,8 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 3/allvanctitr_tx2_barchart_ids.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.allmetro.forlogscale.fig5.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 5/allmetro_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
 abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
 

### 2015-02-03 abxD01_analysis.Rmd
updated rmd html

--- 2015-01-30 abxD01_analysis.Rmd
@@
 date: "September 29, 2014"
-output: word_document
+output: html_document
 ---
@@
 
+For comparison of the CFU's among titration groups: 
+```{r, eval=TRUE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrStrep",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+strep_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p < 0.0001
+#strep_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# all significantly different
+strep.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(strep.aov)
+#TukeyHSD(strep.aov)
+strep_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
+print(strep_ttest)
+# all significantly different
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrCef",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+cef_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p = 0.003
+#cef_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# mid and low dose not different, but high different than mid and low
+cef.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(cef.aov)
+#TukeyHSD(cef.aov)
+cef_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
+print(cef_ttest)
+# all significantly different
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
+file<-file[file$expgroup!="1untrVanc",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+vanc_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p = 0.0016
+#vanc_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# mid and low dose not different, but high different than mid and low
+vanc.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(vanc.aov)
+#TukeyHSD(vanc.aov)
+vanc_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
+print(vanc_ttest)
+# mid and low dose not different, but high different than mid and low
+
+```
+
+For comparisons between family-level phylotypes:
+Performed inside abxD01.barcharts.xOTU.test.r
 ```{r, eval=FALSE}
@@
 
+Statistical comparison of delayed and on time CFU results:
+```{r, eval=TRUE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.allmetro.forlogscale.fig5.csv", header=T)
+file<-file[file$expgroup!="1untr",]
+avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
+metro_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+#kruskal.test(file$nextDayCFU ~ file$expgroup)
+# p < 0.0001
+#strep_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
+# all significantly different
+metro.aov<-aov(file$nextDayCFU ~ file$expgroup)
+anova(metro.aov)
+#TukeyHSD(strep.aov)
+metro_ttest <- t.test(file$nextDayCFU~file$expgroup)
+print(metro_ttest)
+# both significantly different
+```
+
 ## Figure 6
@@
 Then in r: 
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 topdose_logtrans<-read.csv("~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.shared.rf.topdose2.regression.logtrans.filter16mintot.filter16mintot.csv", header=T)

### 2015-02-03 Figure 6/abxD01.leaps.all.R

@@ -17,3 +17,25 @@ coef(leaps.all, 1:2)
 #plot(leaps.all,scale="r2")
 plot(leaps.all,scale="adjr2")
 plot(leaps.all,scale="bic")
+
+
+
+
+
+topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.csv", header=T)
+td<-topdose[,-1]
+attach(td)
+ids<-names(td)
+ids = ids[-1]
+
+leaps.all<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, really.big=T)
+#shows the best models and which variables to include
+summary(leaps.all)
+#shows the value for each parameter in the model along with intercept
+vcov(leaps.all, 50) ##the second parameter corresponds to model number
+#coef
+coef(leaps.all, 1:2)
+#scales=bic, Cp, and adjr2, r2
+#plot(leaps.all,scale="r2")
+plot(leaps.all,scale="adjr2")
+plot(leaps.all,scale="bic")

### 2015-02-04 abxD01_analysis.Rmd
worked w/delayModel, new graphs for modelSelection

@@
 
-ybar = mean(actual_logtrans)
-SSres = sum((actual_logtrans-predicted_logtrans)^2)
-SStot = sum((actual_logtrans-ybar)^2)
-rsq = 1-(SSres/SStot)
-
-#rsq_logtrans <- 1-sum((actual_logtrans-predicted_logtrans)^2)/sum((actual_logtrans-mean(actual_logtrans))^2)
-print(rsq)
-
-ybar2 = mean(actual2_logtrans)
-SSres2 = sum((actual2_logtrans-predicted2_logtrans)^2)
-SStot2 = sum((actual2_logtrans-ybar2)^2)
-rsq2 = 1-(SSres2/SStot2)
-
-print(rsq2)
-
-write.table(cbind(actual_logtrans, predicted_logtrans), file="~/Desktop/mothur/abxD01/rf/RF.regression.topTitr.txt", sep="\t", row.names=FALSE)
-
 ```

### 2015-02-04 Figure 6/abxD01.lmModel.compare2.R

@@ -1,19 +1,60 @@
 library(leaps)
 
-topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.19otus.rfnegpos.csv", header=T)
+#topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.18otus.rfnegpos.csv", header=T)
+#topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.delay.logtrans.filter16mintot.select.csv", header=T)
+topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal2X.topdose2.shared.no6.csv", header=T)
+
+
 actual<-as.data.frame(topdose[,2]) #save the actual results in new df
 row.names(actual) <- topdose[,1] #save the group names
 td<-topdose[,-1] 
 attach(td)
 ids<-names(td)
 ids = ids[-1] #ids of OTUs in topdose
+#names(td)[2:(length(td))] <- paste0("OTU", substr(ids, 6, 8))
+
+#inGroup <- c(which(ids == "Otu00007"), which(ids == "Otu000020"), which(ids == "Otu00039"), which(ids == "Otu00013"), which(ids == "Otu00003"), which(ids == "Otu00015"), which(ids == "Otu00053"))
+inGroup <- NULL
+outGroup <- c(NULL)
+leaps.build<-regsubsets(nextDayCFU ~ ., data=td, nbest=5, nvmax=10, force.in=inGroup, force.out=outGroup, really.big=T)
+plot(leaps.build, scale="adjr2", main="leaps.build")
+plot(leaps.build, scale="bic", main="leaps.build")
+plot(leaps.build, scale="Cp", main="leaps.build")
+
+library(car)
+
+abbrNames <- substr(leaps.build$xnames, 6, 8)
+abbrNames <- abbrNames[-1]
+
+minSubsetSize <- 2
+subsets(leaps.build, names=abbrNames, statistic="adjr2", legend=FALSE, min.size=minSubsetSize, abbrev=6, cex.subsets=.5, las=1, xlim=c(minSubsetSize, 12))
[…]

### 2015-02-15 Figure 6/lm.builds.avg0.005bygroup.r
add SE to bar plots, color by phylum

@@ -1,4 +1,357 @@
 
+
+
+
+
+
+# ############################
+# #Curated Approach 2/13/15, using OTUs above the 1% cutoff, attempted to start.. but decided that the relabund cutoff is fairly reasonable. 
+# ############################
+# 
+# library(randomForest)
+# 
+# topdose<-read.csv("~/Desktop/mothur/abxD01/model/shared.topdose.avg0.01.logtrans.csv", header=T)
+# td<-topdose[,-1] 
+# 
+# #fit the randomforest model
+# rf.model <- randomForest(nextDayCFU~., 
+#                          data = td,  outscale=TRUE,
+#                          importance=TRUE, proximity=TRUE,
+#                          keep.forest=TRUE, ntree=5000
+# )
+# plot(rf.model)
+# print(rf.model) # % Var explained: 87.81
+# 
+# 
+# #what are the important variables (via permutation) #type 1 is mean decrease in accuracy, type 2 is mean decrease in node impurity
+# par(mfrow=c(1, 1)) 
+# varImpPlot(rf.model, type=1)
+# imp<-importance(rf.model)
+# write.table(imp, file="~/Desktop/mothur/abxD01/rf/rf.topdose.avg0.01.txt", sep="\t", row.names=T, col.names=T)
+# partialPlot(rf.model, td, Otu00003)
+# 
+# ############ Then used these values in combination with their correlation values with subsequent C. difficile CFU. 
+# ############ Put those in the following candidatePool.csv:
+# 
+# library(leaps)
+# 
+# topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.candidatePool.csv", header=T)
+# 
+# actual <-NULL
+# actual<-as.data.frame(topdose[,2]) #save the actual results in new df
+# row.names(actual) <- topdose[,1] #save the group names
+# td<-topdose[,-1] 
[…]

### 2015-02-15 abxD01.barcharts.xOTU.test.r

@@ -8,10 +8,12 @@
 #################################################
 # Parameters to change:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.allmetro.forlogscale.fig5.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 5/allmetro_tx2_barchart_ids.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", header=T)
 # Y Labels for each graph: 
-abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
+# abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
+abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
+
 
 # If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
 # The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
@@ -19,25 +21,31 @@ sortbyphyl<-TRUE
 
 # If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
 # If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
-graphbyphyl<-TRUE
-file<-file[file$expgroup!="1untrVanc",]
+graphbyphyl<-FALSE
+# file<-file[file$expgroup!="1untrVanc",]
 
 
 # Highlight all and run!
 #################################################
 
+
+if(graphbyphyl==TRUE)
+{
+  sortbyphyl=TRUE
+}
 file<-file[,-1] #delete the first col of group names
+
+
+
 avgs=NULL
[…]

### 2015-02-17 R scripts/barplotBeside.r
Clean up the bar chart code into smaller functions
Made separate function for determining the statistical letter for each
bar chart. Also added another option for plotting so I could re-do
Figure 3.

@@ -1,0 +1,78 @@
+# The most number of side by side groups should be 3. Not made for others. 
+
+barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids){
+
+  groups <- unique(id.info$abx)
+  numgr <- length(groups)
+
+  #leng<-dim(mavgs)[2]
+  
+  
+  
+  par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
+  par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+  color_transparent <- adjustcolor("black", alpha.f = 0.2) 
+  color <- gray.colors(3, start=0, end=1, alpha=NULL)
+
+  for(j in 1:numgr){
+    
+    graphGroups <- unique(id.info[id.info$abx == groups[j], "expgroup"])
+    mavgs <- matrix.of.avgs[c(graphGroups),]
+    matrix.se <- matrix.of.se[c(graphGroups),]
+      
+    if(j != numgr){
+      bp <- barplot(mavgs, ylab=NULL, col=color,  beside=TRUE, yaxt="n", xaxt="n", ylim=c(0.001, 1), log="y", cex.names=.5)
+
+      # add standard error bars
+      d <- as.data.frame( cbind( c(bp), c(mavgs), c(matrix.se) ) )
+      names(d) <- c("x", "y", "se")
+      with (
+        data = d
+        , expr = errbar(x, y, y+se, y-se, add=T, pch=".", cap=.01)
+      )
+      
+      # FIXXXXXX/TESTTTTT
+      statLetter <- findStatLetters(file, mavgs)
+      #For labeling letters above each bar
+      m <- j
+      for(n in 1:leng){
+        text( label[,n], mavgs[,m], labels=statLetter[,n], cex=1, col="red", pos=3, offset=0.15, xpd=TRUE)
+        m <- m+1
+      }
+      
+      # add labels and pretty the graph
+      axis(2, las=1, at=c(.001, .01, .1, 1), labels=c(0, .01, .1, 1), cex.axis=1.1)
+      mtext(groups[j], side=2, line=6, cex=.8)
+      abline(h=c(0.001, 1), lwd=3) #min/max
+      abline(h=c(0.01, 0.1), col=color_transparent, lty="longdash", lwd=2)
[…]

### 2015-02-17 R scripts/findStatLetters.r

@@ -1,0 +1,72 @@
+
+
+findStatLetters <- function(file, mavgs){
+
+  otus <- dimnames(mavgs)[2][[1]]
+  statLetter <- NULL
+  statLetter <- matrix(c("NA"),nrow=3, ncol=length(otus))
+  
+  for (i in 1:(length(otus))) {      
+    results.wilcox <- pairwise.wilcox.test(file[,which(names(file)==otus[i])], file$expgroup, p.adj="BH")
+    #results.wilcox <- pairwise.t.test(file[,which(names(file)==otus[i])], file$expgroup, p.adj="BH")
+    
+    #double check that all values were calculated, if not put as n.s.
+    if(results.wilcox$p.value[1] == "NaN"){
+      results.wilcox$p.value[1] <- 10 #a value greater than 0.05
+    }
+    if(results.wilcox$p.value[1] == "NaN"){
+      results.wilcox$p.value[1] <- 10
+    }
+    if(results.wilcox$p.value[1] == "NaN"){
+      results.wilcox$p.value[1] <- 10
+    }
+    
+    if( results.wilcox$p.value[1] >= 0.05 ){
+      if( results.wilcox$p.value[2] >= 0.05 ){
+        if( results.wilcox$p.value[4] >= 0.05 ){
+          statLetter[1, i] <- "a"
+          statLetter[2, i] <- "a"
+          statLetter[3, i] <- "a"
+        } else{ 
+          statLetter[1, i] <- "ab"
+          statLetter[2, i] <- "a"
+          statLetter[3, i] <- "b"            
+        }
+      } else{
+        if( results.wilcox$p.value[4] >= 0.05 ){
+          statLetter[1, i] <- "a"
+          statLetter[2, i] <- "ab"
+          statLetter[3, i] <- "b"
+        } else{
+          statLetter[1, i] <- "a"
+          statLetter[2, i] <- "a"
+          statLetter[3, i] <- "b"            
+        }
+      }
+    } else{
+      if( results.wilcox$p.value[2] >= 0.05 ){
+        if( results.wilcox$p.value[4] >= 0.05 ){
+          statLetter[1, i] <- "a"
+          statLetter[2, i] <- "b"
+          statLetter[3, i] <- "ab"
+        } else{
+          statLetter[1, i] <- "a"
+          statLetter[2, i] <- "b"
+          statLetter[3, i] <- "a"
+        }
+      } else{
[…]

### 2015-02-17 R scripts/stacked.barcharts.r

@@ -1,0 +1,537 @@
+#################################################
+#Purpose: Convert csv matrix of columns you want converted into barcharts
+#input: csv file following this heading format format:  sampleID, expgroup, (nextDayCFU,) Otu001, Otu002, Otu003, etc
+#ex:file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.5.subsample.relabund.topdose.forlogscale2.csv
+#Note: It is assumed that anything after the expgroup column are vectors containing data to be split by the groups in 'expgroup' 
+#      and averaged with SD measured. 
+#Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
+#################################################
+# Parameters to change, an example:
+# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
+fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", header=T)
+
+file <- "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv"
+fileIDS <- "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv"
+
+# Y Labels for each graph: 
+# abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
+# abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
+graphLabels<-c("Cefoperazone", "Streptomycin", "Vancomycin")
+
+# If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
+# The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
+sortbyphyl<-TRUE
+
+# If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
[…]

### 2015-02-17 abxD01.barcharts.xOTU.test.r

@@ -1,516 +1,0 @@
-#################################################
-#Purpose: Convert csv matrix of columns you want converted into barcharts
-#input: csv file following this heading format format:  sampleID, expgroup, (nextDayCFU,) Otu001, Otu002, Otu003, etc
-#ex:file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.5.subsample.relabund.topdose.forlogscale2.csv
-#Note: It is assumed that anything after the expgroup column are vectors containing data to be split by the groups in 'expgroup' 
-#      and averaged with SD measured. 
-#Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
-#################################################
-# Parameters to change:
-# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", header=T)
-fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", header=T)
-# Y Labels for each graph: 
-# abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
-abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
-
-
-# If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
-# The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
-sortbyphyl<-TRUE
-
-# If you want individual graphs as each group (FALSE) or as phylums with each OTU (TRUE)
-# If you choose TRUE, then set sortbyphyl as TRUE too... IF you forget to change this it changes automatically in the code.
-graphbyphyl<-FALSE
[…]

### 2015-02-17 abxD01_analysis.Rmd
add SE to bar plots, color by phylum
Update .Rmd to reflect changes in bar chart code

@@
 
-## Figure 1: Bargraph
+## Figure 1
 
@@
 
-This figure was built using the R script called "abxD01.barcharts.xOTU.test.r" in the abxD01 folder.The input files are listed in Github.
-The graph can be sorted 2 ways either by the untreated's decreasing relabund, or by the most abundant within each phylum.
+### Figure 1: Bargraph
+This figure was built using the R script called "stacked.barcharts.r", which contains the stackedbarcharts() funciton in the abxD01 folder.The input files are listed in Github.
+The graph can be sorted 2 ways either by the untreated's decreasing relabund (sortbyphyl=FALSE), or by the most abundant within each phylum (sortbyphyl=TRUE).
 For each treatment, the resulting *C.
@@
 The parameters I changed in the r script were:
-sortbyphyl<-TRUE
-graphbyphyl<-FALSE
+```{r, eval=TRUE}
 abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
 
-File saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf"."
+# Generate the 
+stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+
+stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+
+```
+
+Files saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf".
 
@@
 otuCorrs<-read.csv("~/Documents/Github/abxD01/Figure 2/corr_avgAbove0.001_topdose2_stripchart.csv", header=T)
-#labels=c("Lachnospiraceae", "Ruminococcaceae", "Clostridia", "Lactobacillales", "Firmicutes", "Bacillales", "Porphyromonadaceae", "Bacteroidales", "Bacteroidetes", "Actinobacteria", "Proteobacteria", "Anaeroplasma", "Deinococcus", "Unclassified")
-#ns=c("n=18", "n=9", "n=8", "n=4", "n=2", "n=1", "n=15", "n=4", "n=1", "n=4", "n=3", "n=1", "n=1", "n=3")
 
-#test<-otuCorrs[otuCorrs$phyl=="Bacteroidetes",]
-#length(unique(test$family))
-
-pch <- c(1,1,1,1, 6,6, 6, 6, 6, 6, 6, 6,0, 0, 9, 2, 2, 4)
+pch <- c(6,6,6,6, 1,1, 1, 1, 1, 1, 1, 1,0, 0, 9, 2, 2, 4)
 par(mar=c(12, 7, 2, 2) +0.1, mgp=c(4, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
-stripchart(otuCorrs$corr ~ otuCorrs$graphOrder, vertical=TRUE, ylab="Spearman Correlation", ylim=c(-1, 1), cex.axis=1.1, cex=2, pch=pch, lwd=2, col="black",  method="jitter",  jitter=0.25, cex.lab=1.3, yaxt="n", xaxt="n")
+stripchart(otuCorrs$corr ~ otuCorrs$graphOrder, vertical=TRUE, ylab="Spearman Correlation", ylim=c(-1, 1), cex.axis=1, cex=2, pch=pch, lwd=2, col="black",  method="jitter",  jitter=.2, cex.lab=1.1, yaxt="n", xaxt="n")
 axis(1, cex.axis=1, at=(1:nlevels(otuCorrs$family)), labels=FALSE)
-text(1:nlevels(otuCorrs$family)+.1, par("usr")[3]-0.05, label=unique(otuCorrs$family), xpd=NA, pos=2, srt=45, cex=1.2)
-#text(1:nlevels(otuCorrs$family), par("usr")[3]-0.01, label=ns, xpd=NA, pos=1, cex=.9)
-axis(2, cex.axis=1.1, at=seq(-1,+1, by=.1), las=1)
-abline(v=c(1:nlevels(otuCorrs$family)), col="dark gray")
-#abline(h=c(.3, -.3), col="dark gray", lty="dashed", lwd=4) #change to something else
-abline(h=0, col="black", lwd=2)
-abline(h=c(0.25, 0.5, 0.75, -0.25, -0.5, -0.75), col="light gray", lwd=2, lty="dashed")
-legend("top", legend=unique(otuCorrs$phyl), pch=c(1, 6, 0, 9, 2, 4), cex=0.7, bty="o", horiz=TRUE)
+family.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "family"])
+text(1:nlevels(otuCorrs$family)+.1, par("usr")[3]-0.05, label=(family.labels), xpd=NA, pos=2, srt=45, cex=1)
+axis(2, cex.axis=1, at=seq(-1,+1, by=.1), las=1)
+color_transparent <- adjustcolor("black", alpha.f = 0.3) 
+abline(v=c(1:nlevels(otuCorrs$family)), col=color_transparent)
+abline(h=0, col="dark gray", lwd=2)
+abline(h=c(0.25, 0.5, 0.75, -0.25, -0.5, -0.75), col=color_transparent, lwd=2, lty="dashed")
+phylum.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "phyl"])
+leg<-legend("top", legend=phylum.labels, pch=c(6, 1, 0, 9, 2, 4), cex=0.7, bty="o", horiz=TRUE, pt.cex=1.5, text.width=2.5)
 
@@
 
-Figure saved as "corr_avgAbove0.001_topdose2_10x8.pdf".
+Figure saved as "corr_avgAbove0.001_topdose2_10x10.pdf".
 
@@
 
-The code for this is the script "abxD01.barcharts.xOTU.test.r" in the paper's home Github directory. Opening this file, the parameters I changed in the top were:
-sortbyphyl<-TRUE
-graphbyphyl<-TRUE
+The code for this is the script "stacked.barcharts.r" which contains the stackedbarcharts() function in the paper's home Github directory. Opening this file, the parameters I changed in the top were:
+```{r, eval=FALSE}
+setwd("/Users/alyxschubert/Documents/Github/abxD01/")
+
 abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
-file
-fileIDS
+stackedbarcharts(file = "./Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", fileIDS = "./Figure 3/allceftitr_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = TRUE, excludeGroups = "1untrCef")
+
+abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
+stackedbarcharts(file = "./Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", fileIDS = "./Figure 3/allstreptitr_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = TRUE, excludeGroups = "1untrStrep")
+
+abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
+stackedbarcharts(file = "./Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", fileIDS = "./Figure 3/allvanctitr_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = TRUE, excludeGroups = "1untrVanc")
+
+```
 

### 2015-02-23 R scripts/stacked.barcharts.r
Clean up stacked.barcharts.r comments

@@ -7,26 +7,19 @@
 #Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
 #################################################
 # Parameters to change, an example:
-# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
-#file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
-#fileIDS<-read.csv("~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", header=T)
+# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.001, expgroups #'d by graph order & sorted by first graph)
 
 file <- "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv"
 fileIDS <- "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv"
-
-# Y Labels for each graph: 
-# abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
-# abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
 graphLabels<-c("Cefoperazone", "Streptomycin", "Vancomycin")
 sortbyphyl<-TRUE
 graphbyphyl<-FALSE
 divide<-TRUE
-# file<-file[file$expgroup!="1untrVanc",]
 
 
 # Highlight all and run!
 #################################################
-stackedbarcharts <- function(file, fileIDS, graphLabels, sortbyphyl=TRUE, graphbyphyl=FALSE, excludeGroups="", divide=FALSE) {
+stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, graphbyphyl=FALSE, excludeGroups="", divide=FALSE) {
   file<-read.csv(file=file, header=T)
   file<-file[file$expgroup!=excludeGroups,]
   
@@ -516,12 +509,6 @@ stackedbarcharts <- function(file, fileIDS, graphLabels, sortbyphyl=TRUE, graphb
     
   } # if(divide=TRUE){
   
-  
 }
-###END 2ND PLOT PARAMETERS
 
 
-
-
[…]

### 2015-02-23 R scripts/barplotBeside.r
Change gray color scale of bar charts

@@ -14,7 +14,7 @@ barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
   par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
   par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
   color_transparent <- adjustcolor("black", alpha.f = 0.2) 
-  color <- gray.colors(3, start=0, end=1, alpha=NULL)
+  color <- gray.colors(3, start=0.2, end=1, alpha=NULL)
 
   for(j in 1:numgr){
     

### 2015-02-23 abxD01_analysis.Rmd
Clean comments and set evals to TRUE for rendered file

@@
 
-This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1.
+This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1 using the original set of antibiotic groups.
 
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 
@@
 results <- results[order(results[,3]),]  #order by pvalue column=3
-write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
+#write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
+print(results[1:dim(results)[1],])
 
@@
 
-This figure shows correlation values for OTUs with next day *C. difficile* CFU levels calculated across the original set of antibiotic experiments.
-
 Necessary files are on github, and it was made using the following code:
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 
@@
 
-### Figure 3: Barchart
-
 This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.
 
-The code for this is the script "stacked.barcharts.r" which contains the stackedbarcharts() function in the paper's home Github directory. Opening this file, the parameters I changed in the top were:
-```{r, eval=FALSE}
-setwd("/Users/alyxschubert/Documents/Github/abxD01/")
-
-abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
-stackedbarcharts(file = "./Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", fileIDS = "./Figure 3/allceftitr_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = TRUE, excludeGroups = "1untrCef")
-
-abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
-stackedbarcharts(file = "./Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", fileIDS = "./Figure 3/allstreptitr_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = TRUE, excludeGroups = "1untrStrep")
-
-abx<-c("5 mg/ml", "0.5 mg/ml", "0.1 mg/ml")
-stackedbarcharts(file = "./Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", fileIDS = "./Figure 3/allvanctitr_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = TRUE, excludeGroups = "1untrVanc")
-
-```
+### Figure 3: Barchart
 
@@
 Input files on github:
-1. abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv
-2. abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv
-3. abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv
-4. allceftitr_tx2_barchart_ids.csv
-5. allstreptitr_tx2_barchart_ids.csv
-6. allvanctitr_tx2_barchart_ids.csv
+1. abxD01.final.tx.2.subsample.alltitrations.forlogscale.csv
+2. alltitrations_tx2_barchart_ids.csv
 
-For this graph I decided to take out the untreated communities to highlight the differences among titration groups. So in r I changed the file a bit:
+Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
 
-```{r, eval=FALSE}
-file<-file[file$expgroup!="1untrStrep",]
-#alternatively for the other titrated antibiotics it's "1untrCef", or "1untrVanc"
+```{r, eval=TRUE}
+
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", divide = TRUE)
+
 ```
@@
 ```{r, eval=TRUE}
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
-file<-file[file$expgroup!="1untrStrep",]
-avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
-strep_avgCD<-format(avgCD, scientific=TRUE, digits=2)
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
+file<-file[!grepl("untr", file$expgroup),]
 file$nextDayCFU <- log10(file$nextDayCFU+1)
-#kruskal.test(file$nextDayCFU ~ file$expgroup)
-# p < 0.0001
-#strep_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
-# all significantly different
-strep.aov<-aov(file$nextDayCFU ~ file$expgroup)
-anova(strep.aov)
-#TukeyHSD(strep.aov)
-strep_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
-print(strep_ttest)
-# all significantly different
 
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", header=T)
-file<-file[file$expgroup!="1untrCef",]
-avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
-cef_avgCD<-format(avgCD, scientific=TRUE, digits=2)
-file$nextDayCFU <- log10(file$nextDayCFU+1)
-#kruskal.test(file$nextDayCFU ~ file$expgroup)
-# p = 0.003
-#cef_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
-# mid and low dose not different, but high different than mid and low
-cef.aov<-aov(file$nextDayCFU ~ file$expgroup)
-anova(cef.aov)
-#TukeyHSD(cef.aov)
-cef_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
-print(cef_ttest)
-# all significantly different
+cef <- file[file$abx=="Cefoperazone", 3:4]
+strep <- file[file$abx=="Streptomycin", 3:4]
+vanc <- file[file$abx=="Vancomycin", 3:4]
 
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
-file<-file[file$expgroup!="1untrVanc",]
-avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
-vanc_avgCD<-format(avgCD, scientific=TRUE, digits=2)
-file$nextDayCFU <- log10(file$nextDayCFU+1)
-#kruskal.test(file$nextDayCFU ~ file$expgroup)
-# p = 0.0016
-#vanc_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
-# mid and low dose not different, but high different than mid and low
-vanc.aov<-aov(file$nextDayCFU ~ file$expgroup)
-anova(vanc.aov)
-#TukeyHSD(vanc.aov)
-vanc_ttest <- pairwise.t.test(file$nextDayCFU, file$expgroup, p.adjust="BH")
-print(vanc_ttest)
-# mid and low dose not different, but high different than mid and low
+diffs <- function(data){
+  data.aov<-aov(data$nextDayCFU ~ data$expgroup)
+  anova(data.aov)
+  pairwise.t.test(data$nextDayCFU, data$expgroup, p.adjust="BH")
+}
 
-```
-
-For comparisons between family-level phylotypes:
-Performed inside abxD01.barcharts.xOTU.test.r
-```{r, eval=FALSE}
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allstreptitr.forlogscale.csv", header=T)
-file<-file[file$expgroup!="1untrStrep",]
-kruskal.test(file$Otu001 ~ file$expgroup)
-kruskal.test(file$Otu003 ~ file$expgroup)
-kruskal.test(file$Otu002 ~ file$expgroup)
-kruskal.test(file$Otu005 ~ file$expgroup)
-kruskal.test(file$Otu004 ~ file$expgroup)
-kruskal.test(file$Otu014 ~ file$expgroup)
-kruskal.test(file$Otu006 ~ file$expgroup)
-kruskal.test(file$Otu007 ~ file$expgroup)
-kruskal.test(file$Otu009 ~ file$expgroup)
-kruskal.test(file$Otu015 ~ file$expgroup)
-kruskal.test(file$Otu017 ~ file$expgroup)
-kruskal.test(file$Other ~ file$expgroup)
-
-test <- pairwise.wilcox.test(file$Otu001, file$expgroup, p.adj="BH")
-comparisons <- length(test$p.value[]) - sum(is.na(test$p.value))
-
-pairwise.wilcox.test(file$Otu003, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu005, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu002, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu004, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu014, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu006, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu007, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu009, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu015, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu017, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Other, file$expgroup, p.adj="BH")
-
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allvanctitr.forlogscale.csv", header=T)
-file<-file[file$expgroup!="1untrVanc",]
-kruskal.test(file$Otu001 ~ file$expgroup)
-kruskal.test(file$Otu002 ~ file$expgroup)
-kruskal.test(file$Otu003 ~ file$expgroup)
-kruskal.test(file$Otu005 ~ file$expgroup)
-kruskal.test(file$Otu004 ~ file$expgroup)
-kruskal.test(file$Otu014 ~ file$expgroup)
-kruskal.test(file$Otu006 ~ file$expgroup)
-kruskal.test(file$Otu007 ~ file$expgroup)
-kruskal.test(file$Otu008 ~ file$expgroup)
-kruskal.test(file$Otu019 ~ file$expgroup)
-kruskal.test(file$Otu018 ~ file$expgroup)
-kruskal.test(file$Otu017 ~ file$expgroup)
-kruskal.test(file$Other ~ file$expgroup)
-
-pairwise.wilcox.test(file$Otu001, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu003, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu005, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu002, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu004, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu014, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu006, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu007, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu008, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu019, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu018, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu017, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Other, file$expgroup, p.adj="BH")
-
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx.2.subsample.allceftitr.forlogscale.csv", header=T)
-file<-file[file$expgroup!="1untrCef",]
-kruskal.test(file$Otu001 ~ file$expgroup)
-kruskal.test(file$Otu002 ~ file$expgroup)
-kruskal.test(file$Otu003 ~ file$expgroup)
-kruskal.test(file$Otu005 ~ file$expgroup)
-kruskal.test(file$Otu004 ~ file$expgroup)
-kruskal.test(file$Otu014 ~ file$expgroup)
-kruskal.test(file$Otu006 ~ file$expgroup)
-kruskal.test(file$Otu007 ~ file$expgroup)
-kruskal.test(file$Otu008 ~ file$expgroup)
-kruskal.test(file$Otu009 ~ file$expgroup)
-kruskal.test(file$Otu019 ~ file$expgroup)
-kruskal.test(file$Otu018 ~ file$expgroup)
-kruskal.test(file$Otu016 ~ file$expgroup)
-kruskal.test(file$Otu024 ~ file$expgroup)
-kruskal.test(file$Otu011 ~ file$expgroup)
-kruskal.test(file$Other ~ file$expgroup)
-
-pairwise.wilcox.test(file$Otu001, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu003, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu005, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu002, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu004, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu014, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu006, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu007, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu008, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu009, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu011, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu019, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu018, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu016, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Otu024, file$expgroup, p.adj="BH")
-pairwise.wilcox.test(file$Other, file$expgroup, p.adj="BH")
+diffs(cef)
+diffs(strep)
+diffs(vanc)
 
@@
 varImpPlot(model_tx2, type=1)
-imp<-importance(model_tx2)
-most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
+#imp<-importance(model_tx2)
+#most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
 ```

### 2015-02-25 R scripts/barplotBeside.r
Edit script for figure 5

@@ -1,15 +1,12 @@
 # The most number of side by side groups should be 3. Not made for others. 
 
-
+#run this alyx!! to fix missing stat letter
 barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
 
   library(Hmisc)
   groups <- unique(id.info$abx)
   numgr <- length(groups)
-
-  #leng<-dim(mavgs)[2]
-  
-  
+  leng<-dim(mavgs)[2]
   
   par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
   par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
@@ -86,7 +83,7 @@ barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
   } # for(j in 1:numgr){
  
   plot(0,xaxt='n',yaxt='n',bty='n',pch='',ylab='',xlab='')
-  leg<-c("High Dose", "Middle Dose","Low Dose" )
+  leg<-c("Untreated", "On Time","+5d Recovery" )
   #pch<-c(16, 16, 16)
   colors<-color
   colors[3] <- "black"

### 2015-02-25 R scripts/stacked.barcharts.r

@@ -9,9 +9,9 @@
 # Parameters to change, an example:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.001, expgroups #'d by graph order & sorted by first graph)
 
-file <- "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv"
-fileIDS <- "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv"
-graphLabels<-c("Cefoperazone", "Streptomycin", "Vancomycin")
+file <- "~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv"
+fileIDS <- "~/Documents/Github/abxD01/Figure 5/alldelay_tx2_barchart_ids.csv"
+#graphLabels<-c("Cefoperazone", "Streptomycin", "Vancomycin")
 sortbyphyl<-TRUE
 graphbyphyl<-FALSE
 divide<-TRUE
@@ -19,9 +19,8 @@ divide<-TRUE
 
 # Highlight all and run!
 #################################################
-stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, graphbyphyl=FALSE, excludeGroups="", divide=FALSE) {
+stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, graphbyphyl=FALSE, excludeGroup=NA, divide=FALSE) {
   file<-read.csv(file=file, header=T)
-  file<-file[file$expgroup!=excludeGroups,]
   
   fileIDS<-read.csv(file=fileIDS, header=T)
   abx <- graphLabels
@@ -133,7 +132,6 @@ stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, g
     
     leng<-length(avgs)
     
-    
     #jtime<-NULL
     #itime<-NULL
     #inside<-FALSE
@@ -155,12 +153,9 @@ stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, g
         else{
           j <- j + 1
           #count<-count+1
-          #jtime<-rbind(jtime, j)
-          
-          
+          #jtime<-rbind(jtime, j)    
         }
-      } #WARNING: could cause an error if dont find the ID, theres no check for the end of the list
[…]

### 2015-02-25 Figure 6/modelRfunctions.R
Add graph function for OTUxCD

@@ -2,7 +2,7 @@
 
 ##########################################################################
 # Inputs 
-# Assumes the data input has columns "nextDayCFU  Otu1  Otu2 ..."
+# Assumes the data input has columns "Group nextDayCFU  Otu1  Otu2 ..."
 #
 # Returns 
 # 
@@ -27,7 +27,7 @@ RF.validate <- function(data, iters){
     #plot(rf.trainSet)
     #varImpPlot(rf.trainSet, type=1)
     
-    plot.title <- "RF.model Validation: trained on 2/3 data, tested on 1/3"
+   # plot.title <- "RF.model Validation: trained on 2/3 data, tested on 1/3"
     rsqs <- c(rsqs,rf.predict(data=testSet, descr="rf.trainSet on Test Set Data", rf.model=rf.trainSet, title=plot.title, plotgraph = FALSE))
   } #for(i in i:iters)
   
@@ -36,10 +36,77 @@ RF.validate <- function(data, iters){
   
 }
 
+##########################################################################
+# Inputs 
+# Assumes the data input has columns "nextDayCFU  Otu1  Otu2 ..."
+#
+# Returns 
+# 
+#  
 
-
-
-
+graphOTUxCD <- function(otus, labels, corrs){
+  
+  par(mar=c(0.5,0.5,0.5,0.5))
+  design <- matrix(1:8, nrow=4, byrow=T)
+  design <- cbind(c(9,9,9,9), design)
+  design <- rbind(design, c(0,10,10))
+  design <- cbind(design,c(11, 11, 11, 11, 0))
+  layout(design, widths=c(0.2,1,1, .5), heights=c(1,1, 1, 1,0.3))
+  
+  for(i in 1:(length(otus))){
+    
+    otulog <- toptitdel[,otus[i]]
+    otulog <- otulog/1625
+    zeroes <- otulog == 0
+    x_zeroes <- runif(sum(zeroes),.000075,.00016)
+    
+    #plot controls
+    plot(otulog[!zeroes & expgroup=="control"], actual[!zeroes & expgroup=="control"],
+         log="x", ylim=c(0,10), xlim=c(0.00008, 1), main="",
+         ylab="", 
+         xaxt='n', yaxt="n", xlab="")
+    
+    #plot zeroes in jitter
+    points(x=x_zeroes, actual[zeroes], pch=pch[as.character(results[zeroes,1])],
+           col=colors[as.character(results[zeroes,1])], cex=1.5)
+    
[…]

### 2015-02-25 Figure 6/RFmodels.r
Update RF model script for analysis

@@ -1,3 +1,7 @@
+
+
+
+
 # ############################
 # # Random forest models 2/20/15, using otus=50 (0.875% cutoff) 
 # ############################
@@ -59,18 +63,261 @@ otusAtPercAbund <- function(percThresh, meanOTU){
   return(otus)
 }
 
-otus <- otusAtPercAbund(1, meanOTU)
-
-rf.results<-RF.analysis(dataFile = "~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.csv", 
-            otus = otus, plot.title = paste("RF model, 1% cutoff relabund otus=", length(otus)))
-
-rf.results.perc5 <- rf.results
-# rf.results.0.875perc
-# rf.results.0.9perc
+perc<-1
+otus <- otusAtPercAbund(perc, meanOTU)
+dataFile <- "~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.csv"
+rf.results<-RF.analysis(dataFile=dataFile, 
+            otus = otus, plot.title = paste0("RF model, ", perc, "% cutoff relabund otus=", length(otus)))
+rf.results.perc1  <- rf.results
+write.table(rf.results.perc1, file="~/Desktop/mothur/abxD01/rf/rfmodels.results.1percent.txt", sep="\t", row.names=T, col.names=T)
 # rf.results.perc1
 # rf.results.perc1.5
-# rf.results.perc3
+# rf.results.perc3, has otu3, but not otu 39
 # rf.results.perc.5
+# rf.results.perc5, doesn't contain otu3 or otu39
+# rf.results.perc2, has both otu3 and 39
+
+all <- read.csv(file=dataFile, header=T)
+
+cols<- NULL
+for(i in 1:(length(otus))){
+  cols <-c(cols, which(names(all)==otus[i]))
+}
+cols <- c(which(names(all)=="Group"), which(names(all)=="nextDayCFU"), cols)
+
+topdose <- all[1:99,cols]
+titr <- all[100:153,cols]
+del <- all[154:179, cols]
+toptit <- all[1:153,cols]
+titdel <- all[c(100:153, 154:179),cols]
+topdel <- all[c(1:99, 154:179),cols]
+toptitdel <- all[,cols]
+
+tdOTUs <- c("Otu00007", "Otu00003", "Otu00015")
[…]

### 2015-02-25 abxD01_analysis.Rmd
Update to papers mkdwn pipeline

@@
 
-stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", divide = TRUE)
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", excludeGroup="untr", divide = TRUE)
 
@@
 
+This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole.
+
 ### Figure 5: Random Forest Analysis
@@
 #most.imp<-rownames(imp[order(imp[, "%IncMSE"], decreasing=T)[1:17],])
+
 ```
@@
 
-Input files found on github:
-1. abxD01.final.tx.2.subsample.2.metro.relabund.D0s.csv
+The file/fileIDS should be matched.
+Input files on github folder Figure 5:
+1. abxD01.final.tx.2.subsample.alldelay.forlogscale.csv
+2. alldelay_tx2_barchart_ids.csv
+
+Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
+
+```{r, eval=TRUE}
+
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 5/alldelay_tx2_barchart_ids.csv", divide = TRUE)
+
+```
 

### 2015-02-26 Figure 6/RFmodels.r
Make increase in node purity graph for model

@@ -145,13 +145,17 @@ print(toptitdel.rf)
 plot(toptitdel.rf)
 varImpPlot(toptitdel.rf, type=1)
 imp<-importance(toptitdel.rf)
-imp<-imp[order(imp[,1], decreasing = TRUE),]
-topimp<- imp[1:15,]
+imp1<-imp[order(imp[,1], decreasing = TRUE),]
+imp2<-imp[order(imp[,2], decreasing = TRUE),]
+topimp <- imp1[1:15,]
+toppurity <- imp2[1:15,]
 write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)
+write.table(toppurity, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importanceNode.txt", sep="\t", row.names=TRUE)
+
 
 #Plot importance Plot
 topimp<-topimp[order(topimp[,1], decreasing=FALSE),]
-ids <- read.csv(file = "~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
+ids <- read.csv(file = "~/Documents/Github/abxD01/Figure 6/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
 #ids <- ids[order]
 labels<- paste0(ids$name, " (", ids$otuname, ")")
 par(mfrow=c(1, 1)) #+1 to give extra labeling space
@@ -159,6 +163,18 @@ par(mar=c(5, 15, 0.5, 2) +0.1, mgp=c(3, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/
 plot(topimp[,1], 1:15, xlab="% Increase in Mean Squared Error", yaxt="n", ylab="", pch=16, cex=1.5, xlim=c(20, 80))
 axis(2, at = c(1:15), labels = labels, las=1, cex.axis=.9 )
 abline(h=c(1:15), lty="dashed", col='black')
+
+
+#Plot importance Plot by node purity
+toppurity<-toppurity[order(toppurity[,2], decreasing=FALSE),]
+ids <- read.csv(file = "~/Documents/Github/abxD01/Figure 6/rf.toptitdel.1p.importanceNode.ids.csv", header = TRUE)
+labels<- paste0(ids$name, " (", ids$otuname, ")")
+par(mfrow=c(1, 1)) #+1 to give extra labeling space
[…]

### 2015-02-26 Figure 6/modelRfunctions.R
Edit Figure 6 graphs: labels etc

@@ -43,15 +43,25 @@ RF.validate <- function(data, iters){
 # Returns 
 # 
 #  
+graphOTUxCD(otunames, otulabels, corrs)
 
-graphOTUxCD <- function(otus, labels, corrs){
+pretty_otus <- as.data.frame(gsub("Otu0*", "OTU", otus))
+otulabels <- cbind(ids,pretty_otus)
+names(otulabels) <- c("names", "otus")
+
+
+otus<-otunames
+ids<-otulabels
+
+
+graphOTUxCD <- function(otus, ids, corrs){
   
   par(mar=c(0.5,0.5,0.5,0.5))
   design <- matrix(1:8, nrow=4, byrow=T)
   design <- cbind(c(9,9,9,9), design)
   design <- rbind(design, c(0,10,10))
   design <- cbind(design,c(11, 11, 11, 11, 0))
-  layout(design, widths=c(0.2,1,1, .5), heights=c(1,1, 1, 1,0.3))
+  layout(design, widths=c(0.2,.9,.9, .6), heights=c(1,1, 1, 1,0.3))
   
   for(i in 1:(length(otus))){
     
@@ -79,9 +89,16 @@ graphOTUxCD <- function(otus, labels, corrs){
     abline(v=c(0.005, 0.05, 0.5), lty="dotted", col="gray")
     abline(v=c(0.001, 0.01, 0.1, 1), lty="longdash", col="gray")
     abline(v=0.0005, lty="dotted", lwd=2, col="black")
-    text(x = 0.00005, y = 9.3, labels = paste(labels[i], ", \u03c1 = ",signif(corrs[i, 2], 3)), 
-         cex = 1, pos=4, font=2)
     
+    test <- grep("aceae", ids[i, 1])
+    name<-as.character(ids$names[i])
+    number<-as.character(ids$otus[i])
+    corr<-as.character(signif(corrs$corSpear[i], 2))
+    graphTitle <- bquote(bolditalic(.(name)) *" " *bold(.(number)) *bold(", ") *bold(rho) *bold(" = ") *bold(.(corr)))
+    text(x = 0.00005, y = 9.3, labels = graphTitle, 
+         cex = 1, pos=4, font=2)
+  
+    #paste(labels[i], ", \u03c1 = ",signif(corrs[i, 2], 2))
     #bquote(bold(.(labels[i])) ~ bold(", ") ~ bold(rho) ~ bold(" = ") ~ bold(.(signif(corrs[i, 2], 3))))
[…]

### 2015-02-26 R scripts/stacked.barcharts.r

@@ -295,6 +295,7 @@ stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, g
   #####################################################################################################################
   #####################################################################################################################
   ###PLOT PARAMETERS
+  library(plotrix)
   if(graphbyphyl==FALSE && divide==FALSE){
     par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
     par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)

### 2015-02-26 abxD01_analysis.Rmd
Re-render markdown

@@
 The parameters I changed in the r script were:
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
-
+library(plotrix)
 # Generate the 
@@
 
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 
@@
 ```{r, eval=TRUE}
-file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.allmetro.forlogscale.fig5.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx2.subsample.alldelay.relabund.csv", header=T)
 file<-file[file$expgroup!="1untr",]
-avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
-metro_avgCD<-format(avgCD, scientific=TRUE, digits=2)
 file$nextDayCFU <- log10(file$nextDayCFU+1)
-#kruskal.test(file$nextDayCFU ~ file$expgroup)
-# p < 0.0001
-#strep_CFU_wilcox <- pairwise.wilcox.test(file$nextDayCFU, file$expgroup, p.adj="BH")
-# all significantly different
-metro.aov<-aov(file$nextDayCFU ~ file$expgroup)
-anova(metro.aov)
-#TukeyHSD(strep.aov)
-metro_ttest <- t.test(file$nextDayCFU~file$expgroup)
-print(metro_ttest)
-# both significantly different
+
+metro <- file[file$abx=="Metronidazole", 3:4]
+amp <- file[file$abx=="Ampicillin", 3:4]
+
+diffs <- function(data){
+  data.aov<-aov(data$nextDayCFU ~ data$expgroup)
+  print(anova(data.aov))
+  pairwise.t.test(data$nextDayCFU, data$expgroup, p.adjust="BH")
+}
+
+diffs(metro)
+diffs(amp)
+
 ```

### 2015-02-26 PAPER.Rmd
Add an .Rmd version of the paper

---
title: "abxD01 Paper"
author: "Alyxandria Schubert"
date: "February 26, 2015"
output: html_document
---

# Abstract

# Introduction

The microbiome, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection. This native protection by the microbiome from invasive pathogenic species is termed colonization resistance. It is well established that the gut bacterial microbiome is critical in the hosts’ defense against the pathogen Clostridium difficile. When this indigenous community is perturbed, this often leads to a loss of resistance. This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, chemotherapy, but also exposed to the environmental C. difficile spores. Among hospital-acquired infections in the United States, C. difficile infections (CDI) are the number one reported and are responsible for 14,000 deaths a year. It is not completely understood how different perturbations affect subsequent colonization resistance, nor do we have a comprehensive view of the taxonomic changes associated with a loss in colonization resistance.

There is a need for more knowledge about interactions between members of the community. Know that fmts work [anna’s paper], but don’t know why, except that you recover the diversity and are receiving an established community already that works. Individual probiotics have not worked with complete success [dendukuri 2005], whereas studies find that combinations of bacteria often work better than by themselves [lawley, pamer, stein, hickson 2007]. It’s also common for bacteria to change their behavior in the context of other bacterial species [Crost 2010, quorum sensing-biofilms, recognizing fellow bacterial species vs killing others, synergistic effects (sieuwerts 2010)]. Understanding the scope of relationships between bacteria in a community is necessary to learn about its response, in concert with its host, to perturbations including antibiotic-induced, pathogen invasion, probiotics doses, and diet changes, etc.

The purest way of studying these bacterial interactions is by characterizing them in the context of their whole biome [ecology citation?]. [examples of papers showing microbiomes response to antibiotics in the context of CR]

These observed changes in response to perturbation can be used to inform models and networks to predict future responses (stein, Gordon diet paper) and characterize bacterial interactions (psoriasis/Schubert, sparCC paper). These methods are advantageous in that they can be refined to determine probiotic prophylactics or treatments based on an individual patient’s gut microbial needs.

The purpose of this investigation was to test the effects of an array of perturbations on microbiome CR levels against C. difficile.  We then used this information to design microbiome-based models for estimating pathogen colonization levels. Through the administration of various antibiotic regimens, we altered the murine gut microbiota and observed these changes using 16S rRNA sequencing. The resultant gut communities were then challenged with C. difficile spores. Through correlation and random forest analysis of the starting communities with the subsequent level of C. difficile colonization, we built linear models to predict C. difficile colonization levels based on a subset of the microbiota. We used experimental data to test the ability of the model.

Though we are limited in a resolution at the strain and species level, we have identified groups of related species with antagonistic relationships with C. difficile colonization. We identified several models with equally robust quality statistics and observed that across the top models a subset of 5 species were consistently present. These findings __

# Materials and Methods

## Animal Care

We used 5-8 week old C57Bl/6 mice for all our experiments. These mice were housed in our SPF room in the animal facility at the University of Michigan. All animal-related protocols and experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.

## Antibiotic Administration

Mice were administered one of seven different antibiotics, including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin. The route of administration depended on the specific antibiotic. Cefoperazone (either 0.5mg/ml, 0.3mg/ml, or 0.1mg/ml), vancomycin (0.625mg/ml, 0.3mg/ml, or 0.1mg/ml), streptomycin (5mg/ml, 0.5mg/ml, or 0.1mg/ml), metronidazole (0.5mg/ml), and ampicillin (0.5mg/ml) were all administered in the mouse drinking water for 5 days. Ciprofloxacin (10 mg/kg) was administered via oral gavage, and clindamycin (10 mg/kg) was administered via IP injection. All antibiotic treated mice were given 24 hours of recovery on regular drinking water prior to C. difficile challenge.

## C. difficile Preparation and Challenge

In all our experiments we used C. difficile strain 630Δerm [cite dena lyras]. The spores used were prepared from a single large batch made following previously established protocols [cite paul?]. Spores were quantified within the week prior to C. difficile challenge for each experiment [established protocol I can cite?]. Spores were quantified by boiling a small volume at 65°C for 20 minutes to kill off residual vegetative cells. The volume was then diluted and plated on TCCFA plates. These were incubated at 37°C anaerobically overnight. The next day C. difficile colony forming units (CFU) were counted and the concentration of spores in the working stock was back calculated. On the day of challenge, C. difficile spores, at a final concentration of 103, were administered to mice via oral gavage. Immediately following this challenge, the remaining C. difficile inoculum was diluted in a series and plated to confirm the correct dosage.

## Sample Collection and Plating

Fecal samples were freshly collected for each mouse immediately prior to the administration of any antibiotic, during any antibiotic recovery days, and days 0 and 1 post C. difficile challenge. Each fecal sample on day 1 was weighed and diluted under anaerobic conditions with anaerobic PBS 1:10 based on pellet weight. A dilution series was then plated onto TCCFA plates [cite plates] and incubated at 37°C anaerobically overnight. The colony forming units (CFU) were counted and calculated the subsequent day.

## DNA Extraction and Sequencing

Total bacterial DNA was extracted from each untreated and day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit. Sequencing preparation and sequencing was carried out as previously described (Kozich JJ et al. 2013). Briefly, the V4 region of the 16S rRNA gene was amplified, purified, and pooled in equimolar concentrations. These libraries were then mixed with Illumina-prepared PhiX control libraries. Sequencing was performed using Illumina’s MiSeq platform to produce paired 250-nt reads. Mock/error/error of run?

## Sequence Curation

These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al). The full pipeline is also demonstrated online at the mother website (http://www.mothur.org/wiki/MiSeq_SOP). Sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff. Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 9) requiring an 80% bootstrap confidence score. In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples [Schloss, reducing effects]. For those samples with more than 1,625 sequences per samples, the range was __, the mean was __, the median was __, the median absolute deviation was __. All 16S rRNA gene sequence data and the MIMARKS table are available at __.

## Statistical Analysis & Modeling

Using data from the original antibiotic treatments groups, OTUs were selected for addition into the model following several criteria. First all OTUs were present in abundance at an average of 1% in at least one experimental group.

All statistical analyses were conducted using R version 3.1.2.

# Results

## Levels of colonization resistance are associated with the initial structure of the gut microbiota.

In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to C. difficile colonization. It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] all have decreased resistance against this pathogen. To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (Table 1) to differentially alter the microbiota and assess their resistance to C. difficile colonization. 16S rRNA gene sequencing of the fecal communities sampled at the time of C. difficile inoculation revealed distinct and reproducible microbiota structures (Figure 1). The community structures of the beta-lactam-treated mice (i.e. cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=0.36). Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=0.001) and were characterized by the predominance of Enterobacteriaceae. Of all groups significantly different from the untreated microbiota, the levels of Enterobacteriaceae increased with all treatments except with streptomycin. Vancomycin-treated communities saw the largest bloom in Akkermansia. In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=0.09). Subsequently, 24 hours after inoculation C. difficile was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all of the other treatment groups to similarly high levels (107-108 CFU/g) (Figure 1). We calculated the diversity of all of the communities using the Inverse Simpson (Supplemental Figure 1) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent C. difficile colonization (both p<0.001). These results suggest that multiple microbiota structures are capable of being colonized by C. difficile and that diversity may be an important factor in resistance to invasion.

## Correlation analysis reveals potentially protective bacteria.

To identify bacterial taxa that could be associated with resistance or susceptibility to C. difficile, we measured the correlation between the relative abundance of each species-level OTU on the day of inoculation with the level of C. difficile colonization 24 hours later across all treatment groups (Figure 2). Of the significant OTUs (p<0.05) with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against C. difficile (N=40) and those with a positive correlation were associated with susceptibility (N=7). Among various bacterial families, three were consistently negatively correlated with C. difficile colonization: Porphyromonadaceae (ρaverage= -0.63), Lachnospiraceae (ρaverage= -0.51), Ruminococcaceae (ρaverage= -0.53). OTUs from the Proteobacteria phylum had strong positive relationships with C. difficile colonization. These included OTUs associated with the Pseudomonas (ρ= 0.31) and Escherichia (ρ= 0.56) genera. Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus Lactobacillus (family Lactobacillaceae; OTU23, ρ= -0.73), Alistipes (family Rickenellaceae; OTU20, ρ= -0.78), Enterorhabdus (phylum Actinobacterium, family Coriobacteriaceae; OTU29, ρ= -0.70), and Streptococcus (family Streptococcaceae; OTU78, ρ= 0.36). Overall, these results suggest that several larger groups of related OTUs, particularly within the Firmicutes and Bacteroidetes phyla, may participate in colonization resistance against C. difficile, while members of the Proteobacteria phylum may be involved in susceptibility.

## Smaller perturbations of the microbiota result in maintenance of colonization resistance.

transition
Because most of our original treatments resulted in similarly high levels of C. difficile, we hypothesized that applying smaller antibiotic perturbations to the microbiota would result in reduced sensitivity to C. difficile, as we observed in the ciprofloxacin treated mice. To test this, we titrated the doses of cefoperazone, streptomycin, and vancomycin given to mice (Supplemental Table 1). Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis. C. difficile levels were significantly different between all titration groups within the cefoperazone (p<0.03) and streptomycin treated microbiotas (p<0.01), displaying a greater range of colonization resistance profiles (Figure 3). Next, we hypothesized that the populations we previously identified as being associated with protection against colonization would be elevated in the mice with decreased colonization and those associated with susceptibility would be less abundant.

The titrated antibiotic doses clearly resulted in distinct community structures that were variations of the original dose (Figure 3). The cefoperazone titrations resulted in a shift in dominance from the Bacteroidetes to the Firmicutes phylum with increasing antibiotic dose. Most notable was the increase in OTUs associated with resistance including the Porphyromonadaceae and Alistipes and a reduction in OTUs associated with susceptibility including members of the Proteobacteria, Clostridium, and Staphylococcus. The relative abundance of other OTUs associated with protection including the Lactobacillus, Lachnospiraceae, and Ruminococcaceae did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice. The streptomycin titrations did not result in profound shifts in the overall community structure, but did yield increased resistance. Rather, there were individual populations that were gained or lost across the gradient. Notably, mice receiving the lowest dose retained significantly higher levels of Alistipes and Anaeroplasma (p<0.05), which we previously found to be strongly associated with colonization resistance. The relative abundances of the Porphyromonodaceae, Lachnospiraceae, Ruminococcaceae were unchanged, but were present at similarly high levels to those found in the untreated mice. Although the vancomycin titration yielded a number of large changes in the abundance of individual populations, there was no recovery of colonization resistance. The relative abundance of many of the populations associated with resistance (Porphyromonadaceae, Lachnospiraceae, Anaeroplasma) increased with decreased vancomycin dosage; however, with the exception of the Lactobacillus populations, they did not reach the levels observed in the untreated mice. In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the Akkermansia and Enterobacteriaceae. Together, these results suggest that individual populations are not sufficient to provide colonization resistance. Rather, resistance is likely a product of the overall composition of the community.

## Allowing recovery of specific bacteria restores colonization resistance.

Several antibiotic perturbation studies have demonstrated that given time perturbed communities can return to a “healthy” state in which resistance to C. difficile is restored []. For each antibiotic treatment, a control group of mice receiving antibiotics without inoculation of C. difficile was maintained. The microbiota’s of animals treated with metronidazole underwent a drastic shift in the microbiota structure with an additional 5 days of recovery. We observed changes in relative abundance of several OTUs that we previously observed had strong relationships with subsequent C. difficile colonization (Figure 3). Notably there were significant increases (p<0.05) in overall abundance of the levels of Porphyromonadaceae, Ruminococcaceae, and Lachnospiraceae to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of Enterobacteriaceae. Based on these changes, we hypothesized that delaying C. difficile challenge to allow these potentially important changes in microbiota structure to occur would lead to a decrease in colonization. On Day 1 post challenge of these recovered mice, we observed a significant decrease in C. difficile level of CFU/g feces (p<0.0001). These results strengthen the importance of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Enterobacteriaceae in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of C. difficile colonization.

Additionally, we assessed the resistance of ampicillin treated mice given five extra days of recovery prior to C. difficile challenge. The shift in the microbiome with ampicillin recovery was not as prounounced as seen with recovery of metronidazole treated animals. These communities had significantly reduced levels of bacteria correlated with C. difficile protection, including Bacteroides, Alistipes, and Ruminoccocaceae compared with untreated animals (p<0.05). Importantly, the level of Enterobacteriaceae was not significantly reduced with additional recovery time. Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent C. difficile levels.

## Modeling to predict C. difficile colonization outcomes.

Given our ability to make general predictions of C. difficile colonization based on the microbiota, we next designed a model to calculate expected colonization levels given the relative abundance of a subset of OTUs. Combining the three datasets (Table 1, Supplemental Table 1 and 2), we set a 1% relative abundance threshold, eliminating OTUs that on average were not present at this level in at least one of the 16 experimental groups. In total, there were 44 OTUs utilized for building a random forest regression model. This model fit the data with an R2 of 0.98 (Figure 5A). We identified OTUs that contributed most to increased accuracy of the model using random forest’s feature selection (Figure 5B). Many of the top contributing OTUs includes members of the Porphyromonadaceae family, Alistipes, Lachnospiraceae, Lactobacillus, and Escherichia taxa. The relationship between these OTUs and C. difficile is shown in Figure 5C. Escherichia had the largest positive correlation with C. difficile levels of any bacterial species.  This finding was consistent across susceptible antibiotic treated communities except for streptomycin treated communities, which altogether lacked Enterobacteriaceae spp. OTUs 3 and 39 (Porphyromonadaceae and Lachnospiraceae, respectively), were the two most important features in accurately predicting subsequent C. difficile levels. In fact, a model built with only these two features was able to account for 79% of the variance observed. Together these results suggest that colonization resistance is likely conferred by porphyromonadaceae, Alistipes, and lachnospiraceae and a loss in these bacteria, concurrently with a gain of Escherichia, can result in increased susceptibility to infection.

# Discussion

-community effect—have other things to compensate—highlight examples –streptomycin blue triangles—where low lachno but no cdiff so what else is protecting? Not one otu that explains it all… just cus low levels doesn’t screw you, also see recovered points in 5C-angela, sonnenburg-need other thingsssss
  -fig2—there are like larger groups in protection/susceptibility
	-combo of loss of good bugs, and gain of bad—a model of infection
-tie back to mbio paper, with enterobacteriaceae they are screwed
-modeling—not much done for CDI, link back to humans, shows we could model risk as they come in—put towards end with the larger picture
We designed statistical models based on a data set of varied antibiotic perturbations of the murine microbiota and their subsequent CR levels against C. difficile. Though antibiotic administration induced unique microbiota structures, colonization resistance was similarly low against C. difficile. 
These models incorporated all combinations of top candidate species-level OTUs. We then tested our top model’s ability to use new abundance information on a specific subset of the microbiota to predict the subsequent C. difficile colonization level, which was experimentally validated. These results greatly expand our knowledge of various perturbations and their subsequent effects on microbiota structure and function (colonization resistance ability). Furthermore modeling important/key bacterial subsets of the community we can build models to assess CDI risk.

Through our perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of C. difficile colonization. A large number of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Clostridia, as well as Alistipes were found to have strong potential resistance abilities against C. difficile. Conversely, several members of the Proteobacteria phylum, including Pseudomonas, Streptococcus, and especially E. coli have strong potential positive impacts for C. difficile colonization. These results were supported by their strong correlation values supported across several antibiotic regimens. They are also supported through experiments in which the original doses of cefoperazone, streptomycin, and (vancomycin) were titrated to alter the overall abundances but not membership of the community from the starting dose, as has been observed previously (Sekirov). These subtler structural changes allowed for greater range of colonization levels of C. difficile as well as further/fine-tuned support for our original results. Furthermore, by allowing a known change to occur in important bacterial populations within metronidazole treated mice, we were able to reduce the level of colonization by C. difficile. Lastly we showed their importance through modeling techniques that statistically confirm these results.

Previous work characterizing the microbiome’s of human and mouse hosts in the context of CDI provide glimpses into potentially protective and alternatively susceptible community structures, i.e. which bacteria are present and their abundance. Cross-sectional investigations have characterized the microbiota of patients in the hospital with and without a C. difficile infection (CDI) and compared to healthy controls. These studies found patients with C. difficile tended to have a reduction in numbers of Clostridiales Incertae Sedis XI (Vincent et al), Lachnospiraceae, Ruminococcaceae, and Bacteroides (Schubert et al). These studies however lack the baseline healthy state of the microbiota, which preclude them from assessing risk of C. difficile. Murine models of CDI have the advantage of observing the specific changes that occur before and after controlled perturbations and the introduction of C. difficile. Communities with low levels of CR were associated with blooms in either Lactobacillaceae (cefoperazone-induced) (Theriot, 2014, bassis, 2014), Proteobacteria (induced by either clinda+5abx cocktail, clinda alone, or tigecycline) (Reeves, 2011, Buffie, 2011, Bassis 2014). Lachnospiraceae is associated with resistance to C. difficile infection (Reeves 2011) and has been directly shown to decrease colonization through mono-colonization studies in germfree mice (Reeves 2012).

Even fewer studies attempt to model these observational changes in the microbiome following perturbation. Using the time series data gathered from the C. difficile challenge of clindamycin-treated mice (Buffie 2011), a Lotka-Volterra dyamics-based framework was built which incorporates both time and accounts for external perturbations (Stein 2013). They identified a subset of their network, containing Blautia, Akkermansia, and Coprobacillus, as having a negative relationship with C. difficile, while alternatively Enterococcus had a positive relationship.

------------------

o  How relates to other studies
•	Antonopoulos et al-reproducible community dynamics
•	Antunes et al (finlay)—abx treatment on intestinal metabalome
•	Sekirov—antibiotic induced perturbations of intestinal microbiota
•	 “Change in the compostion alone of the intestinal microbiota (in the absence of significant changes in total numbers of intestinal microbes) prior to Salmonella serovar Typhimurium infection was sufficient to make mice more susceptible to Salmonella serovar Typhimurium, indicating that different subsets of the microbiota may play protective roles or enhance susceptibility to enteric infections.”
•	Streptomycin and vancomycin doses were administered to mice prior to challenge with another gastrointestinal pathogen Salmonella enterica serovar Typhymurium (Sekirov 2008).  This study showed ___. Changes in community composition were dose dependent. 
•	Strep increases bacteroidetes (CFB), vanc increases firms and then proteo (with increasing dose)
•	Reeves/theriot/robinson—clindamycin/cef/cocktail/antbiotic administration alters the community structure
•	Croswell—prolonge impact of abx on intestinal microbial ecology salmonella
•	Rea- effect of  broad and narrow spectrum antimicrobials
•	Ferreira—intestinal microbiota plays a role in salmonella induced colitis
•	Reeves with clone libraries and cef
•	Theriot?—at least the metabolome paper—before and after with metabolites and cef, also does she look at anything taxonomically? yup

Many of the relationships we observed in our study between specific bacteria and C. difficile are supported in the literature. –see my comments in this section
Within the Porphyromonadaceae, only OTU505 had a significant positive relationship with subsequent C. difficile colonization (ρ= +0.21), which by blastn shares 99% identity with Coprobacter fastidiosus. Similarly within the Lachnospiraceae family, only one OTU had a positive relationship with C. difficile colonization (OTU174, ρ= +0.22). The Lachnospiraceae group also contained an OTU that had an exceptionally strong negative correlation (OTU39, ρ= +0.77). The Clostridia class was largely composed of the order Clostridiales (n=6), of which only one had a positive relationship with C. difficile colonization (OTU154, ρ= +0.26).

Several broader groups of bacteria can work together (in some combination) to provide CR …possible that proteobacteria in general are bad …These results indicate that many related species within the larger/broader bacterial groups may be able to serve in providing colonization resistance against C. difficile.
Other taxons in Figure 3 had more species specific correlations. For instance the Actinobacteria phylum included Enterorhabdus (OTU29, ρ= -0.70), Coriobacteriaceae (OTU50, ρ= -0.50), Bifodbacterium (OTU22, ρ= -0.34), Actinomyces (OTU58, ρ= +0.23). The Lactobacillales group is divided into two negatively correlated Lactobacillus OTUs (OTUs 23, 21, ρ= -0.73, -0.40 respectively) and two positively correlated Streptococcus OTUs (OTUs 78, 79). Similarly, the Bacteroidales group included species at both extremes, for example, an Alistipes (OTU20, ρ= -0.78) and a Bacteroides species (OTU44, ρ= +0.34). [END sent, all the other bad bugs look more like random smaller level taxonomic classification divisions, as in it’s a genus/species level effect type deal] 
--functional redundancy? Across related large groups like porphyros?
--mostly Firmicutes—better competitors with cdiff? Because similar nutrients resources? 
Also functional redundancy with a highly diverse community… not just about who possibly but also ab being diverse—emphasizes community as a whole

OTU39—Eubacterium ventriosum (94%), eubacterium rectale (93%), roseburia intestinalis/hominis/faecis (93%) (positively correlated to differences observed in discordant MZ twins for BMI—more abundant in higher BMI siblings… these are all butyrate producers, also capable of degrading fiber)
	FOS—fructooligosaccharides produced by degredation of inulin or polyfructose

Inulin and fructo-oligosaccharides have divergent effects on colitis and commensal microbiota in HLA-B27 transgenic rats.
Abstract
Modulation of intestinal microbiota by non-digestible carbohydrates may reduce inflammation in inflammatory bowel disease (IBD). The aim of the present study was to assess the effects of inulin and fructo-oligosaccharides (FOS) on intestinal microbiota and colitis in HLA-B27 transgenic rats, a well-validated rodent model for IBD. In this study, 4-week-old rats were fed 8 g/kg body weight inulin or FOS for 12 weeks, or not. Faeces were collected at 4 and 16 weeks of age; and caecal samples were collected at necropsy. The effects of inulin and FOS on chronic intestinal inflammation were assessed using a gross gut score, histology score and levels of mucosal IL-1β. Intestinal microbiota were characterised by quantitative PCR and denaturing gradient gel electrophoresis. Colitis was significantly reduced in all FOS-fed rats compared to the control diet, whereas inulin decreased chronic intestinal inflammation in only half the number of animals. Quantitative analysis of caecal microbiota demonstrated that inulin increased the numbers of total bacteria and the Bacteroides-Prevotella-Porphyromonas group, FOS increased bifidobacteria, and both fructans decreased Clostridium cluster XI. In the faecal samples, both inulin and FOS decreased total bacteria, Bacteroides-Prevotella-Porphyromonas group, and Clostridium clusters XI and XIVa. FOS increased Bifidobacterium spp., and mediated a decrease of gene copies of Enterobacteriaceae and Clostridium difficile toxin B in faeces. SCFA concentrations in the faecal and caecal samples were unaffected by the diets. In conclusion, FOS increased the abundance of Bifidobacterium spp., whereas both fructans reduced Clostridium cluster XI and C. difficile toxin gene expression, correlating with a reduction of chronic intestinal inflammation.

Any interesting points about the results?  Mechanisms? Competition—ask matt if they are good?  Proteobacteria could definitely play a role in susceptibility—creating an environment fit for cdiff… what ways could all those other bugs be making a bad environment for cdiff? Using up bile acids/germinants..,  with those bugs gone freeing up sugars, … maybe connecting a lot more to some of caseys metabolome paper, and how mine would fit that

o	Limitations
•	No smaller intervals? Of cdiff levels, or changes in microbiota
•	Porphyromonadaceae is more prevalent in mice… so might not be as relevant for humans
•	Different strains of cdiff have different microbiota requirements to protect?
•	Doesn’t get at function… ?
•	Collect more data to continue to improve our microbiome modeling abilities,

o	Contribution to the field, positive spins
•	PROBIOTICS
•	Possible global effects? CR against other GI pathogens?
•	clinical applications—PCR diagnostic for risk

The final paragraph should highlight the main conclusion(s), and provide some indication of the direction future research should take. (for and I and I mention host)
-therapeutic: continuously supply probiotics  while lowering inflammation of the host (get host side back to normal state in tandem, via drugs? diet)… so attack the issue of CDI from both the microbiota end and host end… microbiota-host-pathogen triangle… …squeeze it out (eg. clindamycin clearance? If compared pathology of host at diff days along 10d for clindamycin vs cef/met/amp (these other ones that are high in cdiff coloniz), is host or toxin diff?... why does clinda clear after being CDhigh but no other ABX?)

The human microbiome is complex; it has been observed that there is no identifiable “core” microbiome that all humans share (Turnbaugh 2009). It’s been observed that there is a high level of interindividual variation among healthy individuals’ microbiomes (Costello 2009). Thus the definition of a healthy microbiome is one that encompasses a wide range of microbiota structures, each one evolved with their host. It’s highly likely there is no one size fits all solution for the treatment or prevention of CDI. Thus characterizing the relationships within the microbiome are important in the context of a community. Need more studies that look at the level of the community as a whole in CR over the effect of single bugs.  Examples of papers looking at combinations/subsets of bacteria in CR

	It makes sense that many bacteria would play a role in colonization resistance and that it is more about the community composition as a whole given that interindividual differences is so great.  Two people share no more than __% of their microbiomes.  They are unique to individuals and yet all these different profiles of microbiotas are capable of colonization resistance.  People do not get C. difficile infections without first being exposed to some major purturbation of their microbiome, namely antibiotics.  Thus many structures of the microbiome, and many distinct bacterial populations must be capable of contributing to resistance.  Identifying/cataloguing a comprehensive profile of those bacteria with potential roles in CR would be an initial step towards personalized/individualized treatments with known mixtures of protective bacteria. 
Bacteria within a community setting may act differently in the context of other bacterial species.

[using this time model for the metro… I wonder if it doesn’t work because of the time difference in the amount of recovery time? I’m picturing sort of an echo/ripple effect where at first the waves, or changes, are really large and then over time that rate of change of the community slows and the changes less pronounced/smaller. In this context the metro communities were at a stage farther along in the echo, meaning changes not happening as fast. Whereas the measurements we’ve taken and are using to make predictions are from the community at a time when you expect more/bigger/faster changes. So then the results are calculated assuming one rate but actually in a different (slower). 
By this image, the initial strength of the perturbation matters too.]

--also somewhere, ecoli are typically on avg (__% of the community, typically wayyyy outnumbered… whats the avg relabund of all those CR OTUs in an untreated mouse, how does that compare to the abundance of those bacteria I’ve implicated in –CR) …point is that Proteobacteria are usually outnumbered… don’t have a chance, gut is suited to bact/firm (which not surprisingly make up the major players in CR).. I bet those other outlier bad bugs are also in low abund in untreated mice (?). but when the gut environment changes for w/e reason, can give those rarer bad ones/proteo/esp e. coli a chance to take hold. They could just be individually opportunistic or work synergistically with cdiff. If individually opportunistic, it might be that they are facing pretty much a common enemy which is the lach/rum/clos/porphs and end up having similar techniques/niches as a result of that common enemy. Working synergistically—eg. b. theta sialidase where it cleaves off resources for cdiff.

### 2015-02-27 PAPER.Rmd
Incorporate R analysis with paper

@@
 
+### CFU stats comparison
+To compare CFU levels: 
+```{r, eval=TRUE}
+
+file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
+
+data <- file[ which(file$expgroup != '1untr'), ] #choose experimental groups
+data$nextDayCFU<-log10(data$nextDayCFU+1) #log transform CFUs
+anova(lm(data$nextDayCFU ~ data$expgroup)) #perform anova across experimental groups for CFU
+
+#Which groups are different? Ciprofloxacin was clearly lower
+data <- data[ which(data$expgroup != '2cipro'), ] #remove ciprofloxacin to see now if there's a difference in remaining groups
+anova(lm(data$nextDayCFU ~ data$expgroup))
+
+```
+
+### Amova to compare resulting community structures
+Amova calculation in mothur to determine the uniqueness of each community structure:
+```
+dist.shared(shared=abxD01.final.an.unique_list.0.03.topdose2.shared, calc=thetayc-jclass, subsample=1625)
+amova(phylip=abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.dist, design=topdose_design_byABX.txt, iters=1000000)
+
+```
+This results in the file called "abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.amova" located in the github folder. The results of these are summarized in "amova8expgroups.xlsx".
+
+### Figure 1: Bargraph
+
+This figure shows the differences in the communities' structures following different antibiotic treatments.
+The average *C.
+difficile* CFU/g Feces is also calculated in this code for each treatment group.
+
+This figure was built using the R script called "stacked.barcharts.r", which contains the stackedbarcharts() funciton in the abxD01 folder.The input files are listed in Github.
+The graph can be sorted 2 ways either by the untreated's decreasing relabund (sortbyphyl=FALSE), or by the most abundant within each phylum (sortbyphyl=TRUE).
+For each treatment, the resulting *C.
+difficile* CFU/g Feces is also labeled on the Y axis below the treatment name.
+The parameters I changed in the r script were:
+```{r, eval=FALSE}
+abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
+library(plotrix)
+# Generate the 
+stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+
+stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+
+```
+
+Files saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf".
+
+### Supplemental Figure 1: correlation plot of diversity with *C. difficile*
+
+For the correlation calculations see the next section of the results.
+Figure making:
+
 ## Correlation analysis reveals potentially protective bacteria.
@@
 
+### Correlation Calculations
+
+The required file is located on Github.
+
+```{r, eval=TRUE}
+
+meta <- read.table('~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
+#meta<-meta[1:96] #change based on number of OTUs above .05%, then add 2 for first two cols
+c<-1
+otu <- c()
+cor.spear <- c()
+pval.spear <- c()
+cor.ken = c()
+pval.ken = c()
+for(i in 3:length(meta)){
+  otu[c] <- colnames(meta[i])
+  cor.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$estimate
+  pval.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$p.value
+  cor.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$estimate #good to see because kendall handles ties
+  pval.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
+  c <- c+1
+}
+
+pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
+pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons
+
+results = NULL
+results <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
+colnames(results) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
+results <- results[order(results[,3]),]  #order by pvalue column=3
+#write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
+print(results[1:dim(results)[1],])
+
+```
+
+### Figure 2: Stipchart
+
+This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1 using the original set of antibiotic groups.
+
+Necessary files are on github, and it was made using the following code:
+
+```{r, eval=TRUE}
+
+##Paper Figure: correlations for topdose 
+otuCorrs<-read.csv("~/Documents/Github/abxD01/Figure 2/corr_avgAbove0.001_topdose2_stripchart.csv", header=T)
+
+pch <- c(6,6,6,6, 1,1, 1, 1, 1, 1, 1, 1,0, 0, 9, 2, 2, 4)
+par(mar=c(12, 7, 2, 2) +0.1, mgp=c(4, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
+stripchart(otuCorrs$corr ~ otuCorrs$graphOrder, vertical=TRUE, ylab="Spearman Correlation", ylim=c(-1, 1), cex.axis=1, cex=2, pch=pch, lwd=2, col="black",  method="jitter",  jitter=.2, cex.lab=1.1, yaxt="n", xaxt="n")
+axis(1, cex.axis=1, at=(1:nlevels(otuCorrs$family)), labels=FALSE)
+family.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "family"])
+text(1:nlevels(otuCorrs$family)+.1, par("usr")[3]-0.05, label=(family.labels), xpd=NA, pos=2, srt=45, cex=1)
+axis(2, cex.axis=1, at=seq(-1,+1, by=.1), las=1)
+color_transparent <- adjustcolor("black", alpha.f = 0.3) 
+abline(v=c(1:nlevels(otuCorrs$family)), col=color_transparent)
+abline(h=0, col="dark gray", lwd=2)
+abline(h=c(0.25, 0.5, 0.75, -0.25, -0.5, -0.75), col=color_transparent, lwd=2, lty="dashed")
+phylum.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "phyl"])
+leg<-legend("top", legend=phylum.labels, pch=c(6, 1, 0, 9, 2, 4), cex=0.7, bty="o", horiz=TRUE, pt.cex=1.5, text.width=2.5)
+
+```
+
+Figure saved as "corr_avgAbove0.001_topdose2_10x10.pdf".
+
 ## Smaller perturbations of the microbiota result in maintenance of colonization resistance.
 
-transition
+<transition>
 Because most of our original treatments resulted in similarly high levels of C.
@@
 
+### CFU stats comparison
+
+For comparison of the CFU's among titration groups: 
+```{r, eval=TRUE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
+file<-file[!grepl("untr", file$expgroup),]
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+
+cef <- file[file$abx=="Cefoperazone", 3:4]
+strep <- file[file$abx=="Streptomycin", 3:4]
+vanc <- file[file$abx=="Vancomycin", 3:4]
+
+diffs <- function(data){
+  data.aov<-aov(data$nextDayCFU ~ data$expgroup)
+  anova(data.aov)
+  pairwise.t.test(data$nextDayCFU, data$expgroup, p.adjust="BH")
+}
+
+diffs(cef)
+diffs(strep)
+diffs(vanc)
+
+```
+
+### Figure 3: Barchart
+
+This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.
+
+The file/fileIDS should be matched by the titration.
+Input files on github:
+1. abxD01.final.tx.2.subsample.alltitrations.forlogscale.csv
+2. alltitrations_tx2_barchart_ids.csv
+
+Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
+
+```{r, eval=FALSE}
+
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", excludeGroup="untr", divide = TRUE)
+
+```
+
 ## Allowing recovery of specific bacteria restores colonization resistance.
@@
 We observed changes in relative abundance of several OTUs that we previously observed had strong relationships with subsequent C.
-difficile colonization (Figure 3).
+difficile colonization (Figure 2).
 Notably there were significant increases (p<0.05) in overall abundance of the levels of Porphyromonadaceae, Ruminococcaceae, and Lachnospiraceae to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of Enterobacteriaceae.
@@
 Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent C.
-difficile levels.
+difficile levels (p=0.12).
 
-## Modeling to predict C. difficile colonization outcomes.
+### Differences between recovery days' thetayc
+
+### Statistical comparison of delayed and on time CFU results
+```{r, eval=TRUE}
+file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx2.subsample.alldelay.relabund.csv", header=T)
+file<-file[file$expgroup!="1untr",]
+file$nextDayCFU <- log10(file$nextDayCFU+1)
+
+metro <- file[file$abx=="Metronidazole", 3:4]
+amp <- file[file$abx=="Ampicillin", 3:4]
+
+diffs <- function(data){
+  data.aov<-aov(data$nextDayCFU ~ data$expgroup)
+  print(anova(data.aov))
+  pairwise.t.test(data$nextDayCFU, data$expgroup, p.adjust="BH")
+}
+
+diffs(metro)
+diffs(amp)
+
+```
+
+### Figure 4: Barchart
+
+This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole and ampicillin.
+
+The file/fileIDS should be matched.
+Input files on github folder Figure 5:
+1. abxD01.final.tx.2.subsample.alldelay.forlogscale.csv
+2. alldelay_tx2_barchart_ids.csv
+
+Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
+
+```{r, eval=FALSE}
+
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 5/alldelay_tx2_barchart_ids.csv", divide = TRUE)
+
+```
+
+## Modeling microbiota data can predict C. difficile colonization outcomes.
 
@@
 Together these results suggest that colonization resistance is likely conferred by porphyromonadaceae, Alistipes, and lachnospiraceae and a loss in these bacteria, concurrently with a gain of Escherichia, can result in increased susceptibility to infection.
+
+### Evaluating cutoffs for OTU inclusion in Random Forest model
+
+```{r, eval=TRUE}
+data<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.csv", header=T)
+
+means <- as.data.frame(matrix(nrow=nlevels(data$expgroup), ncol=0))
+row.names(means) <- levels(data$expgroup)
+for(i in 3:length(data)){
+
+  df <- aggregate( data[,i] ~ data$expgroup, data=data, mean )
+  names(df)[2] <- names(data)[i]
+
+  means <- as.data.frame(cbind( means, df[,2]))
+  names(means)[dim(means)[2]] <- names(data)[i] 
+}
+
+means <- rbind(means, max=-1)
+for(j in 1:length(means)){
+  means["max", j] <- max(means[,j])
+}
+
+means <- rbind(means, relabund=-1)
+for(k in 1:length(means)){
+  means["relabund", k] <- (means[17,k])/1625
+}
+
+means <- rbind(means, percent=-1)
+for(k in 1:length(means)){
+  means["percent", k] <- (means[18,k])*100
+}
+
+meanOTU <- means[,-1] #remove the nextDayCFU column
+
+set <- NULL
+set <- seq(0,3, by=0.01)
+numOTUxThreshold <- NULL
+numOTUxThreshold <- as.data.frame(cbind(numOTUxThreshold, set))
+names(numOTUxThreshold)[1] <- "threshold"
+numOTUxThreshold <- cbind(numOTUxThreshold, numOTU=NA)
+for( i in 1:(length(set)) ) {
+
+  test <- meanOTU[ , meanOTU[19,] >= set[i] ]
+  #test2 <- sum( meanOTU[19,] >= set[i] )
+
+  numOTUxThreshold[i, 2] <- length(test)
+
+}
+
+plot(numOTUxThreshold$numOTU ~ numOTUxThreshold$threshold, xlab="Avg % RelAbund Threshold", ylab="# OTUs included")
+abline(v=seq(0,3, by=.1), col="light gray")
+abline(h=seq(1,300, by=10), col="light gray")
+```
+This feeds into the next analysis section.
+
+### Random forest results
+
+```{r, eval=TRUE}
+
+otusAtPercAbund <- function(percThresh, meanOTU){
+  perc <- meanOTU[ , meanOTU[19,] >= percThresh ]
+  otus <- names(perc)[-(length(perc))]
+  return(otus)
+}
+
+perc<-1
+otus <- otusAtPercAbund(perc, meanOTU)
+
+all <- read.csv(file=dataFile, header=T)
+cols<- NULL
+for(i in 1:(length(otus))){
+  cols <-c(cols, which(names(all)==otus[i]))
+}
+cols <- c(which(names(all)=="Group"), which(names(all)=="nextDayCFU"), cols)
+
+# dose <- all[1:99,cols]
+# titr <- all[100:153,cols]
+# del <- all[154:179, cols]
+# toptit <- all[1:153,cols]
+# titdel <- all[c(100:153, 154:179),cols]
+# topdel <- all[c(1:99, 154:179),cols]
+toptitdel <- all[,cols]
+
+toptitdel.rf <- randomForest(nextDayCFU ~ ., 
+                           data = toptitdel[,-1],  outscale=TRUE,
+                           importance=TRUE, proximity=TRUE,
+                           keep.forest=TRUE, ntree=5000
+)
+print(toptitdel.rf)
+plot(toptitdel.rf)
+varImpPlot(toptitdel.rf, type=1)
+imp<-importance(toptitdel.rf)
+imp1<-imp[order(imp[,1], decreasing = TRUE),]
+#imp2<-imp[order(imp[,2], decreasing = TRUE),]
+topimp <- imp1[1:15,]
+#toppurity <- imp2[1:15,]
+#write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)
+#write.table(toppurity, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importanceNode.txt", sep="\t", row.names=TRUE)
+```
+
+### Figure 5A: Evaluating and plotting the fit of the model
+
+```{r, eval=TRUE}
+predictions <- predict(toptitdel.rf, newdata=toptitdel[,-c(1,2)])
+actual <- toptitdel$nextDayCFU
+expgroup<- as.character(all[,2])
+results<-as.data.frame(cbind(actual, predictions))
+colnames(results) <- c( "actual", "predict")
+
+ybar <- apply(results, 2, mean)
+num<-sum((results$actual-ybar["actual"])*(results$predict-ybar["predict"]))
+denA <- sum((results$actual-ybar["actual"])^2)
+denB <- sum((results$predict-ybar["predict"])^2)
+rsq <- (num^2)/(denA*denB) #calculated from the square of the sample correlation coefficient, little r squared as opposed to big R squared
+
+results <- cbind(expgroup, results)
+
+#Plot observed vs predicted
+#information for plotting
+color<-rainbow(7)
+colors <- c(control = "black",
+          cipro = color[1],
+          clinda = "#FFCC00",
+          vanc.625 = "#00CC00",
+          vanc.3 = "#00CC00",
+          vanc.1 = "#00CC00",
+          strep5 = "#33FFFF",
+          strep.5 = "#33FFFF",
+          strep.1 = "#33FFFF",
+          cef.5 = color[5],
+          cef.3 = color[5],
+          cef.1 = color[5],
+          amp.5 = color[6],
+          amp.5d = color[6],
+          metro1 = color[7],
+          metro1d = color[7])
+
+# topdose is all circles, middle dose is square, 
+# low dose is upward pointing triangle, delayed is starburst
+pch <- c(control = 1,
+         cipro = 16,
+         clinda = 16,
+         vanc.625 = 16,
+         vanc.3 = 15,
+         vanc.1 = 17,
+         strep5 = 16,
+         strep.5 = 15,
+         strep.1 = 17,
+         cef.5 = 16,
+         cef.3 = 15,
+         cef.1 = 17,
+         amp.5 = 16,
+         amp.5d = 8,
+         metro1 = 16,
+         metro1d = 8)
+
+#these are needed for the legend
+legend.labels <- c(control = "Control",
+            cipro = "Ciprofloxacin",
+            clinda = "Clindamycin",
+            vanc.625 = "Vancomycin 0.625 mg/ml",
+            vanc.3 = "Vancomycin 0.3 mg/ml",
+            vanc.1 = "Vancomycin 0.1 mg/ml",
+            strep5 = "Streptomycin 5 mg/ml",
+            strep.5 = "Streptomycin 0.5 mg/ml",
+            strep.1 = "Streptomycin 0.1 mg/ml",
+            cef.5 = "Cefoperazone 0.5 mg/ml",
+            cef.3 = "Cefoperazone 0.3 mg/ml",
+            cef.1 = "Cefoperazone 0.1 mg/ml",
+            amp.5 = "Ampicillin",
+            amp.5d = "Ampicillin +5D",
+            metro1 = "Metronidazole",
+            metro1d = "Metronidazole +5D")
+
+treatments <- c(control = "Control",
+            cipro = "Ciprofloxacin",
+            clinda = "Clindamycin",
+            vanc.625 = "Vancomycin High",
+            vanc.3 = "Vancomycin Medium",
+            vanc.1 = "Vancomycin Low",
+            strep5 = "Streptomycin High",
+            strep.5 = "Streptomycin Medium",
+            strep.1 = "Streptomycin Low",
+            cef.5 = "Cefoperazone High",
+            cef.3 = "Cefoperazone Medium",
+            cef.1 = "Cefoperazone Low",
+            amp.5 = "Ampicillin",
+            amp.5d = "Ampicillin +5D",
+            metro1 = "Metronidazole",
+            metro1d = "Metronidazole +5D")
+
+par(mfrow=c(1, 1)) 
+par(mar=c(5, 5, 4, 2) +0.1, mgp=c(3, 1, 0), las=1) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+plot(results[results[,1]=="control",2], 
+     results[results[,1]=="control",3], 
+     main="", 
+     ylab=expression(paste("Predicted Log ", italic("C. difficile"), " CFU/g Feces")), 
+     xlab=expression(paste("Actual Log ", italic("C. difficile"), " CFU/g Feces")), 
+     ylim=c(0,9), xlim=c(0,9), xaxt='n', yaxt='n', cex=1.5)
+#mtext(bquote("r"^"2" ~ " = " ~ .(signif(rsq, 3))), side=3, line=0)
+abline(a=0, b=1, lty="dashed", lwd=2, col="black")
+axis(1, at = c(0:9), labels=c(0:9))
+axis(2, at = c(0:9), labels=c(0:9))
+
+for(i in 2:(length(legend.labels))){
+  points(results[results[,1]==names(legend.labels)[i],2], 
+         results[results[,1]==names(legend.labels)[i],3], 
+         col=colors[i], pch=pch[i], cex=1.5)
+}
+$legend("bottomright", inset= .07, legend=legend.labels, pch=pch, col=colors, pt.cex=1,  cex=.6, bty="n")
+
+```
+
+### Figure 5B: Importance Plot
+```{r, eval=TRUE}
+topimp<-topimp[order(topimp[,1], decreasing=FALSE),]
+ids <- read.csv(file = "~/Documents/Github/abxD01/Figure 6/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
+labels<- paste0(ids$name, " (", ids$otuname, ")")
+par(mfrow=c(1, 1)) #+1 to give extra labeling space
+par(mar=c(5, 15, 0.5, 2) +0.1, mgp=c(3, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+plot(topimp[,1], 1:15, xlab="% Increase in Mean Squared Error", yaxt="n", ylab="", pch=16, cex=1.5, xlim=c(20, 80))
+axis(2, at = c(1:15), labels = labels, las=1, cex.axis=.9 )
+abline(h=c(1:15), lty="dashed", col='black')
+```
+
+### Testing a random forest model with only OTUs 3 and 39
+
+```{r, eval=TRUE}
+test3.39 <- toptitdel[,c("Group","nextDayCFU","Otu00003", "Otu00039")]
+
+rf.test3.39 <- randomForest(nextDayCFU ~ ., 
+                           data = test3.39[,-1],  outscale=TRUE,
+                           importance=TRUE, proximity=TRUE,
+                           keep.forest=TRUE, ntree=5000
+)
+print(rf.test3.39)
+plot(rf.test3.39)
+
+predictions <- predict(rf.test3.39, newdata=test3.39[,-c(1,2)])
+actual <- test3.39$nextDayCFU
+test.results<-as.data.frame(cbind(actual, predictions))
+colnames(test.results) <- c( "actual", "predict")
+
+ybar <- apply(test.results, 2, mean)
+num<-sum((test.results$actual-ybar["actual"])*(test.results$predict-ybar["predict"]))
+denA <- sum((test.results$actual-ybar["actual"])^2)
+denB <- sum((test.results$predict-ybar["predict"])^2)
+test.rsq <- (num^2)/(denA*denB)
+
+```
+
+### Correlation over all three datasets with *C. difficile* levels
+
+```{r, eval=FALSE}
+
+#First the correlation calculations that were put on the graph:
+c<-1
+otu <- c()
+cor.spear <- c()
+pval.spear <- c()
+cor.ken <- c()
+pval.ken <- c()
+for(i in 3:length(toptitdel)){
+  otu[c] <- colnames(toptitdel[i])
+  cor.spear[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="spearman")$estimate
+  pval.spear[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="spearman")$p.value
+  cor.ken[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="kendall")$estimate #good to see because kendall handles ties
+  pval.ken[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
+  c <- c+1
+}
+pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
+pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons
+
+results = NULL
+results <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
+colnames(results) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
+results <- results[order(results[,3]),]  #order by pvalue column=3
+#write.table(results[1:dim(results)[1],], file="~/Desktop/mothur/abxD01/correlation/toptidel.1p.correl.txt", sep="\t", row.names=FALSE)
+print(results[1:dim(results)[1],])
+```
+
+From this file I selected the top 8 OTUs identified by random forest's increase in mean squared error measurement to graph (next section).
+
+### Figure 5C: OTU abundance vs. *C. difficile* abundance
+
+```{r, eval=FALSE}
+corrs<-read.csv(file="~/Desktop/mothur/abxD01/correlation/toptitdel.correlation.impOTUs.csv", header=TRUE)
+corrs <- corrs[1:8,]
+otunames <-  row.names(topimp)[15:8]
+graphID <- ids[15:8,c(1,3)]
+
+graphOTUxCD(otunames, graphID, corrs)
+```
 

### 2015-03-02 PAPER.Rmd
Change headers to be like Pat's

@@
----
-title: "abxD01 Paper"
-author: "Alyxandria Schubert"
-date: "February 26, 2015"
-output: html_document
----
+**Running title:** Antibiotic induced alterations of the murine gut microbiota and subsequent effects on colonization resistance against *Clostridium difficile*
 
-# Abstract
+**Short title:** Alterations of the Murine Gut Microbiota that Lead to Clostridium difficile Colonization
 
-# Introduction
+Alyxandria M. Schubert<sup>1</sup>, Hamide Sinani<sup>1</sup>, and Patrick D.
+Schloss<sup>1</sup>\*
+
+> \* Correspondence: <pschloss@umich.edu>
+> 734.647.5801
+>
+> Department of Microbiology and Immunology
+>
+> University of Michigan
+>
+> 5618 Medical Sciences II
+>
+> 1500 W. Medical Center Dr.
+>
+> Ann Arbor, MI 48109
+
+1 Department of Microbiology and Immunology, 1500 W. Medical Center,
+University of Michigan, Ann Arbor, MI 48109
+
+### Abstract
+
+### Introduction
 
@@
 When this indigenous community is perturbed, this often leads to a loss of resistance.
-This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, chemotherapy, but also exposed to the environmental C.
-difficile spores.
-Among hospital-acquired infections in the United States, C.
-difficile infections (CDI) are the number one reported and are responsible for 14,000 deaths a year.
+This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, chemotherapy, but also exposed to the environmental *C.
+difficile* spores.
+Among hospital-acquired infections in the United States, *C.
+difficile* infections (CDI) are the number one reported and are responsible for 14,000 deaths a year.
 It is not completely understood how different perturbations affect subsequent colonization resistance, nor do we have a comprehensive view of the taxonomic changes associated with a loss in colonization resistance.
@@
 
-The purpose of this investigation was to test the effects of an array of perturbations on microbiome CR levels against C.
-difficile.
+The purpose of this investigation was to test the effects of an array of perturbations on microbiome CR levels against *C.
+difficile*.
 We then used this information to design microbiome-based models for estimating pathogen colonization levels.
 Through the administration of various antibiotic regimens, we altered the murine gut microbiota and observed these changes using 16S rRNA sequencing.
-The resultant gut communities were then challenged with C.
-difficile spores.
-Through correlation and random forest analysis of the starting communities with the subsequent level of C.
-difficile colonization, we built linear models to predict C.
-difficile colonization levels based on a subset of the microbiota.
+The resultant gut communities were then challenged with *C.
+difficile* spores.
+Through correlation and random forest analysis of the starting communities with the subsequent level of *C.
+difficile* colonization, we built linear models to predict *C.
+difficile* colonization levels based on a subset of the microbiota.
 We used experimental data to test the ability of the model.
 
-Though we are limited in a resolution at the strain and species level, we have identified groups of related species with antagonistic relationships with C.
-difficile colonization.
+Though we are limited in a resolution at the strain and species level, we have identified groups of related species with antagonistic relationships with *C.
+difficile* colonization.
 We identified several models with equally robust quality statistics and observed that across the top models a subset of 5 species were consistently present.
@@
 
-# Materials and Methods
+### Materials and Methods
 
-## Animal Care
-
-We used 5-8 week old C57Bl/6 mice for all our experiments.
+***Animal Care.*** We used 5-8 week old C57Bl/6 mice for all our experiments.
 These mice were housed in our SPF room in the animal facility at the University of Michigan.
@@
 
-## Antibiotic Administration
-
-Mice were administered one of seven different antibiotics, including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
+***Antibiotic Administration.*** Mice were administered one of seven different antibiotics, including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
 The route of administration depended on the specific antibiotic.
@@
 Ciprofloxacin (10 mg/kg) was administered via oral gavage, and clindamycin (10 mg/kg) was administered via IP injection.
-All antibiotic treated mice were given 24 hours of recovery on regular drinking water prior to C.
-difficile challenge.
+All antibiotic treated mice were given 24 hours of recovery on regular drinking water prior to *C.
+difficile* challenge.
 
-## C. difficile Preparation and Challenge
-
-In all our experiments we used C.
-difficile strain 630Δerm [cite dena lyras].
+****C.
+difficile* Preparation and Challenge*** In all our experiments we used *C.
+difficile* strain 630Δerm [cite dena lyras].
 The spores used were prepared from a single large batch made following previously established protocols [cite paul?].
-Spores were quantified within the week prior to C.
-difficile challenge for each experiment [established protocol I can cite?].
+Spores were quantified within the week prior to *C.
+difficile* challenge for each experiment [established protocol I can cite?].
 Spores were quantified by boiling a small volume at 65°C for 20 minutes to kill off residual vegetative cells.
@@
 These were incubated at 37°C anaerobically overnight.
-The next day C.
-difficile colony forming units (CFU) were counted and the concentration of spores in the working stock was back calculated.
-On the day of challenge, C.
-difficile spores, at a final concentration of 103, were administered to mice via oral gavage.
-Immediately following this challenge, the remaining C.
-difficile inoculum was diluted in a series and plated to confirm the correct dosage.
+The next day *C.
+difficile* colony forming units (CFU) were counted and the concentration of spores in the working stock was back calculated.
+On the day of challenge, *C.
+difficile* spores, at a final concentration of 103, were administered to mice via oral gavage.
+Immediately following this challenge, the remaining *C.
+difficile* inoculum was diluted in a series and plated to confirm the correct dosage.
 
-## Sample Collection and Plating
-
-Fecal samples were freshly collected for each mouse immediately prior to the administration of any antibiotic, during any antibiotic recovery days, and days 0 and 1 post C.
-difficile challenge.
+***Sample Collection and Plating*** Fecal samples were freshly collected for each mouse immediately prior to the administration of any antibiotic, during any antibiotic recovery days, and days 0 and 1 post *C.
+difficile* challenge.
 Each fecal sample on day 1 was weighed and diluted under anaerobic conditions with anaerobic PBS 1:10 based on pellet weight.
@@
 
-## DNA Extraction and Sequencing
-
-Total bacterial DNA was extracted from each untreated and day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
+***DNA Extraction and Sequencing*** Total bacterial DNA was extracted from each untreated and day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
 Sequencing preparation and sequencing was carried out as previously described (Kozich JJ et al.
@@
 
-## Sequence Curation
-
-These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
+***Sequence Curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
 The full pipeline is also demonstrated online at the mother website (http://www.mothur.org/wiki/MiSeq_SOP).
@@
 
-## Statistical Analysis & Modeling
-
-Using data from the original antibiotic treatments groups, OTUs were selected for addition into the model following several criteria.
-First all OTUs were present in abundance at an average of 1% in at least one experimental group.
-
+***Statistical Analysis & Modeling*** Across all data, we set a relative abundance threshold of an average of 1% in at least one experimental group.
+The 44 OTUs that met this cutoff were included in the training of the model.
+We used the "randomForest" package in R to build a regression-based random forest model to predict *C.
+difficile* colonization levels on day 1 based on the microbiome data from day 0.
 All statistical analyses were conducted using R version 3.1.2.
 
-# Results
+### Results
 
-## Levels of colonization resistance are associated with the initial structure of the gut microbiota.
-
-In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to C.
-difficile colonization.
+***Levels of colonization resistance are associated with the initial structure of the gut microbiota.*** In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to *C.
+difficile* colonization.
 It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] all have decreased resistance against this pathogen.
-To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (Table 1) to differentially alter the microbiota and assess their resistance to C.
-difficile colonization.
-16S rRNA gene sequencing of the fecal communities sampled at the time of C.
-difficile inoculation revealed distinct and reproducible microbiota structures (Figure 1).
+To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (Table 1) to differentially alter the microbiota and assess their resistance to *C.
+difficile* colonization.
+16S rRNA gene sequencing of the fecal communities sampled at the time of *C.
+difficile* inoculation revealed distinct and reproducible microbiota structures (Figure 1).
 The community structures of the beta-lactam-treated mice (i.e.
@@
 In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=0.09).
-Subsequently, 24 hours after inoculation C.
-difficile was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all of the other treatment groups to similarly high levels (107-108 CFU/g) (Figure 1).
-We calculated the diversity of all of the communities using the Inverse Simpson (Supplemental Figure 1) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent C.
-difficile colonization (both p<0.001).
-These results suggest that multiple microbiota structures are capable of being colonized by C.
-difficile and that diversity may be an important factor in resistance to invasion.
+Subsequently, 24 hours after inoculation *C.
+difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all of the other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g) (Figure 1).
+We calculated the diversity of all of the communities using the Inverse Simpson (Supplemental Figure 1) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.
+difficile* colonization (both p<0.001).
+These results suggest that multiple microbiota structures are capable of being colonized by *C.
+difficile* and that diversity may be an important factor in resistance to invasion.
 
@@
 
-## Correlation analysis reveals potentially protective bacteria.
-
-To identify bacterial taxa that could be associated with resistance or susceptibility to C.
-difficile, we measured the correlation between the relative abundance of each species-level OTU on the day of inoculation with the level of C.
-difficile colonization 24 hours later across all treatment groups (Figure 2).
-Of the significant OTUs (p<0.05) with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against C.
-difficile (N=40) and those with a positive correlation were associated with susceptibility (N=7).
-Among various bacterial families, three were consistently negatively correlated with C.
-difficile colonization: Porphyromonadaceae (ρaverage= -0.63), Lachnospiraceae (ρaverage= -0.51), Ruminococcaceae (ρaverage= -0.53).
-OTUs from the Proteobacteria phylum had strong positive relationships with C.
-difficile colonization.
+***Correlation analysis reveals potentially protective bacteria.*** To identify bacterial taxa that could be associated with resistance or susceptibility to *C.
+difficile*, we measured the correlation between the relative abundance of each species-level OTU on the day of inoculation with the level of *C.
+difficile* colonization 24 hours later across all treatment groups (Figure 2).
+Of the significant OTUs (p<0.05) with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against *C.
+difficile* (N=40) and those with a positive correlation were associated with susceptibility (N=7).
+Among various bacterial families, three were consistently negatively correlated with *C.
+difficile* colonization: Porphyromonadaceae (ρaverage= -0.63), Lachnospiraceae (ρaverage= -0.51), Ruminococcaceae (ρaverage= -0.53).
+OTUs from the Proteobacteria phylum had strong positive relationships with *C.
+difficile* colonization.
 These included OTUs associated with the Pseudomonas (ρ= 0.31) and Escherichia (ρ= 0.56) genera.
 Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus Lactobacillus (family Lactobacillaceae; OTU23, ρ= -0.73), Alistipes (family Rickenellaceae; OTU20, ρ= -0.78), Enterorhabdus (phylum Actinobacterium, family Coriobacteriaceae; OTU29, ρ= -0.70), and Streptococcus (family Streptococcaceae; OTU78, ρ= 0.36).
-Overall, these results suggest that several larger groups of related OTUs, particularly within the Firmicutes and Bacteroidetes phyla, may participate in colonization resistance against C.
-difficile, while members of the Proteobacteria phylum may be involved in susceptibility.
+Overall, these results suggest that several larger groups of related OTUs, particularly within the Firmicutes and Bacteroidetes phyla, may participate in colonization resistance against *C.
+difficile*, while members of the Proteobacteria phylum may be involved in susceptibility.
 
@@
 
-## Smaller perturbations of the microbiota result in maintenance of colonization resistance.
-
-<transition>
-Because most of our original treatments resulted in similarly high levels of C.
-difficile, we hypothesized that applying smaller antibiotic perturbations to the microbiota would result in reduced sensitivity to C.
-difficile, as we observed in the ciprofloxacin treated mice.
+***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** <transition>
+Because most of our original treatments resulted in similarly high levels of *C.
+difficile*, we hypothesized that applying smaller antibiotic perturbations to the microbiota would result in reduced sensitivity to *C.
+difficile*, as we observed in the ciprofloxacin treated mice.
 To test this, we titrated the doses of cefoperazone, streptomycin, and vancomycin given to mice (Supplemental Table 1).
 Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis.
-C.
-difficile levels were significantly different between all titration groups within the cefoperazone (p<0.03) and streptomycin treated microbiotas (p<0.01), displaying a greater range of colonization resistance profiles (Figure 3).
+*C.
+difficile* levels were significantly different between all titration groups within the cefoperazone (p<0.03) and streptomycin treated microbiotas (p<0.01), displaying a greater range of colonization resistance profiles (Figure 3).
 Next, we hypothesized that the populations we previously identified as being associated with protection against colonization would be elevated in the mice with decreased colonization and those associated with susceptibility would be less abundant.
@@
 
-## Allowing recovery of specific bacteria restores colonization resistance.
+***Allowing recovery of specific bacteria restores colonization resistance.*** Several antibiotic perturbation studies have demonstrated that given time perturbed communities can return to a “healthy” state in which resistance to *C.
+difficile* is restored [].
+For each antibiotic treatment, a control group of mice receiving antibiotics without inoculation of *C.
+difficile* was maintained.
+The microbiota’s of animals treated with metronidazole underwent a drastic shift in the microbiota structure with an additional 5 days of recovery.
+We observed changes in relative abundance of several OTUs that we previously observed had strong relationships with subsequent *C.
+difficile* colonization (Figure 2).
+Notably there were significant increases (p<0.05) in overall abundance of the levels of Porphyromonadaceae, Ruminococcaceae, and Lachnospiraceae to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of Enterobacteriaceae.
+Based on these changes, we hypothesized that delaying *C.
+difficile* challenge to allow these potentially important changes in microbiota structure to occur would lead to a decrease in colonization.
+On Day 1 post challenge of these recovered mice, we observed a significant decrease in *C.
+difficile* level of CFU/g feces (p<0.0001).
+These results strengthen the importance of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Enterobacteriaceae in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of *C.
+difficile* colonization.
 
-Several antibiotic perturbation studies have demonstrated that given time perturbed communities can return to a “healthy” state in which resistance to C.
-difficile is restored [].
-For each antibiotic treatment, a control group of mice receiving antibiotics without inoculation of C.
-difficile was maintained.
-The microbiota’s of animals treated with metronidazole underwent a drastic shift in the microbiota structure with an additional 5 days of recovery.
-We observed changes in relative abundance of several OTUs that we previously observed had strong relationships with subsequent C.
-difficile colonization (Figure 2).
-Notably there were significant increases (p<0.05) in overall abundance of the levels of Porphyromonadaceae, Ruminococcaceae, and Lachnospiraceae to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of Enterobacteriaceae.
-Based on these changes, we hypothesized that delaying C.
-difficile challenge to allow these potentially important changes in microbiota structure to occur would lead to a decrease in colonization.
-On Day 1 post challenge of these recovered mice, we observed a significant decrease in C.
-difficile level of CFU/g feces (p<0.0001).
-These results strengthen the importance of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Enterobacteriaceae in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of C.
-difficile colonization.
-
-Additionally, we assessed the resistance of ampicillin treated mice given five extra days of recovery prior to C.
-difficile challenge.
+Additionally, we assessed the resistance of ampicillin treated mice given five extra days of recovery prior to *C.
+difficile* challenge.
 The shift in the microbiome with ampicillin recovery was not as prounounced as seen with recovery of metronidazole treated animals.
-These communities had significantly reduced levels of bacteria correlated with C.
-difficile protection, including Bacteroides, Alistipes, and Ruminoccocaceae compared with untreated animals (p<0.05).
+These communities had significantly reduced levels of bacteria correlated with *C.
+difficile* protection, including Bacteroides, Alistipes, and Ruminoccocaceae compared with untreated animals (p<0.05).
 Importantly, the level of Enterobacteriaceae was not significantly reduced with additional recovery time.
-Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent C.
-difficile levels (p=0.12).
+Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent *C.
+difficile* levels (p=0.12).
 
@@
 
-The file/fileIDS should be matched.
 Input files on github folder Figure 5:
@@
 
-## Modeling microbiota data can predict C. difficile colonization outcomes.
-
-Given our ability to make general predictions of C.
-difficile colonization based on the microbiota, we next designed a model to calculate expected colonization levels given the relative abundance of a subset of OTUs.
+***Modeling microbiota data can predict *C.
+difficile* colonization outcomes.*** Given our ability to make general predictions of *C.
+difficile* colonization based on the microbiota, we next designed a model to calculate expected colonization levels given the relative abundance of a subset of OTUs.
 Combining the three datasets (Table 1, Supplemental Table 1 and 2), we set a 1% relative abundance threshold, eliminating OTUs that on average were not present at this level in at least one of the 16 experimental groups.
@@
 Many of the top contributing OTUs includes members of the Porphyromonadaceae family, Alistipes, Lachnospiraceae, Lactobacillus, and Escherichia taxa.
-The relationship between these OTUs and C.
-difficile is shown in Figure 5C.
-Escherichia had the largest positive correlation with C.
-difficile levels of any bacterial species.
+The relationship between these OTUs and *C.
+difficile* is shown in Figure 5C.
+Escherichia had the largest positive correlation with *C.
+difficile* levels of any bacterial species.
 This finding was consistent across susceptible antibiotic treated communities except for streptomycin treated communities, which altogether lacked Enterobacteriaceae spp.
-OTUs 3 and 39 (Porphyromonadaceae and Lachnospiraceae, respectively), were the two most important features in accurately predicting subsequent C.
-difficile levels.
+OTUs 3 and 39 (Porphyromonadaceae and Lachnospiraceae, respectively), were the two most important features in accurately predicting subsequent *C.
+difficile* levels.
 In fact, a model built with only these two features was able to account for 79% of the variance observed.
@@
 
-# Discussion
+### Discussion
 
@@
 -modeling—not much done for CDI, link back to humans, shows we could model risk as they come in—put towards end with the larger picture
-We designed statistical models based on a data set of varied antibiotic perturbations of the murine microbiota and their subsequent CR levels against C.
-difficile.
-Though antibiotic administration induced unique microbiota structures, colonization resistance was similarly low against C.
-difficile.
+We designed statistical models based on a data set of varied antibiotic perturbations of the murine microbiota and their subsequent CR levels against *C.
+difficile*.
+Though antibiotic administration induced unique microbiota structures, colonization resistance was similarly low against *C.
+difficile*.
 These models incorporated all combinations of top candidate species-level OTUs.
-We then tested our top model’s ability to use new abundance information on a specific subset of the microbiota to predict the subsequent C.
-difficile colonization level, which was experimentally validated.
+We then tested our top model’s ability to use new abundance information on a specific subset of the microbiota to predict the subsequent *C.
+difficile* colonization level, which was experimentally validated.
 These results greatly expand our knowledge of various perturbations and their subsequent effects on microbiota structure and function (colonization resistance ability).
@@
 
-Through our perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of C.
-difficile colonization.
-A large number of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Clostridia, as well as Alistipes were found to have strong potential resistance abilities against C.
-difficile.
+Through our perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of *C.
+difficile* colonization.
+A large number of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Clostridia, as well as Alistipes were found to have strong potential resistance abilities against *C.
+difficile*.
 Conversely, several members of the Proteobacteria phylum, including Pseudomonas, Streptococcus, and especially E.
-coli have strong potential positive impacts for C.
-difficile colonization.
+coli have strong potential positive impacts for *C.
+difficile* colonization.
 These results were supported by their strong correlation values supported across several antibiotic regimens.
 They are also supported through experiments in which the original doses of cefoperazone, streptomycin, and (vancomycin) were titrated to alter the overall abundances but not membership of the community from the starting dose, as has been observed previously (Sekirov).
-These subtler structural changes allowed for greater range of colonization levels of C.
-difficile as well as further/fine-tuned support for our original results.
-Furthermore, by allowing a known change to occur in important bacterial populations within metronidazole treated mice, we were able to reduce the level of colonization by C.
-difficile.
+These subtler structural changes allowed for greater range of colonization levels of *C.
+difficile* as well as further/fine-tuned support for our original results.
+Furthermore, by allowing a known change to occur in important bacterial populations within metronidazole treated mice, we were able to reduce the level of colonization by *C.
+difficile*.
 Lastly we showed their importance through modeling techniques that statistically confirm these results.
@@
 which bacteria are present and their abundance.
-Cross-sectional investigations have characterized the microbiota of patients in the hospital with and without a C.
-difficile infection (CDI) and compared to healthy controls.
-These studies found patients with C.
-difficile tended to have a reduction in numbers of Clostridiales Incertae Sedis XI (Vincent et al), Lachnospiraceae, Ruminococcaceae, and Bacteroides (Schubert et al).
-These studies however lack the baseline healthy state of the microbiota, which preclude them from assessing risk of C.
-difficile.
-Murine models of CDI have the advantage of observing the specific changes that occur before and after controlled perturbations and the introduction of C.
-difficile.
+Cross-sectional investigations have characterized the microbiota of patients in the hospital with and without a *C.
+difficile* infection (CDI) and compared to healthy controls.
+These studies found patients with *C.
+difficile* tended to have a reduction in numbers of Clostridiales Incertae Sedis XI (Vincent et al), Lachnospiraceae, Ruminococcaceae, and Bacteroides (Schubert et al).
+These studies however lack the baseline healthy state of the microbiota, which preclude them from assessing risk of *C.
+difficile*.
+Murine models of CDI have the advantage of observing the specific changes that occur before and after controlled perturbations and the introduction of *C.
+difficile*.
 Communities with low levels of CR were associated with blooms in either Lactobacillaceae (cefoperazone-induced) (Theriot, 2014, bassis, 2014), Proteobacteria (induced by either clinda+5abx cocktail, clinda alone, or tigecycline) (Reeves, 2011, Buffie, 2011, Bassis 2014).
-Lachnospiraceae is associated with resistance to C.
-difficile infection (Reeves 2011) and has been directly shown to decrease colonization through mono-colonization studies in germfree mice (Reeves 2012).
+Lachnospiraceae is associated with resistance to *C.
+difficile* infection (Reeves 2011) and has been directly shown to decrease colonization through mono-colonization studies in germfree mice (Reeves 2012).
 
 Even fewer studies attempt to model these observational changes in the microbiome following perturbation.
-Using the time series data gathered from the C.
-difficile challenge of clindamycin-treated mice (Buffie 2011), a Lotka-Volterra dyamics-based framework was built which incorporates both time and accounts for external perturbations (Stein 2013).
-They identified a subset of their network, containing Blautia, Akkermansia, and Coprobacillus, as having a negative relationship with C.
-difficile, while alternatively Enterococcus had a positive relationship.
+Using the time series data gathered from the *C.
+difficile* challenge of clindamycin-treated mice (Buffie 2011), a Lotka-Volterra dyamics-based framework was built which incorporates both time and accounts for external perturbations (Stein 2013).
+They identified a subset of their network, containing Blautia, Akkermansia, and Coprobacillus, as having a negative relationship with *C.
+difficile*, while alternatively Enterococcus had a positive relationship.
 
@@
 
-Many of the relationships we observed in our study between specific bacteria and C. difficile are supported in the literature. –see my comments in this section
-Within the Porphyromonadaceae, only OTU505 had a significant positive relationship with subsequent C.
-difficile colonization (ρ= +0.21), which by blastn shares 99% identity with Coprobacter fastidiosus.
-Similarly within the Lachnospiraceae family, only one OTU had a positive relationship with C.
-difficile colonization (OTU174, ρ= +0.22).
+Many of the relationships we observed in our study between specific bacteria and *C. difficile* are supported in the literature. –see my comments in this section
+Within the Porphyromonadaceae, only OTU505 had a significant positive relationship with subsequent *C.
+difficile* colonization (ρ= +0.21), which by blastn shares 99% identity with Coprobacter fastidiosus.
+Similarly within the Lachnospiraceae family, only one OTU had a positive relationship with *C.
+difficile* colonization (OTU174, ρ= +0.22).
 The Lachnospiraceae group also contained an OTU that had an exceptionally strong negative correlation (OTU39, ρ= +0.77).
-The Clostridia class was largely composed of the order Clostridiales (n=6), of which only one had a positive relationship with C.
-difficile colonization (OTU154, ρ= +0.26).
+The Clostridia class was largely composed of the order Clostridiales (n=6), of which only one had a positive relationship with *C.
+difficile* colonization (OTU154, ρ= +0.26).
 
-Several broader groups of bacteria can work together (in some combination) to provide CR …possible that proteobacteria in general are bad …These results indicate that many related species within the larger/broader bacterial groups may be able to serve in providing colonization resistance against C.
-difficile.
+Several broader groups of bacteria can work together (in some combination) to provide CR …possible that proteobacteria in general are bad …These results indicate that many related species within the larger/broader bacterial groups may be able to serve in providing colonization resistance against *C.
+difficile*.
 Other taxons in Figure 3 had more species specific correlations.
@@
 SCFA concentrations in the faecal and caecal samples were unaffected by the diets.
-In conclusion, FOS increased the abundance of Bifidobacterium spp., whereas both fructans reduced Clostridium cluster XI and C.
-difficile toxin gene expression, correlating with a reduction of chronic intestinal inflammation.
+In conclusion, FOS increased the abundance of Bifidobacterium spp., whereas both fructans reduced Clostridium cluster XI and *C.
+difficile* toxin gene expression, correlating with a reduction of chronic intestinal inflammation.
 
@@
 They are unique to individuals and yet all these different profiles of microbiotas are capable of colonization resistance.
-People do not get C.
-difficile infections without first being exposed to some major purturbation of their microbiome, namely antibiotics.
+People do not get *C.
+difficile* infections without first being exposed to some major purturbation of their microbiome, namely antibiotics.
 Thus many structures of the microbiome, and many distinct bacterial populations must be capable of contributing to resistance.

### 2015-03-02 PAPER.Rmd
Attempt insert block code and figures

@@
 
-Alyxandria M. Schubert<sup>1</sup>, Hamide Sinani<sup>1</sup>, and Patrick D.
-Schloss<sup>1</sup>\*
+Alyxandria M. Schubert^1^, Hamide Sinani^1^, and Patrick D.
+Schloss^1^\*
 
@@
 It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] all have decreased resistance against this pathogen.
-To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (Table 1) to differentially alter the microbiota and assess their resistance to *C.
+To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (**Table 1**) to differentially alter the microbiota and assess their resistance to *C.
 difficile* colonization.
 16S rRNA gene sequencing of the fecal communities sampled at the time of *C.
-difficile* inoculation revealed distinct and reproducible microbiota structures (Figure 1).
+difficile* inoculation revealed distinct and reproducible microbiota structures (**Figure 1**).
 The community structures of the beta-lactam-treated mice (i.e.
-cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=0.36).
-Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=0.001) and were characterized by the predominance of Enterobacteriaceae.
-Of all groups significantly different from the untreated microbiota, the levels of Enterobacteriaceae increased with all treatments except with streptomycin.
-Vancomycin-treated communities saw the largest bloom in Akkermansia.
+cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=0.-36).
+Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=0.001) and were characterized by the predominance of *Enterobacteriaceae*.
+Of all groups significantly different from the untreated microbiota, the levels of *Enterobacteriaceae* increased with all treatments except with streptomycin.
+Vancomycin-treated communities saw the largest bloom in *Akkermansia*.
 In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=0.09).
 Subsequently, 24 hours after inoculation *C.
-difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all of the other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g) (Figure 1).
-We calculated the diversity of all of the communities using the Inverse Simpson (Supplemental Figure 1) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.
+difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10^7^-10^8^ CFU/g) (**Figure 1**).
+We calculated the diversity of all of the communities using the Inverse Simpson (**Supplemental Figure 1**) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.
 difficile* colonization (both p<0.001).
@@
 
-### CFU stats comparison
-To compare CFU levels: 
+--------
+
+![**Figure 1.**](Figure 1/topdose2_tx2_sorted_10x20.pdf)
+
+**Figure 1.
+Microbiota structures following diverse antibiotic treatments.** This figure shows the average percent relative abundances for family level phylotypes found in each experimental group on the day of *C.
+difficile* challenge.
+Each row depicts a different antibiotic perturbation.
+The level *C.
+difficile* colonization found 24 hours post microbiota sampling is plotted in the top right corner.
+Each represented phylum is a different shade of gray.
+Error bars depict standard error of the mean.
+
+--------
+
+![**Supplemental Figure 1.**](supplemental/FigS1_topdose_inverseSimp.pdf)
+
+**Supplemental Figure 1.
+Diversity is nevatively correlated with *C.
+difficile* levels.** This figure shows the relationship between the inverse simpson diversity index of microbiotas of each experimental group on day 0 and *C.
+difficile* CFU/g feces on day 1.
+Spearman's rho is plotted in the top right corner.
+
+> ### CFU stats comparison
+> To compare CFU levels: 
 ```{r, eval=TRUE}
@@
 ```
-
-### Amova to compare resulting community structures
-Amova calculation in mothur to determine the uniqueness of each community structure:
+> 
+> ### Amova to compare resulting community structures
+> Amova calculation in mothur to determine the uniqueness of each community structure:
+> 
 ```
@@
 ```
-This results in the file called "abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.amova" located in the github folder. The results of these are summarized in "amova8expgroups.xlsx".
-
-### Figure 1: Bargraph
-
-This figure shows the differences in the communities' structures following different antibiotic treatments.
-The average *C.
-difficile* CFU/g Feces is also calculated in this code for each treatment group.
-
-This figure was built using the R script called "stacked.barcharts.r", which contains the stackedbarcharts() funciton in the abxD01 folder.The input files are listed in Github.
+> This results in the file called "abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.amova" located in the github folder. The results of these are summarized in "amova8expgroups.xlsx".
+> 
+> ### Figure 1: Bargraph
+> 
+> This figure was built using the R script called "stacked.barcharts.r", which contains the stackedbarcharts() funciton in the abxD01 folder.The input files are listed in Github.
 The graph can be sorted 2 ways either by the untreated's decreasing relabund (sortbyphyl=FALSE), or by the most abundant within each phylum (sortbyphyl=TRUE).
@@
 difficile* CFU/g Feces is also labeled on the Y axis below the treatment name.
-The parameters I changed in the r script were:
+> The parameters I changed in the r script were:
 ```{r, eval=FALSE}
@@
 ```
-
-Files saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf".
-
-### Supplemental Figure 1: correlation plot of diversity with *C. difficile*
-
-For the correlation calculations see the next section of the results.
-Figure making:
+> 
+> Files saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf".
+> 
+> ### Supplemental Figure 1: correlation plot of diversity with *C. difficile*
+> 
+> For the correlation calculations see the next section of the results.
+> Figure making:
+> 
+--------
 
@@
 difficile*, while members of the Proteobacteria phylum may be involved in susceptibility.
+
+--------
+
+![**Figure 2.**](Figure 2/corr_avgAbove0.001_topdose2_10x10.pdf)
+
+**Figure 2.
+TITLE.** Species level OTUs (3% cutoff) found on day 0 were correlated with the level of *C.
+difficile* CFU/g feces on day 1 using Spearman’s rank based correlation coefficient.
+Only OTUs with an average abundance of at least 0.05% across all original treatments were considered in the correlation analysis.
+All correlations shown in this graph are significant (p<0.05).
+OTUs were grouped by family.
+Asterisk * indicates unclassified at family level.
 
@@
 
-***Modeling microbiota data can predict *C.
-difficile* colonization outcomes.*** Given our ability to make general predictions of *C.
+***Modeling microbiota data can predict C.
+difficile colonization outcomes.*** Given our ability to make general predictions of *C.
 difficile* colonization based on the microbiota, we next designed a model to calculate expected colonization levels given the relative abundance of a subset of OTUs.

### 2015-03-02 PAPER.Rmd
Fix issue with block code

@@
 > To compare CFU levels:
+
 ```{r, eval=TRUE}
@@
 ```
-> 
+
 > ### Amova to compare resulting community structures
@@
 >
+
 ```
@@
 > The parameters I changed in the r script were:
+
 ```{r, eval=FALSE}
@@
 ```
-> 
+
 > Files saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf".
@@
 >
+
 --------

### 2015-03-02 Figure 5/modelRfunctions.R
Fix title error

@@ -42,16 +42,20 @@ RF.validate <- function(data, iters){
 #
 # Returns 
 # 
-#  
-graphOTUxCD(otunames, otulabels, corrs)
-
-pretty_otus <- as.data.frame(gsub("Otu0*", "OTU", otus))
-otulabels <- cbind(ids,pretty_otus)
-names(otulabels) <- c("names", "otus")
 
 
-otus<-otunames
-ids<-otulabels
+
+
+# #  
+# graphOTUxCD(otunames, otulabels, corrs)
+# 
+# pretty_otus <- as.data.frame(gsub("Otu0*", "OTU", otus))
+# otulabels <- cbind(ids,pretty_otus)
+# names(otulabels) <- c("names", "otus")
+# 
+# 
+# otus<-otunames
+# ids<-otulabels
 
 
 graphOTUxCD <- function(otus, ids, corrs){
@@ -91,12 +95,13 @@ graphOTUxCD <- function(otus, ids, corrs){
     abline(v=0.0005, lty="dotted", lwd=2, col="black")
     
     test <- grep("aceae", ids[i, 1])
-    name<-as.character(ids$names[i])
+    name<-as.character(ids$name[i])
     number<-as.character(ids$otus[i])
     corr<-as.character(signif(corrs$corSpear[i], 2))
+    library(base)
     graphTitle <- bquote(bolditalic(.(name)) *" " *bold(.(number)) *bold(", ") *bold(rho) *bold(" = ") *bold(.(corr)))
     text(x = 0.00005, y = 9.3, labels = graphTitle, 
-         cex = 1, pos=4, font=2)
+         cex = 1, pos=4)
   
     #paste(labels[i], ", \u03c1 = ",signif(corrs[i, 2], 2))
     #bquote(bold(.(labels[i])) ~ bold(", ") ~ bold(rho) ~ bold(" = ") ~ bold(.(signif(corrs[i, 2], 3))))

### 2015-03-02 R scripts/barplotBeside.r
Fix subscript out of bounds error for bp[,n]

@@ -31,6 +31,7 @@ barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
       # add standard error bars
       d <- as.data.frame( cbind( c(bp), c(mavgs), c(matrix.se) ) )
       names(d) <- c("x", "y", "se")
+      library(Hmisc)
       with (
         data = d
         , expr = errbar(x, y, y+se, y-se, add=T, pch=".", cap=.01)
@@ -38,6 +39,12 @@ barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
       
       # Statistical differences between groups
       statLetter <- findStatLetters(file, otus, graphGroups)
+      
+      statLetter <- as.data.frame(statLetter)
+      bp <- as.data.frame(bp)
+      mavgs <- as.data.frame(mavgs)
+      matrix.se <- as.data.frame(matrix.se)
+      
       #For labeling letters above each bar
       for(n in 1:leng){
         text( bp[,n], (mavgs[,n]+matrix.se[,n]), labels=statLetter[,n], cex=1, col="red", pos=3, offset=0.15, xpd=TRUE)
@@ -57,6 +64,7 @@ barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
         # add standard error bars
         d <- as.data.frame( cbind( c(bp), c(mavgs), c(matrix.se) ) )
         names(d) <- c("x", "y", "se")
+        library(Hmisc)
         with (
           data = d
           , expr = errbar(x, y, y+se, y-se, add=T, pch=".", cap=.01)
@@ -64,6 +72,12 @@ barplotBeside <- function(id.info, matrix.of.avgs, matrix.of.se, ids, file){
         
         # Statistical differences between groups
         statLetter <- findStatLetters(file, otus, graphGroups)
+        statLetter <- as.data.frame(statLetter)
+        
+        bp <- as.data.frame(bp)
+        mavgs <- as.data.frame(mavgs)
+        matrix.se <- as.data.frame(matrix.se)
+        
         #For labeling letters above each bar
         for(n in 1:leng){
           text( bp[,n], (mavgs[,n]+matrix.se[,n]), labels=statLetter[,n], cex=1, col="red", pos=3, offset=0.15, xpd=TRUE)

### 2015-03-02 R scripts/stacked.barcharts.r
Comment out extra calls

@@ -8,13 +8,15 @@
 #################################################
 # Parameters to change, an example:
 # CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.001, expgroups #'d by graph order & sorted by first graph)
+# 
 
-file <- "~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv"
-fileIDS <- "~/Documents/Github/abxD01/Figure 5/alldelay_tx2_barchart_ids.csv"
-#graphLabels<-c("Cefoperazone", "Streptomycin", "Vancomycin")
-sortbyphyl<-TRUE
-graphbyphyl<-FALSE
-divide<-TRUE
+
+# file <- "~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv"
+# fileIDS <- "~/Documents/Github/abxD01/Figure 5/alldelay_tx2_barchart_ids.csv"
+# #graphLabels<-c("Cefoperazone", "Streptomycin", "Vancomycin")
+# sortbyphyl<-TRUE
+# graphbyphyl<-FALSE
+# divide<-TRUE
 
 
 # Highlight all and run!

### 2015-03-02 PAPER.Rmd
Attempt to make changes to be more like Pat's example
Comment out analysis chunks. Fix accompanying functions. Insert *.png
files with figure legends. Italicize latin bacterial names.

@@
+> Note: The text in block code such as this are not in the original paper. They are there to provide an explanation of the R code and results.
+>
+>
+
 **Running title:** Antibiotic induced alterations of the murine gut microbiota and subsequent effects on colonization resistance against *Clostridium difficile*
@@
 
-Alyxandria M. Schubert^1^, Hamide Sinani^1^, and Patrick D.
-Schloss^1^\*
+Alyxandria M. Schubert<sub>1</sub>, Hamide Sinani<sub>1</sub>, and Patrick D.
+Schloss<sub>1</sub>\*
 
-> \* Correspondence: <pschloss@umich.edu>
-> 734.647.5801
->
-> Department of Microbiology and Immunology
->
-> University of Michigan
->
-> 5618 Medical Sciences II
->
-> 1500 W. Medical Center Dr.
->
-> Ann Arbor, MI 48109
+  \* Correspondence: <pschloss@umich.edu>
+  734.647.5801
+  Department of Microbiology and Immunology
+  University of Michigan
+  5618 Medical Sciences II
+  1500 W. Medical Center Dr.
+  Ann Arbor, MI 48109
 
@@
 Subsequently, 24 hours after inoculation *C.
-difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10^7^-10^8^ CFU/g) (**Figure 1**).
+difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sub>7</sub>-10<sub>8</sub> CFU/g) (**Figure 1**).
 We calculated the diversity of all of the communities using the Inverse Simpson (**Supplemental Figure 1**) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.
@@
 
-![**Figure 1.**](Figure 1/topdose2_tx2_sorted_10x20.pdf)
+![**Figure 1.**](Figure 1/figure1.png)
 
@@
 Spearman's rho is plotted in the top right corner.
+
+--------
 
@@
 abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
-library(plotrix)
-# Generate the 
+
+library(Hmisc)
 stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
 
-stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+# stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
 
@@
 Among various bacterial families, three were consistently negatively correlated with *C.
-difficile* colonization: Porphyromonadaceae (ρaverage= -0.63), Lachnospiraceae (ρaverage= -0.51), Ruminococcaceae (ρaverage= -0.53).
-OTUs from the Proteobacteria phylum had strong positive relationships with *C.
+difficile* colonization: *Porphyromonadaceae* (ρaverage= -0.63), *Lachnospiraceae* (ρaverage= -0.51), *Ruminococcaceae* (ρaverage= -0.53).
+OTUs from the *Proteobacteria* phylum had strong positive relationships with *C.
 difficile* colonization.
-These included OTUs associated with the Pseudomonas (ρ= 0.31) and Escherichia (ρ= 0.56) genera.
-Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus Lactobacillus (family Lactobacillaceae; OTU23, ρ= -0.73), Alistipes (family Rickenellaceae; OTU20, ρ= -0.78), Enterorhabdus (phylum Actinobacterium, family Coriobacteriaceae; OTU29, ρ= -0.70), and Streptococcus (family Streptococcaceae; OTU78, ρ= 0.36).
-Overall, these results suggest that several larger groups of related OTUs, particularly within the Firmicutes and Bacteroidetes phyla, may participate in colonization resistance against *C.
-difficile*, while members of the Proteobacteria phylum may be involved in susceptibility.
+These included OTUs associated with the *Pseudomonas* (ρ= 0.31) and *Escherichia* (ρ= 0.56) genera.
+Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus *Lactobacillus* (family *Lactobacillaceae*; OTU23, ρ= -0.73), *Alistipes* (family *Rickenellaceae*; OTU20, ρ= -0.78), *Enterorhabdus* (phylum *Actinobacterium*, family *Coriobacteriaceae*; OTU29, ρ= -0.70), and *Streptococcus* (family *Streptococcaceae*; OTU78, ρ= 0.36).
+Overall, these results suggest that several larger groups of related OTUs, particularly within the *Firmicutes* and *Bacteroidetes* phyla, may participate in colonization resistance against *C.
+difficile*, while members of the *Proteobacteria* phylum may be involved in susceptibility.
 
@@
 
-![**Figure 2.**](Figure 2/corr_avgAbove0.001_topdose2_10x10.pdf)
+![**Figure 2.**](Figure 2/figure2.png)
 
@@
 
-### Correlation Calculations
+--------
 
-The required file is located on Github.
+> ### Correlation Calculations
+> 
+> The required file is located on Github.
 
@@
 
-### Figure 2: Stipchart
-
-This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1 using the original set of antibiotic groups.
-
-Necessary files are on github, and it was made using the following code:
+> ### Figure 2: Stipchart
+> 
+> This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1 using the original set of antibiotic groups. 
+> 
+> Necessary files are on github, and it was made using the following code:
 
@@
 
-Figure saved as "corr_avgAbove0.001_topdose2_10x10.pdf".
+> Figure saved as "corr_avgAbove0.001_topdose2_10x10.pdf".
 
@@
 The titrated antibiotic doses clearly resulted in distinct community structures that were variations of the original dose (Figure 3).
-The cefoperazone titrations resulted in a shift in dominance from the Bacteroidetes to the Firmicutes phylum with increasing antibiotic dose.
-Most notable was the increase in OTUs associated with resistance including the Porphyromonadaceae and Alistipes and a reduction in OTUs associated with susceptibility including members of the Proteobacteria, Clostridium, and Staphylococcus.
-The relative abundance of other OTUs associated with protection including the Lactobacillus, Lachnospiraceae, and Ruminococcaceae did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.
+The cefoperazone titrations resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
+Most notable was the increase in OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*.
+The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.
 The streptomycin titrations did not result in profound shifts in the overall community structure, but did yield increased resistance.
 Rather, there were individual populations that were gained or lost across the gradient.
-Notably, mice receiving the lowest dose retained significantly higher levels of Alistipes and Anaeroplasma (p<0.05), which we previously found to be strongly associated with colonization resistance.
-The relative abundances of the Porphyromonodaceae, Lachnospiraceae, Ruminococcaceae were unchanged, but were present at similarly high levels to those found in the untreated mice.
+Notably, mice receiving the lowest dose retained significantly higher levels of *Alistipes* and *Anaeroplasma* (p<0.05), which we previously found to be strongly associated with colonization resistance.
+The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice.
 Although the vancomycin titration yielded a number of large changes in the abundance of individual populations, there was no recovery of colonization resistance.
-The relative abundance of many of the populations associated with resistance (Porphyromonadaceae, Lachnospiraceae, Anaeroplasma) increased with decreased vancomycin dosage; however, with the exception of the Lactobacillus populations, they did not reach the levels observed in the untreated mice.
-In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the Akkermansia and Enterobacteriaceae.
+The relative abundance of many of the populations associated with resistance (*Porphyromonadaceae*, *Lachnospiraceae*, *Anaeroplasma*) increased with decreased vancomycin dosage; however, with the exception of the *Lactobacillus* populations, they did not reach the levels observed in the untreated mice.
+In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the *Akkermansia* and *Enterobacteriaceae*.
 Together, these results suggest that individual populations are not sufficient to provide colonization resistance.
@@
 
-### CFU stats comparison
+--------
 
-For comparison of the CFU's among titration groups: 
+![**Figure 3.**](Figure 3/figure3.png)
+
+**Figure 3.
+TITLE.** This figure shows the average percent relative abundances for family level phylotypes found on the day of *C.
+difficile* challenge in each titration group.
+Each row depicts a different antibiotic that was titrated.
+High, medium, and low antibiotic doses are represented by black, grey, and white colored bars respectively.
+The specific dose amount is shown for each antibiotic.
+Error bars show the standard error of the mean.
+
+--------
+
+> ### CFU stats comparison
+> 
+> For comparison of the CFU's among titration groups:
+
 ```{r, eval=TRUE}
@@
 
-### Figure 3: Barchart
-
-This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.
-
-The file/fileIDS should be matched by the titration.
-Input files on github:
-1. abxD01.final.tx.2.subsample.alltitrations.forlogscale.csv
-2. alltitrations_tx2_barchart_ids.csv
-
-Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
+> ### Figure 3: Barchart
+> 
+> This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.  
+> 
+> The file/fileIDS should be matched by the titration.
+> Input files on github:
+> 1. abxD01.final.tx.2.subsample.alltitrations.forlogscale.csv
+> 2. alltitrations_tx2_barchart_ids.csv
+> 
+> Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
 
@@
 difficile* colonization (Figure 2).
-Notably there were significant increases (p<0.05) in overall abundance of the levels of Porphyromonadaceae, Ruminococcaceae, and Lachnospiraceae to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of Enterobacteriaceae.
+Notably there were significant increases (p<0.05) in overall abundance of the levels of *Porphyromonadaceae*, *Ruminococcaceae*, and *Lachnospiraceae* to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of *Enterobacteriaceae*.
 Based on these changes, we hypothesized that delaying *C.
@@
 difficile* level of CFU/g feces (p<0.0001).
-These results strengthen the importance of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Enterobacteriaceae in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of *C.
+These results strengthen the importance of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae* in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of *C.
 difficile* colonization.
@@
 These communities had significantly reduced levels of bacteria correlated with *C.
-difficile* protection, including Bacteroides, Alistipes, and Ruminoccocaceae compared with untreated animals (p<0.05).
-Importantly, the level of Enterobacteriaceae was not significantly reduced with additional recovery time.
+difficile* protection, including *Bacteroides*, *Alistipes*, and *Ruminoccocaceae* compared with untreated animals (p<0.05).
+Importantly, the level of *Enterobacteriaceae* was not significantly reduced with additional recovery time.
 Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent *C.
@@
 
-### Differences between recovery days' thetayc
+--------
 
-### Statistical comparison of delayed and on time CFU results
+![**Figure 4.**](Figure 4/figure4.png)
+
+**Figure 4.
+TITLE.** This figure shows the average percent relative abundances for family level phylotypes found on the day of *C.
+difficile* challenge in each experimental group.
+Each row depicts a different antibiotic that was allowed to recover.
+Black bars represent untreated animals.
+Grey bars represent animals given 1 day of recovery off antibiotics as in Figure 1.
+White bars represent animals given a total of 6 recovery days (5 additional days).
+Error bars show standard error of the mean.
+
+--------
+
+> ### Differences between recovery days' thetayc
+> 
+> ### Statistical comparison of delayed and on time CFU results
+
 ```{r, eval=TRUE}
@@
 
-### Figure 4: Barchart
-
-This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole and ampicillin.
-
-Input files on github folder Figure 5:
-1. abxD01.final.tx.2.subsample.alldelay.forlogscale.csv
-2. alldelay_tx2_barchart_ids.csv
-
-Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
+> ### Figure 4: Barchart
+> 
+> This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole and ampicillin.
+> 
+> Input files on github folder Figure 5:
+> 1. abxD01.final.tx.2.subsample.alldelay.forlogscale.csv
+> 2. alldelay_tx2_barchart_ids.csv
+> 
+> Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
 
@@
 
-stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 5/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 5/alldelay_tx2_barchart_ids.csv", divide = TRUE)
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 4/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 4/alldelay_tx2_barchart_ids.csv", divide = TRUE)
 
@@
 We identified OTUs that contributed most to increased accuracy of the model using random forest’s feature selection (Figure 5B).
-Many of the top contributing OTUs includes members of the Porphyromonadaceae family, Alistipes, Lachnospiraceae, Lactobacillus, and Escherichia taxa.
+Many of the top contributing OTUs includes members of the *Porphyromonadaceae* family, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia* taxa.
 The relationship between these OTUs and *C.
 difficile* is shown in Figure 5C.
-Escherichia had the largest positive correlation with *C.
+*Escherichia* had the largest positive correlation with *C.
 difficile* levels of any bacterial species.
-This finding was consistent across susceptible antibiotic treated communities except for streptomycin treated communities, which altogether lacked Enterobacteriaceae spp.
-OTUs 3 and 39 (Porphyromonadaceae and Lachnospiraceae, respectively), were the two most important features in accurately predicting subsequent *C.
+This finding was consistent across susceptible antibiotic treated communities except for streptomycin treated communities, which altogether lacked *Enterobacteriaceae* spp.
+OTUs 3 and 39 (*Porphyromonadaceae* and *Lachnospiraceae*, respectively), were the two most important features in accurately predicting subsequent *C.
 difficile* levels.
 In fact, a model built with only these two features was able to account for 79% of the variance observed.
-Together these results suggest that colonization resistance is likely conferred by porphyromonadaceae, Alistipes, and lachnospiraceae and a loss in these bacteria, concurrently with a gain of Escherichia, can result in increased susceptibility to infection.
+Together these results suggest that colonization resistance is likely conferred by *porphyromonadaceae*, *Alistipes*, and *lachnospiraceae* and a loss in these bacteria, concurrently with a gain of *Escherichia*, can result in increased susceptibility to infection.
 
-### Evaluating cutoffs for OTU inclusion in Random Forest model
+--------
+
+![**Figure 5A.**](Figure 5/figure5a_9x9.png)
+
+![**Figure 5B.**](Figure 5/figure5b_6x4.5.png)
+
+![**Figure 5C.**](Figure 5/figure5c.png)
+
+**Figure 5.
+Random forest model to predict *C.
+difficile* colonization levels based on the microbiota.** a) The fit of the random forest model by experimental group.
+The dashed line represents the y=x line.
+b) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model.
+c) The percent relative abundance of the top 8 OTUs from (b) on day 0 are plotted against the log<sub>10</sub> *C.
+difficile* CFU/g feces.
+
+--------
+
+> ### Evaluating cutoffs for OTU inclusion in Random Forest model
 
@@
 ```
-This feeds into the next analysis section.
-
-### Random forest results
+> This feeds into the next analysis section.
+> 
+> ### Random forest results
 
@@
 toptitdel <- all[,cols]
+library(randomForest)
 
@@
 
-### Figure 5A: Evaluating and plotting the fit of the model
+> ### Figure 5A: Evaluating and plotting the fit of the model
 
@@
 }
-$legend("bottomright", inset= .07, legend=legend.labels, pch=pch, col=colors, pt.cex=1,  cex=.6, bty="n")
+legend(x=7, y=3, legend=legend.labels, pch=pch, col=colors, pt.cex=1,  cex=.6, bty="n")
 
@@
 topimp<-topimp[order(topimp[,1], decreasing=FALSE),]
-ids <- read.csv(file = "~/Documents/Github/abxD01/Figure 6/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
+ids <- read.csv(file = "~/Documents/Github/abxD01/Figure 5/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
 labels<- paste0(ids$name, " (", ids$otuname, ")")
@@
 
-### Testing a random forest model with only OTUs 3 and 39
+> ### Testing a random forest model with only OTUs 3 and 39
 
@@
 
-### Correlation over all three datasets with *C. difficile* levels
+> ### Correlation over all three datasets with *C. difficile* levels
 
@@
 
-From this file I selected the top 8 OTUs identified by random forest's increase in mean squared error measurement to graph (next section).
-
-### Figure 5C: OTU abundance vs. *C. difficile* abundance
+> From this file I selected the top 8 OTUs identified by random forest's increase in mean squared error measurement to graph (next section).
+> 
+> ### Figure 5C: OTU abundance vs. *C. difficile* abundance
 
@@
 
-graphOTUxCD(otunames, graphID, corrs)
+graphOTUxCD(otus = otunames, ids = graphID, corrs = corrs)
 ```

### 2015-03-02 PAPER.Rmd
Fix correspondence indent
Fix correspondence

@@
 
-  \* Correspondence: <pschloss@umich.edu>
-  734.647.5801
-  Department of Microbiology and Immunology
-  University of Michigan
-  5618 Medical Sciences II
-  1500 W. Medical Center Dr.
-  Ann Arbor, MI 48109
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;\* Correspondence: <pschloss@umich.edu>
+
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;734.647.5801
+
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;Department of Microbiology and Immunology
+
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;University of Michigan
+
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;5618 Medical Sciences II
+
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;1500 W. Medical Center Dr.
+
+$nbsp;$nbsp;$nbsp;$nbsp;$nbsp;Ann Arbor, MI 48109
 

### 2015-03-02 PAPER.Rmd
Fix correspondence

--- 2015-03-02 PAPER.Rmd
@@
 
-  \* Correspondence: <pschloss@umich.edu>
-  734.647.5801
-  Department of Microbiology and Immunology
-  University of Michigan
-  5618 Medical Sciences II
-  1500 W. Medical Center Dr.
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\* Correspondence: <pschloss@umich.edu>
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;734.647.5801
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Department of Microbiology and Immunology
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;University of Michigan
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;5618 Medical Sciences II
+
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;1500 W. Medical Center Dr.
+
 Ann Arbor, MI 48109

### 2015-03-02 PAPER.Rmd
Fix correspondence
Fix superscripts

@@
 
-Alyxandria M. Schubert<sub>1</sub>, Hamide Sinani<sub>1</sub>, and Patrick D.
-Schloss<sub>1</sub>\*
+Alyxandria M. Schubert<sup>1</sup>, Hamide Sinani<sup>1</sup>, and Patrick D.
+Schloss<sup>1</sup>\*
 
@@
 
-Ann Arbor, MI 48109
+&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Ann Arbor, MI 48109
 
@@
 Subsequently, 24 hours after inoculation *C.
-difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sub>7</sub>-10<sub>8</sub> CFU/g) (**Figure 1**).
+difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g) (**Figure 1**).
 We calculated the diversity of all of the communities using the Inverse Simpson (**Supplemental Figure 1**) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.

### 2015-03-03 Figure5/abxD01.optimizingModel.Rmd
Eliminate space from figure folder names

--- 2015-01-30 Figure 6/abxD01.optimizingModel.Rmd
@@
 
+Going along with this I did correlation calculations between the delay data and nextDayCFU:
+```{r, eval=FALSE}
+
+meta <- read.csv('~/Desktop/mothur/abxD01/rf/abxD01.final.an.unique_list.0.03.subsample.0.03.pick.shared.rf.delay.regression.logtrans.filter16mintot.noUntr.csv',header=T)
+#meta<-meta[1:96] #change based on number of OTUs above .05%, then add 2 for first two cols
+meta$nextDayCFU <- 10^(meta$nextDayCFU)
+c<-1
+otu <- c()
+cor.spear <- c()
+pval.spear <- c()
+cor.ken <- c()
+pval.ken <- c()
+for(i in 3:length(meta)){
+  otu[c] <- colnames(meta[i])
+  cor.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$estimate
+  pval.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$p.value
+  cor.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$estimate #good to see because kendall handles ties
+  pval.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
+  c <- c+1
+}
+
+pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
+pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons
+
+results = NULL
+results <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
+colnames(results) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
+results <- results[order(results[,3]),]  #order by pvalue column=3
+write.table(results[1:dim(results)[1],], file="~/Desktop/mothur/abxD01/correlation/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.delay.noUntr.results.txt", sep="\t", row.names=FALSE)
+
+```
+
 I wanted to use the information in the titration data set to improve the model.
@@
 ```
-
-The % variance explained was 90.95. Otu13, followed by 3 and 1 were the most obvious influences on the model.
-
-I've also performed this same script to a file without the untreated titration groups.
-This yields 82.69% variance explained.
-Otu3 is the most important OTU in this dataset, followed by 13, 1, 20, 42.
-In fact all the top OTUs are mostly porphyromonadaceae, a lachnospiraceae, and alistipes.
-It almost seems from this data like resistance is more about the loss of these protective bugs as opposed to growth of *C.
-difficile* promoting bacteria.
 

### 2015-03-03 PAPER.Rmd
Change file paths within dmd, re-render html

@@
 
-****C.
-difficile* Preparation and Challenge*** In all our experiments we used *C.
+***C.
+difficile Preparation and Challenge*** In all our experiments we used *C.
 difficile* strain 630Δerm [cite dena lyras].
@@
 
-![**Figure 1.**](Figure 1/figure1.png)
+![**Figure 1.**](Figure1/figure1.png)
 
@@
 
-file<-read.csv("~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
 
@@
 library(Hmisc)
-stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+stackedbarcharts(file="~/Documents/Github/abxD01/Figure1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
 
-# stackedbarcharts(file="~/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
+# stackedbarcharts(file="~/Documents/Github/abxD01/Figure1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
 
@@
 
-![**Figure 2.**](Figure 2/figure2.png)
+![**Figure 2.**](Figure2/figure2.png)
 
@@
 
-```{r, eval=TRUE}
+```{r, eval=TRUE, warning=FALSE}
 
-meta <- read.table('~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
+meta <- read.table('~/Documents/Github/abxD01/Figure2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
 #meta<-meta[1:96] #change based on number of OTUs above .05%, then add 2 for first two cols
@@
 results <- results[order(results[,3]),]  #order by pvalue column=3
-#write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure 2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
+#write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
 print(results[1:dim(results)[1],])
@@
 
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 
 ##Paper Figure: correlations for topdose 
-otuCorrs<-read.csv("~/Documents/Github/abxD01/Figure 2/corr_avgAbove0.001_topdose2_stripchart.csv", header=T)
+otuCorrs<-read.csv("~/Documents/Github/abxD01/Figure2/corr_avgAbove0.001_topdose2_stripchart.csv", header=T)
 
@@
 
-![**Figure 3.**](Figure 3/figure3.png)
+![**Figure 3.**](Figure3/figure3.png)
 
@@
 ```{r, eval=TRUE}
-file<-read.csv("~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
 file<-file[!grepl("untr", file$expgroup),]
@@
 
-stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 3/alltitrations_tx2_barchart_ids.csv", excludeGroup="untr", divide = TRUE)
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure3/alltitrations_tx2_barchart_ids.csv", excludeGroup="untr", divide = TRUE)
 
@@
 
-![**Figure 4.**](Figure 4/figure4.png)
+![**Figure 4.**](Figure4/figure4.png)
 
@@
 ```{r, eval=TRUE}
-file<-read.csv("~/Documents/Github/abxD01/Figure 5/abxD01.final.tx2.subsample.alldelay.relabund.csv", header=T)
+file<-read.csv("~/Documents/Github/abxD01/Figure4/abxD01.final.tx2.subsample.alldelay.relabund.csv", header=T)
 file<-file[file$expgroup!="1untr",]
@@
 
-stackedbarcharts(file = "~/Documents/Github/abxD01/Figure 4/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure 4/alldelay_tx2_barchart_ids.csv", divide = TRUE)
+stackedbarcharts(file = "~/Documents/Github/abxD01/Figure4/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure4/alldelay_tx2_barchart_ids.csv", divide = TRUE)
 
@@
 
-![**Figure 5A.**](Figure 5/figure5a_9x9.png)
+![**Figure 5A.**](Figure5/figure5a_9x9.png)
 
-![**Figure 5B.**](Figure 5/figure5b_6x4.5.png)
+![**Figure 5B.**](Figure5/figure5b_6x4.5.png)
 
-![**Figure 5C.**](Figure 5/figure5c.png)
+![**Figure 5C.**](Figure5/figure5c.png)
 
@@
 ```
+
 > This feeds into the next analysis section.
@@
 
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 predictions <- predict(toptitdel.rf, newdata=toptitdel[,-c(1,2)])
@@
 
-### Figure 5B: Importance Plot
-```{r, eval=TRUE}
+> ### Figure 5B: Importance Plot
+
+```{r, eval=FALSE}
 topimp<-topimp[order(topimp[,1], decreasing=FALSE),]
-ids <- read.csv(file = "~/Documents/Github/abxD01/Figure 5/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
+ids <- read.csv(file = "~/Documents/Github/abxD01/Figure5/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
 labels<- paste0(ids$name, " (", ids$otuname, ")")

### 2015-03-03 Figure5/RFmodels.r
Clean up unused files

@@ -1,796 +1,0 @@
-
-
-
-
-# ############################
-# # Random forest models 2/20/15, using otus=50 (0.875% cutoff) 
-# ############################
-#
-
-data<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.csv", header=T)
-
-means <- as.data.frame(matrix(nrow=nlevels(data$expgroup), ncol=0))
-row.names(means) <- levels(data$expgroup)
-for(i in 3:length(data)){
-  
-  df <- aggregate( data[,i] ~ data$expgroup, data=data, mean )
-  names(df)[2] <- names(data)[i]
-  
-  means <- as.data.frame(cbind( means, df[,2]))
-  names(means)[dim(means)[2]] <- names(data)[i]
-  
-}
-
-means <- rbind(means, max=-1)
-for(j in 1:length(means)){
-  means["max", j] <- max(means[,j])
-}
-
-means <- rbind(means, relabund=-1)
-for(k in 1:length(means)){
-  means["relabund", k] <- (means[17,k])/1625
-}
-
-means <- rbind(means, percent=-1)
-for(k in 1:length(means)){
-  means["percent", k] <- (means[18,k])*100
-}
-
-meanOTU <- means[,-1] #remove the nextDayCFU column
-
-set <- NULL
-set <- seq(0,3, by=0.01)
-numOTUxThreshold <- NULL
-numOTUxThreshold <- as.data.frame(cbind(numOTUxThreshold, set))
-names(numOTUxThreshold)[1] <- "threshold"
-numOTUxThreshold <- cbind(numOTUxThreshold, numOTU=NA)
-for( i in 1:(length(set)) ) {
-  
-  test <- meanOTU[ , meanOTU[19,] >= set[i] ]
-  #test2 <- sum( meanOTU[19,] >= set[i] )
-  
-  numOTUxThreshold[i, 2] <- length(test)
-  
-}
-
-plot(numOTUxThreshold$numOTU ~ numOTUxThreshold$threshold, xlab="Avg % RelAbund Threshold", ylab="# OTUs included")
-abline(v=seq(0,3, by=.1), col="light gray")
-abline(h=seq(1,300, by=10), col="light gray")
-
-otusAtPercAbund <- function(percThresh, meanOTU){
[…]

### 2015-03-03 Figure5/abxD01.leaps.all.R

@@ -1,41 +1,0 @@
-library(leaps)
-
-topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.csv", header=T)
-td<-topdose[,-1]
-attach(td)
-ids<-names(td)
-ids = ids[-1]
-
-leaps.all<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, really.big=T)
-#shows the best models and which variables to include
-summary(leaps.all)
-#shows the value for each parameter in the model along with intercept
-vcov(leaps.all, 50) ##the second parameter corresponds to model number
-#coef
-coef(leaps.all, 1:2)
-#scales=bic, Cp, and adjr2, r2
-#plot(leaps.all,scale="r2")
-plot(leaps.all,scale="adjr2")
-plot(leaps.all,scale="bic")
-
-
-
-
-
-topdose<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.topdose2.logtrans.csv", header=T)
-td<-topdose[,-1]
-attach(td)
-ids<-names(td)
-ids = ids[-1]
-
-leaps.all<-regsubsets(nextDayCFU ~ ., data=td, nbest=3, nvmax=10, really.big=T)
-#shows the best models and which variables to include
-summary(leaps.all)
-#shows the value for each parameter in the model along with intercept
-vcov(leaps.all, 50) ##the second parameter corresponds to model number
-#coef
-coef(leaps.all, 1:2)
-#scales=bic, Cp, and adjr2, r2
-#plot(leaps.all,scale="r2")
-plot(leaps.all,scale="adjr2")
-plot(leaps.all,scale="bic")

### 2015-03-03 Figure5/abxD01.lmModel.compare2.R

@@ -1,1256 +1,0 @@
-#######################
-# 2/9/15
-# Looked for OTUs above different threshold cutoffs. 
-# If at least 1 in 16 experimental groups had it at this cutoff avg abundance,
-# then it was included. 
-# Made graph showing number of OTUs included as function of avg abundance cutoff. 
-# Based on this going to try using a 0.005 relabund cutoff or 0.5%, which leaves 81 OTUs included.
-
-data<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.csv", header=T)
-
-means <- as.data.frame(matrix(nrow=nlevels(data$expgroup), ncol=0))
-row.names(means) <- levels(data$expgroup)
-for(i in 3:length(data)){
-  
-  df <- aggregate( data[,i] ~ data$expgroup, data=data, mean )
-  names(df)[2] <- names(data)[i]
-  
-  means <- as.data.frame(cbind( means, df[,2]))
-  names(means)[dim(means)[2]] <- names(data)[i]
-  
-}
-
-means <- rbind(means, max=-1)
-for(j in 1:length(means)){
-  means["max", j] <- max(means[,j])
-}
-
-means <- rbind(means, relabund=-1)
-for(k in 1:length(means)){
-  means["relabund", k] <- (means[17,k])/1625
-}
-
-means <- rbind(means, percent=-1)
-for(k in 1:length(means)){
-  means["percent", k] <- (means[18,k])*100
-}
-
-meanOTU <- means[,-1] #remove the nextDayCFU column
-
-set <- NULL
-set <- seq(0,3, by=0.01)
-numOTUxThreshold <- NULL
-numOTUxThreshold <- as.data.frame(cbind(numOTUxThreshold, set))
-names(numOTUxThreshold)[1] <- "threshold"
-numOTUxThreshold <- cbind(numOTUxThreshold, numOTU=NA)
-for( i in 1:(length(set)) ) {
-   
-  test <- meanOTU[ , meanOTU[19,] >= set[i] ]
-  #test2 <- sum( meanOTU[19,] >= set[i] )
-
-  numOTUxThreshold[i, 2] <- length(test)
-  
-}
-
[…]

### 2015-03-03 Figure5/lm.builds.avg0.005bygroup.r

@@ -1,758 +1,0 @@
-
-
-############################
-# Uncurated Approach 2/18/15, using new shared file with top 20 OTUs by toptitdelrf
-############################
-library(leaps)
-
-topdose<-read.csv("~/Desktop/mothur/abxD01/model/shared.topdose.noNewUntr.logtrans.20OTUs.csv", header=T)
-
-actual <-NULL
-actual<-as.data.frame(topdose[,2]) #save the actual results in new df
-row.names(actual) <- topdose[,1] #save the group names
-td<-topdose[,-1] 
-attach(td)
-#detach(td)
-ids<-names(td)
-ids = ids[-1] #ids of OTUs in topdose
-
-#Perform an exhaustive search for the best linear models with different numbers of parameters.
-#inGroup <- c(which(ids == "Otu00007"), which(ids == "Otu000020"), which(ids == "Otu00039"), which(ids == "Otu00013"), which(ids == "Otu00003"), which(ids == "Otu00015"), which(ids == "Otu00053"))
-inGroup <- c(NULL)
-outGroup <- c(which(ids == "Otu00006"))
-leaps.build<-regsubsets(nextDayCFU ~ ., data=td, nbest=5, nvmax=10, force.in=inGroup, force.out=outGroup)
-td.models <- buildAllLM(all.leaps.models = summary(leaps.build)$which, data = toptitdel, actual=actual)
-
-
-
-leaps.plots(leaps.build, 5, 11)
-models6<-buildLM(leaps.models = (getModels(leaps.build, 6)), data = td, actual=actual)
-
-
-library(plotrix)
-abbrNames <- substr(leaps.build$xnames, 6, 8)
-abbrNames <- abbrNames[-1]
-subsets(leaps.build, names=abbrNames, statistic="bic", legend=FALSE, min.size=5, abbrev=6, cex.subsets=.5, las=1, xlim=c(5, 11), main=leaps.build$call)
-addtable2plot(x = 6, y = -150, table = lms, cex=.5, xpad=.5)
-#mtext("topdose.uncrurated.avg0.01")
-
-detach(td)
-
-####################### now try to use toptit data
-####################### 
-
[…]

### 2015-03-03 R scripts/stacked.barcharts.r
Change layout to include one overall y axis

@@ -291,16 +291,24 @@ stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, g
   matrix.se<-as.matrix(sort_stderr)
   
   
-  
-  
   #####################################################################################################################
   #####################################################################################################################
   #####################################################################################################################
   ###PLOT PARAMETERS
   library(plotrix)
+  library(Hmisc)
   if(graphbyphyl==FALSE && divide==FALSE){
-    par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
-    par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+    #par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
+  #  par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0), las=2) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+   
+    par(mar=c(0.5,0.5,0.5,0.5))
+    design <- matrix(1:numgr, nrow=numgr, byrow=T)
+    design <- cbind(rep((numgr+1), numgr), design)
+    design <- rbind(design, c(0, numgr+2))
+    layout(design, widths=c(0.1,1), heights=c(rep(1, numgr), 1))
+    
+    
+    
     color_transparent <- adjustcolor("black", alpha.f = 0.1) 
     color <- gray.colors((numphyla+1), start=0, end=1, alpha=NULL)
     
@@ -332,12 +340,15 @@ stackedbarcharts <- function(file, fileIDS, graphLabels=NULL, sortbyphyl=TRUE, g
           , expr = errbar(x, y, y+se, y-se, add=T, pch=".", cap=.01)
         )
         
-        axis(2, las=1, at=c(.001, .01, .1, 1), labels=c(0, .01, .1, 1), cex.axis=1.1)
-        mtext(abx[j], side=2, line=6, cex=.8)
-        mtext(avgCD[j], side=2, line=4.5, cex=.8)
-        abline(h=c(0.001, 1), lwd=3) #min/max
[…]

### 2015-03-03 PAPER.Rmd
Make small adjustments to figures

@@
 
-```{r, eval=FALSE}
+```{r, eval=FALSE, warning=FALSE}
 abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
@@
 stackedbarcharts(file="~/Documents/Github/abxD01/Figure1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
-
-# stackedbarcharts(file="~/Documents/Github/abxD01/Figure1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure1/topdose_tx1_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
 
@@
 
-![**Figure 2.**](Figure2/figure2.png)
+![**Figure 2.**](Figure2/figure2_9x6.5.png)
 
 **Figure 2.
-TITLE.** Species level OTUs (3% cutoff) found on day 0 were correlated with the level of *C.
+Broader taxonomic groups share similar relationships with *C.
+difficile*.** Species level OTUs (3% cutoff) found on day 0 were correlated with the level of *C.
 difficile* CFU/g feces on day 1 using Spearman’s rank based correlation coefficient.
@@
 pch <- c(6,6,6,6, 1,1, 1, 1, 1, 1, 1, 1,0, 0, 9, 2, 2, 4)
-par(mar=c(12, 7, 2, 2) +0.1, mgp=c(4, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
+par(mar=c(8, 6, 2, 2) +0.1, mgp=c(4, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
 stripchart(otuCorrs$corr ~ otuCorrs$graphOrder, vertical=TRUE, ylab="Spearman Correlation", ylim=c(-1, 1), cex.axis=1, cex=2, pch=pch, lwd=2, col="black",  method="jitter",  jitter=.2, cex.lab=1.1, yaxt="n", xaxt="n")
@@
 family.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "family"])
-text(1:nlevels(otuCorrs$family)+.1, par("usr")[3]-0.05, label=(family.labels), xpd=NA, pos=2, srt=45, cex=1)
+text(1:nlevels(otuCorrs$family)+.3, par("usr")[3]-0.05, label=(family.labels), xpd=NA, pos=2, srt=45, cex=1)
 axis(2, cex.axis=1, at=seq(-1,+1, by=.1), las=1)
@@
 phylum.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "phyl"])
-leg<-legend("top", legend=phylum.labels, pch=c(6, 1, 0, 9, 2, 4), cex=0.7, bty="o", horiz=TRUE, pt.cex=1.5, text.width=2.5)
+leg<-legend("top",  legend=phylum.labels, pch=c(6, 1, 0, 9, 2, 4), cex=1, bty="o", horiz=TRUE, pt.cex=1, text.width=2.5)
 

### 2015-03-03 PAPER.Rmd
Merge Fig 5ab

@@
 
-![**Figure 5A.**](Figure5/figure5a_9x9.png)
-
-![**Figure 5B.**](Figure5/figure5b_6x4.5.png)
+![**Figure 5AB.**](Figure5/figure5ab.png)
 
@@
 
-# dose <- all[1:99,cols]
-# titr <- all[100:153,cols]
-# del <- all[154:179, cols]
-# toptit <- all[1:153,cols]
-# titdel <- all[c(100:153, 154:179),cols]
-# topdel <- all[c(1:99, 154:179),cols]
 toptitdel <- all[,cols]
@@
 imp1<-imp[order(imp[,1], decreasing = TRUE),]
-#imp2<-imp[order(imp[,2], decreasing = TRUE),]
 topimp <- imp1[1:15,]
-#toppurity <- imp2[1:15,]
 #write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)
-#write.table(toppurity, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importanceNode.txt", sep="\t", row.names=TRUE)
+
 ```
@@
 
-par(mfrow=c(1, 1)) 
-par(mar=c(5, 5, 4, 2) +0.1, mgp=c(3, 1, 0), las=1) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+#par(mfrow=c(1,2))
+design<-as.matrix(1:2)
+design<-cbind(design, c(3,3))
+layout(design, widths=c(1,1), heights=c(.3, 1))
+
+plot.new()
+par(mar=c(rep(0.5, 4)))
+legend("bottom", legend=legend.labels, pch=pch, col=colors, pt.cex=1,  cex=1, bty="n", ncol = 2)
+
+par(mar=c(5, 5, 0.5, 0.5) +0.1, mgp=c(3, 1, 0), las=1) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
 plot(results[results[,1]=="control",2], 
@@
 }
-legend(x=7, y=3, legend=legend.labels, pch=pch, col=colors, pt.cex=1,  cex=.6, bty="n")
 
@@
 ids <- read.csv(file = "~/Documents/Github/abxD01/Figure5/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
-labels<- paste0(ids$name, " (", ids$otuname, ")")
-par(mfrow=c(1, 1)) #+1 to give extra labeling space
-par(mar=c(5, 15, 0.5, 2) +0.1, mgp=c(3, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+#par(mfrow=c(1, 1)) #+1 to give extra labeling space
+par(mar=c(5, 15, 0.5, 1) +0.1, mgp=c(3, 1, 0), las=0) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
 plot(topimp[,1], 1:15, xlab="% Increase in Mean Squared Error", yaxt="n", ylab="", pch=16, cex=1.5, xlim=c(20, 80))
-axis(2, at = c(1:15), labels = labels, las=1, cex.axis=.9 )
+axis(2, at = c(1:15), labels = rep("", 15), las=1, cex.axis=.9 )
+for(z in 1:(dim(ids)[1])){
+  name<-as.character(ids$name[z])
+  otuname<- as.character(ids$otuname[z])
+  mtext(text = bquote(italic(.(name))*" ("*.(otuname)*")"), at=z, side = 2, line = 1, las=2)
+}
 abline(h=c(1:15), lty="dashed", col='black')
+
 ```

### 2015-03-04 PAPER.Rmd
Fix code/section for supplemental figure1

@@
 
-![**Supplemental Figure 1.**](supplemental/FigS1_topdose_inverseSimp.pdf)
+![**Supplemental Figure 1.**](supplemental/sfigure1.png)
 
@@
 > Figure making:
->
+
+```
+# In mothur:
+summary.single(calc=invsimpson-shannon, shared=abxD01.final.an.unique_list.shared, subsample=1625)
+
+```
+> Using these results, I selected the samples within the original antibiotic set.
+Then I calculated the spearman correlation of *C.
+difficile* CFU/g feces on day 1 with the level of diversity found in the community on day 0 (day of challenge).
+Both inverse simpson and shannon indices were included.
+
+```{r, eval=FALSE}
+
+diversity<-read.csv("~/Documents/Github/abxD01/supplemental/topdose_diversity.csv",  header=T)
+diversity$nextDayCFU <- log((diversity$nextDayCFU+1), 10)
+
+#correlation of top dose
+pval.spear <- c()
+cor.spear <- c()
+count <- 1
+for(i in 4:5){
+  cor.spear[count] <- cor.test(diversity[,3],diversity[,i], method="spearman")$estimate
+  pval.spear[count] <- cor.test(diversity[,3],diversity[,i], method="spearman")$p.value
+  count <- count + 1 
+}
+corr.results <- as.matrix(cbind(cor.spear, pval.spear))
+rownames(corr.results) <- names(diversity)[4:5]
+
+# Plot inverse simpson vs. C. difficile levels
+color<-rainbow(7)
+colors.top <- c(control = "black",
+          cipro = color[1],
+          clinda = "#FFCC00",
+          vanc.625 = "#00CC00",
+          strep5 = "#33FFFF",
+          cef.5 = color[5],
+          amp.5 = "#6600FF",
+          metro1 = color[7])
+
+# shapes
+pch.top <- c(control = 1,
+         cipro = 16,
+         clinda = 16,
+         vanc.625 = 16,
+         strep5 = 16,
+         cef.5 = 16,
+         amp.5 = 16,
+         metro1 = 16)
+
+#these are needed for the legend
+legend.labels.top <- c(control = "Control",
+            cipro = "Ciprofloxacin",
+            clinda = "Clindamycin",
+            vanc.625 = "Vancomycin",
+            strep5 = "Streptomycin",
+            cef.5 = "Cefoperazone",
+            amp.5 = "Ampicillin",
+            metro1 = "Metronidazole")
+
+par(mar=c(5, 5, 1, 1) +0.1, mgp=c(3, 1, 0)) 
+par(mfrow=c(1,1))
+
+plot(diversity$invsimpson[diversity$expgroup=="control"], 
+     diversity$nextDayCFU[diversity$expgroup=="control"], 
+     pch=pch.top[1], col=colors.top[1], xlim=c(0, 20), ylim=c(0,10),  
+     cex=1.5, yaxt="n",  xaxt="n", cex.lab=1,
+     ylab=as.expression("Log " ~ italic("C. difficile") ~ " CFU/g feces"), xlab="Inverse Simpson Index")
+
+axis(1, las=1, cex.axis=1, at=seq(0,20, by=2))
+axis(2, las=1, cex.axis=1, at=seq(0,10, by=1))
+
+for(i in 2:(length(unique(diversity$expgroup)))){
+  points(diversity$invsimpson[diversity$expgroup==names(legend.labels.top)[i]],
+         diversity$nextDayCFU[diversity$expgroup==names(legend.labels.top)[i]], 
+         pch=pch.top[i], col=colors.top[i], cex=1.5)
+}
+
+legend(16, 6, legend=legend.labels.top, pch=pch.top, col=colors.top,  pt.cex=1.5, cex=1, bty="n", y.intersp=1.5)
+
+```
 
@@
           cef.1 = color[5],
-          amp.5 = color[6],
-          amp.5d = color[6],
+          amp.5 = "#6600FF",
+          amp.5d = "#6600FF",
           metro1 = color[7],

### 2015-03-04 R scripts/modelRfunctions.R

@@ -96,7 +96,7 @@ graphOTUxCD <- function(otus, ids, corrs){
     
     test <- grep("aceae", ids[i, 1])
     name<-as.character(ids$name[i])
-    number<-as.character(ids$otus[i])
+    number<-as.character(ids$otuname[i])
     corr<-as.character(signif(corrs$corSpear[i], 2))
     library(base)
     graphTitle <- bquote(bolditalic(.(name)) *" " *bold(.(number)) *bold(", ") *bold(rho) *bold(" = ") *bold(.(corr)))

### 2015-03-04 PAPER.Rmd
Adjust legend in sfig1

@@
      diversity$nextDayCFU[diversity$expgroup=="control"], 
-     pch=pch.top[1], col=colors.top[1], xlim=c(0, 20), ylim=c(0,10),  
+     pch=pch.top[1], col=colors.top[1], xlim=c(0, 25), ylim=c(0,10),  
      cex=1.5, yaxt="n",  xaxt="n", cex.lab=1,
@@
 
-axis(1, las=1, cex.axis=1, at=seq(0,20, by=2))
+axis(1, las=1, cex.axis=1, at=seq(0,25, by=5))
 axis(2, las=1, cex.axis=1, at=seq(0,10, by=1))
@@
 
-legend(16, 6, legend=legend.labels.top, pch=pch.top, col=colors.top,  pt.cex=1.5, cex=1, bty="n", y.intersp=1.5)
+cor<-signif(results[1,1],2)
+legend("topright", inset=0.05, legend=legend.labels.top, pch=pch.top, col=colors.top,  pt.cex=1.5, cex=1, bty="n", y.intersp=1.5)
 

### 2015-03-06 PAPER.Rmd
Update sections to supfig2

@@
 
-cor<-signif(results[1,1],2)
 legend("topright", inset=0.05, legend=legend.labels.top, pch=pch.top, col=colors.top,  pt.cex=1.5, cex=1, bty="n", y.intersp=1.5)
@@
 
+![**Supplemental Figure 2.**](supplemental/sfigure2.png)
+
+**Supplemental Figure 2.
+Structural distances between untreated and recovery days.** This figure shows the average theta YC distance between pretreatment and recovery days by antibiotic group.
+Day -11 represent untreated mice.
+Day 0 represents day of *C.
+difficile* challenge.
+Error bars show standard error of the mean.
+
+--------
+
 > ### Differences between recovery days' thetayc
-> 
+
+```{r, eval=FALSE}
+file = "~/Desktop/mothur/abxD01/abxD01.final.an.unique_list.0.03.recovery.vancomycin.thetayc.0.03.lt.ave.dist"
+diffs <- getDistDiffs(file=file)
+
+metadata <- read.delim(file="~/Desktop/mothur/abxD01/searchme.txt", header = T, stringsAsFactors=F)
+metadata$sample[1951] <- "007-1D1" # ID was messed up
+names(metadata)[3] <- "group"
+
+plot.dist.from(dist.from = diffs)
+
+```
+
 > ### Statistical comparison of delayed and on time CFU results

### 2015-03-06 R scripts/modelRfunctions.R

@@ -482,12 +482,9 @@ getDistDiffs<-function(file, setRange=NULL){
   } # end for(i in 1:(length(uniq.ids)))
   
   dist.from <- dist.from[,-1]
+  
   return(dist.from)
 }
-
-metadata <- read.delim(file="~/Desktop/mothur/abxD01/searchme.txt", header = T, stringsAsFactors=F)
-metadata$sample[1951] <- "007-1D1"
-names(metadata)[3] <- "group"
 
 plot.dist.from <- function(dist.from){
   matchPos <- match(colnames(dist.from), metadata$sample)
@@ -500,10 +497,10 @@ plot.dist.from <- function(dist.from){
   delay.group <- group[time=="delayed" & day == "-11"]
   ids <- cbind(delay.ids, delay.abx, delay.group)
   
-  dist.from <- dist.from[ids, ]
-  na.ind <- which(is.na(dist.from))
-  max <- max(c(dist.from)[-na.ind])
-  
+#   dist.from <- dist.from[ids, ]
+#   na.ind <- which(is.na(dist.from))
+#   max <- max(c(dist.from)[-na.ind])
+#   
   colors<-rainbow(length(unique(delay.abx)))
   colors <- c(amp="#6600FF",
               metro="#FF00DBFF")
@@ -512,14 +509,37 @@ plot.dist.from <- function(dist.from){
            ylim=c(.2,1), xlim=c(-5, 0))
   # Fill remaining points
   for(i in 1:(dim(ids)[1])){
-    notNApts <- !is.na(dist.from[6:11,ids[i]])
+    notNApts <- !is.na(dist.from[6:11,ids[i,1]])
     
     points(row.names(dist.from)[which(notNApts)+5], dist.from[which(notNApts)+5,ids[i]], 
            col=colors[delay.abx[i]], pch=which(unique(delay.group)==delay.group[i]), cex=1)
     lines(row.names(dist.from)[which(notNApts)+5], dist.from[which(notNApts)+5,ids[i]], 
           col=colors[delay.abx[i]], pch=which(unique(delay.group)==delay.group[i]), cex=1)
   }
-  legend("bottom", legend=names(colors), col=colors, pch=1)
+  legend("bottom", legend=names(colors), col=colors, pch=1, bty="n")
+  
+  subdist <- dist.from[6:11, ids[,1]]
+  ampdist <- subdist[,ids[,2]=="amp"]
[…]

### 2015-03-07 PAPER.Rmd
Change header
Add diversity to sfigure 2. Update code in paper.

--- 2015-03-04 PAPER.Rmd
@@
 
-> ### Differences between recovery days' thetayc
-> 
+![**Supplemental Figure 2.**](supplemental/sfigure2ab.png)
+
+**Supplemental Figure 2.
+Changes between untreated and recovery days.** A) The average thetaYC distance between pretreatment (day -11) and recovery days (days -5 through -1) is represented by circles connected by solid lines.
+Dashed lines with the same symbol indicate individual mice from the same cage.
+Antibiotic treatment is indicated by color.
+B) The average level of the inverse simpson diversity index is plotted by antibiotic treatment across time.
+Error bars show standard error of the mean.
+Mice were challenged with *C.
+difficile* on day 0.
+
+--------
+
+> ### Supplemental Figure 2A: Differences between recovery days' thetayc
+
+```{r, eval=FALSE}
+file <- "~/Desktop/mothur/abxD01/abxD01.final.an.unique_list.0.03.recovery.vancomycin.thetayc.0.03.lt.ave.dist"
+diffs <- getDistDiffs(file=file)
+
+#metadata <- read.delim(file="~/Desktop/mothur/abxD01/searchme.txt", header = T, stringsAsFactors=F)
+metadata$sample[1951] <- "007-1D1" # ID was messed up
+names(metadata)[3] <- "group"
+
+par(mfrow=c(2, 1))
+par(mar=c(3, 5, 0.5, 0.5), las=2, mgp=c(3, 1, 0))
+plot.dist.from(dist.from = diffs)
+
+# Arrange diversity data for supplemental figure 2B
+
+diversity <- read.delim(file="~/Desktop/mothur/abxD01/abxD01.final.an.unique_list.groups.ave-std.summary", sep="\t", header=T, stringsAsFactors=F)
+
+diversity$group[2046] <- "007-1D1"
+diversity$group[124] <- "007-1D1"
+
+div.ind <- match(diversity$group[diversity$method=="ave"], metadata$sample)
+div.group <- metadata$group[div.ind]
+div.time <- metadata$time[div.ind]
+div.abx <- metadata$abx[div.ind]
+div.day <- metadata$day[div.ind]
+
+delay.invsimp <- diversity$invsimpson[div.time=="delayed" & diversity$method=="ave"]
+delay.shannon <- diversity$shannon[div.time=="delayed" & diversity$method=="ave"]
+delay.abx <- div.abx[div.time=="delayed" & diversity$method=="ave"]
+delay.day <- div.day[div.time=="delayed" & diversity$method=="ave"]
+delay.group <- div.group[div.time=="delayed" & diversity$method=="ave"]
+
+delay.diversity <- cbind(group=delay.group, day=delay.day, abx=delay.abx, invsimp=delay.invsimp, shannon=delay.shannon)
+
+recov.diversity <- matrix(nrow=7)
+row.names(recov.diversity) <- c(-11, -5:0)
+
+amp.div <- matrix(nrow=7, ncol=2)
+row.names(amp.div) <- c(-11, -5:0)
+colnames(amp.div) <- c("amp.avg", "amp.se")
+library(plotrix)
+for(j in 1:7){
+  matches <- delay.diversity[,3]=="amp" & delay.diversity[,2]==row.names(amp.div)[j]
+  amp.div[j, "amp.avg"] <- mean(as.numeric(delay.diversity[matches, "invsimp"]))
+  amp.div[j, "amp.se"] <- std.error(as.numeric(delay.diversity[matches, "invsimp"]))
+}
+
+metro.div <- matrix(nrow=7, ncol=2)
+row.names(metro.div) <- c(-11, -5:0)
+colnames(metro.div) <- c("metro.avg", "metro.se")
+for(j in 1:7){
+  matches <- delay.diversity[,3]=="metro" & delay.diversity[,2]==row.names(metro.div)[j]
+  metro.div[j, "metro.avg"] <- mean(as.numeric(delay.diversity[matches, "invsimp"]))
+  metro.div[j, "metro.se"] <- std.error(as.numeric(delay.diversity[matches, "invsimp"]))
+}
+
+recov.diversity <- cbind(amp.div, metro.div)
+
+# Plot for second part of supplemental figure 2
+transpose <-t(recov.diversity)
+colors <- c(amp="#6600FF",
+            metro="#FF00DBFF")
+par(mar=c(5, 5, 0.5, 0.5), las = 2, mgp=c(3, 1, 0))
+bp <- barplot(transpose[c("amp.avg", "metro.avg"), ], beside=T, ylim=c(0, 15), 
+              ylab="Inverse Simpson Diversity", xlab="", 
+              col=colors, xaxt='n' )
+xticks <- apply(bp, 2, mean)
+axis(1, at=xticks, label=c(-11, -5:0), las=1)
+mtext(text = expression(paste("Days Pre-", italic("C. difficile"), " Challenge")), side = 1, line = 3, cex = 1.5, las=0)
+abline(v=mean(xticks[1:2]), lty="dashed", lwd=3)
+box()
+
+x <- c(bp[1,], bp[2,])
+y <- as.numeric(c(transpose[1,], transpose[3,]))
+se <- as.numeric(c(transpose[2,], transpose[4,]))
+d <- as.data.frame( cbind(x=x, y=y, se=se  ) )
+
+library(Hmisc)
+with (
+  data = d
+  , expr = errbar(x, y, y+se, y-se, add=T, pch=".", cap=.01)
+)
+
+legend("topright",inset=0.08, legend=c("Ampicillin", "Metronidazole"), 
+       col=colors, pch=15, bty="n", y.intersp=1.5, pt.cex = 2.5, cex=1.2)
+
+```
+
 > ### Statistical comparison of delayed and on time CFU results

### 2015-03-07 R scripts/modelRfunctions.R

@@ -486,6 +486,9 @@ getDistDiffs<-function(file, setRange=NULL){
   return(dist.from)
 }
 
+
+
+# This function is pretty tailored to the ampicillin/metronidazole recovery data
 plot.dist.from <- function(dist.from){
   matchPos <- match(colnames(dist.from), metadata$sample)
   abx <- metadata$abx[matchPos]
@@ -496,51 +499,69 @@ plot.dist.from <- function(dist.from){
   delay.abx <- abx[time=="delayed" & day == "-11"]
   delay.group <- group[time=="delayed" & day == "-11"]
   ids <- cbind(delay.ids, delay.abx, delay.group)
-  
-#   dist.from <- dist.from[ids, ]
-#   na.ind <- which(is.na(dist.from))
-#   max <- max(c(dist.from)[-na.ind])
-#   
+
   colors<-rainbow(length(unique(delay.abx)))
   colors <- c(amp="#6600FF",
               metro="#FF00DBFF")
+  # xlab=expression(paste("Days Pre-", italic("C. difficile"), " Challenge")),
+  plot(x="", y="", 
+       ylab=paste("ThetaYC distance from Day", start), 
+       ylim=c(0,1), xlim=c(-6, 0.25), yaxt='n', xaxt='n')
+  axis(1, at=seq(-6, 0, by=1), label=c(-11, seq(-5, 0, by=1)), las= 1)
+  axis(2, at=seq(0, 1, by=0.2))
   
-  plot(x="", y="",xlab="Day", ylab=paste("ThetaYC distance from Day", start), 
-           ylim=c(.2,1), xlim=c(-5, 0))
-  # Fill remaining points
-  for(i in 1:(dim(ids)[1])){
-    notNApts <- !is.na(dist.from[6:11,ids[i,1]])
-    
-    points(row.names(dist.from)[which(notNApts)+5], dist.from[which(notNApts)+5,ids[i]], 
-           col=colors[delay.abx[i]], pch=which(unique(delay.group)==delay.group[i]), cex=1)
-    lines(row.names(dist.from)[which(notNApts)+5], dist.from[which(notNApts)+5,ids[i]], 
-          col=colors[delay.abx[i]], pch=which(unique(delay.group)==delay.group[i]), cex=1)
+  # "range" to jitter about
+  # "number" of values needed to jitter about x
+  jitter.about <- function(range=c(0,0), number=1) {
[…]

### 2015-03-07 R scripts/modelRfunctions.R
Make sfig2 symbols consistent with figure 5

@@ -555,10 +555,10 @@ plot.dist.from <- function(dist.from){
     return(result)
   }
 
-  points(-6:0, c(0, apply(ampdist, 1, avg)), pch=16, col=colors[1], cex=1.7)
-  lines(-6:0, c(0, apply(ampdist, 1, avg)), pch=16, col=colors[1], lwd=3)
-  points(-6:0, c(0, apply(metrodist, 1, avg)), pch=16, col=colors[2], cex=1.7)
-  lines(-6:0, c(0, apply(metrodist, 1, avg)), pch=16, col=colors[2], lwd=3)
+  points(-6:0, c(0, apply(ampdist, 1, avg)), pch=8, col=colors[1], cex=2, lwd=4)
+  lines(-6:0, c(0, apply(ampdist, 1, avg)), col=colors[1], lwd=3)
+  points(-6:0, c(0, apply(metrodist, 1, avg)), pch=8, col=colors[2], cex=2, lwd=4)
+  lines(-6:0, c(0, apply(metrodist, 1, avg)),  col=colors[2], lwd=3)
   abline(v=-5.5, lty="dashed", lwd=3)
   #legend("bottom", legend=c("Ampicillin", "Metronidazole"), col=colors, pch=16, bty="n", horiz=TRUE, text.width=1, pt.cex = 1.5)
 

### 2015-03-07 PAPER.Rmd
Add Akkermansia to 5C

@@
 topimp <- imp1[1:15,]
-#write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)
+write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)
 
@@
 
-results <- cbind(expgroup, results)
+fit.results <- cbind(expgroup, results)
 
@@
             amp.5 = "Ampicillin",
-            amp.5d = "Ampicillin +5D",
+            amp.5d = "Ampicillin +5 days",
             metro1 = "Metronidazole",
-            metro1d = "Metronidazole +5D")
-
-treatments <- c(control = "Control",
-            cipro = "Ciprofloxacin",
-            clinda = "Clindamycin",
-            vanc.625 = "Vancomycin High",
-            vanc.3 = "Vancomycin Medium",
-            vanc.1 = "Vancomycin Low",
-            strep5 = "Streptomycin High",
-            strep.5 = "Streptomycin Medium",
-            strep.1 = "Streptomycin Low",
-            cef.5 = "Cefoperazone High",
-            cef.3 = "Cefoperazone Medium",
-            cef.1 = "Cefoperazone Low",
-            amp.5 = "Ampicillin",
-            amp.5d = "Ampicillin +5D",
-            metro1 = "Metronidazole",
-            metro1d = "Metronidazole +5D")
-
-#par(mfrow=c(1,2))
+            metro1d = "Metronidazole +5 days")
+
 design<-as.matrix(1:2)
 design<-cbind(design, c(3,3))
-layout(design, widths=c(1,1), heights=c(.3, 1))
+layout(design, widths=c(1,1), heights=c(.35, 1))
 
 plot.new()
-par(mar=c(rep(0.5, 4)))
-legend("bottom", legend=legend.labels, pch=pch, col=colors, pt.cex=1,  cex=1, bty="n", ncol = 2)
-
-par(mar=c(5, 5, 0.5, 0.5) +0.1, mgp=c(3, 1, 0), las=1) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
+par(mar=c(0.5, 0.5, 0.5, 0.5))
+legend("bottomright", legend=legend.labels, pch=pch, col=colors, pt.cex=1.5,  cex=1, bty="n", ncol = 2, text.width=0.4)
+
+par(mar=c(5, 5, 0.5, 0.5) +0.1, mgp=c(3, 1, 0), las=1) 
 plot(results[results[,1]=="control",2], 
@@
      ylim=c(0,9), xlim=c(0,9), xaxt='n', yaxt='n', cex=1.5)
-#mtext(bquote("r"^"2" ~ " = " ~ .(signif(rsq, 3))), side=3, line=0)
 abline(a=0, b=1, lty="dashed", lwd=2, col="black")
@@
   otuname<- as.character(ids$otuname[z])
-  mtext(text = bquote(italic(.(name))*" ("*.(otuname)*")"), at=z, side = 2, line = 1, las=2)
+  mtext(text = bquote(italic(.(name))*" ("*.(otuname)*")"), at=z, side = 2, line = 1, las=2, cex=0.9)
 }
@@
 
-results = NULL
-results <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
-colnames(results) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
-results <- results[order(results[,3]),]  #order by pvalue column=3
-#write.table(results[1:dim(results)[1],], file="~/Desktop/mothur/abxD01/correlation/toptidel.1p.correl.txt", sep="\t", row.names=FALSE)
-print(results[1:dim(results)[1],])
+correl = NULL
+correl <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
+colnames(correl) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
+correl <- correl[order(correl[,3]),]  #order by pvalue column=3
+#write.table(correl, file="~/Desktop/mothur/abxD01/correlation/toptidel.1p.correl.txt", sep="\t", row.names=FALSE)
+print(correl)
 ```
@@
 ```{r, eval=FALSE}
-corrs<-read.csv(file="~/Desktop/mothur/abxD01/correlation/toptitdel.correlation.impOTUs.csv", header=TRUE)
-corrs <- corrs[1:8,]
-otunames <-  row.names(topimp)[15:8]
-graphID <- ids[15:8,c(1,3)]
-
-graphOTUxCD(otus = otunames, ids = graphID, corrs = corrs)
+topimp.otus <-  row.names(topimp)[c(15:7)]
+temp.match <- match(topimp.otus, ids.master[,1])
+graphID <- ids.master[temp.match,c(4,3)]
+
+temp.match <- match(topimp.otus, correl[,"otu"])
+topimp.corrs <- correl[temp.match, c("corSpear", "pvalSpear")]
+row.names(topimp.corrs) <- topimp.otus
+
+graphOTUxCD(otus = topimp.otus, ids = graphID, corrs = topimp.corrs)
 ```

### 2015-03-07 R scripts/modelRfunctions.R

@@ -57,15 +57,17 @@ RF.validate <- function(data, iters){
 # otus<-otunames
 # ids<-otulabels
 
-
+# this is actually dependent on expgroup, fit.results, and legend.labels 
+# existing correctly outside of the function... 
 graphOTUxCD <- function(otus, ids, corrs){
   
   par(mar=c(0.5,0.5,0.5,0.5))
-  design <- matrix(1:8, nrow=4, byrow=T)
-  design <- cbind(c(9,9,9,9), design)
-  design <- rbind(design, c(0,10,10))
-  design <- cbind(design,c(11, 11, 11, 11, 0))
-  layout(design, widths=c(0.2,.9,.9, .6), heights=c(1,1, 1, 1,0.3))
+  design <- matrix(1:9, nrow=3, byrow=T)
+  design <- cbind(c(10,10,10), design)
+  design <- rbind(design, c(0,11,11,11))
+  layout(design, widths=c(0.2,1,1,1), heights=c(1,1, 1, 0.3))
+  
+  corrs <- as.data.frame(corrs)
   
   for(i in 1:(length(otus))){
     
@@ -81,8 +83,8 @@ graphOTUxCD <- function(otus, ids, corrs){
          xaxt='n', yaxt="n", xlab="")
     
     #plot zeroes in jitter
-    points(x=x_zeroes, actual[zeroes], pch=pch[as.character(results[zeroes,1])],
-           col=colors[as.character(results[zeroes,1])], cex=1.5)
+    points(x=x_zeroes, actual[zeroes], pch=pch[as.character(fit.results[zeroes,1])],
+           col=colors[as.character(fit.results[zeroes,1])], cex=1.5)
     
     #add other treatment points
     for(j in 2:(length(legend.labels))){
@@ -94,10 +96,9 @@ graphOTUxCD <- function(otus, ids, corrs){
     abline(v=c(0.001, 0.01, 0.1, 1), lty="longdash", col="gray")
     abline(v=0.0005, lty="dotted", lwd=2, col="black")
     
-    test <- grep("aceae", ids[i, 1])
     name<-as.character(ids$name[i])
     number<-as.character(ids$otuname[i])
-    corr<-as.character(signif(corrs$corSpear[i], 2))
+    corr<-as.character(signif(as.numeric(as.character(corrs$corSpear[i])), 2))
     library(base)
[…]

### 2015-03-07 PAPER.Rmd
Edit Methods section

@@
 
-The microbiome, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection.
-This native protection by the microbiome from invasive pathogenic species is termed colonization resistance.
-It is well established that the gut bacterial microbiome is critical in the hosts’ defense against the pathogen Clostridium difficile.
+The microbiome, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection [ref].
+This native protection by the microbiome from invasive pathogenic species is termed colonization resistance [ref].
+It is well established that the gut bacterial microbiome is critical in the hosts’ defense against the pathogen Clostridium difficile [ref].
 When this indigenous community is perturbed, this often leads to a loss of resistance.
 This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, chemotherapy, but also exposed to the environmental *C.
-difficile* spores.
+difficile* spores [ref].
 Among hospital-acquired infections in the United States, *C.
-difficile* infections (CDI) are the number one reported and are responsible for 14,000 deaths a year.
-It is not completely understood how different perturbations affect subsequent colonization resistance, nor do we have a comprehensive view of the taxonomic changes associated with a loss in colonization resistance.
-
+difficile* infections (CDI) are the number one reported and are responsible for 14,000 deaths a year [ref].
+It is not completely understood how different perturbations to the taxonomic composition of the gut microbiota result in a loss of colonization resistance.
+
+<not done>
 There is a need for more knowledge about interactions between members of the community.
@@
 These findings __
+</not done>
 
@@
 ***Animal Care.*** We used 5-8 week old C57Bl/6 mice for all our experiments.
-These mice were housed in our SPF room in the animal facility at the University of Michigan.
+These mice were reared under SPF conditions within the animal facility at the University of Michigan.
 All animal-related protocols and experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
@@
 The route of administration depended on the specific antibiotic.
-Cefoperazone (either 0.5mg/ml, 0.3mg/ml, or 0.1mg/ml), vancomycin (0.625mg/ml, 0.3mg/ml, or 0.1mg/ml), streptomycin (5mg/ml, 0.5mg/ml, or 0.1mg/ml), metronidazole (0.5mg/ml), and ampicillin (0.5mg/ml) were all administered in the mouse drinking water for 5 days.
-Ciprofloxacin (10 mg/kg) was administered via oral gavage, and clindamycin (10 mg/kg) was administered via IP injection.
-All antibiotic treated mice were given 24 hours of recovery on regular drinking water prior to *C.
+Cefoperazone (either 0.5 mg/ml, 0.3mg/ml, or 0.1 mg/ml), vancomycin (0.625 mg/ml, 0.3 mg/ml, or 0.1 mg/ml), streptomycin (5 mg/ml, 0.5 mg/ml, or 0.1 mg/ml), metronidazole (0.5 mg/ml), and ampicillin (0.5 mg/ml) were all administered in the mouse drinking water for 5 days.
+Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via IP injection.
+Mice that did not receive antibiotics were used as negative controls for these experiments.
+
+***C.
+difficile Preparation and Challenge*** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
 difficile* challenge.
-
-***C.
-difficile Preparation and Challenge*** In all our experiments we used *C.
-difficile* strain 630Δerm [cite dena lyras].
-The spores used were prepared from a single large batch made following previously established protocols [cite paul?].
-Spores were quantified within the week prior to *C.
-difficile* challenge for each experiment [established protocol I can cite?].
-Spores were quantified by boiling a small volume at 65°C for 20 minutes to kill off residual vegetative cells.
-The volume was then diluted and plated on TCCFA plates.
-These were incubated at 37°C anaerobically overnight.
-The next day *C.
-difficile* colony forming units (CFU) were counted and the concentration of spores in the working stock was back calculated.
-On the day of challenge, *C.
-difficile* spores, at a final concentration of 103, were administered to mice via oral gavage.
+*C.
+difficile* strain 630Δerm spores were used in all experiments [cite dena lyras].
+Spores were prepared from a single large batch whose concentration was determined within the week prior to each *C.
+difficile* challenge [cite paul?].
+On the day of challenge 10^3 *C.
+difficile* spores were administered to mice via oral gavage.
 Immediately following this challenge, the remaining *C.
@@
 
-***Sample Collection and Plating*** Fecal samples were freshly collected for each mouse immediately prior to the administration of any antibiotic, during any antibiotic recovery days, and days 0 and 1 post *C.
-difficile* challenge.
-Each fecal sample on day 1 was weighed and diluted under anaerobic conditions with anaerobic PBS 1:10 based on pellet weight.
-A dilution series was then plated onto TCCFA plates [cite plates] and incubated at 37°C anaerobically overnight.
-The colony forming units (CFU) were counted and calculated the subsequent day.
-
-***DNA Extraction and Sequencing*** Total bacterial DNA was extracted from each untreated and day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
+***Sample Collection and Plating*** Fecal samples were freshly collected for each mouse on the day of *C.
+difficile* challenge (day 0).
+On the day after the challenge (day 1), another fecal sample was weighed and diluted under anaerobic conditions with anaerobic PBS.
+The number of colony forming units (CFU) were counted following 24 hrs growth on TCCFA plates at 37°C under anaerobic conditions [cite plates].
+
+***DNA Extraction and Sequencing*** Total bacterial DNA was extracted from each  day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
 Sequencing preparation and sequencing was carried out as previously described (Kozich JJ et al.
@@
 These libraries were then mixed with Illumina-prepared PhiX control libraries.
-Sequencing was performed using Illumina’s MiSeq platform to produce paired 250-nt reads.
-Mock/error/error of run?
 
 ***Sequence Curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
-The full pipeline is also demonstrated online at the mother website (http://www.mothur.org/wiki/MiSeq_SOP).
 Sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
 Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 9) requiring an 80% bootstrap confidence score.
+In parallel to the fecal samples, we also sequeced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
+Analysis of the mock community data indicated that the error rate following our cuation procedure was 0.0X%.
 In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples [Schloss, reducing effects].
-For those samples with more than 1,625 sequences per samples, the range was __, the mean was __, the median was __, the median absolute deviation was __.
-All 16S rRNA gene sequence data and the MIMARKS table are available at __.
-
-***Statistical Analysis & Modeling*** Across all data, we set a relative abundance threshold of an average of 1% in at least one experimental group.
-The 44 OTUs that met this cutoff were included in the training of the model.
-We used the "randomForest" package in R to build a regression-based random forest model to predict *C.
-difficile* colonization levels on day 1 based on the microbiome data from day 0.
-All statistical analyses were conducted using R version 3.1.2.
+All 16S rRNA gene sequence data and metadata are available through the
+Sequence Read Archive under accession XXXXXX.
+
+***Statistical Analysis & Modeling*** Random forest regression models were constructed using the randomForest R package using XX,000 trees [ref]. The
+model was constructed using the relative abundance of those OTUs that had an
+average relative abundance within any antibiotic treatment group of 1%. This
+resulted in 44 OTUs. The regression was performed using the log10 transformation
+of the number of CFU per gram of fecal material as the dependent variable. All analyses were conducted using R version 3.1.2. Complete analysis scripts are
+available at the online repository for this study (https://github.com/SchlossLab/Schubert_AbxD01_mBio_2015).
 

### 2015-03-07 PAPER.Rmd
Edit first paragraph of Results and Fig 1 and S1 captions

@@
 difficile* colonization.
-It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] all have decreased resistance against this pathogen.
+It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] had decreased resistance against this pathogen.
 To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (**Table 1**) to differentially alter the microbiota and assess their resistance to *C.
@@
 The community structures of the beta-lactam-treated mice (i.e.
-cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=0.-36).
-Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=0.001) and were characterized by the predominance of *Enterobacteriaceae*.
+cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=XXX0.36XXX).
+Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=XXX0.001XXX) and were characterized by the predominance of *Enterobacteriaceae*.
 Of all groups significantly different from the untreated microbiota, the levels of *Enterobacteriaceae* increased with all treatments except with streptomycin.
 Vancomycin-treated communities saw the largest bloom in *Akkermansia*.
-In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=0.09).
+In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, XXXp=0.09XXX).
 Subsequently, 24 hours after inoculation *C.
 difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g) (**Figure 1**).
-We calculated the diversity of all of the communities using the Inverse Simpson (**Supplemental Figure 1**) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.
-difficile* colonization (both p<0.001).
-These results suggest that multiple microbiota structures are capable of being colonized by *C.
-difficile* and that diversity may be an important factor in resistance to invasion.
+We calculated the diversity of all of the communities using the Inverse Simpson and Shannon indices and observed a significant negative correlation with subsequent *C.
+difficile* colonization (both **Supplemental Figure 1**; p<XXX0.001XXX).
+These results demonstrate that a variety of non-overlapping community structures can be sensitive to *C.
+difficile* colonization.
 
@@
 The level *C.
-difficile* colonization found 24 hours post microbiota sampling is plotted in the top right corner.
-Each represented phylum is a different shade of gray.
-Error bars depict standard error of the mean.
+difficile* colonization found 24 hours post microbiota sampling is indicated in the top right for each treatment.
+Error bars indicate standard error of the mean.
 
@@
 **Supplemental Figure 1.
-Diversity is nevatively correlated with *C.
-difficile* levels.** This figure shows the relationship between the inverse simpson diversity index of microbiotas of each experimental group on day 0 and *C.
-difficile* CFU/g feces on day 1.
-Spearman's rho is plotted in the top right corner.
+The inverse of Simpson's diversity index for the community on day 0 is negatively correlated with *C.
+difficile* colonization on day 1.** Spearman's rho is plotted in the top right corner.
 

### 2015-03-07 PAPER.Rmd
Edit correlation analysis section

@@
 difficile*, we measured the correlation between the relative abundance of each species-level OTU on the day of inoculation with the level of *C.
-difficile* colonization 24 hours later across all treatment groups (Figure 2).
-Of the significant OTUs (p<0.05) with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against *C.
-difficile* (N=40) and those with a positive correlation were associated with susceptibility (N=7).
-Among various bacterial families, three were consistently negatively correlated with *C.
-difficile* colonization: *Porphyromonadaceae* (ρaverage= -0.63), *Lachnospiraceae* (ρaverage= -0.51), *Ruminococcaceae* (ρaverage= -0.53).
+difficile* colonization 24 hours later (Figure 2).
+<needs help> Of the OTUs with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against *C.
+difficile* (N=XXX40XXX) and those with a positive correlation were associated with susceptibility (N=XXX7XXX) (p<XXX0.05XXX).</needs help> Among various bacterial families, three were consistently negatively correlated with *C.
+difficile* colonization: *Porphyromonadaceae* (ρ<sub>average</sub>=XXX-0.63XXX), *Lachnospiraceae* (ρ<sub>average</sub>=XXX-0.51XXX), *Ruminococcaceae* (ρ<sub>average</sub>=XXX-0.53XXX).
 OTUs from the *Proteobacteria* phylum had strong positive relationships with *C.
 difficile* colonization.
-These included OTUs associated with the *Pseudomonas* (ρ= 0.31) and *Escherichia* (ρ= 0.56) genera.
-Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus *Lactobacillus* (family *Lactobacillaceae*; OTU23, ρ= -0.73), *Alistipes* (family *Rickenellaceae*; OTU20, ρ= -0.78), *Enterorhabdus* (phylum *Actinobacterium*, family *Coriobacteriaceae*; OTU29, ρ= -0.70), and *Streptococcus* (family *Streptococcaceae*; OTU78, ρ= 0.36).
+These included OTUs associated with the genera *Pseudomonas* (ρ=XXX0.31XXX) and *Escherichia* (ρ=XXX0.56XXX).
+Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus *Lactobacillus* (family *Lactobacillaceae*; OTU23, ρ=XXX-0.73XXX), *Alistipes* (family *Rickenellaceae*; OTU20, ρ=XXX-0.78XXX), *Enterorhabdus* (phylum *Actinobacterium*, family *Coriobacteriaceae*; OTU29, ρ=XXX-0.70XXX), and *Streptococcus* (family *Streptococcaceae*; OTU78, ρ=XXX0.36XXX).
 Overall, these results suggest that several larger groups of related OTUs, particularly within the *Firmicutes* and *Bacteroidetes* phyla, may participate in colonization resistance against *C.
@@
 difficile* CFU/g feces on day 1 using Spearman’s rank based correlation coefficient.
-Only OTUs with an average abundance of at least 0.05% across all original treatments were considered in the correlation analysis.
-All correlations shown in this graph are significant (p<0.05).
-OTUs were grouped by family.
-Asterisk * indicates unclassified at family level.
+Only OTUs with an average abundance of at least <needs help>0.05%<needs help> across all original treatments were considered in the correlation analysis.
+All correlations shown in this graph are significant after controlling for multiple comparisons using the Benjimani-Hochberg correction (p<0.05).
+OTUs are grouped by family.
+An asterisk (*) indicates that the OTU was unclassified at the family level.
 

### 2015-03-07 PAPER.Rmd
Edit titration analysis

@@
 In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples [Schloss, reducing effects].
-All 16S rRNA gene sequence data and metadata are available through the
-Sequence Read Archive under accession XXXXXX.
-
-***Statistical Analysis & Modeling*** Random forest regression models were constructed using the randomForest R package using XX,000 trees [ref]. The
-model was constructed using the relative abundance of those OTUs that had an
-average relative abundance within any antibiotic treatment group of 1%. This
-resulted in 44 OTUs. The regression was performed using the log10 transformation
-of the number of CFU per gram of fecal material as the dependent variable. All analyses were conducted using R version 3.1.2. Complete analysis scripts are
-available at the online repository for this study (https://github.com/SchlossLab/Schubert_AbxD01_mBio_2015).
+All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession XXXXXX.
+
+***Statistical Analysis & Modeling*** All analyses were conducted using R version 3.1.2.
+Complete analysis scripts are available at the online repository for this study (https://github.com/SchlossLab/Schubert_AbxD01_mBio_2015).
+
+Need text about...
+* correlation analysis
+* comparison of titration levels
+* For the random forest you use 1% within each treatment group, for the correlation you state using 0.1 and 0.01% across all samples. Need to stay consistent.
+* statistical methods used to differentiate the abundances of OTUs in the titrations
+
+Random forest regression models were constructed using the randomForest R package using XX,000 trees [ref].
+The model was constructed using the relative abundance of those OTUs that had an average relative abundance within any antibiotic treatment group of 1%.
+This resulted in 44 OTUs.
+The regression was performed using the log10 transformation of the number of CFU per gram of fecal material as the dependent variable.
 
@@
 
-***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** <transition>
-Because most of our original treatments resulted in similarly high levels of *C.
-difficile*, we hypothesized that applying smaller antibiotic perturbations to the microbiota would result in reduced sensitivity to *C.
-difficile*, as we observed in the ciprofloxacin treated mice.
-To test this, we titrated the doses of cefoperazone, streptomycin, and vancomycin given to mice (Supplemental Table 1).
-Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis.
+***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** All but the ciprofloxacin treatment resulted in similarly high levels of *C.
+difficile*.
+Therefore, we hypothesized that titrating the dose of antibiotics given to the mice would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C.
+difficile*.
+We titrated the doses of cefoperazone, streptomycin, and vancomycin given to the mice (Supplemental Table 1).
+Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis (Figure 3).
 *C.
-difficile* levels were significantly different between all titration groups within the cefoperazone (p<0.03) and streptomycin treated microbiotas (p<0.01), displaying a greater range of colonization resistance profiles (Figure 3).
-Next, we hypothesized that the populations we previously identified as being associated with protection against colonization would be elevated in the mice with decreased colonization and those associated with susceptibility would be less abundant.
-
-The titrated antibiotic doses clearly resulted in distinct community structures that were variations of the original dose (Figure 3).
-The cefoperazone titrations resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
-Most notable was the increase in OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*.
+difficile* levels were significantly different between all titration groups within the cefoperazone (p<XXX0.03XXX) and streptomycin treated mice (p<XXX0.01XXX).
+For each of the antibiotics we titrated, reduced dosages resulted in reduced colonization by *C.
+difficile*.
+By sequencing the 16S rRNA genes from feces collected on day 0 from these mice, we sought to validate the relationships we observed with the full dosage.
+Titrating cefoperazone resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
+Most notable was the increase in the relative abundance of OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*.
 The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.
-The streptomycin titrations did not result in profound shifts in the overall community structure, but did yield increased resistance.
-Rather, there were individual populations that were gained or lost across the gradient.
-Notably, mice receiving the lowest dose retained significantly higher levels of *Alistipes* and *Anaeroplasma* (p<0.05), which we previously found to be strongly associated with colonization resistance.
+The at the lowest streptomycin titration, mice retained significantly higher abundances of *Alistipes* and *Anaeroplasma* (p<0.05), which we found to be strongly associated with colonization resistance.
 The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice.
@@
 **Figure 3.
-TITLE.** This figure shows the average percent relative abundances for family level phylotypes found on the day of *C.
-difficile* challenge in each titration group.
-Each row depicts a different antibiotic that was titrated.
+The average percent relative abundances for family level phylotypes found on the day of *C.
+difficile* challenge in each titration group.** Each row depicts a different antibiotic that was titrated.
 High, medium, and low antibiotic doses are represented by black, grey, and white colored bars respectively.

### 2015-03-07 PAPER.Rmd
Edit recovery paragraph

@@
 
-***Allowing recovery of specific bacteria restores colonization resistance.*** Several antibiotic perturbation studies have demonstrated that given time perturbed communities can return to a “healthy” state in which resistance to *C.
-difficile* is restored [].
-For each antibiotic treatment, a control group of mice receiving antibiotics without inoculation of *C.
-difficile* was maintained.
-The microbiota’s of animals treated with metronidazole underwent a drastic shift in the microbiota structure with an additional 5 days of recovery.
-We observed changes in relative abundance of several OTUs that we previously observed had strong relationships with subsequent *C.
-difficile* colonization (Figure 2).
-Notably there were significant increases (p<0.05) in overall abundance of the levels of *Porphyromonadaceae*, *Ruminococcaceae*, and *Lachnospiraceae* to levels similar to untreated microbiota, and significant decreases (p<0.05) in levels of *Enterobacteriaceae*.
-Based on these changes, we hypothesized that delaying *C.
-difficile* challenge to allow these potentially important changes in microbiota structure to occur would lead to a decrease in colonization.
-On Day 1 post challenge of these recovered mice, we observed a significant decrease in *C.
-difficile* level of CFU/g feces (p<0.0001).
-These results strengthen the importance of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae* in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of *C.
-difficile* colonization.
-
-Additionally, we assessed the resistance of ampicillin treated mice given five extra days of recovery prior to *C.
-difficile* challenge.
-The shift in the microbiome with ampicillin recovery was not as prounounced as seen with recovery of metronidazole treated animals.
-These communities had significantly reduced levels of bacteria correlated with *C.
-difficile* protection, including *Bacteroides*, *Alistipes*, and *Ruminoccocaceae* compared with untreated animals (p<0.05).
+***Allowing recovery of specific bacterial populations restores colonization resistance.*** Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C.
+difficile* is restored [refs].
+We allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover an additional five days.
+We again sought to test the relationships that we identified in the correlation-based analysis using animals with the partially recovered microbiota.
+On Day 1 post challenge of the metronidazole-treated mice who were allowed to recover five extra days, we observed a significant decrease in *C.
+difficile* level of CFU per g feces (p<XXX0.0001XXX).
+Within their microbiota there were significant increases (p<XXX0.05XXX) in overall abundances of *Porphyromonadaceae*, *Ruminococcaceae*, and *Lachnospiraceae* to abundances similar to those of untreated mice, and significant decreases (p<XXX0.05XXX) in levels of *Enterobacteriaceae*.
+In contrast, the ampicillin-treated mice that were given five extra days of recovery prior to *C.
+difficile* challenge did not experience an increase in colonization resistance.
+Inspection of their microbiota indicated that they still had significantly reduced levels of bacteria correlated with *C.
+difficile* protection, including *Bacteroides*, *Alistipes*, and *Ruminoccocaceae* compared with untreated animals (p<XXX0.05XXX).
 Importantly, the level of *Enterobacteriaceae* was not significantly reduced with additional recovery time.
 Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent *C.
-difficile* levels (p=0.12).
+difficile* levels (p=XXX0.12XXX).
+Together, these results strengthen the importance of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae* in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of *C.
+difficile* colonization.
 
@@
 **Figure 4.
-TITLE.** This figure shows the average percent relative abundances for family level phylotypes found on the day of *C.
-difficile* challenge in each experimental group.
-Each row depicts a different antibiotic that was allowed to recover.
+Changes in the average relative abundance for family level phylotypes durin the period of recovery prior to challenge with *C.
+difficile*.** Each row depicts a different antibiotic that was allowed to recover.
 Black bars represent untreated animals.
 Grey bars represent animals given 1 day of recovery off antibiotics as in Figure 1.
-White bars represent animals given a total of 6 recovery days (5 additional days).
-Error bars show standard error of the mean.
-
---------
+Error bars indicate the standard error of the mean.
+
+--------
+
+> How is this relevant? - PDS
 
@@
 --------
-
+> How is this relevant? - PDS
 > ### Supplemental Figure 2A: Differences between recovery days' thetayc

### 2015-03-07 PAPER.Rmd
Edit random forest paragraph

@@
 
-***Modeling microbiota data can predict C.
-difficile colonization outcomes.*** Given our ability to make general predictions of *C.
-difficile* colonization based on the microbiota, we next designed a model to calculate expected colonization levels given the relative abundance of a subset of OTUs.
-Combining the three datasets (Table 1, Supplemental Table 1 and 2), we set a 1% relative abundance threshold, eliminating OTUs that on average were not present at this level in at least one of the 16 experimental groups.
-In total, there were 44 OTUs utilized for building a random forest regression model.
-This model fit the data with an R2 of 0.98 (Figure 5A).
-We identified OTUs that contributed most to increased accuracy of the model using random forest’s feature selection (Figure 5B).
-Many of the top contributing OTUs includes members of the *Porphyromonadaceae* family, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia* taxa.
-The relationship between these OTUs and *C.
-difficile* is shown in Figure 5C.
-*Escherichia* had the largest positive correlation with *C.
-difficile* levels of any bacterial species.
-This finding was consistent across susceptible antibiotic treated communities except for streptomycin treated communities, which altogether lacked *Enterobacteriaceae* spp.
-OTUs 3 and 39 (*Porphyromonadaceae* and *Lachnospiraceae*, respectively), were the two most important features in accurately predicting subsequent *C.
+***The composition of the microbiota on day 0 is predictive of C.
+difficile colonization levels on day 1.*** Given our ability to make qualitative predictions of *C.
+difficile* colonization based on the microbiota, we next designed a statistical model to calculate expected colonization levels given the relative abundance of a subset of OTUs.
+We used a regression-based random forest machine learning algorithm to predict the level of *C.
+difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge.
+The model explained XXX98.0XXX% of the variation in the observed *C.
+difficile* colonization levels (Figure 5).
+Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia*.
+These results further validate the observations from the correlation-based analysis.
+Together these results suggest that colonization resistance is likely conferred by *porphyromonadaceae*, *Alistipes*, and *lachnospiraceae* and a loss in these bacteria, concurrently with a gain of *Escherichia*, can result in increased susceptibility to infection.
+<needs help>Interestingly, the relationship between these OTUs and colonization resistance was not a simple dichotomous or linear relationship.
+For example, among the streptomycin-treated mice which altogether lacked the *Enterobacteriaceae*, *Porphyromonadaceae* (OTU 3) and *Lachnospiraceae* (OTU 39), were the two most important features in accurately predicting subsequent *C.
 difficile* levels.
 In fact, a model built with only these two features was able to account for 79% of the variance observed.
-Together these results suggest that colonization resistance is likely conferred by *porphyromonadaceae*, *Alistipes*, and *lachnospiraceae* and a loss in these bacteria, concurrently with a gain of *Escherichia*, can result in increased susceptibility to infection.
+</needs help>
 
@@
 Random forest model to predict *C.
-difficile* colonization levels based on the microbiota.** a) The fit of the random forest model by experimental group.
+difficile* colonization levels based on the microbiota.** (A) The fit of the random forest model by experimental group.
 The dashed line represents the y=x line.
-b) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model.
-c) The percent relative abundance of the top 8 OTUs from (b) on day 0 are plotted against the log<sub>10</sub> *C.
+(B) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model.
+(C) The percent relative abundance of the top 8 OTUs from (B) on day 0 are plotted against the log<sub>10</sub> *C.
 difficile* CFU/g feces.

### 2015-03-09 PAPER.Rmd
Add to discussion

--- 2015-03-07 PAPER.Rmd
@@
 
-The microbiome, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection [ref].
-This native protection by the microbiome from invasive pathogenic species is termed colonization resistance [ref].
-It is well established that the gut bacterial microbiome is critical in the hosts’ defense against the pathogen Clostridium difficile [ref].
+The microbiome, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection.
+This native protection by the microbiome from invasive pathogenic species is termed colonization resistance.
+It is well established that the gut bacterial microbiome is critical in the hosts’ defense against the pathogen Clostridium difficile.
 When this indigenous community is perturbed, this often leads to a loss of resistance.
 This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, chemotherapy, but also exposed to the environmental *C.
-difficile* spores [ref].
+difficile* spores.
 Among hospital-acquired infections in the United States, *C.
-difficile* infections (CDI) are the number one reported and are responsible for 14,000 deaths a year [ref].
-It is not completely understood how different perturbations to the taxonomic composition of the gut microbiota result in a loss of colonization resistance.
-
-<not done>
+difficile* infections (CDI) are the number one reported and are responsible for 14,000 deaths a year.
+It is not completely understood how different perturbations affect subsequent colonization resistance, nor do we have a comprehensive view of the taxonomic changes associated with a loss in colonization resistance.
+
 There is a need for more knowledge about interactions between members of the community.
@@
 These findings __
-</not done>
 
@@
 ***Animal Care.*** We used 5-8 week old C57Bl/6 mice for all our experiments.
-These mice were reared under SPF conditions within the animal facility at the University of Michigan.
+These mice were housed in our SPF room in the animal facility at the University of Michigan.
 All animal-related protocols and experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
@@
 The route of administration depended on the specific antibiotic.
-Cefoperazone (either 0.5 mg/ml, 0.3mg/ml, or 0.1 mg/ml), vancomycin (0.625 mg/ml, 0.3 mg/ml, or 0.1 mg/ml), streptomycin (5 mg/ml, 0.5 mg/ml, or 0.1 mg/ml), metronidazole (0.5 mg/ml), and ampicillin (0.5 mg/ml) were all administered in the mouse drinking water for 5 days.
-Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via IP injection.
-Mice that did not receive antibiotics were used as negative controls for these experiments.
+Cefoperazone (either 0.5mg/ml, 0.3mg/ml, or 0.1mg/ml), vancomycin (0.625mg/ml, 0.3mg/ml, or 0.1mg/ml), streptomycin (5mg/ml, 0.5mg/ml, or 0.1mg/ml), metronidazole (0.5mg/ml), and ampicillin (0.5mg/ml) were all administered in the mouse drinking water for 5 days.
+Ciprofloxacin (10 mg/kg) was administered via oral gavage, and clindamycin (10 mg/kg) was administered via IP injection.
+All antibiotic treated mice were given 24 hours of recovery on regular drinking water prior to *C.
+difficile* challenge.
 
 ***C.
-difficile Preparation and Challenge*** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
-difficile* challenge.
-*C.
-difficile* strain 630Δerm spores were used in all experiments [cite dena lyras].
-Spores were prepared from a single large batch whose concentration was determined within the week prior to each *C.
-difficile* challenge [cite paul?].
-On the day of challenge 10^3 *C.
-difficile* spores were administered to mice via oral gavage.
+difficile Preparation and Challenge*** In all our experiments we used *C.
+difficile* strain 630Δerm [cite dena lyras].
+The spores used were prepared from a single large batch made following previously established protocols [cite paul?].
+Spores were quantified within the week prior to *C.
+difficile* challenge for each experiment [established protocol I can cite?].
+Spores were quantified by boiling a small volume at 65°C for 20 minutes to kill off residual vegetative cells.
+The volume was then diluted and plated on TCCFA plates.
+These were incubated at 37°C anaerobically overnight.
+The next day *C.
+difficile* colony forming units (CFU) were counted and the concentration of spores in the working stock was back calculated.
+On the day of challenge, *C.
+difficile* spores, at a final concentration of 103, were administered to mice via oral gavage.
 Immediately following this challenge, the remaining *C.
@@
 
-***Sample Collection and Plating*** Fecal samples were freshly collected for each mouse on the day of *C.
-difficile* challenge (day 0).
-On the day after the challenge (day 1), another fecal sample was weighed and diluted under anaerobic conditions with anaerobic PBS.
-The number of colony forming units (CFU) were counted following 24 hrs growth on TCCFA plates at 37°C under anaerobic conditions [cite plates].
-
-***DNA Extraction and Sequencing*** Total bacterial DNA was extracted from each  day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
+***Sample Collection and Plating*** Fecal samples were freshly collected for each mouse immediately prior to the administration of any antibiotic, during any antibiotic recovery days, and days 0 and 1 post *C.
+difficile* challenge.
+Each fecal sample on day 1 was weighed and diluted under anaerobic conditions with anaerobic PBS 1:10 based on pellet weight.
+A dilution series was then plated onto TCCFA plates [cite plates] and incubated at 37°C anaerobically overnight.
+The colony forming units (CFU) were counted and calculated the subsequent day.
+
+***DNA Extraction and Sequencing*** Total bacterial DNA was extracted from each untreated and day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
 Sequencing preparation and sequencing was carried out as previously described (Kozich JJ et al.
@@
 These libraries were then mixed with Illumina-prepared PhiX control libraries.
+Sequencing was performed using Illumina’s MiSeq platform to produce paired 250-nt reads.
+Mock/error/error of run?
 
 ***Sequence Curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
+The full pipeline is also demonstrated online at the mother website (http://www.mothur.org/wiki/MiSeq_SOP).
 Sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
 Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 9) requiring an 80% bootstrap confidence score.
-In parallel to the fecal samples, we also sequeced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
-Analysis of the mock community data indicated that the error rate following our cuation procedure was 0.0X%.
 In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples [Schloss, reducing effects].
-All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession XXXXXX.
-
-***Statistical Analysis & Modeling*** All analyses were conducted using R version 3.1.2.
-Complete analysis scripts are available at the online repository for this study (https://github.com/SchlossLab/Schubert_AbxD01_mBio_2015).
-
-Need text about...
-* correlation analysis
-* comparison of titration levels
-* For the random forest you use 1% within each treatment group, for the correlation you state using 0.1 and 0.01% across all samples. Need to stay consistent.
-* statistical methods used to differentiate the abundances of OTUs in the titrations
-
-Random forest regression models were constructed using the randomForest R package using XX,000 trees [ref].
-The model was constructed using the relative abundance of those OTUs that had an average relative abundance within any antibiotic treatment group of 1%.
-This resulted in 44 OTUs.
-The regression was performed using the log10 transformation of the number of CFU per gram of fecal material as the dependent variable.
+For those samples with more than 1,625 sequences per samples, the range was __, the mean was __, the median was __, the median absolute deviation was __.
+All 16S rRNA gene sequence data and the MIMARKS table are available at __.
+
+***Statistical Analysis & Modeling*** Across all data, we set a relative abundance threshold of an average of 1% in at least one experimental group.
+The 44 OTUs that met this cutoff were included in the training of the model.
+We used the "randomForest" package in R to build a regression-based random forest model to predict *C.
+difficile* colonization levels on day 1 based on the microbiome data from day 0.
+All statistical analyses were conducted using R version 3.1.2.
 
@@
 difficile* colonization.
-It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] had decreased resistance against this pathogen.
+It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] all have decreased resistance against this pathogen.
 To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (**Table 1**) to differentially alter the microbiota and assess their resistance to *C.
@@
 The community structures of the beta-lactam-treated mice (i.e.
-cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=XXX0.36XXX).
-Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=XXX0.001XXX) and were characterized by the predominance of *Enterobacteriaceae*.
+cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=0.-36).
+Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=0.001) and were characterized by the predominance of *Enterobacteriaceae*.
 Of all groups significantly different from the untreated microbiota, the levels of *Enterobacteriaceae* increased with all treatments except with streptomycin.
 Vancomycin-treated communities saw the largest bloom in *Akkermansia*.
-In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, XXXp=0.09XXX).
+In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=0.09).
 Subsequently, 24 hours after inoculation *C.
 difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g) (**Figure 1**).
-We calculated the diversity of all of the communities using the Inverse Simpson and Shannon indices and observed a significant negative correlation with subsequent *C.
-difficile* colonization (both **Supplemental Figure 1**; p<XXX0.001XXX).
-These results demonstrate that a variety of non-overlapping community structures can be sensitive to *C.
-difficile* colonization.
+We calculated the diversity of all of the communities using the Inverse Simpson (**Supplemental Figure 1**) and Shannon indices (data not shown) and observed a significant negative correlation with subsequent *C.
+difficile* colonization (both p<0.001).
+These results suggest that multiple microbiota structures are capable of being colonized by *C.
+difficile* and that diversity may be an important factor in resistance to invasion.
 
@@
 The level *C.
-difficile* colonization found 24 hours post microbiota sampling is indicated in the top right for each treatment.
-Error bars indicate standard error of the mean.
+difficile* colonization found 24 hours post microbiota sampling is plotted in the top right corner.
+Each represented phylum is a different shade of gray.
+Error bars depict standard error of the mean.
 
@@
 **Supplemental Figure 1.
-The inverse of Simpson's diversity index for the community on day 0 is negatively correlated with *C.
-difficile* colonization on day 1.** Spearman's rho is plotted in the top right corner.
+Diversity is nevatively correlated with *C.
+difficile* levels.** This figure shows the relationship between the inverse simpson diversity index of microbiotas of each experimental group on day 0 and *C.
+difficile* CFU/g feces on day 1.
+Spearman's rho is plotted in the top right corner.
 
@@
 difficile*, we measured the correlation between the relative abundance of each species-level OTU on the day of inoculation with the level of *C.
-difficile* colonization 24 hours later (Figure 2).
-<needs help> Of the OTUs with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against *C.
-difficile* (N=XXX40XXX) and those with a positive correlation were associated with susceptibility (N=XXX7XXX) (p<XXX0.05XXX).</needs help> Among various bacterial families, three were consistently negatively correlated with *C.
-difficile* colonization: *Porphyromonadaceae* (ρ<sub>average</sub>=XXX-0.63XXX), *Lachnospiraceae* (ρ<sub>average</sub>=XXX-0.51XXX), *Ruminococcaceae* (ρ<sub>average</sub>=XXX-0.53XXX).
+difficile* colonization 24 hours later across all treatment groups (Figure 2).
+Of the significant OTUs (p<0.05) with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against *C.
+difficile* (N=40) and those with a positive correlation were associated with susceptibility (N=7).
+Among various bacterial families, three were consistently negatively correlated with *C.
+difficile* colonization: *Porphyromonadaceae* (ρaverage= -0.63), *Lachnospiraceae* (ρaverage= -0.51), *Ruminococcaceae* (ρaverage= -0.53).
 OTUs from the *Proteobacteria* phylum had strong positive relationships with *C.
 difficile* colonization.
-These included OTUs associated with the genera *Pseudomonas* (ρ=XXX0.31XXX) and *Escherichia* (ρ=XXX0.56XXX).
-Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus *Lactobacillus* (family *Lactobacillaceae*; OTU23, ρ=XXX-0.73XXX), *Alistipes* (family *Rickenellaceae*; OTU20, ρ=XXX-0.78XXX), *Enterorhabdus* (phylum *Actinobacterium*, family *Coriobacteriaceae*; OTU29, ρ=XXX-0.70XXX), and *Streptococcus* (family *Streptococcaceae*; OTU78, ρ=XXX0.36XXX).
+These included OTUs associated with the *Pseudomonas* (ρ= 0.31) and *Escherichia* (ρ= 0.56) genera.
+Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus *Lactobacillus* (family *Lactobacillaceae*; OTU23, ρ= -0.73), *Alistipes* (family *Rickenellaceae*; OTU20, ρ= -0.78), *Enterorhabdus* (phylum *Actinobacterium*, family *Coriobacteriaceae*; OTU29, ρ= -0.70), and *Streptococcus* (family *Streptococcaceae*; OTU78, ρ= 0.36).
 Overall, these results suggest that several larger groups of related OTUs, particularly within the *Firmicutes* and *Bacteroidetes* phyla, may participate in colonization resistance against *C.
@@
 difficile* CFU/g feces on day 1 using Spearman’s rank based correlation coefficient.
-Only OTUs with an average abundance of at least <needs help>0.05%<needs help> across all original treatments were considered in the correlation analysis.
-All correlations shown in this graph are significant after controlling for multiple comparisons using the Benjimani-Hochberg correction (p<0.05).
-OTUs are grouped by family.
-An asterisk (*) indicates that the OTU was unclassified at the family level.
+Only OTUs with an average abundance of at least 0.05% across all original treatments were considered in the correlation analysis.
+All correlations shown in this graph are significant (p<0.05).
+OTUs were grouped by family.
+Asterisk * indicates unclassified at family level.
 
@@
 
-***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** All but the ciprofloxacin treatment resulted in similarly high levels of *C.
-difficile*.
-Therefore, we hypothesized that titrating the dose of antibiotics given to the mice would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C.
-difficile*.
-We titrated the doses of cefoperazone, streptomycin, and vancomycin given to the mice (Supplemental Table 1).
-Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis (Figure 3).
+***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** <transition>
+Because most of our original treatments resulted in similarly high levels of *C.
+difficile*, we hypothesized that applying smaller antibiotic perturbations to the microbiota would result in reduced sensitivity to *C.
+difficile*, as we observed in the ciprofloxacin treated mice.
+To test this, we titrated the doses of cefoperazone, streptomycin, and vancomycin given to mice (Supplemental Table 1).
+Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis.
 *C.
-difficile* levels were significantly different between all titration groups within the cefoperazone (p<XXX0.03XXX) and streptomycin treated mice (p<XXX0.01XXX).
-For each of the antibiotics we titrated, reduced dosages resulted in reduced colonization by *C.
-difficile*.
-By sequencing the 16S rRNA genes from feces collected on day 0 from these mice, we sought to validate the relationships we observed with the full dosage.
-Titrating cefoperazone resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
-Most notable was the increase in the relative abundance of OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*.
+difficile* levels were significantly different between all titration groups within the cefoperazone (p<0.03) and streptomycin treated microbiotas (p<0.01), displaying a greater range of colonization resistance profiles (Figure 3).
+Next, we hypothesized that the populations we previously identified as being associated with protection against colonization would be elevated in the mice with decreased colonization and those associated with susceptibility would be less abundant.
+
+The titrated antibiotic doses clearly resulted in distinct community structures that were variations of the original dose (Figure 3).
+The cefoperazone titrations resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
+Most notable was the increase in OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*.
 The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.
-The at the lowest streptomycin titration, mice retained significantly higher abundances of *Alistipes* and *Anaeroplasma* (p<0.05), which we found to be strongly associated with colonization resistance.
+The streptomycin titrations did not result in profound shifts in the overall community structure, but did yield increased resistance.
+Rather, there were individual populations that were gained or lost across the gradient.
+Notably, mice receiving the lowest dose retained significantly higher levels of *Alistipes* and *Anaeroplasma* (p<0.05), which we previously found to be strongly associated with colonization resistance.
 The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice.
@@
 **Figure 3.
-The average percent relative abundances for family level phylotypes found on the day of *C.
-difficile* challenge in each titration group.** Each row depicts a different antibiotic that was titrated.
+TITLE.** This figure shows the average percent relative abundances for family level phylotypes found on the day of *C.
+difficile* challenge in each titration group.
+Each row depicts a different antibiotic that was titrated.
 High, medium, and low antibiotic doses are represented by black, grey, and white colored bars respectively.
@@
 
-Through our perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of *C.
+-----------
+
+Through our antibiotic perturbation experiments we identified a panel of bacteria with either a strong positive or negative relationship with subsequent levels of *C.
 difficile* colonization.
-A large number of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Clostridia, as well as Alistipes were found to have strong potential resistance abilities against *C.
+A large number of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, other *Clostridia*, and *Alistipes* were found to have strong potential resistance abilities against *C.
 difficile*.
-Conversely, several members of the Proteobacteria phylum, including Pseudomonas, Streptococcus, and especially E.
-coli have strong potential positive impacts for *C.
+Conversely, several members of the *Proteobacteria* phylum, including *Pseudomonas*, *Streptococcus*, and especially *Escherichia* have strong potential positive influences on *C.
 difficile* colonization.
-These results were supported by their strong correlation values supported across several antibiotic regimens.
-They are also supported through experiments in which the original doses of cefoperazone, streptomycin, and (vancomycin) were titrated to alter the overall abundances but not membership of the community from the starting dose, as has been observed previously (Sekirov).
-These subtler structural changes allowed for greater range of colonization levels of *C.
-difficile* as well as further/fine-tuned support for our original results.
-Furthermore, by allowing a known change to occur in important bacterial populations within metronidazole treated mice, we were able to reduce the level of colonization by *C.
+These results were supported by their significant correlation values across several antibiotic regimens.
+By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had relatively increased resistance to *C.
+difficile* colonization.
+Using this collective microbiota data spanning a wide range of colonization resistance levels, we next moved beyond correlation to prediction of *C.
+difficile* colonization based on the composition of the microbiota.
+Through modeling we identified bacteria within the community which contributed greatest to accurately predicting *C.
+difficile* levels.
+This is the first analysis measuring and predicting colonization resistance levels based on the microbiota structure.
+
+Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance.
+Correlation analysis identified whole families with significant negative or positive relationships with *C.
 difficile*.
-Lastly we showed their importance through modeling techniques that statistically confirm these results.
+However, correlation analysis alone was insuficient in identifying all informative bacteria.
+Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
+difficile* yet improved model accuracy.
+*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (Figure 5bc).
+While these bacteria may not have important direct effects on *C.
+difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
+difficile*.
+Through the model analysis, we observed that no single bacteria was able to consistently explain the subsequent level of *C.
+difficile* observed.
+For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C.
+difficile*.
+The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C.
+difficile* (*Figure 5C*).
+However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C.
+difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces.
+Similarly, while having high amounts of *Porphyromonadaceae* OTU13 in many instances resulted in low *C.
+difficile* colonization, 3 ampicillin treated, 2 cefoperazone (both 0.1 mg/ml and 0.3 mg/ml) treated, and 2 streptomycin (0.1 mg/ml) treated microbiotas (OTU13 abundance at least ~5%), had levels of *C.
+difficile* ranging from 10<sup>2</sup> to upwards of 10<sup>8</sup>.
+[Can say a similar thing about *Porphyromonadaceae* OTU6] Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).
+
+Furthermore, the ability of *C.
+difficile* to colonize a community does not appear to be dependent soley on the loss of protective species.
+<> Also having high levels of an OTU with high positive correlation with *C.
+difficile* is not a requirement for *C.
+difficile* colonization.
+For example,  *Escherichia* OTU11 was most commonly found among the most positively correlated OTUs with *C.
+difficile* across antibiotic treatments.
+However, following streptomycin treatment, these microbiotas harbored less than 0.5% relative abundance of OTU11 and were still colonized with as much as 10<sup>8</sup> *C.
+difficile* CFU/g feces.
+
+ Understanding how individual bacterial populations live in the context of a community is important for not only determining mechanisms but also for informing novel therapies.
+
+In the context of antibiotic perturbation however, widespread loss of protective bacteria allows the growth of bacteria which otherwise are outcompeted, including many *Proteobacteria* (*E. coli*).
+
+Furthermore, the ability of *C.
+difficile* to colonize a community does not appear to be dependent soley on the loss of protective species.
+Rather, it seems likely that as protective species are lost, advantageous bacteria such as *E.
+coli* can colonize quickly and adapt, which in turn also promotes a favorable environment for *C.
+dfficile* colonization.
+Many members of the proteobacteria---attributes: grow quickly, wide array of nutrient sources, more generalists.
+Examples of e.coli causing inflammation, which can help pathogens in some instances where bad for resident good bugs.
+Also e.
+coli using nutrients within mucus (?) or at least contributing to loss of oxygen near epithelium, which can also let pathogens get closer and colonize..
+cause further damage (by toxins, etc)...
 

### 2015-03-09 PAPER.Rmd

--- 2015-03-07 PAPER.Rmd
@@
 
-Through our perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of *C.
+-----------
+
+Through our antibiotic perturbation experiments we identified a panel of bacteria with either a strong positive or negative relationship with subsequent levels of *C.
 difficile* colonization.
-A large number of Porphyromonadaceae, Lachnospiraceae, Ruminococcaceae, and Clostridia, as well as Alistipes were found to have strong potential resistance abilities against *C.
-difficile*.
-Conversely, several members of the Proteobacteria phylum, including Pseudomonas, Streptococcus, and especially E.
-coli have strong potential positive impacts for *C.
+A large number of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, other *Clostridia*, and *Alistipes* were found to have strong potential resistance abilities against *C.
+difficile*.
+Conversely, several members of the *Proteobacteria* phylum, including *Pseudomonas*, *Streptococcus*, and especially *Escherichia* have strong potential positive influences on *C.
 difficile* colonization.
-These results were supported by their strong correlation values supported across several antibiotic regimens.
-They are also supported through experiments in which the original doses of cefoperazone, streptomycin, and (vancomycin) were titrated to alter the overall abundances but not membership of the community from the starting dose, as has been observed previously (Sekirov).
-These subtler structural changes allowed for greater range of colonization levels of *C.
-difficile* as well as further/fine-tuned support for our original results.
-Furthermore, by allowing a known change to occur in important bacterial populations within metronidazole treated mice, we were able to reduce the level of colonization by *C.
-difficile*.
-Lastly we showed their importance through modeling techniques that statistically confirm these results.
+These results were supported by their significant correlation values across several antibiotic regimens.
+By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had relatively increased resistance to *C.
+difficile* colonization.
+Using this collective microbiota data spanning a wide range of colonization resistance levels, we next moved beyond correlation to prediction of *C.
+difficile* colonization based on the composition of the microbiota.
+Through modeling we identified bacteria within the community which contributed greatest to accurately predicting *C.
+difficile* levels.
+This is the first analysis measuring and predicting colonization resistance levels based on the microbiota structure.
+
+Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance.
+Correlation analysis identified whole families with significant negative or positive relationships with *C.
+difficile*.
+However, correlation analysis alone was insuficient in identifying all informative bacteria.
+Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
+difficile* yet improved model accuracy.
+*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (Figure 5bc).
+While these bacteria may not have important direct effects on *C.
+difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
+difficile*.
+Through the model analysis, we observed that no single bacteria was able to consistently explain the subsequent level of *C.
+difficile* observed.
+For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C.
+difficile*.
+The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C.
+difficile* (*Figure 5C*).
+However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C.
+difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces.
+Similarly, while having high amounts of *Porphyromonadaceae* OTU13 in many instances resulted in low *C.
+difficile* colonization, 3 ampicillin treated, 2 cefoperazone (both 0.1 mg/ml and 0.3 mg/ml) treated, and 2 streptomycin (0.1 mg/ml) treated microbiotas (OTU13 abundance at least ~5%), had levels of *C.
+difficile* ranging from 10<sup>2</sup> to upwards of 10<sup>8</sup>.
+[Can say a similar thing about *Porphyromonadaceae* OTU6] Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).
+
+Furthermore, the ability of *C.
+difficile* to colonize a community does not appear to be dependent soley on the loss of protective species.
+<> Also having high levels of an OTU with high positive correlation with *C.
+difficile* is not a requirement for *C.
+difficile* colonization.
+For example,  *Escherichia* OTU11 was most commonly found among the most positively correlated OTUs with *C.
+difficile* across antibiotic treatments.
+However, following streptomycin treatment, these microbiotas harbored less than 0.5% relative abundance of OTU11 and were still colonized with as much as 10<sup>8</sup> *C.
+difficile* CFU/g feces.
+
+ Understanding how individual bacterial populations live in the context of a community is important for not only determining mechanisms but also for informing novel therapies.
+
+In the context of antibiotic perturbation however, widespread loss of protective bacteria allows the growth of bacteria which otherwise are outcompeted, including many *Proteobacteria* (*E. coli*).
+
+Furthermore, the ability of *C.
+difficile* to colonize a community does not appear to be dependent soley on the loss of protective species.
+Rather, it seems likely that as protective species are lost, advantageous bacteria such as *E.
+coli* can colonize quickly and adapt, which in turn also promotes a favorable environment for *C.
+dfficile* colonization.
+Many members of the proteobacteria---attributes: grow quickly, wide array of nutrient sources, more generalists.
+Examples of e.coli causing inflammation, which can help pathogens in some instances where bad for resident good bugs.
+Also e.
+coli using nutrients within mucus (?) or at least contributing to loss of oxygen near epithelium, which can also let pathogens get closer and colonize..
+cause further damage (by toxins, etc)...
 

### 2015-03-09 PAPER.Rmd

@@
 difficile* levels.
-This is the first analysis measuring and predicting colonization resistance levels based on the microbiota structure.
+This is the first analysis both measuring and predicting colonization resistance levels based on the microbiota structure.
 
@@
 difficile* yet improved model accuracy.
-*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (Figure 5bc).
+*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*).
 While these bacteria may not have important direct effects on *C.
@@
 difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces.
-Similarly, while having high amounts of *Porphyromonadaceae* OTU13 in many instances resulted in low *C.
-difficile* colonization, 3 ampicillin treated, 2 cefoperazone (both 0.1 mg/ml and 0.3 mg/ml) treated, and 2 streptomycin (0.1 mg/ml) treated microbiotas (OTU13 abundance at least ~5%), had levels of *C.
-difficile* ranging from 10<sup>2</sup> to upwards of 10<sup>8</sup>.
-[Can say a similar thing about *Porphyromonadaceae* OTU6] Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).
-
-Furthermore, the ability of *C.
-difficile* to colonize a community does not appear to be dependent soley on the loss of protective species.
-<> Also having high levels of an OTU with high positive correlation with *C.
-difficile* is not a requirement for *C.
-difficile* colonization.
-For example,  *Escherichia* OTU11 was most commonly found among the most positively correlated OTUs with *C.
-difficile* across antibiotic treatments.
-However, following streptomycin treatment, these microbiotas harbored less than 0.5% relative abundance of OTU11 and were still colonized with as much as 10<sup>8</sup> *C.
-difficile* CFU/g feces.
+Similarly, having low amounts of one "protective" bacteria such as *Lachnospriaceae* OTU39 does not automatically render the host susceptible to colonization.
+For instance streptomycin treated (0.1 mg/ml) mice did not harbor *Lachnospiraceae* OTU39, yet had undetectable levels of *C.
+difficile*.
+Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).
+
+Furthermore, the dynamics of *C.
+difficile* colonization does not appear to be dependent soley on the loss of protective species.
+Both *Escherichia* OTU11 and *Streptococcus* OTU78 were positively correlated with *C.
+difficile* colonization and important in the random forest model.
+Notably, *Escherichia* OTU11 was among the strongest positively correlated OTUs with *C.
+difficile* across all antibiotic treatments, except for streptomycin.
+In fact, while streptomycin treatment (5 mg/ml) lead to high *C.
+difficile* colonization, these communities had very little, if any, *Escherichia* OTU11.
+This demostrates that while *Escherichia* may aid in colonization of *C.
+difficile*, it is not essential.
+These observations lead to a model of colonization by *C.
+difficile* in which antibiotic perturbation leads to loss of protective species.
 

### 2015-03-09 code/make_design.R
Initial commit

@@ -1,0 +1,16 @@
+################################################################################
+#
+# make_design.R
+#
+# This script will generate the various design files that are needed to run the
+# amova command from within mothur
+#
+# Dependencies...
+# * The shared file that will be used in the amova command
+# * The table that maps the sample names to the treatments (data/raw/abxD01_IDS.txt)
+# * Information about the type of design file to make.
+#
+# Output...
+# * The relevant design file
+#
+################################################################################

### 2015-03-09 PAPER.Rmd
Update discussion

@@
 Not one otu that explains it all… just cus low levels doesn’t screw you, also see recovered points in 5C-angela, sonnenburg-need other thingsssss
-  -fig2—there are like larger groups in protection/susceptibility
-	-combo of loss of good bugs, and gain of bad—a model of infection
 -tie back to mbio paper, with enterobacteriaceae they are screwed
@@
 difficile* in which antibiotic perturbation leads to loss of protective species.
+This in turn eliminates competition, frees resources, and eliminates direct or indirect inhibitory effects of bacterial products (e.g.
+bacteriocins or short chain fatty acids) made by these protective bacteria.
+While this may be suitable enough for *C.
+difficile* to colonize, other opportunistic species such as *Escherichia coli* and other *Proteobacteria* may also take advantage of these microbiota perturbations.
+Increased colonization by these bacteria can increase inflammation, produce a new profile of metabolites and bacterial products, reduce the mucus layer,  and deoxygenate the microoxic tiers nearing the epithelium.
+These changes in the gut milieu may be advantageous for *C.
+difficile* and other gut pathogens.
+Inflammation, for instance, has been shown to be advantageous for *Salmonella tyhpumurium* to further disrupt the indigenous microbiota [].
+Inflammation also leads to a shift in the availability of resources, such as nitrites, which can be used as an alternative resource by *Salmonella typhimurium* [].
+*E.coli* is capable of using resources found in the mucus layer while simultaneously deoxygenating this region [].
+This reduces the spatial protection of the mucus layer, allowing bacteria, including strictly anaerobic bacteria, and bacterial products (e.g.
+toxins) to more readily reach the epithelial layer.
 
 Understanding how individual bacterial populations live in the context of a community is important for not only determining mechanisms but also for informing novel therapies.
-
-In the context of antibiotic perturbation however, widespread loss of protective bacteria allows the growth of bacteria which otherwise are outcompeted, including many *Proteobacteria* (*E. coli*).
-
-Furthermore, the ability of *C.
-difficile* to colonize a community does not appear to be dependent soley on the loss of protective species.
-Rather, it seems likely that as protective species are lost, advantageous bacteria such as *E.
-coli* can colonize quickly and adapt, which in turn also promotes a favorable environment for *C.
-dfficile* colonization.
-Many members of the proteobacteria---attributes: grow quickly, wide array of nutrient sources, more generalists.
-Examples of e.coli causing inflammation, which can help pathogens in some instances where bad for resident good bugs.
-Also e.
-coli using nutrients within mucus (?) or at least contributing to loss of oxygen near epithelium, which can also let pathogens get closer and colonize..
-cause further damage (by toxins, etc)...
 

### 2015-03-10 code/extract_abxD01_IDS_xlsx.R
Have discrepancy btwn xlsx and fastq files identified

@@ -77,3 +77,51 @@ sum(grepl("^\\d{1,2}-", rownames(spread_sheet)))
 
 #remove the dash between the animal number and the D that appears in some smaple names
 rownames(spread_sheet) <- gsub("^(\\d{3}-\\d)-D", "\\1D", rownames(spread_sheet))
+
+
+
+#Read in fastq file names so that we can make sure that we have all of the
+#samples and that they are named the same in the metadata (here) and the
+#sequence data
+
+fastq <- scan("data/raw/clean_fastqs.txt", what="")
+stub <- gsub("(.*)_S.*", "\\1", fastq)
+unique_stubs <- unique(stub)
+mocks <- grepl("mock", unique_stubs)
+
+sample_stub <- unique_stubs[!mocks]
+mock_stub <- unique_stubs[mocks]
+
+
+#all of the sample names in the spreadsheet have a fastq file
+rownames(spread_sheet)[!(rownames(spread_sheet) %in% sample_stub)]
+#[1] "021-3D1"   "021-3D3"   "021-3D5"   "021-4D6"   "106-1D1"   "106-1D7"
+#[7] "109-1D5"   "109-5D1"   "109-5D7"   "111-1D1"   "111-1D10"  "111-1D4"
+#[13] "111-1D7"   "111-2D1"   "111-2D4"   "111-2D7"   "111-4D1"   "111-4D4"
+#[19] "117-4D5"   "117-5D1"   "600-1D2"   "600-2D6"   "600-2D7"   "600-2D8"
+#[25] "600-2D9"   "600-3D1"   "600-3D2"   "600-3D3"   "600-3D5"   "600-4D2"
+#[31] "600-4D9"   "102-1D8"   "102-2D-11" "102-2D-2"  "102-5D4"   "103-2D3"
+#[37] "103-3D1"   "103-3D2"   "103-3D3"   "103-3D6"   "107-1D3"   "107-1D9"
+#[43] "107-3D3"   "107-3D9"   "108-1D10"  "108-1D4"   "108-2D10"  "108-2D4"
+#[49] "116-1D7"   "116-2D7"   "088-1D1"   "088-1D10"  "088-1D2"   "088-1D3"
+#[55] "088-2D2"   "088-2D3"   "088-2D4"   "088-2D5"   "088-2D6"   "088-2D8"
[…]

### 2015-03-10 PAPER.Rmd

@@
 ```{r, eval=TRUE}
-data<-read.csv("~/Desktop/mothur/abxD01/model/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.csv", header=T)
+
+dataFile <- "~/Documents/Github/abxD01/results/figures/Figure5/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.highvancD0.csv"
+
+data<-read.csv(file=dataFile, header=T)
 
@@
 par(mar=c(5, 5, 0.5, 0.5) +0.1, mgp=c(3, 1, 0), las=1)
-plot(results[results[,1]=="control",2],
-     results[results[,1]=="control",3],
+plot(fit.results[fit.results[,1]=="control",2],
+     fit.results[fit.results[,1]=="control",3],
      main="",
@@
 for(i in 2:(length(legend.labels))){
-  points(results[results[,1]==names(legend.labels)[i],2],
-         results[results[,1]==names(legend.labels)[i],3],
+  points(fit.results[fit.results[,1]==names(legend.labels)[i],2],
+         fit.results[fit.results[,1]==names(legend.labels)[i],3],
          col=colors[i], pch=pch[i], cex=1.5)

### 2015-03-10 PAPER.Rmd
Add to discussion relate to humans

@@
 
-Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance.
 Correlation analysis identified whole families with significant negative or positive relationships with *C.
 difficile*.
+Many of the correlaions we observed were not surprising given other studies looking at the relationship between *C.
+difficile* and the microbiome.
+For instance, monocolonization of germfree mice with either a *Lachnospiraceae* species or *E.
+coli* were tested for their resistance to *C.
+difficile* colonization [Reeves].
+Lachnospiraceae was able to decrease the levels of *C.
+difficile* observed, while *E.
+coli* had similar colonization levels compared to germfree mice given *C.
+difficile* alone.
+Association studies in humans with and without CDI in hospitals have also highlighted several bacteria found in this study, including *Lachnospiraceae*, *Rumicoccaceae*, *Clostridia*, and *Alistipes* as decreased and *E.
+coli* and *Streptococcus* as increased in subjects with CDI [Schubert, Antharam].
+While we observe overlaps in bacteria with potentially protective roles in both mice and humans at broad taxonomy levels, there are notable differences at refined levels.
+*Porphyromonadaceae*, for example, are more common among mice than humans, where instead there are strong negative associations of *C.
+difficile* with *Bacteroides* [Schubert].
 However, correlation analysis alone was insuficient in identifying all informative bacteria.
@@
 difficile*.
-Through the model analysis, we observed that no single bacteria was able to consistently explain the subsequent level of *C.
+Few studies attempt to model the microbiome to predict disease or estimate health [Pamer, clostridium paper, ].
+
+Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance.
+No single bacteria was able to consistently explain the subsequent level of *C.
 difficile* observed.
@@
 bacteriocins or short chain fatty acids) made by these protective bacteria.
+[can talk about change in gut with abx--loss of some, gain of others that are resistant to abx or grow up quickly after, and direct effects of abx...
+last weeks jclub paper; sonnenburg has 2 papers with nutrients/metbolites and cdiff related to the microbiome]
+
 While this may be suitable enough for *C.

### 2015-03-11 PAPER.Rmd
Add to discussion

@@
 
-Through our antibiotic perturbation experiments we identified a panel of bacteria with either a strong positive or negative relationship with subsequent levels of *C.
+Through our antibiotic perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of *C.
 difficile* colonization.
@@
 These results were supported by their significant correlation values across several antibiotic regimens.
-By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had relatively increased resistance to *C.
+By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had increased resistance to *C.
 difficile* colonization.
@@
 
-Correlation analysis identified whole families with significant negative or positive relationships with *C.
+Correlation analysis identified major families with significant negative or positive relationships with *C.
 difficile*.
@@
 difficile* colonization [Reeves].
-Lachnospiraceae was able to decrease the levels of *C.
+Pre-colonization with *Lachnospiraceae* was able to decrease the levels of *C.
 difficile* observed, while *E.
@@
 While we observe overlaps in bacteria with potentially protective roles in both mice and humans at broad taxonomy levels, there are notable differences at refined levels.
-*Porphyromonadaceae*, for example, are more common among mice than humans, where instead there are strong negative associations of *C.
-difficile* with *Bacteroides* [Schubert].
-However, correlation analysis alone was insuficient in identifying all informative bacteria.
+*Porphyromonadaceae* (Class *Bacteroidales*), for example, are more common among mice than humans, where instead there are strong negative associations of *C.
+difficile* with *Bacteroides* (Class *Bacteroidales*) [Schubert].
+
+Correlation analysis alone was insufficient in identifying all informative bacteria.
 Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
@@
 difficile*.
-Few studies attempt to model the microbiome to predict disease or estimate health [Pamer, clostridium paper, ].
+Modeling of the microbiome allows us to characterize *C.
+difficile* colonization in terms of a community, which is necessary to capture the complex relationships among its members.
+Few studies attempt to model the microbiome to predict disease or estimate health [Pamer, clostridium paper, psoriases, schubert].
 
@@
 •	Ferreira—intestinal microbiota plays a role in salmonella induced colitis
-•	Reeves with clone libraries and cef
 •	Theriot?—at least the metabolome paper—before and after with metabolites and cef, also does she look at anything taxonomically? yup
 
-Many of the relationships we observed in our study between specific bacteria and *C. difficile* are supported in the literature. –see my comments in this section
-Within the Porphyromonadaceae, only OTU505 had a significant positive relationship with subsequent *C.
-difficile* colonization (ρ= +0.21), which by blastn shares 99% identity with Coprobacter fastidiosus.
-Similarly within the Lachnospiraceae family, only one OTU had a positive relationship with *C.
-difficile* colonization (OTU174, ρ= +0.22).
-The Lachnospiraceae group also contained an OTU that had an exceptionally strong negative correlation (OTU39, ρ= +0.77).
-The Clostridia class was largely composed of the order Clostridiales (n=6), of which only one had a positive relationship with *C.
-difficile* colonization (OTU154, ρ= +0.26).
-
-Several broader groups of bacteria can work together (in some combination) to provide CR …possible that proteobacteria in general are bad …These results indicate that many related species within the larger/broader bacterial groups may be able to serve in providing colonization resistance against *C.
-difficile*.
-Other taxons in Figure 3 had more species specific correlations.
-For instance the Actinobacteria phylum included Enterorhabdus (OTU29, ρ= -0.70), Coriobacteriaceae (OTU50, ρ= -0.50), Bifodbacterium (OTU22, ρ= -0.34), Actinomyces (OTU58, ρ= +0.23).
-The Lactobacillales group is divided into two negatively correlated Lactobacillus OTUs (OTUs 23, 21, ρ= -0.73, -0.40 respectively) and two positively correlated Streptococcus OTUs (OTUs 78, 79).
-Similarly, the Bacteroidales group included species at both extremes, for example, an Alistipes (OTU20, ρ= -0.78) and a Bacteroides species (OTU44, ρ= +0.34).
-[END sent, all the other bad bugs look more like random smaller level taxonomic classification divisions, as in it’s a genus/species level effect type deal]
---functional redundancy? Across related large groups like porphyros?
---mostly Firmicutes—better competitors with cdiff? Because similar nutrients resources?
 Also functional redundancy with a highly diverse community… not just about who possibly but also ab being diverse—emphasizes community as a whole
@@
 Bacteria within a community setting may act differently in the context of other bacterial species.
-
-[using this time model for the metro… I wonder if it doesn’t work because of the time difference in the amount of recovery time?
-I’m picturing sort of an echo/ripple effect where at first the waves, or changes, are really large and then over time that rate of change of the community slows and the changes less pronounced/smaller.
-In this context the metro communities were at a stage farther along in the echo, meaning changes not happening as fast.
-Whereas the measurements we’ve taken and are using to make predictions are from the community at a time when you expect more/bigger/faster changes.
-So then the results are calculated assuming one rate but actually in a different (slower).
-By this image, the initial strength of the perturbation matters too.]
-
---also somewhere, ecoli are typically on avg (__% of the community, typically wayyyy outnumbered… whats the avg relabund of all those CR OTUs in an untreated mouse, how does that compare to the abundance of those bacteria I’ve implicated in –CR) …point is that Proteobacteria are usually outnumbered… don’t have a chance, gut is suited to bact/firm (which not surprisingly make up the major players in CR)..
-I bet those other outlier bad bugs are also in low abund in untreated mice (?).
-but when the gut environment changes for w/e reason, can give those rarer bad ones/proteo/esp e.
-coli a chance to take hold.
-They could just be individually opportunistic or work synergistically with cdiff.
-If individually opportunistic, it might be that they are facing pretty much a common enemy which is the lach/rum/clos/porphs and end up having similar techniques/niches as a result of that common enemy.
-Working synergistically—eg.
-b.
-theta sialidase where it cleaves off resources for cdiff.

### 2015-03-11 PAPER.Rmd
Add to discussion

@@
 
-Correlation analysis alone was insufficient in identifying all informative bacteria.
-Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
-difficile* yet improved model accuracy.
-*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*).
-While these bacteria may not have important direct effects on *C.
-difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
-difficile*.
-Modeling of the microbiome allows us to characterize *C.
-difficile* colonization in terms of a community, which is necessary to capture the complex relationships among its members.
-Few studies attempt to model the microbiome to predict disease or estimate health [Pamer, clostridium paper, psoriases, schubert].
-
 Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance.
@@
 
+Modeling of the microbiome allows us to characterize *C.
+difficile* colonization in terms of a community, which is necessary to capture these complex relationships among its members.
+Correlation analysis alone was insufficient in identifying all informative bacteria.
+Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
+difficile* yet improved model accuracy.
+*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*).
+While these bacteria may not have important direct effects on *C.
+difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
+difficile*.
+Although the applications for microbiome modelling are diverse, given the microbiome's relationship to many aspects of human health and disease, there are currently few studies attempt to model the microbiome to predict disease or estimate health [Pamer, psoriases, schubert].
+Microbiome modeling has been successfully applied to the prediction of psoriasis diagnosis, in which a microbiota signature defining subjects with and without psoriasis was discovered [Statnikov et al 2013].
+In a time-series model of microbiome dynamics, the microbiota response to clindamycin or clindamycin followed by *C.
+difficile* challenge was predicted.
+
+--------------
+
 Understanding how individual bacterial populations live in the context of a community is important for not only determining mechanisms but also for informing novel therapies.

### 2015-03-11 code/fix_abxD01_IDS_xlsx.R
Update results to reflect all files

@@ -62,7 +62,7 @@ spread_sheet[spread_sheet$sample=="021-3D-1" & spread_sheet$day==1,"sample"] <- 
 spread_sheet[spread_sheet$sample=="117-4-D4" & spread_sheet$day==5,"sample"] <- "117-4-D5"
 
 spread_sheet$sample[duplicated(spread_sheet$sample)]
-#spread_sheet$sample[duplicated(spread_sheet$sample)]
+#character(0)
 
 #make rownames the same as the sample names
 rownames(spread_sheet) <- spread_sheet$sample
@@ -95,25 +95,19 @@ mock_stub <- unique_stubs[mocks]
 
 #all of the sample names in the spreadsheet have a fastq file
 rownames(spread_sheet)[!(rownames(spread_sheet) %in% sample_stub)]
-#[1] "021-3D1"   "021-3D3"   "021-3D5"   "021-4D6"   "106-1D1"   "106-1D7"
-#[7] "109-1D5"   "109-5D1"   "109-5D7"   "111-1D1"   "111-1D10"  "111-1D4"
-#[13] "111-1D7"   "111-2D1"   "111-2D4"   "111-2D7"   "111-4D1"   "111-4D4"
-#[19] "117-4D5"   "117-5D1"   "600-1D2"   "600-2D6"   "600-2D7"   "600-2D8"
-#[25] "600-2D9"   "600-3D1"   "600-3D2"   "600-3D3"   "600-3D5"   "600-4D2"
-#[31] "600-4D9"   "102-1D8"   "102-2D-11" "102-2D-2"  "102-5D4"   "103-2D3"
-#[37] "103-3D1"   "103-3D2"   "103-3D3"   "103-3D6"   "107-1D3"   "107-1D9"
-#[43] "107-3D3"   "107-3D9"   "108-1D10"  "108-1D4"   "108-2D10"  "108-2D4"
-#[49] "116-1D7"   "116-2D7"   "088-1D1"   "088-1D10"  "088-1D2"   "088-1D3"
-#[55] "088-2D2"   "088-2D3"   "088-2D4"   "088-2D5"   "088-2D6"   "088-2D8"
-#[61] "088-3D5"   "088-3D8"   "088-4D2"   "088-4D3"   "088-4D4"   "089-1D5"
[…]

### 2015-03-11 PAPER.Rmd

@@
 
--community effect—have other things to compensate—highlight examples –streptomycin blue triangles—where low lachno but no cdiff so what else is protecting?
-Not one otu that explains it all… just cus low levels doesn’t screw you, also see recovered points in 5C-angela, sonnenburg-need other thingsssss
--tie back to mbio paper, with enterobacteriaceae they are screwed
--modeling—not much done for CDI, link back to humans, shows we could model risk as they come in—put towards end with the larger picture
-We designed statistical models based on a data set of varied antibiotic perturbations of the murine microbiota and their subsequent CR levels against *C.
-difficile*.
-Though antibiotic administration induced unique microbiota structures, colonization resistance was similarly low against *C.
-difficile*.
-These models incorporated all combinations of top candidate species-level OTUs.
-We then tested our top model’s ability to use new abundance information on a specific subset of the microbiota to predict the subsequent *C.
-difficile* colonization level, which was experimentally validated.
-These results greatly expand our knowledge of various perturbations and their subsequent effects on microbiota structure and function (colonization resistance ability).
-Furthermore modeling important/key bacterial subsets of the community we can build models to assess CDI risk.
-
------------
-
 Through our antibiotic perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of *C.
@@
 difficile* colonization.
-Using this collective microbiota data spanning a wide range of colonization resistance levels, we next moved beyond correlation to prediction of *C.
+Using this collective microbiota data spanning a wide range of colonization resistance levels, we moved beyond correlation to prediction of *C.
 difficile* colonization based on the composition of the microbiota.
@@
 difficile* levels.
-This is the first analysis both measuring and predicting colonization resistance levels based on the microbiota structure.
+This is the first comprehensive analysis both measuring and predicting colonization resistance levels based on the microbiota structure.
 
@@
 toxins) to more readily reach the epithelial layer.
+< EDIT PARAGRAPH.
+TRANSITION >
 
@@
 Microbiome modeling has been successfully applied to the prediction of psoriasis diagnosis, in which a microbiota signature defining subjects with and without psoriasis was discovered [Statnikov et al 2013].
-In a time-series model of microbiome dynamics, the microbiota response to clindamycin or clindamycin followed by *C.
-difficile* challenge was predicted.
-
---------------
-
-Understanding how individual bacterial populations live in the context of a community is important for not only determining mechanisms but also for informing novel therapies.
-
-Previous work characterizing the microbiome’s of human and mouse hosts in the context of CDI provide glimpses into potentially protective and alternatively susceptible community structures, i.e.
-which bacteria are present and their abundance.
-Cross-sectional investigations have characterized the microbiota of patients in the hospital with and without a *C.
-difficile* infection (CDI) and compared to healthy controls.
-These studies found patients with *C.
-difficile* tended to have a reduction in numbers of Clostridiales Incertae Sedis XI (Vincent et al), Lachnospiraceae, Ruminococcaceae, and Bacteroides (Schubert et al).
-These studies however lack the baseline healthy state of the microbiota, which preclude them from assessing risk of *C.
-difficile*.
-Murine models of CDI have the advantage of observing the specific changes that occur before and after controlled perturbations and the introduction of *C.
-difficile*.
-Communities with low levels of CR were associated with blooms in either Lactobacillaceae (cefoperazone-induced) (Theriot, 2014, bassis, 2014), Proteobacteria (induced by either clinda+5abx cocktail, clinda alone, or tigecycline) (Reeves, 2011, Buffie, 2011, Bassis 2014).
-Lachnospiraceae is associated with resistance to *C.
-difficile* infection (Reeves 2011) and has been directly shown to decrease colonization through mono-colonization studies in germfree mice (Reeves 2012).
-
-Even fewer studies attempt to model these observational changes in the microbiome following perturbation.
-Using the time series data gathered from the *C.
-difficile* challenge of clindamycin-treated mice (Buffie 2011), a Lotka-Volterra dyamics-based framework was built which incorporates both time and accounts for external perturbations (Stein 2013).
-They identified a subset of their network, containing Blautia, Akkermansia, and Coprobacillus, as having a negative relationship with *C.
-difficile*, while alternatively Enterococcus had a positive relationship.
+Similarly, incorporating microbiota data into epidemiological based models that classify *C.
+difficile* diagnosis significantly enhances predictive power [Schubert et al.
+2013].
+In a time-series model of microbiome dynamics, the microbiota response to *C.
+difficile* alone, clindamycin alone, or clindamycin followed by *C.
+difficile* challenge was predicted [Stein et al.
+2013].
+This model had a spearman correlation of 0.62 between actual and predicted data along time.
+
+Expanding on these studies, we have shown across several different types of perturbations that we can accurately predict the resistance levels of these communities to *C. 
+This has great applications for predicting patient risk of CDI in the hospital.
+Furthermore, this analysis can be used towards the design of personalized medicine, whether through tailoring probiotic or antibiotic regimens.
 
@@
 
-Any interesting points about the results?
-Mechanisms?
-Competition—ask matt if they are good?
-Proteobacteria could definitely play a role in susceptibility—creating an environment fit for cdiff… what ways could all those other bugs be making a bad environment for cdiff?
-Using up bile acids/germinants..,  with those bugs gone freeing up sugars, … maybe connecting a lot more to some of caseys metabolome paper, and how mine would fit that
-
-o	Limitations
-•	No smaller intervals? Of cdiff levels, or changes in microbiota
-•	Porphyromonadaceae is more prevalent in mice… so might not be as relevant for humans
-•	Different strains of cdiff have different microbiota requirements to protect?
-•	Doesn’t get at function… ?
-•	Collect more data to continue to improve our microbiome modeling abilities,
-
-o	Contribution to the field, positive spins
-•	PROBIOTICS
-•	Possible global effects? CR against other GI pathogens?
-•	clinical applications—PCR diagnostic for risk
-
-The final paragraph should highlight the main conclusion(s), and provide some indication of the direction future research should take. (for and I and I mention host)
--therapeutic: continuously supply probiotics  while lowering inflammation of the host (get host side back to normal state in tandem, via drugs?
-diet)… so attack the issue of CDI from both the microbiota end and host end… microbiota-host-pathogen triangle… …squeeze it out (eg.
-clindamycin clearance?
-If compared pathology of host at diff days along 10d for clindamycin vs cef/met/amp (these other ones that are high in cdiff coloniz), is host or toxin diff?...
-why does clinda clear after being CDhigh but no other ABX?)
-
 The human microbiome is complex; it has been observed that there is no identifiable “core” microbiome that all humans share (Turnbaugh 2009).

### 2015-03-11 PAPER.Rmd
Update intro

@@
 
-<not done>
-There is a need for more knowledge about interactions between members of the community.
-Know that fmts work [anna’s paper], but don’t know why, except that you recover the diversity and are receiving an established community already that works.
-Individual probiotics have not worked with complete success [dendukuri 2005], whereas studies find that combinations of bacteria often work better than by themselves [lawley, pamer, stein, hickson 2007].
+There is a clear need for more knowledge on *C.
+difficile* and its interactions between members of the bacterial community in understanding colonization resistance.
+Fecal microbiota transplants, which have been shown to increase *Bacteroidetes* and decrease *Proteobacteria* levels in the recipient, have successfully restored colonization resistance in patients and effectively resolved their severe CDI [Seekatz et al.
+2014].
+Precisely how this occurs is not fully understood, but it emphasizes the strength of the community as a whole in its protective ability.
+Individual probiotics have not worked with complete success [dendukuri 2005].
+In fact, several studies using probiotic mixtures of several bacteria find these members often work better together than by themselves [lawley, pamer, stein, hickson 2007].
+
 It’s also common for bacteria to change their behavior in the context of other bacterial species [Crost 2010, quorum sensing-biofilms, recognizing fellow bacterial species vs killing others, synergistic effects (sieuwerts 2010)].
 Understanding the scope of relationships between bacteria in a community is necessary to learn about its response, in concert with its host, to perturbations including antibiotic-induced, pathogen invasion, probiotics doses, and diet changes, etc.
-
-The purest way of studying these bacterial interactions is by characterizing them in the context of their whole biome [ecology citation?].
-[examples of papers showing microbiomes response to antibiotics in the context of CR]
 

### 2015-03-11 code/fix_abxD01_IDS_xlsx.R
Output text version of file

@@ -95,19 +95,11 @@ mock_stub <- unique_stubs[mocks]
 
 #all of the sample names in the spreadsheet have a fastq file
 rownames(spread_sheet)[!(rownames(spread_sheet) %in% sample_stub)]
-#[1] "021-3D1"  "021-3D3"  "021-3D5"  "021-4D6"  "111-1D1"  "111-1D10"
-#[7] "111-1D4"  "111-1D7"  "111-2D1"  "111-2D4"  "111-2D7"  "111-4D1"
-#[13] "111-4D4"  "117-4D5"  "117-5D1"  "103-3D2"  "116-1D7"  "116-2D7"
-#[19] "088-2D4"  "0071D1"   "0081D1"
+#[1] "021-3D1" "103-3D2" "088-2D4" "0071D1"  "0081D1"
 
 
 sample_stub[!(sample_stub %in% rownames(spread_sheet))]
-#[1] "007-1D1"   "008-1D1"   "087-2D6"   "088-1D-6"  "088-2D-6"  "088-4D-6"
-#[7] "089-1D-6"  "102-1D-11" "102-5D0"   "107-1D-5"  "108-2D-1"  "109-1D-5"
-#[13] "109-2D5"   "109-3D-11" "109-3D6"   "109-4D-11" "600-1D0"   "600-1D-1"
-#[19] "600-2D0"   "600-2D-3"  "600-2D-4"  "600-2D-5"  "600-2D-6"  "600-4D-1"
-#[25] "600-4D-3"  "600-4D-5"  "603-1D3"   "603-1D6"   "603-2D-1"  "626-2D10"
-#[31] "626-2D4"
+[1] "007-1D1" "008-1D1"
 
 
 # we can take care of these last two...
@@ -115,15 +107,10 @@ rownames(spread_sheet)[rownames(spread_sheet) == "0071D1"] <- "007-1D1"
 rownames(spread_sheet)[rownames(spread_sheet) == "0081D1"] <- "008-1D1"
 
 sample_stub[!(sample_stub %in% rownames(spread_sheet))]
-#[1] "087-2D6"   "088-1D-6"  "088-2D-6"  "088-4D-6"  "089-1D-6"  "102-1D-11"
-#[7] "102-5D0"   "107-1D-5"  "108-2D-1"  "109-1D-5"  "109-2D5"   "109-3D-11"
[…]

### 2015-03-11 code/fix_abxD01_IDS_xlsx.R
Entered # for comment lines

@@ -84,7 +84,7 @@ rownames(spread_sheet) <- gsub("^(\\d{3}-\\d)-D", "\\1D", rownames(spread_sheet)
 #samples and that they are named the same in the metadata (here) and the
 #sequence data
 
-fastq <- scan("data/raw/clean_fastqs.txt", what="")
+fastq <- scan("data/raw/clean_fastqs.txt", what="", quiet=TRUE)
 stub <- gsub("(.*)_S.*", "\\1", fastq)
 unique_stubs <- unique(stub)
 mocks <- grepl("mock", unique_stubs)
@@ -99,7 +99,7 @@ rownames(spread_sheet)[!(rownames(spread_sheet) %in% sample_stub)]
 
 
 sample_stub[!(sample_stub %in% rownames(spread_sheet))]
-[1] "007-1D1" "008-1D1"
+#[1] "007-1D1" "008-1D1"
 
 
 # we can take care of these last two...
@@ -107,7 +107,7 @@ rownames(spread_sheet)[rownames(spread_sheet) == "0071D1"] <- "007-1D1"
 rownames(spread_sheet)[rownames(spread_sheet) == "0081D1"] <- "008-1D1"
 
 sample_stub[!(sample_stub %in% rownames(spread_sheet))]
-character(0)
+#character(0)
 
 rownames(spread_sheet)[!(rownames(spread_sheet) %in% sample_stub)]
 #[1] "021-3D1" "103-3D2" "088-2D4"

### 2015-03-11 code/fix_abxD01_IDS_xlsx.R
Change the name of the output file in the header comment box

@@ -10,7 +10,7 @@
 # * data/raw/abx_D01IDS.xlsx
 #
 # Output...
-# * data/process/abxD01_IDS.txt
+# * data/process/abx_cdiff_metadata.tsv
 #
 ################################################################################
 

### 2015-03-11 code/fix_abxD01_IDS_xlsx.R
Fix information in comment header

@@ -4,7 +4,7 @@
 #
 # This script will convert Alyx's original abxD01_IDS.xlsx file into a text file
 # where the sample names correspond to the fastq sequence names. We are using
-# the file that Alyx sent to Pat on 2015-02-03
+# the file that Alyx sent to Pat on 2015-03-11
 #
 # Dependencies...
 # * data/raw/abx_D01IDS.xlsx

### 2015-03-11 PAPER.Rmd
Move paragraph from disc to intro

@@
 
-It’s also common for bacteria to change their behavior in the context of other bacterial species [Crost 2010, quorum sensing-biofilms, recognizing fellow bacterial species vs killing others, synergistic effects (sieuwerts 2010)].
+<not done>
 Understanding the scope of relationships between bacteria in a community is necessary to learn about its response, in concert with its host, to perturbations including antibiotic-induced, pathogen invasion, probiotics doses, and diet changes, etc.
@@
 These observed changes in response to perturbation can be used to inform models and networks to predict future responses (stein, Gordon diet paper) and characterize bacterial interactions (psoriasis/Schubert, sparCC paper).
-These methods are advantageous in that they can be refined to determine probiotic prophylactics or treatments based on an individual patient’s gut microbial needs.
-
-The purpose of this investigation was to test the effects of an array of perturbations on microbiome CR levels against *C.
-difficile*.
-We then used this information to design microbiome-based models for estimating pathogen colonization levels.
-Through the administration of various antibiotic regimens, we altered the murine gut microbiota and observed these changes using 16S rRNA sequencing.
+
+</not done>
+
+Although the applications for microbiome modelling are diverse, given the microbiome's relationship to many aspects of human health and disease, there are currently few studies attempt to model the microbiome to predict disease or estimate health [Pamer, psoriases, schubert].
+Microbiome modeling has been successfully applied to the prediction of psoriasis diagnosis, in which a microbiota signature defining subjects with and without psoriasis was discovered [Statnikov et al 2013].
+Similarly, incorporating microbiota data into epidemiological based models that classify *C.
+difficile* diagnosis significantly enhances predictive power [Schubert et al.
+2013].
+In a time-series model of microbiome dynamics, the microbiota response to *C.
+difficile* alone, clindamycin alone, or clindamycin followed by *C.
+difficile* challenge was predicted (spearman correlation of 0.62) [Stein et al.
+2013].
+
+The purpose of this investigation was to expand our current knowledge of the effects of various perturbations on the microbiota's colonization resistance abilities against *C.
+difficile* and identify important members for this function.
+Through the administration of different antibiotic types, doses, and varying recovery time, we altered the murine gut microbiota and observed these structural changes using 16S rRNA sequencing.
 The resultant gut communities were then challenged with *C.
 difficile* spores.
-Through correlation and random forest analysis of the starting communities with the subsequent level of *C.
-difficile* colonization, we built linear models to predict *C.
-difficile* colonization levels based on a subset of the microbiota.
-We used experimental data to test the ability of the model.
-
-Though we are limited in a resolution at the strain and species level, we have identified groups of related species with antagonistic relationships with *C.
+Using this data, we built a random forest model to predict *C.
+difficile* colonization levels based on the microbiota structure.
+Through this analysis, we have identified groups of related species with antagonistic relationships with *C.
 difficile* colonization.
-We identified several models with equally robust quality statistics and observed that across the top models a subset of 5 species were consistently present.
-These findings __
-</not done>
+Modeling identified bacteria that did not have significant correlations with *C.
+difficile* but were important in prediction accuracy.
+These findings show we can successfully apply modeling techniques to accurately measure the colonization resistance ability of a given microbiome.
 
@@
 
+< Clean >
 Furthermore, the dynamics of *C.
@@
 
+< Clean >
 While this may be suitable enough for *C.
@@
 toxins) to more readily reach the epithelial layer.
-< EDIT PARAGRAPH.
-TRANSITION >
-
+
+< TRANSITION >
 Modeling of the microbiome allows us to characterize *C.
@@
 difficile*.
-Although the applications for microbiome modelling are diverse, given the microbiome's relationship to many aspects of human health and disease, there are currently few studies attempt to model the microbiome to predict disease or estimate health [Pamer, psoriases, schubert].
-Microbiome modeling has been successfully applied to the prediction of psoriasis diagnosis, in which a microbiota signature defining subjects with and without psoriasis was discovered [Statnikov et al 2013].
-Similarly, incorporating microbiota data into epidemiological based models that classify *C.
-difficile* diagnosis significantly enhances predictive power [Schubert et al.
-2013].
-In a time-series model of microbiome dynamics, the microbiota response to *C.
-difficile* alone, clindamycin alone, or clindamycin followed by *C.
-difficile* challenge was predicted [Stein et al.
-2013].
-This model had a spearman correlation of 0.62 between actual and predicted data along time.
-
+
+< Transition >
 Expanding on these studies, we have shown across several different types of perturbations that we can accurately predict the resistance levels of these communities to *C. 
@@
 Identifying/cataloguing a comprehensive profile of those bacteria with potential roles in CR would be an initial step towards personalized/individualized treatments with known mixtures of protective bacteria.
-Bacteria within a community setting may act differently in the context of other bacterial species.
+
+It’s also common for bacteria to change their behavior in the context of other bacterial species [Crost 2010, quorum sensing-biofilms, recognizing fellow bacterial species vs killing others, synergistic effects (sieuwerts 2010)].

### 2015-03-11 code/make_files_file.R
Initial commit

@@ -1,0 +1,49 @@
+################################################################################
+#
+# make_files_file.R
+#
+# This script will build the files file that will be used in the make.contigs
+# command from within mothur.
+#
+# Dependencies...
+# * data/process/abx_cdiff_metadata.tsv
+# * fastq files stored in ../../cdiff_fastqs/
+#
+# Output...
+# * data/process/abxD0.files
+#
+################################################################################
+
+metadata <- read.table(file="data/process/abx_cdiff_metadata.tsv", header=T)
+
+#did the sample come from a mouse that eventually received Cdiff?
+cdiff_challenged <- metadata[metadata$cdiff==TRUE,]
+
+#get the day 0 samples from the abx treated mice
+abx_treated <- cdiff_challenged[cdiff_challenged$day == 0, ]
+
+#get the pre-abx samples from mice that would receive abx
+untreated <-  cdiff_challenged[cdiff_challenged$preAbx & cdiff_challenged$abx != "none", ]
+
+#concatenate the sample names together
+samples <- c(rownames(abx_treated), rownames(untreated))
+
+#get the list of fastq file names
+fastqs <- list.files("../../cdiff_fastqs/", pattern="*.fastq")
+
+
+#need a function that will build the files file line for each stub
+find_files <- function(stub, files=fastqs){
+    pattern <- paste0(stub, ".*R1.*fastq")  #use the stub to create a search pattern
+    r1_file <- files[grep(pattern, files)]  #find all of the possible R1 files
+    r2_file <- gsub("R1", "R2", r1_file)    #find all of the possible R2 files
+    label <- rep(stub, length(r1_file))     #make a vector of the sample name
+
+    #this all is necessary because some samples were sequenced multiple times
+    #the output of the above will be a series of vectors that here we will bind
+    #as columns, and paste across the columns for each row
+    apply(cbind(label, r1_file, r2_file), 1, paste, collapse="\t")
+}
+
+files_lines <- lapply(samples, find_files)
+write(unlist(files_lines), "data/process/abxD0.files")

### 2015-03-11 code/make_files_file.R
Add mock community samples

@@ -47,3 +47,11 @@ find_files <- function(stub, files=fastqs){
 
 files_lines <- lapply(samples, find_files)
 write(unlist(files_lines), "data/process/abxD0.files")
+
+
+r1_mock <- fastqs[grep("mock.*R1.*fastq", fastqs)]
+r2_mock <- fastqs[grep("mock.*R2.*fastq", fastqs)]
+stub_mock <- gsub("(.*)_S.*", "\\1", r1_mock)
+
+lines_mock <- apply(cbind(stub_mock, r1_mock, r2_mock), 1, paste, collapse="\t")
+write(lines_mock, "data/process/abxD0.files", append=T)

### 2015-03-11 PAPER.Rmd
Fix middle paragraphs of intro

@@
 
-There is a clear need for more knowledge on *C.
-difficile* and its interactions between members of the bacterial community in understanding colonization resistance.
+In understanding colonization resistance, there is a clear need for more knowledge on *C.
+difficile* and its interactions between members of the bacterial community.
 Fecal microbiota transplants, which have been shown to increase *Bacteroidetes* and decrease *Proteobacteria* levels in the recipient, have successfully restored colonization resistance in patients and effectively resolved their severe CDI [Seekatz et al.
@@
 In fact, several studies using probiotic mixtures of several bacteria find these members often work better together than by themselves [lawley, pamer, stein, hickson 2007].
-
-<not done>
-Understanding the scope of relationships between bacteria in a community is necessary to learn about its response, in concert with its host, to perturbations including antibiotic-induced, pathogen invasion, probiotics doses, and diet changes, etc.
-
-These observed changes in response to perturbation can be used to inform models and networks to predict future responses (stein, Gordon diet paper) and characterize bacterial interactions (psoriasis/Schubert, sparCC paper).
-
-</not done>
+Thus, studying *C.
+difficile* in the context of the community it is exposed to is essential for building a complete picture of the variables at play in its ability to colonize.
+One way of capturing these variables is through statistical modeling.
+Using models of the microbiome, we can not only identify bacteria with positive and negative roles in *C.
+difficile* colonization but also predict levels of colonization based on the structure of the microbiota, which *C.
+difficile* encountered.
 

### 2015-03-12 PAPER.Rmd
Finish edit Intro and Discussion
Update figure links. Delete intro/discussion notes.

@@
 difficile* and its interactions between members of the bacterial community.
-Fecal microbiota transplants, which have been shown to increase *Bacteroidetes* and decrease *Proteobacteria* levels in the recipient, have successfully restored colonization resistance in patients and effectively resolved their severe CDI [Seekatz et al.
+Fecal microbiota transplants, which have been shown to increase *Bacteroidetes* and decrease *Proteobacteria* levels in recipients, have successfully restored colonization resistance in patients and effectively resolved their severe CDI [Seekatz et al.
 2014].
@@
 Individual probiotics have not worked with complete success [dendukuri 2005].
-In fact, several studies using probiotic mixtures of several bacteria find these members often work better together than by themselves [lawley, pamer, stein, hickson 2007].
+In fact, several studies using probiotic mixtures of several bacteria find these members often work better together than by themselves [lawley, Buffie CG et al 2015, stein, hickson 2007].
 Thus, studying *C.
@@
 Using models of the microbiome, we can not only identify bacteria with positive and negative roles in *C.
-difficile* colonization but also predict levels of colonization based on the structure of the microbiota, which *C.
+difficile* colonization but also predict levels of colonization based on the structure of the microbiota that *C.
 difficile* encountered.
 
-Although the applications for microbiome modelling are diverse, given the microbiome's relationship to many aspects of human health and disease, there are currently few studies attempt to model the microbiome to predict disease or estimate health [Pamer, psoriases, schubert].
+Although the applications for microbiome modeling are diverse, given the microbiome's relationship to many aspects of human health and disease, there are currently few studies that attempt to model the microbiome to predict disease or estimate health [Pamer, psoriases, schubert].
 Microbiome modeling has been successfully applied to the prediction of psoriasis diagnosis, in which a microbiota signature defining subjects with and without psoriasis was discovered [Statnikov et al 2013].
@@
 2013].
+Lastly, the microbiota response to *Citrobacter rodentium* infection over time was modeled to capture time-dependent signatures of the microbiota [Belzer et al 2014].
 
@@
 
-![**Figure 1.**](Figure1/figure1.png)
+![**Figure 1.**](results/figures/Figure1/figure1.png)
 
@@
 
-![**Supplemental Figure 1.**](supplemental/sfigure1.png)
+![**Supplemental Figure 1.**](results/figures/supplemental/sfigure1.png)
 
@@
 
-![**Figure 2.**](Figure2/figure2_9x6.5.png)
+![**Figure 2.**](results/figures/Figure2/figure2_9x6.5.png)
 
@@
 
-![**Figure 3.**](Figure3/figure3.png)
+![**Figure 3.**](results/figures/Figure3/figure3.png)
 
@@
 
-![**Figure 4.**](Figure4/figure4.png)
+![**Figure 4.**](results/figures/Figure4/figure4.png)
 
@@
 
-![**Supplemental Figure 2.**](supplemental/sfigure2ab.png)
+![**Supplemental Figure 2.**](results/figures/supplemental/sfigure2ab.png)
 
@@
 
-![**Figure 5AB.**](Figure5/figure5ab.png)
-
-![**Figure 5C.**](Figure5/figure5c.png)
+![**Figure 5AB.**](results/figures/Figure5/figure5ab.png)
+
+![**Figure 5C.**](results/figures/Figure5/figure5c.png)
 
@@
 Pre-colonization with *Lachnospiraceae* was able to decrease the levels of *C.
-difficile* observed, while *E.
-coli* had similar colonization levels compared to germfree mice given *C.
-difficile* alone.
+difficile*, while *E.
+coli* pre-colonization and completely germfree mice had similar *C.
+difficile* levels.
 Association studies in humans with and without CDI in hospitals have also highlighted several bacteria found in this study, including *Lachnospiraceae*, *Rumicoccaceae*, *Clostridia*, and *Alistipes* as decreased and *E.
 coli* and *Streptococcus* as increased in subjects with CDI [Schubert, Antharam].
-While we observe overlaps in bacteria with potentially protective roles in both mice and humans at broad taxonomy levels, there are notable differences at refined levels.
-*Porphyromonadaceae* (Class *Bacteroidales*), for example, are more common among mice than humans, where instead there are strong negative associations of *C.
-difficile* with *Bacteroides* (Class *Bacteroidales*) [Schubert].
+While we observe overlaps in potentially protective bacteria between mice and humans at broad taxonomy levels, there are notable differences at finer taxonomic levels.
+Two different *Bacteroidales* species, a *Porphyromonadaceae* and a *Bacteroides*, have the strongest negative correlations with *C.
+difficile* in mice and humans, respectively [Schubert].
+*Porphyromonadaceae* is more commonly found in mice than in humans, indicating that there are host specific bacteria in colonization resistance.
 
@@
 difficile* (*Figure 5C*).
-However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C.
+However, a small group of microbiotas with greater than 1% OTU3 abundance were not fully protected by the presence of this OTU alone and ranged in *C.
 difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces.
+These included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas.
 Similarly, having low amounts of one "protective" bacteria such as *Lachnospriaceae* OTU39 does not automatically render the host susceptible to colonization.
@@
 difficile*.
-Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).
-
-< Clean >
+Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with protective capacities.
+
 Furthermore, the dynamics of *C.
-difficile* colonization does not appear to be dependent soley on the loss of protective species.
+difficile* colonization does not appear to be influenced solely by the loss of protective species.
 Both *Escherichia* OTU11 and *Streptococcus* OTU78 were positively correlated with *C.
@@
 Notably, *Escherichia* OTU11 was among the strongest positively correlated OTUs with *C.
-difficile* across all antibiotic treatments, except for streptomycin.
-In fact, while streptomycin treatment (5 mg/ml) lead to high *C.
-difficile* colonization, these communities had very little, if any, *Escherichia* OTU11.
-This demostrates that while *Escherichia* may aid in colonization of *C.
-difficile*, it is not essential.
-These observations lead to a model of colonization by *C.
-difficile* in which antibiotic perturbation leads to loss of protective species.
-This in turn eliminates competition, frees resources, and eliminates direct or indirect inhibitory effects of bacterial products (e.g.
-bacteriocins or short chain fatty acids) made by these protective bacteria.
-[can talk about change in gut with abx--loss of some, gain of others that are resistant to abx or grow up quickly after, and direct effects of abx...
-last weeks jclub paper; sonnenburg has 2 papers with nutrients/metbolites and cdiff related to the microbiome]
-
-< Clean >
+difficile* across the majority of antibiotic treatments.
+Though *Escherichia* may aid in colonization of *C.
+difficile*, our data suggests it is not essential.
+*Escherichia* OTU11 was near absent in streptomycin treated (5 mg/ml) mice, but *C.
+difficile* colonized to high levels (*Figure 1*).
+Together These observations lead to a model of colonization by *C.
+difficile* in which antibiotic perturbation leads to a significant loss of protective species.
+This in turn leads to a loss of immune homeostasis [Smith et al.
+2013 Science], an increase in available resources [Theriot 2014, Ferreyra et al 2014, Ng KM et al  2013, Koenigsknecht et al 2015] and eliminates direct or indirect inhibitory effects of bacterial products made by these protective bacteria (e.g.
+bacteriocins or short chain fatty acids) [Buffie CG et al 2015, Theriot 2014].
 While this may be suitable enough for *C.
 difficile* to colonize, other opportunistic species such as *Escherichia coli* and other *Proteobacteria* may also take advantage of these microbiota perturbations.
-Increased colonization by these bacteria can increase inflammation, produce a new profile of metabolites and bacterial products, reduce the mucus layer,  and deoxygenate the microoxic tiers nearing the epithelium.
-These changes in the gut milieu may be advantageous for *C.
-difficile* and other gut pathogens.
-Inflammation, for instance, has been shown to be advantageous for *Salmonella tyhpumurium* to further disrupt the indigenous microbiota [].
-Inflammation also leads to a shift in the availability of resources, such as nitrites, which can be used as an alternative resource by *Salmonella typhimurium* [].
-*E.coli* is capable of using resources found in the mucus layer while simultaneously deoxygenating this region [].
-This reduces the spatial protection of the mucus layer, allowing bacteria, including strictly anaerobic bacteria, and bacterial products (e.g.
-toxins) to more readily reach the epithelial layer.
-
-< TRANSITION >
+The bloom in these bacteria may further change the gut milieu in favor of *C.
+difficile* colonization.
+Increased growth by these bacteria can trigger inflammation, which has been shown to be advantageous for *Salmonella tyhpumurium* to further disrupt and suppress the indigenous microbiota [Stecher et al 2007 Plos Biol].
+Increased *Proteobacteria* abundances were also found among mice with increased mucus layer penetrability [Jakobsson et al.
+2015 EMBO reports].
+
 Modeling of the microbiome allows us to characterize *C.
@@
 difficile*.
-
-< Transition >
-Expanding on these studies, we have shown across several different types of perturbations that we can accurately predict the resistance levels of these communities to *C. 
+<Awkward placement, maybe after line 797>
+
+We have shown across several different types of perturbations that we can accurately predict the resistance levels of these communities to *C.
+difficile*.
 This has great applications for predicting patient risk of CDI in the hospital.
-Furthermore, this analysis can be used towards the design of personalized medicine, whether through tailoring probiotic or antibiotic regimens.
-
-------------------
-
-o  How relates to other studies
-•	Antonopoulos et al-reproducible community dynamics
-•	Antunes et al (finlay)—abx treatment on intestinal metabalome
-•	Sekirov—antibiotic induced perturbations of intestinal microbiota
-•	 “Change in the compostion alone of the intestinal microbiota (in the absence of significant changes in total numbers of intestinal microbes) prior to Salmonella serovar Typhimurium infection was sufficient to make mice more susceptible to Salmonella serovar Typhimurium, indicating that different subsets of the microbiota may play protective roles or enhance susceptibility to enteric infections.”
-•	Streptomycin and vancomycin doses were administered to mice prior to challenge with another gastrointestinal pathogen Salmonella enterica serovar Typhymurium (Sekirov 2008).
-This study showed ___.
-Changes in community composition were dose dependent.
-•	Strep increases bacteroidetes (CFB), vanc increases firms and then proteo (with increasing dose)
-•	Reeves/theriot/robinson—clindamycin/cef/cocktail/antbiotic administration alters the community structure
-•	Croswell—prolonge impact of abx on intestinal microbial ecology salmonella
-•	Rea- effect of  broad and narrow spectrum antimicrobials
-•	Ferreira—intestinal microbiota plays a role in salmonella induced colitis
-•	Theriot?—at least the metabolome paper—before and after with metabolites and cef, also does she look at anything taxonomically? yup
-
-Also functional redundancy with a highly diverse community… not just about who possibly but also ab being diverse—emphasizes community as a whole
-
-OTU39—Eubacterium ventriosum (94%), eubacterium rectale (93%), roseburia intestinalis/hominis/faecis (93%) (positively correlated to differences observed in discordant MZ twins for BMI—more abundant in higher BMI siblings… these are all butyrate producers, also capable of degrading fiber)
-	FOS—fructooligosaccharides produced by degredation of inulin or polyfructose
-
-Inulin and fructo-oligosaccharides have divergent effects on colitis and commensal microbiota in HLA-B27 transgenic rats.
-Abstract
-Modulation of intestinal microbiota by non-digestible carbohydrates may reduce inflammation in inflammatory bowel disease (IBD).
-The aim of the present study was to assess the effects of inulin and fructo-oligosaccharides (FOS) on intestinal microbiota and colitis in HLA-B27 transgenic rats, a well-validated rodent model for IBD.
-In this study, 4-week-old rats were fed 8 g/kg body weight inulin or FOS for 12 weeks, or not.
-Faeces were collected at 4 and 16 weeks of age; and caecal samples were collected at necropsy.
-The effects of inulin and FOS on chronic intestinal inflammation were assessed using a gross gut score, histology score and levels of mucosal IL-1β.
-Intestinal microbiota were characterised by quantitative PCR and denaturing gradient gel electrophoresis.
-Colitis was significantly reduced in all FOS-fed rats compared to the control diet, whereas inulin decreased chronic intestinal inflammation in only half the number of animals.
-Quantitative analysis of caecal microbiota demonstrated that inulin increased the numbers of total bacteria and the Bacteroides-Prevotella-Porphyromonas group, FOS increased bifidobacteria, and both fructans decreased Clostridium cluster XI.
-In the faecal samples, both inulin and FOS decreased total bacteria, Bacteroides-Prevotella-Porphyromonas group, and Clostridium clusters XI and XIVa.
-FOS increased Bifidobacterium spp., and mediated a decrease of gene copies of Enterobacteriaceae and Clostridium difficile toxin B in faeces.
-SCFA concentrations in the faecal and caecal samples were unaffected by the diets.
-In conclusion, FOS increased the abundance of Bifidobacterium spp., whereas both fructans reduced Clostridium cluster XI and *C.
-difficile* toxin gene expression, correlating with a reduction of chronic intestinal inflammation.
-
-The human microbiome is complex; it has been observed that there is no identifiable “core” microbiome that all humans share (Turnbaugh 2009).
-It’s been observed that there is a high level of interindividual variation among healthy individuals’ microbiomes (Costello 2009).
-Thus the definition of a healthy microbiome is one that encompasses a wide range of microbiota structures, each one evolved with their host.
-It’s highly likely there is no one size fits all solution for the treatment or prevention of CDI.
-Thus characterizing the relationships within the microbiome are important in the context of a community.
-Need more studies that look at the level of the community as a whole in CR over the effect of single bugs.
-Examples of papers looking at combinations/subsets of bacteria in CR
-
-	It makes sense that many bacteria would play a role in colonization resistance and that it is more about the community composition as a whole given that interindividual differences is so great.
-Two people share no more than __% of their microbiomes.
-They are unique to individuals and yet all these different profiles of microbiotas are capable of colonization resistance.
-People do not get *C.
-difficile* infections without first being exposed to some major purturbation of their microbiome, namely antibiotics.
-Thus many structures of the microbiome, and many distinct bacterial populations must be capable of contributing to resistance.
-Identifying/cataloguing a comprehensive profile of those bacteria with potential roles in CR would be an initial step towards personalized/individualized treatments with known mixtures of protective bacteria.
-
-It’s also common for bacteria to change their behavior in the context of other bacterial species [Crost 2010, quorum sensing-biofilms, recognizing fellow bacterial species vs killing others, synergistic effects (sieuwerts 2010)].
+Furthermore, this analysis can be used to inform the design of personalized medicine, whether through tailoring probiotic or antibiotic regimens.
+Future studies are needed to examine of the human-specific bacteria in colonization resistance and also to examine the changes in models necessary to account for physiological differences between *C.
+difficile* strains.
+<End sentence is weak.>

### 2015-03-12 PAPER.Rmd
Update statistical methods

@@
 ***Statistical Analysis & Modeling*** All analyses were conducted using R version 3.1.2.
+OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (43 OTUs).
+Spearman rank correlation analysis was performed with a Benjamini and Hochberg adjustment between OTU counts or diversity levels and *C.
+difficile* CFU/g feces.
+Comparison of bacterial levels among titration groups or delayed groups of the same antibiotic was performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests with a Benjamini and Hochberg adjustment.
+Comparison of log10 transformed *C.
+difficile* CFU/g feces between experimental groups was calculated using ANOVA and subsequently pairwise T tests with a Benjamini and Hochberg adjustment.
+Random forest regression models were constructed using the randomForest R package using 5,000 trees [ref].
+The regression was performed using the log10 transformation of the number of CFU/g fecal material as the dependent variable and the 43 OTUs as predictor variables.
 Complete analysis scripts are available at the online repository for this study (https://github.com/SchlossLab/Schubert_AbxD01_mBio_2015).
-
-Need text about...
-* correlation analysis
-* comparison of titration levels
-* For the random forest you use 1% within each treatment group, for the correlation you state using 0.1 and 0.01% across all samples. Need to stay consistent.
-* statistical methods used to differentiate the abundances of OTUs in the titrations
-
-Random forest regression models were constructed using the randomForest R package using XX,000 trees [ref].
-The model was constructed using the relative abundance of those OTUs that had an average relative abundance within any antibiotic treatment group of 1%.
-This resulted in 44 OTUs.
-The regression was performed using the log10 transformation of the number of CFU per gram of fecal material as the dependent variable.
 

### 2015-03-16 PAPER.Rmd
Add pre-amble code

@@
-> Note: The text in block code such as this are not in the original paper. They are there to provide an explanation of the R code and results.
->
->
+```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
+opts_chunk$set("dev" = c("png", "cairo_pdf"))
+opts_chunk$set(results = "hold")
+opts_chunk$set(fig.show = "hold")
+opts_chunk$set(fig.align = "center")
+opts_chunk$set(cache = FALSE)
+opts_chunk$set(echo = FALSE)
+opts_chunk$set(warning = FALSE)
+
+load_package <- function(package){
+    if(!(package %in% rownames(installed.packages()))){
+        install.packages(package)
+    }
+    library(package, quietly=TRUE)
+}
+
+load_package("randomForest")
+load_package("png")
+
+set.seed(6201976)
+```
 
@@
 
-**Short title:** Alterations of the Murine Gut Microbiota that Lead to Clostridium difficile Colonization
-
-Alyxandria M. Schubert<sup>1</sup>, Hamide Sinani<sup>1</sup>, and Patrick D.
-Schloss<sup>1</sup>\*
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\* Correspondence: <pschloss@umich.edu>
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;734.647.5801
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Department of Microbiology and Immunology
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;University of Michigan
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;5618 Medical Sciences II
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;1500 W. Medical Center Dr.
-
-&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Ann Arbor, MI 48109
-
-1 Department of Microbiology and Immunology, 1500 W. Medical Center,
-University of Michigan, Ann Arbor, MI 48109
+**Short title:** Alterations of the murine gut microbiota lead to *C. difficile* colonization
+
+Alyxandria M. Schubert, Hamide Sinani, and Patrick D.
+Schloss\*
+
+\* Correspondence:
+> Department of Microbiology and Immunology
+
+> University of Michigan
+
+> 1520A Medical Science Research Bldg. I
+
+> 1500 W. Medical Center Dr.
+
+> Ann Arbor, MI 48109
+
+> pschloss@umich.edu
+
+> 734.647.5801
 
@@
 
-***Animal Care.*** We used 5-8 week old C57Bl/6 mice for all our experiments.
+***Animal care.*** We used 5-8 week old C57Bl/6 mice for all our experiments.
 These mice were reared under SPF conditions within the animal facility at the University of Michigan.
@@
 
-***Antibiotic Administration.*** Mice were administered one of seven different antibiotics, including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
+***Antibiotic administration.*** Mice were administered one of seven different antibiotics, including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
 The route of administration depended on the specific antibiotic.
@@
 ***C.
-difficile Preparation and Challenge*** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
+difficile preparation and challenge*** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
 difficile* challenge.
@@
 
-***Sample Collection and Plating*** Fecal samples were freshly collected for each mouse on the day of *C.
+***Sample collection and plating*** Fecal samples were freshly collected for each mouse on the day of *C.
 difficile* challenge (day 0).
@@
 
-***DNA Extraction and Sequencing*** Total bacterial DNA was extracted from each  day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
+***DNA extraction and sequencing*** Total bacterial DNA was extracted from each  day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
 Sequencing preparation and sequencing was carried out as previously described (Kozich JJ et al.
@@
 
-***Sequence Curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
+***Sequence curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
 Sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
@@
 
-***Statistical Analysis & Modeling*** All analyses were conducted using R version 3.1.2.
+***Statistical analysis and modeling*** All analyses were conducted using R version 3.1.2.
 OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (43 OTUs).
@@
 *Porphyromonadaceae* is more commonly found in mice than in humans, indicating that there are host specific bacteria in colonization resistance.
+
+Modeling of the microbiome allows us to characterize *C.
+difficile* colonization in terms of a community, which is necessary to capture these complex relationships among its members.
+Correlation analysis alone was insufficient in identifying all informative bacteria.
+Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
+difficile* yet improved model accuracy.
+*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*).
+While these bacteria may not have important direct effects on *C.
+difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
+difficile*.
 
@@
 
-Modeling of the microbiome allows us to characterize *C.
-difficile* colonization in terms of a community, which is necessary to capture these complex relationships among its members.
-Correlation analysis alone was insufficient in identifying all informative bacteria.
-Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
-difficile* yet improved model accuracy.
-*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*).
-While these bacteria may not have important direct effects on *C.
-difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
-difficile*.
-<Awkward placement, maybe after line 797>
-
 We have shown across several different types of perturbations that we can accurately predict the resistance levels of these communities to *C.

### 2015-03-16 PAPER.Rmd
Add amova analysis

@@
 
+```{r amova_analysis}
+amova <- scan(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.topdose.thetayc.0.03.lt.ave.amova", what="", sep="\n", quiet=T)
+
+comparison_line <- amova[(1:length(amova)) %% 6 == 1]
+comparison <- gsub("\t.*", "", comparison_line)
+
+pvalue_line <- amova[(1:length(amova)) %% 6 == 0]
+pvalue <- gsub("p-value: ", "", pvalue_line)
+pvalue <- gsub("\\*", "", pvalue)
+pvalue <- gsub("<", "", pvalue)
+pvalue <- as.numeric(pvalue)
+sig <- p.adjust(pvalue, method="BH")
+```
+
 ***Levels of colonization resistance are associated with the initial structure of the gut microbiota.*** In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to *C.
@@
 difficile* inoculation revealed distinct and reproducible microbiota structures (**Figure 1**).
-The community structures of the beta-lactam-treated mice (i.e.
-cefoperazone and ampicillin) were not significantly different from each other (AMOVA, p=XXX0.36XXX).
-Clindamycin-treated microbiotas were significantly different from all other treatment groups (AMOVA, p=XXX0.001XXX) and were characterized by the predominance of *Enterobacteriaceae*.
-Of all groups significantly different from the untreated microbiota, the levels of *Enterobacteriaceae* increased with all treatments except with streptomycin.
-Vancomycin-treated communities saw the largest bloom in *Akkermansia*.
-In addition, the ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, XXXp=0.09XXX).
+The ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=`r round(sig[comparison=="cipro-no_abx"], 2)`).
 Subsequently, 24 hours after inoculation *C.
-difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g) (**Figure 1**).
+difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g).
+With the exception of the community structures found in the beta-lactam-treated mice (i.e.
+cefoperazone and ampicillin; AMOVA, p=`r round(sig[comparison=="amp-cef"], 2)`) all of the other comparisons were significantly different from each other and from the untreated controls (AMOVA, all p<`r round(max(sig[sig<0.05]), 2)`).
+
 We calculated the diversity of all of the communities using the Inverse Simpson and Shannon indices and observed a significant negative correlation with subsequent *C.
@@
 
-> ### Amova to compare resulting community structures
-> Amova calculation in mothur to determine the uniqueness of each community structure:
->
-
-```
-dist.shared(shared=abxD01.final.an.unique_list.0.03.topdose2.shared, calc=thetayc-jclass, subsample=1625)
-amova(phylip=abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.dist, design=topdose_design_byABX.txt, iters=1000000)
-
-```
-> This results in the file called "abxD01.final.an.unique_list.0.03.topdose2.thetayc.0.03.lt.ave.amova" located in the github folder. The results of these are summarized in "amova8expgroups.xlsx".
->
 > ### Figure 1: Bargraph

### 2015-03-25 PAPER.Rmd
Fix top dose descriptive paragraph

@@
 
+```{r top_dose_cfu_analysis}
+cfu <- read.table(file="data/process/abxD1.counts")
+cfu_top_dose <- cfu[cfu$experiment == "top_dose" & cfu$abx != "control",]
+cfu_top_dose_summary <-aggregate(cfu_top_dose$CFU, by=list(cfu_top_dose$abx), summary)
+treatments <- cfu_top_dose_summary$Group.1
+cfu_top_dose_summary <- format(cfu_top_dose_summary$x, scientific=T, digits=2)
+rownames(cfu_top_dose_summary) <- treatments
+```
+
 ```{r amova_analysis}
@@
 pvalue <- as.numeric(pvalue)
+
+comparison <- comparison[-1]
+pvalue <- pvalue[-1]
+
 sig <- p.adjust(pvalue, method="BH")
@@
 difficile* colonization.
-It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] had decreased resistance against this pathogen.
-To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics (**Table 1**) to differentially alter the microbiota and assess their resistance to *C.
-difficile* colonization.
-16S rRNA gene sequencing of the fecal communities sampled at the time of *C.
-difficile* inoculation revealed distinct and reproducible microbiota structures (**Figure 1**).
-The ciprofloxacin-treated microbiota were the only group that was not significantly different from untreated mice (AMOVA, p=`r round(sig[comparison=="cipro-no_abx"], 2)`).
-Subsequently, 24 hours after inoculation *C.
-difficile* was undetectable in the untreated mice or the ciprofloxacin-treated mice; however it had successfully colonized the mice in all other treatment groups to similarly high levels (10<sup>7</sup>-10<sup>8</sup> CFU/g).
-With the exception of the community structures found in the beta-lactam-treated mice (i.e.
-cefoperazone and ampicillin; AMOVA, p=`r round(sig[comparison=="amp-cef"], 2)`) all of the other comparisons were significantly different from each other and from the untreated controls (AMOVA, all p<`r round(max(sig[sig<0.05]), 2)`).
-
-We calculated the diversity of all of the communities using the Inverse Simpson and Shannon indices and observed a significant negative correlation with subsequent *C.
-difficile* colonization (both **Supplemental Figure 1**; p<XXX0.001XXX).
-These results demonstrate that a variety of non-overlapping community structures can be sensitive to *C.
-difficile* colonization.
+It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] had decreased colonization resistance.
+To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics to differentially alter the microbiota and assess their resistance to *C.
+difficile* colonization (**Table 1**).
+On the day after the mice were challenged with *C.
+difficile* we enumerated the density of *C.
+difficile* in the animals' feces.
+We observed reproducibly high levels of *C.
+difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin.
+We observed variable levels of *C.
+difficile* colonization in mice treated with amipicilin.
+Only one of six mice receiving vancomycin was colonized with *C.
+difficile* and none of the mice that received ciprofloxacin were colonized.
+We next sequenced the 16S rRNA genes from the fecal communities sampled at the time of *C.
+difficile* challenge to characterize the populations within untreated mice that provided for colonization resistance.
+The antibiotic treatments resulted in distinct and reproducible microbiota structures (**Figure 1**; AMOVA, P<`r round(max(sig[grepl("control", comparison)]), 2)`).
+In addition, all of the other comparisons were significantly different from each other (AMOVA, P<`r round(max(sig[!grepl("control", comparison)]), 2)`).
+These results indicate that perturbing the gut microbiota with antibiotics resulted in non-overlapping community structures that yielded significant variation in susceptibility to colonization when challenged with *C.
+difficile*.
 
@@
 --------
-
-> ### CFU stats comparison
-> To compare CFU levels:
-
-```{r, eval=TRUE}
-
-file<-read.csv("~/Documents/Github/abxD01/Figure1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
-
-data <- file[ which(file$expgroup != '1untr'), ] #choose experimental groups
-data$nextDayCFU<-log10(data$nextDayCFU+1) #log transform CFUs
-anova(lm(data$nextDayCFU ~ data$expgroup)) #perform anova across experimental groups for CFU
-
-#Which groups are different? Ciprofloxacin was clearly lower
-data <- data[ which(data$expgroup != '2cipro'), ] #remove ciprofloxacin to see now if there's a difference in remaining groups
-anova(lm(data$nextDayCFU ~ data$expgroup))
-
-```
 

### 2015-03-25 code/build_figure1.R
Rewrite function to build Figure 1

@@ -1,0 +1,139 @@
+#data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.subsample.shared
+#data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.cons.taxonomy
+#data/process/abxD0.counts
+
+
+# read in the metadata file
+counts_file <- read.table(file="data/process/abxD1.counts", header=T)
+top_dose <- counts_file[counts_file$experiment=="top_dose" | counts_file$abx=="control",]
+top_dose_med <- aggregate(top_dose$CFU, by=list(top_dose$abx), median)
+abx_cfu <- format(top_dose_med$x, scientific=T, digits=2)
+abx_cfu <- gsub("e\\+0", "E", abx_cfu)
+abx_cfu <- gsub("0.0E0", "0", abx_cfu)
+names(abx_cfu) <- top_dose_med$Group.1
+
+# read in the shared file and get the relative abundance
+shared_file <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.subsample.shared", header=T, row.names=2)
+shared_file <- shared_file[,!(colnames(shared_file) %in% c("label", "numOtus"))]
+n_seqs <- apply(shared_top_dose, 1, sum)[1]
+rel_abund <- shared_top_dose/n_seqs
+
+
+# need to figure out which samples made it through the pipeline and look at those
+overlap <- which(rownames(rel_abund) %in% rownames(counts_file))
+rel_abund <- rel_abund[overlap,]
+rel_abund <- 100 * rel_abund
+top_dose <- top_dose[overlap,]
+
+# let's get the relative abundances for those phyla that have at least one
+# sample where they are more than 10% o the community
+good_phyla <- apply(rel_abund, 2, max) > 10
+rel_abund_good <- rel_abund[,good_phyla]
+
+
+# let's get the taxonomy data
[…]

### 2015-03-25 PAPER.Rmd
Edit up to correlation analysis

@@
 
-![**Figure 1.**](results/figures/Figure1/figure1.png)
-
-**Figure 1.
-Microbiota structures following diverse antibiotic treatments.** This figure shows the average percent relative abundances for family level phylotypes found in each experimental group on the day of *C.
-difficile* challenge.
-Each row depicts a different antibiotic perturbation.
-The level *C.
-difficile* colonization found 24 hours post microbiota sampling is indicated in the top right for each treatment.
-Error bars indicate standard error of the mean.
-
 --------
@@
 --------
-
-> ### Figure 1: Bargraph
->
-> This figure was built using the R script called "stacked.barcharts.r", which contains the stackedbarcharts() funciton in the abxD01 folder.The input files are listed in Github.
-The graph can be sorted 2 ways either by the untreated's decreasing relabund (sortbyphyl=FALSE), or by the most abundant within each phylum (sortbyphyl=TRUE).
-For each treatment, the resulting *C.
-difficile* CFU/g Feces is also labeled on the Y axis below the treatment name.
-> The parameters I changed in the r script were:
-
-```{r, eval=FALSE, warning=FALSE}
-abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")
-
-library(Hmisc)
-stackedbarcharts(file="~/Documents/Github/abxD01/Figure1/abxD01.final.tx.2.subsample.relabund.topdose2.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure1/topdose_tx2_barchart_ids.csv", graphLabels = abx, sortbyphyl = TRUE, graphbyphyl = FALSE)
-
-```
-
-> Files saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf".
->
-> ### Supplemental Figure 1: correlation plot of diversity with *C. difficile*
->
-> For the correlation calculations see the next section of the results.
-> Figure making:
-
-```
-# In mothur:
-summary.single(calc=invsimpson-shannon, shared=abxD01.final.an.unique_list.shared, subsample=1625)
-
-```
-> Using these results, I selected the samples within the original antibiotic set.
-Then I calculated the spearman correlation of *C.
-difficile* CFU/g feces on day 1 with the level of diversity found in the community on day 0 (day of challenge).
-Both inverse simpson and shannon indices were included.
 
@@
 <End sentence is weak.>
+
+**Figure 1.
+Antibiotic treatments result in significant alterations to the structure of the microbiome and variation in colonization resistance.** The median percent relative abundances of the most abundant phyla found in each experimental group on the day of *C.
+difficile* challenge.
+Each bargraph depicts a different antibiotic perturbation.
+The median level *C.
+difficile* colonization found 24 hours post microbiota sampling is indicated in the top right for each treatment.
+Error bars indicate the interquartile range.

### 2015-03-25 code/build_figure1.R
Fix indexing bug

@@ -15,12 +15,12 @@ names(abx_cfu) <- top_dose_med$Group.1
 # read in the shared file and get the relative abundance
 shared_file <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.subsample.shared", header=T, row.names=2)
 shared_file <- shared_file[,!(colnames(shared_file) %in% c("label", "numOtus"))]
-n_seqs <- apply(shared_top_dose, 1, sum)[1]
-rel_abund <- shared_top_dose/n_seqs
+n_seqs <- apply(shared_file, 1, sum)[1]
+rel_abund <- shared_file/n_seqs
 
 
 # need to figure out which samples made it through the pipeline and look at those
-overlap <- which(rownames(rel_abund) %in% rownames(counts_file))
+overlap <- rownames(rel_abund)[which(rownames(rel_abund) %in% rownames(top_dose))]
 rel_abund <- rel_abund[overlap,]
 rel_abund <- 100 * rel_abund
 top_dose <- top_dose[overlap,]

### 2015-03-26 PAPER.Rmd
Fix order of x-axis labels; output correlation values

@@
 
---------
-
---------
-
-![**Supplemental Figure 1.**](results/figures/supplemental/sfigure1.png)
-
-**Supplemental Figure 1.
-The inverse of Simpson's diversity index for the community on day 0 is negatively correlated with *C.
-difficile* colonization on day 1.** Spearman's rho is plotted in the top right corner.
-
---------
-
-```{r, eval=FALSE}
-
-diversity<-read.csv("~/Documents/Github/abxD01/supplemental/topdose_diversity.csv",  header=T)
-diversity$nextDayCFU <- log((diversity$nextDayCFU+1), 10)
-
-#correlation of top dose
-pval.spear <- c()
-cor.spear <- c()
-count <- 1
-for(i in 4:5){
-  cor.spear[count] <- cor.test(diversity[,3],diversity[,i], method="spearman")$estimate
-  pval.spear[count] <- cor.test(diversity[,3],diversity[,i], method="spearman")$p.value
-  count <- count + 1
-}
-corr.results <- as.matrix(cbind(cor.spear, pval.spear))
-rownames(corr.results) <- names(diversity)[4:5]
-
-# Plot inverse simpson vs. C. difficile levels
-color<-rainbow(7)
-colors.top <- c(control = "black",
-          cipro = color[1],
-          clinda = "#FFCC00",
-          vanc.625 = "#00CC00",
-          strep5 = "#33FFFF",
-          cef.5 = color[5],
-          amp.5 = "#6600FF",
-          metro1 = color[7])
-
-# shapes
-pch.top <- c(control = 1,
-         cipro = 16,
-         clinda = 16,
-         vanc.625 = 16,
-         strep5 = 16,
-         cef.5 = 16,
-         amp.5 = 16,
-         metro1 = 16)
-
-#these are needed for the legend
-legend.labels.top <- c(control = "Control",
-            cipro = "Ciprofloxacin",
-            clinda = "Clindamycin",
-            vanc.625 = "Vancomycin",
-            strep5 = "Streptomycin",
-            cef.5 = "Cefoperazone",
-            amp.5 = "Ampicillin",
-            metro1 = "Metronidazole")
-
-par(mar=c(5, 5, 1, 1) +0.1, mgp=c(3, 1, 0))
-par(mfrow=c(1,1))
-
-plot(diversity$invsimpson[diversity$expgroup=="control"],
-     diversity$nextDayCFU[diversity$expgroup=="control"],
-     pch=pch.top[1], col=colors.top[1], xlim=c(0, 25), ylim=c(0,10),  
-     cex=1.5, yaxt="n",  xaxt="n", cex.lab=1,
-     ylab=as.expression("Log " ~ italic("C. difficile") ~ " CFU/g feces"), xlab="Inverse Simpson Index")
-
-axis(1, las=1, cex.axis=1, at=seq(0,25, by=5))
-axis(2, las=1, cex.axis=1, at=seq(0,10, by=1))
-
-for(i in 2:(length(unique(diversity$expgroup)))){
-  points(diversity$invsimpson[diversity$expgroup==names(legend.labels.top)[i]],
-         diversity$nextDayCFU[diversity$expgroup==names(legend.labels.top)[i]],
-         pch=pch.top[i], col=colors.top[i], cex=1.5)
-}
-
-legend("topright", inset=0.05, legend=legend.labels.top, pch=pch.top, col=colors.top,  pt.cex=1.5, cex=1, bty="n", y.intersp=1.5)
-
-```
-
---------
+```{r top_dose_correlation}
+
+```
 
 ***Correlation analysis reveals potentially protective bacteria.*** To identify bacterial taxa that could be associated with resistance or susceptibility to *C.
-difficile*, we measured the correlation between the relative abundance of each species-level OTU on the day of inoculation with the level of *C.
+difficile*, we measured the correlation between the relative abundance of each OTU that had an average abundance greater than 0.1% on the day of inoculation with the level of *C.
 difficile* colonization 24 hours later (Figure 2).
-<needs help> Of the OTUs with an average abundance greater than 0.1%, OTUs with negative a correlation were associated with providing resistance against *C.
+
+OTUs with negative a correlation were associated with providing resistance against *C.
 difficile* (N=XXX40XXX) and those with a positive correlation were associated with susceptibility (N=XXX7XXX) (p<XXX0.05XXX).</needs help> Among various bacterial families, three were consistently negatively correlated with *C.

### 2015-03-26 code/build_figure2.R

@@ -1,8 +1,3 @@
-#data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared
-#data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy
-#data/process/abxD1.counts
-
-rm(list=ls())
 
 # read in the metadata file
 counts_file <- read.table(file="data/process/abxD1.counts", header=T)
@@ -30,34 +25,59 @@ names(taxonomy) <- rownames(taxonomy_file)
 taxonomy <- gsub("\\(\\d*\\)", "", taxonomy)
 taxonomy <- gsub(";[^;]*;$", "", taxonomy)
 taxonomy <- gsub(";unclassified", "", taxonomy)
-taxonomy <- gsub(".*;", "", taxonomy)
-
+taxonomy <- gsub("_1", "", taxonomy)
 taxonomy <- taxonomy[colnames(rel_abund_good)]
 
 
 get_CFU_cor <- function(OTU){
-
     cor.test(top_dose$CFU, rel_abund_good[,OTU], method="spearman")[c("estimate","p.value")]
-
 }
 
-otu_cfu_corrs <- matrix(unlist(sapply(1:ncol(rel_abund_good), get_CFU_cor)), ncol=2, byrow=T)
-rownames(otu_cfu_corrs) <- colnames(rel_abund_good)
-colnames(otu_cfu_corrs) <- c("rho", "pvalue")
-sig <- p.adjust(otu_cfu_corrs[,"pvalue"], method="BH") < 0.05
+otu_cfu_corrs <- sapply(1:ncol(rel_abund_good), get_CFU_cor)
 
-otu_cfu_corrs <- otu_cfu_corrs[sig,]
-taxonomy <- taxonomy[sig]
+colnames(otu_cfu_corrs) <- colnames(rel_abund_good)
 
+sig <- p.adjust(otu_cfu_corrs["p.value",], method="BH") < 0.05
+otu_cfu_corrs <- otu_cfu_corrs[,sig]
+
+sig_taxonomy <- taxonomy[sig]
+sig_corrs <- unlist(otu_cfu_corrs["estimate",])
+
+
+sig_taxonomy <- gsub("Actinobacteria", "5Actinobacteria", sig_taxonomy)
+sig_taxonomy <- gsub("Bacteroidetes", "1Bacteroidetes", sig_taxonomy)
[…]

### 2015-03-26 code/build_figure1.R
Add header and comments

@@ -1,6 +1,22 @@
-#data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.subsample.shared
-#data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.cons.taxonomy
-#data/process/abxD0.counts
+################################################################################
+#
+# build_figure1.R
+#
+# This script builds Figure 1, which is a barchart of the median relative
+# abundance of each phylum found in the mice treated with the top dose of
+# antibiotics as well as the untreated control mice. Also included is the
+# number of CFU per gram of feces.
+#
+# Dependencies...
+#   * data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.subsample.shared
+#   * data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.v4.wang.pick.pick.tx.5.cons.taxonomy
+#   * data/process/abxD1.counts
+#
+# Output...
+#   * results/figures/figure1.pdf
+#
+################################################################################
+
 
 
 # read in the metadata file

### 2015-03-26 code/build_figure2.R

@@ -1,7 +1,29 @@
+################################################################################
+#
+# build_figure2.R
+#
+# This script builds Figure 2. The first thing it does is to calculate the
+# correlation between each OTU that has an average relative abundance over 1%
+# and the number of CFU per gram of feces. We then generate a stripchart that
+# indicates the correlation for each family of OTUs. Finally, the correlation
+# data is outputed so that it can be reported in the main paper. The p-values
+# for each correlation are corrected using the Benjimani-Hochberg correction.
+#
+#
+# Dependencies...
+#   * data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared
+#   * data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy
+#   * data/process/abxD1.counts
+#
+# Output...
+#   * results/figures/figure1.pdf
+#
+################################################################################
 
 # read in the metadata file
 counts_file <- read.table(file="data/process/abxD1.counts", header=T)
 top_dose <- counts_file[counts_file$experiment=="top_dose" | counts_file$abx=="control",]
+
 
 # read in the shared file and get the relative abundance
 shared_file <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared", header=T, row.names=2)
@@ -9,16 +31,22 @@ shared_file <- shared_file[,!(colnames(shared_file) %in% c("label", "numOtus"))]
 n_seqs <- apply(shared_file, 1, sum)[1]
 rel_abund <- shared_file/n_seqs
 
+
 # need to figure out which samples made it through the pipeline and look at those
 overlap <- rownames(rel_abund)[which(rownames(rel_abund) %in% rownames(top_dose))]
 rel_abund <- rel_abund[overlap,]
 top_dose <- top_dose[overlap,]
 
+
+# limit the analysis to those OTUs that have an average relative abundance over
+# 1% across all samples
[…]

### 2015-03-26 code/build_figure2.R
Fix delimeter in output file

@@ -126,4 +126,4 @@ dev.off()
 
 
 # output the correlation data
-write.table(file="data/process/top_dose_corr.tsv", cbind(taxonomy=gsub("\\d", "", sig_taxonomy), sig_corrs), quote=F, row.names=F, sep="")
+write.table(file="data/process/top_dose_corr.tsv", cbind(taxonomy=gsub("\\d", "", sig_taxonomy), sig_corrs), quote=F, row.names=F, sep="\t")

### 2015-03-26 code/build_figure2.R
Set random seed so stripchart doesn't keep changing

@@ -105,6 +105,9 @@ tax_labels <- paste(family_levels, phylum_levels)
 
 
 pdf(file="results/figures/figure2.pdf", width=6.5, height=4.0)
+
+    set.seed("6201976")
+
     par(mar=c(10,6,0.5,0.5))
 
     plot(NA, xlim=c(1,length(tax_labels)), ylim=c(-0.8, 0.8), axes=F, xlab="", ylab="")

### 2015-03-26 PAPER.Rmd
Add correlation analysis and figure legend

@@
 ```{r top_dose_correlation}
+corr_data <- read.table(file="data/process/top_dose_corr.tsv", header=T)
+families <- gsub(".*;", "", corr_data$taxonomy)
+ave_corr_family_agg <- aggregate(corr_data$sig_corrs, by=list(families), function(x){ c(mean(x), length(x))})
+ave_corr_family <- round(ave_corr_family_agg$x[,1], 2)
+names(ave_corr_family) <- ave_corr_family_agg$Group.1
 
+N_family <- round(ave_corr_family_agg$x[,2], 2)
+names(N_family) <- ave_corr_family_agg$Group.1
 ```
@@
 difficile* colonization 24 hours later (Figure 2).
-
-OTUs with negative a correlation were associated with providing resistance against *C.
-difficile* (N=XXX40XXX) and those with a positive correlation were associated with susceptibility (N=XXX7XXX) (p<XXX0.05XXX).</needs help> Among various bacterial families, three were consistently negatively correlated with *C.
-difficile* colonization: *Porphyromonadaceae* (ρ<sub>average</sub>=XXX-0.63XXX), *Lachnospiraceae* (ρ<sub>average</sub>=XXX-0.51XXX), *Ruminococcaceae* (ρ<sub>average</sub>=XXX-0.53XXX).
-OTUs from the *Proteobacteria* phylum had strong positive relationships with *C.
+OTUs associated with providing resistance against *C.
+difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
+Among various bacterial families, three were consistently associated with *C.
+difficile* resistance: *Porphyromonadaceae* (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs), *Lachnospiraceae* (ρ<sub>average</sub>=`r ave_corr_family["Lachnospiraceae"]`, N=`r N_family["Lachnospiraceae"]` OTUs), and *Ruminococcaceae* (ρ<sub>average</sub>=`r ave_corr_family["Ruminococcaceae"]`, N=`r N_family["Ruminococcaceae"]` OTUs).
+Two OTUs from the *Proteobacteria* had a significant association with *C.
 difficile* colonization.
-These included OTUs associated with the genera *Pseudomonas* (ρ=XXX0.31XXX) and *Escherichia* (ρ=XXX0.56XXX).
-Other notable OTUs that were found at the extremes of the distribution of correlation scores included those affiliated with the genus *Lactobacillus* (family *Lactobacillaceae*; OTU23, ρ=XXX-0.73XXX), *Alistipes* (family *Rickenellaceae*; OTU20, ρ=XXX-0.78XXX), *Enterorhabdus* (phylum *Actinobacterium*, family *Coriobacteriaceae*; OTU29, ρ=XXX-0.70XXX), and *Streptococcus* (family *Streptococcaceae*; OTU78, ρ=XXX0.36XXX).
-Overall, these results suggest that several larger groups of related OTUs, particularly within the *Firmicutes* and *Bacteroidetes* phyla, may participate in colonization resistance against *C.
+These included OTUs associated with the genera *Pseudomonas* (ρ=`r round(corr_data[families=="Pseudomonadaceae","sig_corrs"], 2)`) and *Escherichia* (ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
+Overall, these results suggest that several large groups of related OTUs, particularly within the *Firmicutes* and *Bacteroidetes*, may participate in colonization resistance against *C.
 difficile*, while members of the *Proteobacteria* phylum may be involved in susceptibility.
-
---------
-
-![**Figure 2.**](results/figures/Figure2/figure2_9x6.5.png)
-
-**Figure 2.
-Broader taxonomic groups share similar relationships with *C.
-difficile*.** Species level OTUs (3% cutoff) found on day 0 were correlated with the level of *C.
-difficile* CFU/g feces on day 1 using Spearman’s rank based correlation coefficient.
-Only OTUs with an average abundance of at least <needs help>0.05%<needs help> across all original treatments were considered in the correlation analysis.
-All correlations shown in this graph are significant after controlling for multiple comparisons using the Benjimani-Hochberg correction (p<0.05).
-OTUs are grouped by family.
-An asterisk (*) indicates that the OTU was unclassified at the family level.
-
---------
-
-> ### Correlation Calculations
->
-> The required file is located on Github.
-
-```{r, eval=TRUE, warning=FALSE}
-
-meta <- read.table('~/Documents/Github/abxD01/Figure2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.txt',header=T)
-#meta<-meta[1:96] #change based on number of OTUs above .05%, then add 2 for first two cols
-c<-1
-otu <- c()
-cor.spear <- c()
-pval.spear <- c()
-cor.ken = c()
-pval.ken = c()
-for(i in 3:length(meta)){
-  otu[c] <- colnames(meta[i])
-  cor.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$estimate
-  pval.spear[c] <- cor.test(meta[,2],meta[,i], method="spearman")$p.value
-  cor.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$estimate #good to see because kendall handles ties
-  pval.ken[c] <- cor.test(meta[,2],meta[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
-  c <- c+1
-}
-
-pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
-pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons
-
-results = NULL
-results <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
-colnames(results) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
-results <- results[order(results[,3]),]  #order by pvalue column=3
-#write.table(results[1:dim(results)[1],], file="~/Documents/Github/abxD01/Figure2/abxD01.final.an.unique_list.0.03.subsample.filter16mintotal.shared.correl.topdose2.results.txt", sep="\t", row.names=FALSE)
-print(results[1:dim(results)[1],])
-
-```
-
-> ### Figure 2: Stipchart
->
-> This figure shows the correlation analysis of bacterial species present on Day 0 with *C. difficile* levels on Day 1 using the original set of antibiotic groups.
->
-> Necessary files are on github, and it was made using the following code:
-
-```{r, eval=FALSE}
-
-##Paper Figure: correlations for topdose
-otuCorrs<-read.csv("~/Documents/Github/abxD01/Figure2/corr_avgAbove0.001_topdose2_stripchart.csv", header=T)
-
-pch <- c(6,6,6,6, 1,1, 1, 1, 1, 1, 1, 1,0, 0, 9, 2, 2, 4)
-par(mar=c(8, 6, 2, 2) +0.1, mgp=c(4, 1, 0)) #default is 5.1 4.1 4.1 2.1 [bottom, left, top, right space], mgp=c(3, 1, 0) [label line location for x/y location labels, tick mark labels location, tick mark locations]
-stripchart(otuCorrs$corr ~ otuCorrs$graphOrder, vertical=TRUE, ylab="Spearman Correlation", ylim=c(-1, 1), cex.axis=1, cex=2, pch=pch, lwd=2, col="black",  method="jitter",  jitter=.2, cex.lab=1.1, yaxt="n", xaxt="n")
-axis(1, cex.axis=1, at=(1:nlevels(otuCorrs$family)), labels=FALSE)
-family.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "family"])
-text(1:nlevels(otuCorrs$family)+.3, par("usr")[3]-0.05, label=(family.labels), xpd=NA, pos=2, srt=45, cex=1)
-axis(2, cex.axis=1, at=seq(-1,+1, by=.1), las=1)
-color_transparent <- adjustcolor("black", alpha.f = 0.3)
-abline(v=c(1:nlevels(otuCorrs$family)), col=color_transparent)
-abline(h=0, col="dark gray", lwd=2)
-abline(h=c(0.25, 0.5, 0.75, -0.25, -0.5, -0.75), col=color_transparent, lwd=2, lty="dashed")
-phylum.labels <- unique(otuCorrs[order(otuCorrs$graphOrder), "phyl"])
-leg<-legend("top",  legend=phylum.labels, pch=c(6, 1, 0, 9, 2, 4), cex=1, bty="o", horiz=TRUE, pt.cex=1, text.width=2.5)
-
-```
-
-> Figure saved as "corr_avgAbove0.001_topdose2_10x10.pdf".
 
@@
 
+**Figure Legends**
+
 **Figure 1.
@@
 Error bars indicate the interquartile range.
+
+**Figure 2.
+Broader taxonomic groups share similar relationships with *C.
+difficile* colonization.** The relative abundance of OTUs found on the day that mice were challenged with *C.
+difficile* spores were correlated with the level of C.
+difficile CFU/g feces 24 hours later.
+Only OTUs with an average abundance of at least 0.1% across all original treatments were considered in the correlation analysis.
+All correlations shown in this graph are significant after controlling for multiple comparisons using the Benjimani-Hochberg correction (P<0.05).
+OTUs are grouped by the taxonomic family that its sequences classified as or the next deeper classification that provided a classification.
+The letters in the parentheses correspond to the phylum that the taxa belongs to.
+B: Bacteroidetes, F: Firmicutes, P: Proteobacteria, V: Verrucomicrobia, A: Actinobacteria, T: Tenericutes.

### 2015-03-26 PAPER.Rmd
Fix formatting

@@
 difficile* challenge to characterize the populations within untreated mice that provided for colonization resistance.
-The antibiotic treatments resulted in distinct and reproducible microbiota structures (**Figure 1**; AMOVA, P<`r round(max(sig[grepl("control", comparison)]), 2)`).
+The antibiotic treatments resulted in distinct and reproducible microbiota structures (Figure 1; AMOVA, P<`r round(max(sig[grepl("control", comparison)]), 2)`).
 In addition, all of the other comparisons were significantly different from each other (AMOVA, P<`r round(max(sig[!grepl("control", comparison)]), 2)`).

### 2015-03-26 PAPER.Rmd
Add hyp tests to compare CFU for dilutions

@@
 
-***Smaller perturbations of the microbiota result in maintenance of colonization resistance.*** All but the ciprofloxacin treatment resulted in similarly high levels of *C.
+```{r titration}
+cfu <- read.table(file="data/process/abxD1.counts")
+titration <- cfu[cfu$abx %in% c("vanc", "cef", "strep"),]
+
+vanc <- titration[titration$abx=="vanc",]
+vanc.test <- pairwise.t.test(log(vanc$CFU+1), factor(vanc$dose), method="BH")
+
+strep <- titration[titration$abx=="strep",]
+strep.test <- pairwise.t.test(log(strep$CFU+1), factor(strep$dose), method="BH")
+
+cef <- titration[titration$abx=="cef",]
+cef.test <- pairwise.t.test(log(cef$CFU+1), factor(cef$dose), method="BH")
+
+```
+
+***Reduced perturbation of the microbiota increases colonization resistance.*** Based on the *C.
+difficile* colonization levels in our seven antibiotic treatments, we hypothesized that reducing the dose of antibiotics that the mice received would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C.
 difficile*.
-Therefore, we hypothesized that titrating the dose of antibiotics given to the mice would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C.
-difficile*.
-We titrated the doses of cefoperazone, streptomycin, and vancomycin given to the mice (Supplemental Table 1).
-Colonization levels decreased significantly in all mice receiving titrated doses in support of our hypothesis (Figure 3).
-*C.
-difficile* levels were significantly different between all titration groups within the cefoperazone (p<XXX0.03XXX) and streptomycin treated mice (p<XXX0.01XXX).
-For each of the antibiotics we titrated, reduced dosages resulted in reduced colonization by *C.
-difficile*.
-By sequencing the 16S rRNA genes from feces collected on day 0 from these mice, we sought to validate the relationships we observed with the full dosage.
+In addition to the previous treatments, we treated mice with cefoperazone (0.3 and 0.1 mg/mL), streptomycin (0.5 and 0.1 mg/mL), and vancomycin (0.3 and 0.1 mg/mL).
+These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e.
+cefoperazone), Gram-negative (i.e.
+streptomycin) and Gram-positive (i.e.
+vancomycin) bacteria.
+Vancomycin was also selected because although the community was quite different from untreated mice, we did not observe high levels of colonization and we anticipated that lower doses might result in a community structure that would result in colonization.
+As expected, colonization levels decreased significantly in all mice receiving reduced doses of cefoperazone (P<`r round(max(cef.test$p.value, na.rm=T), 2)`) and streptomycin (P<`r round(max(strep.test$p.value, na.rm=T), 2)`; Figure 3).
+Although the 0.625 mg/mL dose of vancomycin did not result in colonization, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`).
+
 Titrating cefoperazone resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
@@
 The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.
+
 The at the lowest streptomycin titration, mice retained significantly higher abundances of *Alistipes* and *Anaeroplasma* (p<0.05), which we found to be strongly associated with colonization resistance.
 The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice.
+
 Although the vancomycin titration yielded a number of large changes in the abundance of individual populations, there was no recovery of colonization resistance.
@@
 In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the *Akkermansia* and *Enterobacteriaceae*.
-Together, these results suggest that individual populations are not sufficient to provide colonization resistance.
-Rather, resistance is likely a product of the overall composition of the community.
 
---------
-
-![**Figure 3.**](results/figures/Figure3/figure3.png)
-
-**Figure 3.
-The average percent relative abundances for family level phylotypes found on the day of *C.
-difficile* challenge in each titration group.** Each row depicts a different antibiotic that was titrated.
-High, medium, and low antibiotic doses are represented by black, grey, and white colored bars respectively.
-The specific dose amount is shown for each antibiotic.
-Error bars show the standard error of the mean.
-
---------
+Together, these results suggest that individual populations are not sufficient to provide colonization resistance. Rather, resistance is likely a product of the overall composition of the community.
 
@@
 B: Bacteroidetes, F: Firmicutes, P: Proteobacteria, V: Verrucomicrobia, A: Actinobacteria, T: Tenericutes.
+
+**Figure 3.
+The average percent relative abundances for family level phylotypes found on the day of *C.
+difficile* challenge in each titration group.** Each row depicts a different antibiotic that was titrated.
+High, medium, and low antibiotic doses are represented by black, grey, and white colored bars respectively.
+The specific dose amount is shown for each antibiotic.
+Error bars show the standard error of the mean.

### 2015-04-01 PAPER.Rmd
Remove extra code from titration section

@@
 vancomycin) bacteria.
-Vancomycin was also selected because although the community was quite different from untreated mice, we did not observe high levels of colonization and we anticipated that lower doses might result in a community structure that would result in colonization.
-As expected, colonization levels decreased significantly in all mice receiving reduced doses of cefoperazone (P<`r round(max(cef.test$p.value, na.rm=T), 2)`) and streptomycin (P<`r round(max(strep.test$p.value, na.rm=T), 2)`; Figure 3).
+Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
+We anticipated that lower doses might result in a community structure that would result in colonization.
+As expected, colonization levels decreased significantly in all mice receiving reduced doses of cefoperazone (P<`r round(max(cef.test$p.value, na.rm=T), 2)`).
+Reducing the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 4) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 9) and an OTU associated with the genus *Pseudomonas* (OTU 65; Figure 3).
+The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis.
+Reducing the dose of streptomycin significantly reduced the colonization levels (P<`r round(max(strep.test$p.value, na.rm=T), 2)`; Figure 3).
+These changes matched the correlations that we observed across all of the antibiotic treatments.
+Reducing the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Allistipes* (OTU 11), and *Bacteroidales* (OTU 17).
+Interestingly, the OTUs that classified as members of the *Porphyromonadaceae* family did not all follow the same dose response.
+Of the 8 OTUs that changed with dose of streptomycin, 5 followed the response predicted from the correlation-based analysis (OTUs 5, 9, 10, 11, and 13), 1 contradicted the expected response (OTU 3), and 2 had a mixed response (OTUs 2 and 17).
 Although the 0.625 mg/mL dose of vancomycin did not result in colonization, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`).
-
-Titrating cefoperazone resulted in a shift in dominance from the *Bacteroidetes* to the *Firmicutes* phylum with increasing antibiotic dose.
-Most notable was the increase in the relative abundance of OTUs associated with resistance including the *Porphyromonadaceae* and *Alistipes* and a reduction in OTUs associated with susceptibility including members of the *Proteobacteria*, *Clostridium*, and *Staphylococcus*.
-The relative abundance of other OTUs associated with protection including the *Lactobacillus*, *Lachnospiraceae*, and *Ruminococcaceae* did not change as expected based on the initial correlation analysis; however, these OTUs were present at levels similar to the untreated mice.
-
-The at the lowest streptomycin titration, mice retained significantly higher abundances of *Alistipes* and *Anaeroplasma* (p<0.05), which we found to be strongly associated with colonization resistance.
-The relative abundances of the *Porphyromonodaceae*, *Lachnospiraceae*, *Ruminococcaceae* were unchanged, but were present at similarly high levels to those found in the untreated mice.
-
-Although the vancomycin titration yielded a number of large changes in the abundance of individual populations, there was no recovery of colonization resistance.
-The relative abundance of many of the populations associated with resistance (*Porphyromonadaceae*, *Lachnospiraceae*, *Anaeroplasma*) increased with decreased vancomycin dosage; however, with the exception of the *Lactobacillus* populations, they did not reach the levels observed in the untreated mice.
-In addition, all of the vancomycin-treated mice had high levels of OTUs affiliated with the *Akkermansia* and *Enterobacteriaceae*.
-
-Together, these results suggest that individual populations are not sufficient to provide colonization resistance. Rather, resistance is likely a product of the overall composition of the community.
-
-> ### CFU stats comparison
->
-> For comparison of the CFU's among titration groups:
-
-```{r, eval=TRUE}
-file<-read.csv("~/Documents/Github/abxD01/Figure3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", header=T)
-file<-file[!grepl("untr", file$expgroup),]
-file$nextDayCFU <- log10(file$nextDayCFU+1)
-
-cef <- file[file$abx=="Cefoperazone", 3:4]
-strep <- file[file$abx=="Streptomycin", 3:4]
-vanc <- file[file$abx=="Vancomycin", 3:4]
-
-diffs <- function(data){
-  data.aov<-aov(data$nextDayCFU ~ data$expgroup)
-  anova(data.aov)
-  pairwise.t.test(data$nextDayCFU, data$expgroup, p.adjust="BH")
-}
-
-diffs(cef)
-diffs(strep)
-diffs(vanc)
-
-```
-
-> ### Figure 3: Barchart
->
-> This figure shows the effect of titrating antibiotics on the community as well as the resulting *C. difficile* levels.  
->
-> The file/fileIDS should be matched by the titration.
-> Input files on github:
-> 1. abxD01.final.tx.2.subsample.alltitrations.forlogscale.csv
-> 2. alltitrations_tx2_barchart_ids.csv
->
-> Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
-
-```{r, eval=FALSE}
-
-stackedbarcharts(file = "~/Documents/Github/abxD01/Figure3/abxD01.final.tx2.subsample.alltitrations.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure3/alltitrations_tx2_barchart_ids.csv", excludeGroup="untr", divide = TRUE)
-
-```
+Seven OTUs were differentially represented across the three vancomycin doses.
+Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
+Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
+Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
+Together, these results suggest that although individual populations were occasionally predictive of colonization levels, they were not sufficient to consistently predict colonization resistance.
+In light of such results, resistance is likely a product of the overall composition of the community.
 

### 2015-04-01 code/build_figure4.R
Use one-sided test

@@ -166,7 +166,7 @@ cairo_pdf(file="results/figures/figure4.pdf", width=7.5, height=4.125)
     arrows(x0=q, y0=as.matrix(metro_cfu_med), y1=as.matrix(metro_cfu_lci), angle=90, length=0.05)
     axis(4, las=1, at=c(1, 1e2, 1e4, 1e6, 1e8), label=c(0, expression(10^2), expression(10^4), expression(10^6), expression(10^8)))
     box()
-    #wilcox.test(metro_metadata$CFU~metro_metadata$experiment)
+    #wilcox.test(metro_metadata$CFU~metro_metadata$experiment, alternative="less")
     text(x=2, y=rep(4e8), labels=c("*"), cex=2)
 
 
@@ -180,8 +180,8 @@ cairo_pdf(file="results/figures/figure4.pdf", width=7.5, height=4.125)
     arrows(x0=q, y0=as.matrix(amp_cfu_med), y1=as.matrix(amp_cfu_lci), angle=90, length=0.05)
     axis(4, las=1, at=c(1, 1e2, 1e4, 1e6, 1e8), label=c(0, expression(10^2), expression(10^4), expression(10^6), expression(10^8)))
     box()
-    #wilcox.test(amp_metadata$CFU~amp_metadata$experiment)
-    text(x=2, y=4e8, labels=c("NS"))
+    #wilcox.test(amp_metadata$CFU~amp_metadata$experiment, alternative="less")
+    text(x=2, y=4e8, labels=c("*"), cex=2)
 
 
     mtext(side=4, "C. difficile colonization (CFU/g)", line=3, at=1e10)

### 2015-04-01 code/build_figure4.R
Fix color problem

@@ -114,7 +114,7 @@ cairo_pdf(file="results/figures/figure4.pdf", width=7.5, height=4.125)
     metro_uci <- aggregate(sig_metro, by=list(metro_metadata$experiment), function(x){quantile(x, prob=0.75)})[,-1]
     metro_lci <- aggregate(sig_metro, by=list(metro_metadata$experiment), function(x){quantile(x, prob=0.25)})[,-1]
 
-    z <- barplot(as.matrix(metro_med), beside=T, names.arg=rep("", ncol(metro_med)), ylim=c(0,80), xlim=c(1.5,x_max), axes=F, col=c("black", "gray", "white"))
+    z <- barplot(as.matrix(metro_med), beside=T, names.arg=rep("", ncol(metro_med)), ylim=c(0,80), xlim=c(1.5,x_max), axes=F, col=c("gray", "white"))
     arrows(x0=z, y0=as.matrix(metro_med), y1=as.matrix(metro_uci), angle=90, length=0.02)
     arrows(x0=z, y0=as.matrix(metro_med), y1=as.matrix(metro_lci), angle=90, length=0.02)
 
@@ -133,7 +133,7 @@ cairo_pdf(file="results/figures/figure4.pdf", width=7.5, height=4.125)
     amp_uci <- aggregate(sig_amp, by=list(amp_metadata$experiment), function(x){quantile(x, prob=0.75)})[,-1]
     amp_lci <- aggregate(sig_amp, by=list(amp_metadata$experiment), function(x){quantile(x, prob=0.25)})[,-1]
 
-    z <- barplot(as.matrix(amp_med), beside=T, names.arg=rep("", ncol(amp_med)), ylim=c(0,60), xlim=c(1.5,x_max), axes=F, col=c("black", "gray", "white"))
+    z <- barplot(as.matrix(amp_med), beside=T, names.arg=rep("", ncol(amp_med)), ylim=c(0,60), xlim=c(1.5,x_max), axes=F, col=c("gray", "white"))
     arrows(x0=z, y0=as.matrix(amp_med), y1=as.matrix(amp_uci), angle=90, length=0.02)
     arrows(x0=z, y0=as.matrix(amp_med), y1=as.matrix(amp_lci), angle=90, length=0.02)
 
@@ -161,7 +161,7 @@ cairo_pdf(file="results/figures/figure4.pdf", width=7.5, height=4.125)
[…]

### 2015-04-01 PAPER.Rmd
Rough in random forest section

@@
 
-***Allowing recovery of specific bacterial populations restores colonization resistance.*** Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C.
+```{r delay}
+cfu <- read.table(file="data/process/abxD1.counts")
+delay <- cfu[cfu$abx %in% c("metro", "amp"),]
+
+metro <- delay[delay$abx=="metro",]
+metro_test <- wilcox.test(metro$CFU~factor(metro$experiment), alternative="less")
+metro_med_cfu <- aggregate(metro$CFU, by=list(metro$experiment), median)$x
+metro_delay <- metro[metro$experiment=="delay",]
+
+amp <- delay[delay$abx=="amp",]
+amp_test <- wilcox.test(amp$CFU~factor(amp$experiment), alternative="less")
+```
+
+***Allowing recovery of specific bacterial populations restores colonization resistance.***
+In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hrs before challenging them with *C.
+difficile*.
+Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C.
 difficile* is restored [refs].
-We allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover an additional five days.
-We again sought to test the relationships that we identified in the correlation-based analysis using animals with the partially recovered microbiota.
-On Day 1 post challenge of the metronidazole-treated mice who were allowed to recover five extra days, we observed a significant decrease in *C.
-difficile* level of CFU per g feces (p<XXX0.0001XXX).
-Within their microbiota there were significant increases (p<XXX0.05XXX) in overall abundances of *Porphyromonadaceae*, *Ruminococcaceae*, and *Lachnospiraceae* to abundances similar to those of untreated mice, and significant decreases (p<XXX0.05XXX) in levels of *Enterobacteriaceae*.
-In contrast, the ampicillin-treated mice that were given five extra days of recovery prior to *C.
-difficile* challenge did not experience an increase in colonization resistance.
-Inspection of their microbiota indicated that they still had significantly reduced levels of bacteria correlated with *C.
-difficile* protection, including *Bacteroides*, *Alistipes*, and *Ruminoccocaceae* compared with untreated animals (p<XXX0.05XXX).
-Importantly, the level of *Enterobacteriaceae* was not significantly reduced with additional recovery time.
-Given the smaller changes towards a protective baseline state, it was not all that surprising that we were unable to detect significant changes in subsequent *C.
-difficile* levels (p=XXX0.12XXX).
-Together, these results strengthen the importance of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae* in setting up the right environment of negatively and positively correlated bacteria to affect the outcome of *C.
-difficile* colonization.
+To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover an additional five days.
+Among the metronidazole-treated mice, those that recovered longer had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)` reduction in colonization (P<0.001; Figure 4).
+In addition, of the `r nrow(metro_delay)` mice that experienced the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C.
+difficile* 24 hours after challenge.
+We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 4).
+Most notable among these was an OTU associated with the the family *Porphyromonadaceae* (OTU 3) and the genus *Escherichia* (OTU 4).
+The relative abundance of the former increased with the delay and the relative abundance of the latter decreased.
+Based on the correlation analysis, the trajectories of the relative abundances for these OTUs matched the observed colonization levels.
+Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in colonization (P=`r round(amp_test$p.value, 2)`).
+Again, we observed a similar increase and decrease in relative abundance for the *Porphyromonadaceae* (OTU 3) and *Escherichia* (OTU 4) OTUs.
 
---------
+```{r random_forest}
 
-![**Figure 4.**](results/figures/Figure4/figure4.png)
-
-**Figure 4.
-Changes in the average relative abundance for family level phylotypes durin the period of recovery prior to challenge with *C.
-difficile*.** Each row depicts a different antibiotic that was allowed to recover.
-Black bars represent untreated animals.
-Grey bars represent animals given 1 day of recovery off antibiotics as in Figure 1.
-Error bars indicate the standard error of the mean.
-
---------
-
-> How is this relevant? - PDS
-
-![**Supplemental Figure 2.**](results/figures/supplemental/sfigure2ab.png)
-
-**Supplemental Figure 2.
-Changes between untreated and recovery days.** A) The average thetaYC distance between pretreatment (day -11) and recovery days (days -5 through -1) is represented by circles connected by solid lines.
-Dashed lines with the same symbol indicate individual mice from the same cage.
-Antibiotic treatment is indicated by color.
-B) The average level of the inverse simpson diversity index is plotted by antibiotic treatment across time.
-Error bars show standard error of the mean.
-Mice were challenged with *C.
-difficile* on day 0.
-
---------
-> How is this relevant? - PDS
-> ### Supplemental Figure 2A: Differences between recovery days' thetayc
-
-```{r, eval=FALSE}
-file <- "~/Desktop/mothur/abxD01/abxD01.final.an.unique_list.0.03.recovery.vancomycin.thetayc.0.03.lt.ave.dist"
-diffs <- getDistDiffs(file=file)
-
-#metadata <- read.delim(file="~/Desktop/mothur/abxD01/searchme.txt", header = T, stringsAsFactors=F)
-metadata$sample[1951] <- "007-1D1" # ID was messed up
-names(metadata)[3] <- "group"
-
-par(mfrow=c(2, 1))
-par(mar=c(3, 5, 0.5, 0.5), las=2, mgp=c(3, 1, 0))
-plot.dist.from(dist.from = diffs)
-
-# Arrange diversity data for supplemental figure 2B
-
-diversity <- read.delim(file="~/Desktop/mothur/abxD01/abxD01.final.an.unique_list.groups.ave-std.summary", sep="\t", header=T, stringsAsFactors=F)
-
-diversity$group[2046] <- "007-1D1"
-diversity$group[124] <- "007-1D1"
-
-div.ind <- match(diversity$group[diversity$method=="ave"], metadata$sample)
-div.group <- metadata$group[div.ind]
-div.time <- metadata$time[div.ind]
-div.abx <- metadata$abx[div.ind]
-div.day <- metadata$day[div.ind]
-
-delay.invsimp <- diversity$invsimpson[div.time=="delayed" & diversity$method=="ave"]
-delay.shannon <- diversity$shannon[div.time=="delayed" & diversity$method=="ave"]
-delay.abx <- div.abx[div.time=="delayed" & diversity$method=="ave"]
-delay.day <- div.day[div.time=="delayed" & diversity$method=="ave"]
-delay.group <- div.group[div.time=="delayed" & diversity$method=="ave"]
-
-delay.diversity <- cbind(group=delay.group, day=delay.day, abx=delay.abx, invsimp=delay.invsimp, shannon=delay.shannon)
-
-recov.diversity <- matrix(nrow=7)
-row.names(recov.diversity) <- c(-11, -5:0)
-
-amp.div <- matrix(nrow=7, ncol=2)
-row.names(amp.div) <- c(-11, -5:0)
-colnames(amp.div) <- c("amp.avg", "amp.se")
-library(plotrix)
-for(j in 1:7){
-  matches <- delay.diversity[,3]=="amp" & delay.diversity[,2]==row.names(amp.div)[j]
-  amp.div[j, "amp.avg"] <- mean(as.numeric(delay.diversity[matches, "invsimp"]))
-  amp.div[j, "amp.se"] <- std.error(as.numeric(delay.diversity[matches, "invsimp"]))
 }
 
-metro.div <- matrix(nrow=7, ncol=2)
-row.names(metro.div) <- c(-11, -5:0)
-colnames(metro.div) <- c("metro.avg", "metro.se")
-for(j in 1:7){
-  matches <- delay.diversity[,3]=="metro" & delay.diversity[,2]==row.names(metro.div)[j]
-  metro.div[j, "metro.avg"] <- mean(as.numeric(delay.diversity[matches, "invsimp"]))
-  metro.div[j, "metro.se"] <- std.error(as.numeric(delay.diversity[matches, "invsimp"]))
-}
-
-recov.diversity <- cbind(amp.div, metro.div)
-
-# Plot for second part of supplemental figure 2
-transpose <-t(recov.diversity)
-colors <- c(amp="#6600FF",
-            metro="#FF00DBFF")
-par(mar=c(5, 5, 0.5, 0.5), las = 2, mgp=c(3, 1, 0))
-bp <- barplot(transpose[c("amp.avg", "metro.avg"), ], beside=T, ylim=c(0, 15),
-              ylab="Inverse Simpson Diversity", xlab="",
-              col=colors, xaxt='n' )
-xticks <- apply(bp, 2, mean)
-axis(1, at=xticks, label=c(-11, -5:0), las=1)
-mtext(text = expression(paste("Days Pre-", italic("C. difficile"), " Challenge")), side = 1, line = 3, cex = 1.5, las=0)
-abline(v=mean(xticks[1:2]), lty="dashed", lwd=3)
-box()
-
-x <- c(bp[1,], bp[2,])
-y <- as.numeric(c(transpose[1,], transpose[3,]))
-se <- as.numeric(c(transpose[2,], transpose[4,]))
-d <- as.data.frame( cbind(x=x, y=y, se=se  ) )
-
-library(Hmisc)
-with (
-  data = d
-  , expr = errbar(x, y, y+se, y-se, add=T, pch=".", cap=.01)
-)
-
-legend("topright",inset=0.08, legend=c("Ampicillin", "Metronidazole"),
-       col=colors, pch=15, bty="n", y.intersp=1.5, pt.cex = 2.5, cex=1.2)
-
-```
-
-> ### Statistical comparison of delayed and on time CFU results
-
-```{r, eval=TRUE}
-file<-read.csv("~/Documents/Github/abxD01/Figure4/abxD01.final.tx2.subsample.alldelay.relabund.csv", header=T)
-file<-file[file$expgroup!="1untr",]
-file$nextDayCFU <- log10(file$nextDayCFU+1)
-
-metro <- file[file$abx=="Metronidazole", 3:4]
-amp <- file[file$abx=="Ampicillin", 3:4]
-
-diffs <- function(data){
-  data.aov<-aov(data$nextDayCFU ~ data$expgroup)
-  print(anova(data.aov))
-  pairwise.t.test(data$nextDayCFU, data$expgroup, p.adjust="BH")
-}
-
-diffs(metro)
-diffs(amp)
-
-```
-
-> ### Figure 4: Barchart
->
-> This figure shows the difference between the Day 0's in the delayed and original treatments for metronidazole and ampicillin.
->
-> Input files on github folder Figure 5:
-> 1. abxD01.final.tx.2.subsample.alldelay.forlogscale.csv
-> 2. alldelay_tx2_barchart_ids.csv
->
-> Uses functions in R scripts folder: stacked.barcharts.r, findStatLetters.r, barplotBeside.r
-
-```{r, eval=FALSE}
-
-stackedbarcharts(file = "~/Documents/Github/abxD01/Figure4/abxD01.final.tx.2.subsample.alldelay.forlogscale.csv", fileIDS = "~/Documents/Github/abxD01/Figure4/alldelay_tx2_barchart_ids.csv", divide = TRUE)
-
-```
-
-***The composition of the microbiota on day 0 is predictive of C.
-difficile colonization levels on day 1.*** Given our ability to make qualitative predictions of *C.
-difficile* colonization based on the microbiota, we next designed a statistical model to calculate expected colonization levels given the relative abundance of a subset of OTUs.
-We used a regression-based random forest machine learning algorithm to predict the level of *C.
+***The composition of the microbiota on day 0 is predictive of *C.
+difficile* colonization levels on day 1.*** The three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C.
+difficile* colonization, but in other contexts they were not.
+For example, the OTU affiliated with *Escherichia* (OTU 4) was associated with high levels of colonization in the ampicillin, metronidazole, and cefoperazone-treated mice; however, it was not detected in the streptomycin-treated mice although *C.
+difficile was able to colonize those mice to high levels.
+This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization.
+Therefore, we used a regression-based random forest machine learning algorithm to predict the level of *C.
 difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge.
@@
 difficile* colonization levels (Figure 5).
+
 Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia*.
@@
 Error bars show the standard error of the mean.
+
+**Figure 4.
+Changes in the average relative abundance for family level phylotypes durin the period of recovery prior to challenge with *C.
+difficile*.** Each row depicts a different antibiotic that was allowed to recover.
+Black bars represent untreated animals.
+Grey bars represent animals given 1 day of recovery off antibiotics as in Figure 1.
+Error bars indicate the standard error of the mean.
+
+**Supplemental Figure 2.
+Changes between untreated and recovery days.** A) The average thetaYC distance between pretreatment (day -11) and recovery days (days -5 through -1) is represented by circles connected by solid lines.
+Dashed lines with the same symbol indicate individual mice from the same cage.
+Antibiotic treatment is indicated by color.
+B) The average level of the inverse simpson diversity index is plotted by antibiotic treatment across time.
+Error bars show standard error of the mean.
+Mice were challenged with *C.
+difficile* on day 0.

### 2015-04-04 code/build_figure5_6.R
Adjust page size and pch size

@@ -132,7 +132,7 @@ dev.off()
 
 
 # let's build Figure 6
-cairo_pdf(file="results/figures/figure6.pdf", width=7.5, height=7.5)
+cairo_pdf(file="results/figures/figure6.pdf", width=7.5, height=9.0)
 
     #want to jitter the relative abundance for those mice that had no Cdiff
     #colonization
@@ -164,7 +164,7 @@ cairo_pdf(file="results/figures/figure6.pdf", width=7.5, height=7.5)
         #can't plot zeroes on a log scale
         plot(otu_abund,logCFU, log="x",
             pch=19,
-            cex=0.5,
+            cex=0.8,
             ylab="", xlab="",
             xlim=c(1e-2, 100), ylim=c(0,9),
             yaxt="n", xaxt="n"

### 2015-04-04 code/build_figure5_6.R
Adjust location of 'B' in Fig. 5

@@ -126,7 +126,7 @@ cairo_pdf(file="results/figures/figure5.pdf", width=7.5, height=3.5)
     box()
     mtext(side=1, text="Observed colonization (log CFU)", line=2.0)
     mtext(side=2, text="Predicted colonization (log CFU)", line=2.5)
-    mtext(text="B", line=2.5, side=2, at=10, las=2, cex=2, font=2)
+    mtext(text="B", line=2.5, side=2, at=10.25, las=2, cex=2, font=2)
 dev.off()
 
 

### 2015-04-04 code/build_figure5_6.R
Add alternative RF plots; output RF data

@@ -79,6 +79,8 @@ n_trees <- 1e4
 set.seed("6201976")
 rf_full <- randomForest(logCFU ~ ., data=abund_good, importance=TRUE, ntree=n_trees)
 
+write(c(ncol(abund_good), rf_full$rsq[n_trees]), file="data/process/random_forest.data")
+
 # sort the features by their effect on % increase in the standard error
 importance_sorted <- sort(importance(rf_full)[,1], decreasing=T)
 
@@ -99,6 +101,7 @@ names(tax_otu_imp_labels) <- names(taxonomy[names(importance_subset)])
 # let's get Rsq for the subset
 set.seed("6201976")
 rf_partial <- randomForest(logCFU ~ ., data=abund_good[,names(importance_sorted)[1:n_features]], importance=TRUE, ntree=n_trees)
+write(c(n_features, rf_partial$rsq[n_trees]), file="data/process/random_forest.data", append=)
 
 
 # let's build Figure 5
@@ -200,7 +203,161 @@ cairo_pdf(file="results/figures/figure6.pdf", width=7.5, height=9.0)
 
 dev.off()
 
-#To do:
-# * supplemental figure: full feature importance plot
-# * supplemental figure: color fit by treatment group
-# * output model fits to put information into Rmd file
+
+# supplemental figure 5A: full feature importance plot
+cairo_pdf(file="results/figures/figure5A_full.pdf", width=3.5, height=5.0)
+
+    par(mar=c(3,8,0.5,0.5))
+    plot(NA, yaxt="n", xlab="", ylab="", xlim=c(min(importance_sorted), max(importance_sorted)),
+        ylim=c(1, length(importance_sorted)), axes=F)
+
+    abline(h=1:length(importance_sorted), lty=3, col="gray")
+    points(x=rev(importance_sorted), y=1:length(importance_sorted), pch=19, cex=0.8)
+    axis(1, at=seq(0,100,25), label=c("0", "", "50", "", "100"), cex=0.8)
+    box()
+    mtext(side=2, line=7.5, adj=0, at=1:length(importance_sorted), text=rev(tax_otu_labels[names(importance_sorted)]), las=2, cex=0.6)
[…]

### 2015-04-04 code/build_figure5_6.R
Add color & pch'd version of Fig. 6

@@ -361,3 +361,76 @@ cairo_pdf(file="results/figures/figure5B_full.pdf", width=5.0, height=6.0)
     par(mar=rep(0.1,4))
     text(x=0.5,y=0.2, label="Predicted colonization (log CFU)", cex=1.2)
 dev.off()
+
+
+
+
+# let's build Figure 6 (w/ color & pch)
+cairo_pdf(file="results/figures/figure6_col.pdf", width=7.5, height=9.0)
+
+    #want to jitter the relative abundance for those mice that had no Cdiff
+    #colonization
+    cd_zeroes <- logCFU == 0
+    logCFU[cd_zeroes] <- runif(sum(cd_zeroes),0,1)
+
+    par(mar=c(0.5,0.5,0.5,0.5))
+
+    design <- matrix(1:n_features, nrow=4, byrow=T)
+    design <- cbind(c(rep(13,4)), design)
+    design <- rbind(design, c(0,14,14,14))
+    layout(design, widths=c(0.3,1,1,1), heights=c(1,1,1,1,0.3))
+
+    for(i in 1:n_features){
+        #get the row and column number for each spot in the layout
+        row <- ceiling(i/3)
+        column <- ((i-1) %% 3) + 1
+
+        #extract the relative abundance data for this OTU
+        otu_abund <- rel_abund[,names(importance_subset)[i]]
+
+        #want to jitter the number of tumors for those mice that had a zero
+        #relative abundance
+        ra_zeroes <- otu_abund == 0
+        otu_abund[ra_zeroes] <- runif(sum(ra_zeroes),1.0e-2,1.5e-2)
+
+        #plot the relative abundance with the number of tumors for each animal. plot
+        #on consistent log scaled x-axis for all OTUs. will throw errors because it
+        #can't plot zeroes on a log scale
+        plot(otu_abund,logCFU, log="x",
+            pch=pch[grouping],
+            col=clrs[grouping],
+            cex=0.8,
+            ylab="", xlab="",
+            xlim=c(1e-2, 100), ylim=c(0,9),
+            yaxt="n", xaxt="n"
+        )
+
+        #create a vertical line to denote the limit of detection
+        abline(v=2.2e-2, col="gray")
+
+        #create a horizontal line to denote the limit of detection
+        abline(h=1.5, col="gray")
+
+        #put the OTU label in the upper left corner of the plot
[…]

### 2015-04-04 PAPER.Rmd
Add flag to end of delay paragraph
Remove package loading code
Remove excess code

@@
 opts_chunk$set(warning = FALSE)
-
-load_package <- function(package){
-    if(!(package %in% rownames(installed.packages()))){
-        install.packages(package)
-    }
-    library(package, quietly=TRUE)
-}
-
-load_package("randomForest")
-load_package("png")
 
@@
 Again, we observed a similar increase and decrease in relative abundance for the *Porphyromonadaceae* (OTU 3) and *Escherichia* (OTU 4) OTUs.
+<Need a conclusion sentence here>
 
@@
 </needs help>
-
---------
-
-![**Figure 5AB.**](results/figures/Figure5/figure5ab.png)
-
-![**Figure 5C.**](results/figures/Figure5/figure5c.png)
-
-**Figure 5.
-Random forest model to predict *C.
-difficile* colonization levels based on the microbiota.** (A) The fit of the random forest model by experimental group.
-The dashed line represents the y=x line.
-(B) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model.
-(C) The percent relative abundance of the top 8 OTUs from (B) on day 0 are plotted against the log<sub>10</sub> *C.
-difficile* CFU/g feces.
-
---------
-
-> ### Evaluating cutoffs for OTU inclusion in Random Forest model
-
-```{r, eval=TRUE}
-
-dataFile <- "~/Documents/Github/abxD01/results/figures/Figure5/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.highvancD0.csv"
-
-data<-read.csv(file=dataFile, header=T)
-
-means <- as.data.frame(matrix(nrow=nlevels(data$expgroup), ncol=0))
-row.names(means) <- levels(data$expgroup)
-for(i in 3:length(data)){
-
-  df <- aggregate( data[,i] ~ data$expgroup, data=data, mean )
-  names(df)[2] <- names(data)[i]
-
-  means <- as.data.frame(cbind( means, df[,2]))
-  names(means)[dim(means)[2]] <- names(data)[i]
-}
-
-means <- rbind(means, max=-1)
-for(j in 1:length(means)){
-  means["max", j] <- max(means[,j])
-}
-
-means <- rbind(means, relabund=-1)
-for(k in 1:length(means)){
-  means["relabund", k] <- (means[17,k])/1625
-}
-
-means <- rbind(means, percent=-1)
-for(k in 1:length(means)){
-  means["percent", k] <- (means[18,k])*100
-}
-
-meanOTU <- means[,-1] #remove the nextDayCFU column
-
-set <- NULL
-set <- seq(0,3, by=0.01)
-numOTUxThreshold <- NULL
-numOTUxThreshold <- as.data.frame(cbind(numOTUxThreshold, set))
-names(numOTUxThreshold)[1] <- "threshold"
-numOTUxThreshold <- cbind(numOTUxThreshold, numOTU=NA)
-for( i in 1:(length(set)) ) {
-
-  test <- meanOTU[ , meanOTU[19,] >= set[i] ]
-  #test2 <- sum( meanOTU[19,] >= set[i] )
-
-  numOTUxThreshold[i, 2] <- length(test)
-
-}
-
-plot(numOTUxThreshold$numOTU ~ numOTUxThreshold$threshold, xlab="Avg % RelAbund Threshold", ylab="# OTUs included")
-abline(v=seq(0,3, by=.1), col="light gray")
-abline(h=seq(1,300, by=10), col="light gray")
-```
-
-> This feeds into the next analysis section.
->
-> ### Random forest results
-
-```{r, eval=TRUE}
-
-otusAtPercAbund <- function(percThresh, meanOTU){
-  perc <- meanOTU[ , meanOTU[19,] >= percThresh ]
-  otus <- names(perc)[-(length(perc))]
-  return(otus)
-}
-
-perc<-1
-otus <- otusAtPercAbund(perc, meanOTU)
-
-all <- read.csv(file=dataFile, header=T)
-cols<- NULL
-for(i in 1:(length(otus))){
-  cols <-c(cols, which(names(all)==otus[i]))
-}
-cols <- c(which(names(all)=="Group"), which(names(all)=="nextDayCFU"), cols)
-
-toptitdel <- all[,cols]
-library(randomForest)
-
-toptitdel.rf <- randomForest(nextDayCFU ~ .,
-                           data = toptitdel[,-1],  outscale=TRUE,
-                           importance=TRUE, proximity=TRUE,
-                           keep.forest=TRUE, ntree=5000
-)
-print(toptitdel.rf)
-plot(toptitdel.rf)
-varImpPlot(toptitdel.rf, type=1)
-imp<-importance(toptitdel.rf)
-imp1<-imp[order(imp[,1], decreasing = TRUE),]
-topimp <- imp1[1:15,]
-write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)
-
-```
-
-> ### Figure 5A: Evaluating and plotting the fit of the model
-
-```{r, eval=FALSE}
-predictions <- predict(toptitdel.rf, newdata=toptitdel[,-c(1,2)])
-actual <- toptitdel$nextDayCFU
-expgroup<- as.character(all[,2])
-results<-as.data.frame(cbind(actual, predictions))
-colnames(results) <- c( "actual", "predict")
-
-ybar <- apply(results, 2, mean)
-num<-sum((results$actual-ybar["actual"])*(results$predict-ybar["predict"]))
-denA <- sum((results$actual-ybar["actual"])^2)
-denB <- sum((results$predict-ybar["predict"])^2)
-rsq <- (num^2)/(denA*denB) #calculated from the square of the sample correlation coefficient, little r squared as opposed to big R squared
-
-fit.results <- cbind(expgroup, results)
-
-#Plot observed vs predicted
-#information for plotting
-color<-rainbow(7)
-colors <- c(control = "black",
-          cipro = color[1],
-          clinda = "#FFCC00",
-          vanc.625 = "#00CC00",
-          vanc.3 = "#00CC00",
-          vanc.1 = "#00CC00",
-          strep5 = "#33FFFF",
-          strep.5 = "#33FFFF",
-          strep.1 = "#33FFFF",
-          cef.5 = color[5],
-          cef.3 = color[5],
-          cef.1 = color[5],
-          amp.5 = "#6600FF",
-          amp.5d = "#6600FF",
-          metro1 = color[7],
-          metro1d = color[7])
-
-# topdose is all circles, middle dose is square,
-# low dose is upward pointing triangle, delayed is starburst
-pch <- c(control = 1,
-         cipro = 16,
-         clinda = 16,
-         vanc.625 = 16,
-         vanc.3 = 15,
-         vanc.1 = 17,
-         strep5 = 16,
-         strep.5 = 15,
-         strep.1 = 17,
-         cef.5 = 16,
-         cef.3 = 15,
-         cef.1 = 17,
-         amp.5 = 16,
-         amp.5d = 8,
-         metro1 = 16,
-         metro1d = 8)
-
-#these are needed for the legend
-legend.labels <- c(control = "Control",
-            cipro = "Ciprofloxacin",
-            clinda = "Clindamycin",
-            vanc.625 = "Vancomycin 0.625 mg/ml",
-            vanc.3 = "Vancomycin 0.3 mg/ml",
-            vanc.1 = "Vancomycin 0.1 mg/ml",
-            strep5 = "Streptomycin 5 mg/ml",
-            strep.5 = "Streptomycin 0.5 mg/ml",
-            strep.1 = "Streptomycin 0.1 mg/ml",
-            cef.5 = "Cefoperazone 0.5 mg/ml",
-            cef.3 = "Cefoperazone 0.3 mg/ml",
-            cef.1 = "Cefoperazone 0.1 mg/ml",
-            amp.5 = "Ampicillin",
-            amp.5d = "Ampicillin +5 days",
-            metro1 = "Metronidazole",
-            metro1d = "Metronidazole +5 days")
-
-design<-as.matrix(1:2)
-design<-cbind(design, c(3,3))
-layout(design, widths=c(1,1), heights=c(.35, 1))
-
-plot.new()
-par(mar=c(0.5, 0.5, 0.5, 0.5))
-legend("bottomright", legend=legend.labels, pch=pch, col=colors, pt.cex=1.5,  cex=1, bty="n", ncol = 2, text.width=0.4)
-
-par(mar=c(5, 5, 0.5, 0.5) +0.1, mgp=c(3, 1, 0), las=1)
-plot(fit.results[fit.results[,1]=="control",2],
-     fit.results[fit.results[,1]=="control",3],
-     main="",
-     ylab=expression(paste("Predicted Log ", italic("C. difficile"), " CFU/g Feces")),
-     xlab=expression(paste("Actual Log ", italic("C. difficile"), " CFU/g Feces")),
-     ylim=c(0,9), xlim=c(0,9), xaxt='n', yaxt='n', cex=1.5)
-abline(a=0, b=1, lty="dashed", lwd=2, col="black")
-axis(1, at = c(0:9), labels=c(0:9))
-axis(2, at = c(0:9), labels=c(0:9))
-
-for(i in 2:(length(legend.labels))){
-  points(fit.results[fit.results[,1]==names(legend.labels)[i],2],
-         fit.results[fit.results[,1]==names(legend.labels)[i],3],
-         col=colors[i], pch=pch[i], cex=1.5)
-}
-
-```
-
-> ### Figure 5B: Importance Plot
-
-```{r, eval=FALSE}
-topimp<-topimp[order(topimp[,1], decreasing=FALSE),]
-ids <- read.csv(file = "~/Documents/Github/abxD01/Figure5/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
-#par(mfrow=c(1, 1)) #+1 to give extra labeling space
-par(mar=c(5, 15, 0.5, 1) +0.1, mgp=c(3, 1, 0), las=0) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
-plot(topimp[,1], 1:15, xlab="% Increase in Mean Squared Error", yaxt="n", ylab="", pch=16, cex=1.5, xlim=c(20, 80))
-axis(2, at = c(1:15), labels = rep("", 15), las=1, cex.axis=.9 )
-for(z in 1:(dim(ids)[1])){
-  name<-as.character(ids$name[z])
-  otuname<- as.character(ids$otuname[z])
-  mtext(text = bquote(italic(.(name))*" ("*.(otuname)*")"), at=z, side = 2, line = 1, las=2, cex=0.9)
-}
-abline(h=c(1:15), lty="dashed", col='black')
-
-```
-
-> ### Testing a random forest model with only OTUs 3 and 39
-
-```{r, eval=TRUE}
-test3.39 <- toptitdel[,c("Group","nextDayCFU","Otu00003", "Otu00039")]
-
-rf.test3.39 <- randomForest(nextDayCFU ~ .,
-                           data = test3.39[,-1],  outscale=TRUE,
-                           importance=TRUE, proximity=TRUE,
-                           keep.forest=TRUE, ntree=5000
-)
-print(rf.test3.39)
-plot(rf.test3.39)
-
-predictions <- predict(rf.test3.39, newdata=test3.39[,-c(1,2)])
-actual <- test3.39$nextDayCFU
-test.results<-as.data.frame(cbind(actual, predictions))
-colnames(test.results) <- c( "actual", "predict")
-
-ybar <- apply(test.results, 2, mean)
-num<-sum((test.results$actual-ybar["actual"])*(test.results$predict-ybar["predict"]))
-denA <- sum((test.results$actual-ybar["actual"])^2)
-denB <- sum((test.results$predict-ybar["predict"])^2)
-test.rsq <- (num^2)/(denA*denB)
-
-```
-
-> ### Correlation over all three datasets with *C. difficile* levels
-
-```{r, eval=FALSE}
-
-#First the correlation calculations that were put on the graph:
-c<-1
-otu <- c()
-cor.spear <- c()
-pval.spear <- c()
-cor.ken <- c()
-pval.ken <- c()
-for(i in 3:length(toptitdel)){
-  otu[c] <- colnames(toptitdel[i])
-  cor.spear[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="spearman")$estimate
-  pval.spear[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="spearman")$p.value
-  cor.ken[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="kendall")$estimate #good to see because kendall handles ties
-  pval.ken[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
-  c <- c+1
-}
-pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
-pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons
-
-correl = NULL
-correl <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
-colnames(correl) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
-correl <- correl[order(correl[,3]),]  #order by pvalue column=3
-#write.table(correl, file="~/Desktop/mothur/abxD01/correlation/toptidel.1p.correl.txt", sep="\t", row.names=FALSE)
-print(correl)
-```
-
-> From this file I selected the top 8 OTUs identified by random forest's increase in mean squared error measurement to graph (next section).
->
-> ### Figure 5C: OTU abundance vs. *C. difficile* abundance
-
-```{r, eval=FALSE}
-topimp.otus <-  row.names(topimp)[c(15:7)]
-temp.match <- match(topimp.otus, ids.master[,1])
-graphID <- ids.master[temp.match,c(4,3)]
-
-temp.match <- match(topimp.otus, correl[,"otu"])
-topimp.corrs <- correl[temp.match, c("corSpear", "pvalSpear")]
-row.names(topimp.corrs) <- topimp.otus
-
-graphOTUxCD(otus = topimp.otus, ids = graphID, corrs = topimp.corrs)
-```
 
@@
 
+**Figure 5.
+Random forest model to predict *C.
+difficile* colonization levels based on the microbiota.** (A) The fit of the random forest model by experimental group.
+The dashed line represents the y=x line.
+(B) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model.
+(C) The percent relative abundance of the top 8 OTUs from (B) on day 0 are plotted against the log<sub>10</sub> *C.
+difficile* CFU/g feces.
+
 **Supplemental Figure 2.

### 2015-04-04 code/build_figure1.R
Rename barplot function

@@ -78,13 +78,14 @@ u_qtr <- u_qtr[,-1]
 rownames(u_qtr) <- drugs
 u_qtr <- t(u_qtr)
 
-make_barplot <- function(drug, label){
-    pos <- barplot(as.vector(med_ra[,drug]), ylim=c(0,100), axes=F)
+fulldose_phylum_barplot <- function(drug, label){
+    pos <- barplot(as.vector(med_ra[,drug]), ylim=c(0,105), axes=F)
     arrows(x0=pos, x1=pos, y0=med_ra[,drug], y1=u_qtr[,drug], angle=90, length=0.1)
     arrows(x0=pos, x1=pos, y0=med_ra[,drug], y1=l_qtr[,drug], angle=90, length=0.1)
     axis(2, at=seq(0,100,25), label=seq(0,100,25), las=1)
     box()
     text(x=4.9, y=70, label=paste(label, "\n", abx_cfu[drug] ,"CFU/g"),adj = c(1,0))
+    pos
 }
 
 pdf(height=9, width=3.75, file="results/figures/figure1.pdf")
@@ -103,14 +104,14 @@ pdf(height=9, width=3.75, file="results/figures/figure1.pdf")
 
     par(mar=c(0.5, 0.5, 1.0, 0.5))
 
-    make_barplot("control", "No antibiotics")
-    make_barplot("cipro", "Ciprofloxacin")
-    make_barplot("vanc", "Vancomycin")
-    make_barplot("amp", "Ampicillin")
-    make_barplot("clinda", "Clindamycin")
-    make_barplot("strep", "Streptomycin")
-    make_barplot("cef", "Cefoperazone")
-    make_barplot("metro", "Metronidazol")
+    pos <- fulldose_phylum_barplot("control", "No antibiotics")
+    pos <- fulldose_phylum_barplot("cipro", "Ciprofloxacin")
+    pos <- fulldose_phylum_barplot("vanc", "Vancomycin")
+    pos <- fulldose_phylum_barplot("amp", "Ampicillin")
+    pos <- fulldose_phylum_barplot("clinda", "Clindamycin")
+    pos <- fulldose_phylum_barplot("strep", "Streptomycin")
+    pos <- fulldose_phylum_barplot("cef", "Cefoperazone")
+    pos <- fulldose_phylum_barplot("metro", "Metronidazol")
 
 
     plot.new()

### 2015-04-04 code/build_figure3.R
Change family name to genus names

@@ -96,15 +96,15 @@ sig_otus <- sort(unique(c(strep_sig_otus, cef_sig_otus, vanc_sig_otus)))
 x_max <- length(sig_otus)*4-0.5
 
 # let's get the taxonomy data so that we have the string from the kingdom to
-# the family level name or whatever the next level up is that provided a robust
+# the genus level name or whatever the next level up is that provided a robust
 # classification.
 taxonomy_file <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy", header=T, row.names=1)
 taxonomy <- taxonomy_file$Taxonomy
 names(taxonomy) <- rownames(taxonomy_file)
 taxonomy <- gsub("\\(\\d*\\)", "", taxonomy)
-taxonomy <- gsub(";[^;]*;$", "", taxonomy)
 taxonomy <- gsub(";unclassified", "", taxonomy)
-taxonomy <- gsub("_1", "", taxonomy)
+taxonomy <- gsub("/.*", "", taxonomy)
+taxonomy <- gsub(";$", "", taxonomy)
 taxonomy <- gsub(".*;", "", taxonomy)
 taxonomy <- taxonomy[sig_otus]
 

### 2015-04-04 code/build_figure4.R

@@ -81,15 +81,15 @@ x_max <- length(sig_otus)*3-0.5
 
 
 # let's get the taxonomy data so that we have the string from the kingdom to
-# the family level name or whatever the next level up is that provided a robust
+# the genus level name or whatever the next level up is that provided a robust
 # classification.
 taxonomy_file <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy", header=T, row.names=1)
 taxonomy <- taxonomy_file$Taxonomy
 names(taxonomy) <- rownames(taxonomy_file)
 taxonomy <- gsub("\\(\\d*\\)", "", taxonomy)
-taxonomy <- gsub(";[^;]*;$", "", taxonomy)
 taxonomy <- gsub(";unclassified", "", taxonomy)
-taxonomy <- gsub("_1", "", taxonomy)
+taxonomy <- gsub("/.*", "", taxonomy)
+taxonomy <- gsub(";$", "", taxonomy)
 taxonomy <- gsub(".*;", "", taxonomy)
 taxonomy <- taxonomy[sig_otus]
 

### 2015-04-04 PAPER.Rmd
Add rf code to text

@@
 ```{r random_forest}
+data <- read.table(file="data/process/random_forest.data")
+n_full_features <- data[1,1]
+rsq_full_features <- data[1,2]
 
-}
+n_part_features <- data[2,1]
+rsq_part_features <- data[2,2]
+```
 
@@
 difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge.
-The model explained XXX98.0XXX% of the variation in the observed *C.
+The model explained `r round(rsq_full_features*100, 1)`% of the variation in the observed *C.
 difficile* colonization levels (Figure 5).
+Noticing a natural jump following the twelfth OTU in the percent increase in the mean squared error when each OTU was removed, we regenerated the model using only the top 12 OTUs.
+This model explained `r round(rsq_part_features*100, 1)`% of the variation in the observed *C.
+difficile* colonization levels.
 
@@
 For example, among the streptomycin-treated mice which altogether lacked the *Enterobacteriaceae*, *Porphyromonadaceae* (OTU 3) and *Lachnospiraceae* (OTU 39), were the two most important features in accurately predicting subsequent *C.
-difficile* levels.
-In fact, a model built with only these two features was able to account for 79% of the variance observed.
-</needs help>
+difficile* levels</needs help>
 

### 2015-04-04 code/build_figure5_6.R
Add append=T to second write statment

@@ -101,7 +101,7 @@ names(tax_otu_imp_labels) <- names(taxonomy[names(importance_subset)])
 # let's get Rsq for the subset
 set.seed("6201976")
 rf_partial <- randomForest(logCFU ~ ., data=abund_good[,names(importance_sorted)[1:n_features]], importance=TRUE, ntree=n_trees)
-write(c(n_features, rf_partial$rsq[n_trees]), file="data/process/random_forest.data", append=)
+write(c(n_features, rf_partial$rsq[n_trees]), file="data/process/random_forest.data", append=T)
 
 
 # let's build Figure 5

### 2015-04-06 Schubert_abxD01_XXXXXX_2015.Rmd
Rename PAPER.Rmd file

--- 2015-04-04 PAPER.Rmd
@@
 
+```{r error}
+error_table <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.error.summary", header=T)
+
+error_nochim <- error[error$numparents==1,]
+
+error_rate <- 100 * sum(error$weight * error$error)/sum(error$weight)
+```
+
 ***Sequence curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
@@
 In parallel to the fecal samples, we also sequeced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
-Analysis of the mock community data indicated that the error rate following our cuation procedure was 0.0X%.
+Analysis of the mock community data indicated that the error rate following our cuation procedure was `r round(error_rate, 2)`%.
 In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples [Schloss, reducing effects].
@@
 Comparison of bacterial levels among titration groups or delayed groups of the same antibiotic was performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests with a Benjamini and Hochberg adjustment.
-Comparison of log10 transformed *C.
+Comparison of log (base 10) transformed *C.
 difficile* CFU/g feces between experimental groups was calculated using ANOVA and subsequently pairwise T tests with a Benjamini and Hochberg adjustment.

### 2015-04-07 code/make_files_file.R
Limit control mice to those from cages that received top dose

@@ -22,8 +22,13 @@ cdiff_challenged <- metadata[metadata$cdiff==TRUE,]
 #get the day 0 samples from the abx treated mice
 abx_treated <- cdiff_challenged[cdiff_challenged$day == 0, ]
 
-#get the pre-abx samples from mice that would receive abx
-untreated <-  cdiff_challenged[cdiff_challenged$preAbx & cdiff_challenged$abx != "none", ]
+#get the pre-abx samples from mice that would receive abx and treat as the
+#control group
+
+control_groups <- c(21, 24, 50, 600, 7, 39, 40, 103, 47, 627, 2, 4, 15, 52, 97, 98)
+
+untreated <-  cdiff_challenged[cdiff_challenged$preAbx &
+                                cdiff_challenged$group %in% control_groups, ]
 
 #concatenate the sample names together
 samples <- c(rownames(abx_treated), rownames(untreated))

### 2015-04-08 code/build_figure4.R
Filter to be 1% per treatment grouping

@@ -22,25 +22,26 @@
 
 # read in the metadata file
 counts_file <- read.table(file="data/process/abxD1.counts", header=T)
-top_dose <- counts_file[counts_file$experiment=="top_dose" | counts_file$abx=="control",]
 
 
 # read in the shared file and get the relative abundance
 shared_file <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared", header=T, row.names=2)
 shared_file <- shared_file[,!(colnames(shared_file) %in% c("label", "numOtus"))]
 n_seqs <- apply(shared_file, 1, sum)[1]
-rel_abund <- shared_file/n_seqs
+rel_abund <- 100 * shared_file/n_seqs
 
 
 # need to figure out which samples made it through the pipeline and look at those
-overlap <- rownames(rel_abund)[which(rownames(rel_abund) %in% rownames(top_dose))]
+overlap <- rownames(rel_abund)[which(rownames(rel_abund) %in% rownames(counts_file))]
 rel_abund <- rel_abund[overlap,]
-top_dose <- top_dose[overlap,]
+counts_file <- counts_file[overlap,]
 
 
 # limit the analysis to those OTUs that have an average relative abundance over
 # 1% across all samples
-abund <- apply(rel_abund, 2, mean) > 0.001
+grouping <- paste(counts_file$abx, counts_file$dose, counts_file$experiment, sep="-")
+medians <- aggregate(rel_abund, by=list(grouping), median)[,-1]
+abund <- apply(medians, 2, max) > 1.0
 rel_abund_good <- rel_abund[,abund]
 
 
@@ -60,7 +61,7 @@ taxonomy <- taxonomy[colnames(rel_abund_good)]
 # calculate the Spearman correlation value and its P-value for each OTU against
 # the number of Cdiff in the feces.
 get_CFU_cor <- function(OTU){
-    cor.test(top_dose$CFU, rel_abund_good[,OTU], method="spearman")[c("estimate","p.value")]
+    cor.test(counts_file$CFU, rel_abund_good[,OTU], method="spearman")[c("estimate","p.value")]
 }
 
 otu_cfu_corrs <- sapply(1:ncol(rel_abund_good), get_CFU_cor)
@@ -129,4 +130,4 @@ dev.off()
 
 
 # output the correlation data
[…]

### 2015-04-08 code/build_figure4.R
Rename output file for correlation analysis in Rmd file

@@ -130,4 +130,4 @@ dev.off()
 
 
 # output the correlation data
-#write.table(file="data/process/top_dose_corr.tsv", cbind(taxonomy=gsub("\\d", "", sig_taxonomy), sig_corrs), quote=F, row.names=T, sep="\t")
+write.table(file="data/process/correlation_analysis.tsv", cbind(taxonomy=gsub("\\d", "", sig_taxonomy), sig_corrs), quote=F, row.names=T, sep="\t")

### 2015-04-08 Schubert_abxD01_mBio_2015.Rmd
Rename files to indicate mBio as journal
Update for new correlation file name

--- 2015-04-06 Schubert_abxD01_XXXXXX_2015.Rmd
@@
 
-error_nochim <- error[error$numparents==1,]
+error_nochim <- error_table[error_table$numparents==1,]
 
-error_rate <- 100 * sum(error$weight * error$error)/sum(error$weight)
+error_rate <- 100 * sum(error_nochim$weight * error_nochim$error)/sum(error_nochim$weight)
 ```
@@
 ### Results
-
-```{r top_dose_cfu_analysis}
-cfu <- read.table(file="data/process/abxD1.counts")
-cfu_top_dose <- cfu[cfu$experiment == "top_dose" & cfu$abx != "control",]
-cfu_top_dose_summary <-aggregate(cfu_top_dose$CFU, by=list(cfu_top_dose$abx), summary)
-treatments <- cfu_top_dose_summary$Group.1
-cfu_top_dose_summary <- format(cfu_top_dose_summary$x, scientific=T, digits=2)
-rownames(cfu_top_dose_summary) <- treatments
-```
 
@@
 ```{r top_dose_correlation}
-corr_data <- read.table(file="data/process/top_dose_corr.tsv", header=T)
+corr_data <- read.table(file="data/process/correlation_analysis.tsv", header=T)
 families <- gsub(".*;", "", corr_data$taxonomy)

### 2015-04-08 code/build_figure1.R
Make threshold 1% within each treatment group

@@ -54,11 +54,11 @@ top_dose <- top_dose[overlap,]
 
 
 # limit the analysis to those OTUs that have an median relative abundance over
-# 3% within each antibiotic dose
+# 1% within each antibiotic dose
 control_rabund <- rel_abund[top_dose$abx == "control",]
 control_metadata <- top_dose[top_dose$abx == "control",]
 control_median <- apply(control_rabund, 2, median)
-control_otus <- control_median > 3.0
+control_otus <- control_median > 1.0
 
 otu_hyp_test <- function(drug){
 

### 2015-04-08 code/build_figure2.R

@@ -46,7 +46,7 @@ cef <- rel_abund[titration$abx == "cef",]
 cef_metadata <- titration[titration$abx == "cef",]
 
 cef_med <- aggregate(cef, by=list(cef_metadata$dose), median)[,-1]
-cef_abund <- apply(cef_med, 2, max) > 3.0
+cef_abund <- apply(cef_med, 2, max) > 1.0
 
 cef_abund_good <- cef[,cef_abund]
 cef_p_value <- rep(NA, ncol(cef_abund_good))

### 2015-04-08 code/build_figure3.R

@@ -15,8 +15,6 @@
 #
 ################################################################################
 
-top_experiment_corr <- read.table(file="data/process/top_dose_corr.tsv", header=T)
-
 # read in the metadata file
 counts_file <- read.table(file="data/process/abxD1.counts", header=T)
 delay <- counts_file[counts_file$abx=="amp" | counts_file$abx=="metro",]
@@ -46,7 +44,7 @@ amp <- rel_abund[delay$abx == "amp",]
 amp_metadata <- delay[delay$abx == "amp",]
 
 amp_med <- aggregate(amp, by=list(amp_metadata$experiment), median)[,-1]
-amp_abund <- apply(amp_med, 2, max) > 3.0
+amp_abund <- apply(amp_med, 2, max) > 1.0
 
 amp_abund_good <- amp[,amp_abund]
 amp_p_value <- rep(NA, ncol(amp_abund_good))
@@ -93,12 +91,9 @@ taxonomy <- gsub(";$", "", taxonomy)
 taxonomy <- gsub(".*;", "", taxonomy)
 taxonomy <- taxonomy[sig_otus]
 
-corr <- round(top_experiment_corr[sig_otus,"sig_corrs"], digits=2)
-corr[is.na(corr)] <- "NS"
-
 otu <- gsub("Otu0*", "", names(taxonomy))
 
-label <- paste0(taxonomy, "\n(OTU ", otu, ", ρ=", corr, ")")
+label <- paste0(taxonomy, "\n(OTU ", otu, ")")
 label <- gsub("\\(.=NS\\)", "(NS)", label)
 
 cairo_pdf(file="results/figures/figure3.pdf", width=7.5, height=4.125)

### 2015-04-08 code/build_figure5_6_S4.R

@@ -88,7 +88,7 @@ importance_sorted <- sort(importance(rf_full)[,1], decreasing=T)
 counts_file$fit_full <- predict(rf_full, abund_good)
 
 # there appears to be a natural break after OTU 11 (the 12th feature)
-n_features <- 12
+n_features <- 9
 importance_subset <- importance_sorted[1:n_features]
 
 tax_otu_imp_labels <- paste0(taxonomy[names(importance_subset)],
@@ -277,7 +277,7 @@ cairo_pdf(file="results/figures/figure6.pdf", width=7.5, height=7.5)
     par(mar=c(0.5,0.5,0.5,0.5))
 
     design <- matrix(1:n_features, nrow=3, byrow=T)
-    design <- cbind(c(rep(10,4)), design)
+    design <- cbind(c(rep(10,3)), design)
     design <- rbind(design, c(0,11,11,11))
     layout(design, widths=c(0.3,1,1,1), heights=c(1,1,1,0.3))
 

### 2015-04-08 code/build_figure5_6_S4.R
Remove MSE values

@@ -92,7 +92,7 @@ n_features <- 9
 importance_subset <- importance_sorted[1:n_features]
 
 tax_otu_imp_labels <- paste0(taxonomy[names(importance_subset)],
-                        " (MSE: ", round(importance_subset, 1), "; ",
+                        " (",
                         otus[names(importance_subset)], ")")
 names(tax_otu_imp_labels) <- names(taxonomy[names(importance_subset)])
 

### 2015-04-08 code/build_figure1.R
Adjust x-axis labels

@@ -148,7 +148,7 @@ single_drug_bars <- function(drug, drug_sig_otus, drug_label){
 
 
 
-cairo_pdf(file="results/figures/figure1.pdf", width=3.5, height=10.0)
+cairo_pdf(file="results/figures/figure1.pdf", width=4.5, height=10.0)
     par(cex=1.2)
 
     layout(matrix(c(

### 2015-04-08 code/build_figure2.R

@@ -106,17 +106,17 @@ taxonomy <- gsub(";unclassified", "", taxonomy)
 taxonomy <- gsub("/.*", "", taxonomy)
 taxonomy <- gsub(";$", "", taxonomy)
 taxonomy <- gsub(".*;", "", taxonomy)
+taxonomy <- gsub("_.*", "", taxonomy)
 taxonomy <- taxonomy[sig_otus]
 
 otu <- gsub("Otu0*", "", names(taxonomy))
 
-label <- paste0(taxonomy, "\n(OTU ", otu, ")")
-label <- gsub("\\(.=NS\\)", "(NS)", label)
+label <- paste0(taxonomy, " (OTU ", otu, ")")
 
-cairo_pdf(file="results/figures/figure2.pdf", width=7.5, height=5.5)
+cairo_pdf(file="results/figures/figure2.pdf", width=7.5, height=5.75)
     par(cex=1.2)
 
-    layout(matrix(c(1,5,2,6,3,7,4,8), nrow=4, byrow=T), width=c(1,0.25), height=c(1,1,1,1))
+    layout(matrix(c(1,5,2,6,3,7,4,8), nrow=4, byrow=T), width=c(1,0.25), height=c(1,1,1,1.2))
 
     par(mar=c(0.5,5,1.5,0.5))
     sig_cef <- cef[,sig_otus]
@@ -158,6 +158,8 @@ cairo_pdf(file="results/figures/figure2.pdf", width=7.5, height=5.5)
     legend(x=x_max*0.85, y=60, legend=paste(levels(factor(strep_metadata$dose)), "mg/mL"), fill=c("black", "gray", "white"), bg="white")
 
 
+
+
     sig_vanc <- vanc[,sig_otus]
     vanc_med <- aggregate(sig_vanc, by=list(vanc_metadata$dose), median)[,-1]
     vanc_uci <- aggregate(sig_vanc, by=list(vanc_metadata$dose), function(x){quantile(x, prob=0.75)})[,-1]
@@ -176,7 +178,11 @@ cairo_pdf(file="results/figures/figure2.pdf", width=7.5, height=5.5)
 
     legend(x=x_max*0.83, y=60, legend=paste(levels(factor(vanc_metadata$dose)), "mg/mL"), fill=c("black", "gray", "white"), bg="white")
 
[…]

### 2015-04-08 code/build_figure3.R

@@ -93,13 +93,12 @@ taxonomy <- taxonomy[sig_otus]
 
 otu <- gsub("Otu0*", "", names(taxonomy))
 
-label <- paste0(taxonomy, "\n(OTU ", otu, ")")
-label <- gsub("\\(.=NS\\)", "(NS)", label)
+label <- paste0(taxonomy, " (OTU ", otu, ")")
 
-cairo_pdf(file="results/figures/figure3.pdf", width=7.5, height=4.125)
+cairo_pdf(file="results/figures/figure3.pdf", width=7.5, height=4.5)
     par(cex=1.2)
 
-    layout(matrix(c(1,4,2,5,3,6), nrow=3, byrow=T), width=c(1,0.25), height=c(1,1,1))
+    layout(matrix(c(1,4,2,5,3,6), nrow=3, byrow=T), width=c(1,0.25), height=c(1,1,1.1))
 
     par(mar=c(0.5,5,1.5,0.5))
 
@@ -140,7 +139,7 @@ cairo_pdf(file="results/figures/figure3.pdf", width=7.5, height=4.125)
 
 
 
-    text(x=apply(z, 2, mean)+0.5, y=par("usr")[1]-10, xpd=NA, label=label, pos=2, srt=70, cex=1.2)
+    text(x=apply(z, 2, mean)+0.75, y=par("usr")[1]-8, xpd=NA, label=label, pos=2, srt=70, cex=1)
 
     mtext(side=2, "Relative abundance (%)", line=3, at=68)
 

### 2015-04-09 Schubert_abxD01_mBio_2015.Rmd
Add Table 1 to manuscript

@@
 
+Table 1. Description of Antibiotics used in this study
+
+Antibiotic | Administration | Class | Mechanism | Target
+:---------:|:--------------:|:-----:|:---------:|:---------:
+Ciprofloxacin | Oral gavage, one time | Fluoroquinolone | Inhibits DNA gyrase | Gram +/-
+Clindamycin | IP injection, one time | Lincosamide | Inhibits protein synthesis | Anaerobes
+Vancomycin | *Ad libitum* in drinking water, five days | Glycopeptide | Inhibits peptidoglycan synthesis | Gram +
+Streptomycin | *Ad libitum* in drinking water, five days | Aminoglycoside | Inhibits protein synthesis | Gram +/-
+Cefoperazone | *Ad libitum* in drinking water, five days | β-lactam: Cephalosporin | Inhibits peptidoglycan synthesis | Gram +/-
+Ampicillin | *Ad libitum* in drinking water, five days | β-lactam: Penicillin | Inhibits peptidoglycan synthesis | Gram +/-
+Metronidazole | *Ad libitum* in drinking water, five days | Nitromidazole | Destabilizes bacterial DNA | Anaerobes
+
 **Figure Legends**

### 2015-04-09 code/build_figure1.R
Add N and dose information

@@ -100,11 +100,6 @@ o <- order(control_median[sig_otus], decreasing=T)
 rel_abund_sig <- rel_abund[,sig_otus[o]]
 
 
-#needed?
-#top_dose_corr <- read.table(file="data/process/top_dose_corr.tsv", header=T)
-#corr <- round(top_dose_corr[sig_otus,"sig_corrs"], digits=2)
-#corr[is.na(corr)] <- "NS"
-
 otu <- gsub("Otu0*", "OTU ", names(taxonomy))
 names(otu) <- names(taxonomy)
 
@@ -114,6 +109,8 @@ single_drug_bars <- function(drug, drug_sig_otus, drug_label){
 
     drug_rabund <- rel_abund_sig[top_dose$abx == drug,]
     drug_metadata <- top_dose[top_dose$abx == drug,]
+
+    n <- nrow(drug_rabund)
 
     drug_med <- apply(drug_rabund, 2, median)
     drug_uci <- apply(drug_rabund, 2, function(x){quantile(x, prob=0.75)})
@@ -138,7 +135,7 @@ single_drug_bars <- function(drug, drug_sig_otus, drug_label){
                             prob=c(0.25, 0.50, 0.75)), scientific=T, digits=2)
 
     summary_string <- paste0(summary_stats[2], " (",
-                        summary_stats[1], "-", summary_stats[3], ")")
+                        summary_stats[1], "-", summary_stats[3], "; N=", n, ")")
     summary_string <- gsub("e\\+0", "x10^", summary_string)
 
     text(x=par("usr")[2], y=1.05*par("usr")[4], labels=summary_string,
@@ -167,15 +164,15 @@ cairo_pdf(file="results/figures/figure1.pdf", width=4.5, height=10.0)
     par(mar=c(0.5,5,1.5,0.5))
 
     z <- single_drug_bars("control", "", "No antibiotics")
-    z <- single_drug_bars("amp", amp_sig_otus, "Ampicillin")
-    z <- single_drug_bars("cef", cef_sig_otus, "Cefoperazone")
-    z <- single_drug_bars("cipro", cipro_sig_otus, "Ciprofloxacin")
-    z <- single_drug_bars("clinda", clinda_sig_otus, "Clindamycin")
+    z <- single_drug_bars("amp", amp_sig_otus, "Ampicillin (0.5 mg/mL)")
[…]

### 2015-04-09 Schubert_abxD01_mBio_2015.Rmd
Incorporate revisions from docx file

@@
 ```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
-opts_chunk$set("dev" = c("png", "cairo_pdf"))
+opts_chunk$set("dev" = c("png", "pdf"))
 opts_chunk$set(results = "hold")
@@
 opts_chunk$set(warning = FALSE)
-
-set.seed(6201976)
 ```
 
-**Running title:** Antibiotic induced alterations of the murine gut microbiota and subsequent effects on colonization resistance against *Clostridium difficile*
+Antibiotic induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against *Clostridium
+difficile*
 
-**Short title:** Alterations of the murine gut microbiota lead to *C. difficile* colonization
+**Running title:** Colonization resistance against *C. difficile*
 
-Alyxandria M. Schubert, Hamide Sinani, and Patrick D.
-Schloss\*
+Alyxandria M. Schubert, Hamide Sinani, and Patrick D. Schloss\*
 
-\* Correspondence:
+Correspondence:
+
 > Department of Microbiology and Immunology
+> University of Michigan
+> 1520A Medical Science Research Bldg. I
+> 1500 W. Medical Center Dr.
+> Ann Arbor, MI 48109
+> pschloss@umich.edu
+> 734.647.5801
 
-> University of Michigan
+```{r random_forest}
+data <- read.table(file="data/process/random_forest.data")
+n_full_features <- data[1,1]
+rsq_full_features <- data[1,2] * 100
 
-> 1520A Medical Science Research Bldg. I
-
-> 1500 W. Medical Center Dr.
-
-> Ann Arbor, MI 48109
-
-> pschloss@umich.edu
-
-> 734.647.5801
+n_part_features <- data[2,1]
+rsq_part_features <- data[2,2] * 100
+```
 
@@
 
+Perturbations to the gut microbiota result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*.
+Although *C.
+difficile* infection is commonly associated with antibiotic use, the precise alterations to the microbiota associated with this loss in function are unknown.
+We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which we then challenged with *C.
+difficile* spores.
+Across these treatments we observed that *C.
+difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner.
+Using relative abundance data, we built machine learning regression models to predict the levels *C.
+difficile* that were found 24 hours after challenging the perturbed communities.
+This model was able to explain `r round(rsq_full_features, 1)`% of the variation in the observed number of *C.
+difficile* per gram of feces.
+This model revealed important bacterial populations within the microbiota, which correlation analysis alone did not detect.
+Specifically, we observed that populations associated with the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes* were protective and populations associated with *Escherichia* and *Streptococcus* were associated with high levels of colonization.
+In addition, a population affiliated with *Akkermansia* indicated a strong context dependency on other members of the microbiota.
+Together, these results indicate that individual bacterial populations do not drive colonization resistance to *C.
+difficile*.
+Rather, resistance can be achieved with multiple diverse assemblages.
+
+###  Importance
+
+The gastrointestinal tract harbors a complex community of bacteria, known as the microbiota, which plays an integral role in resisting the invasion by gut pathogens.
+This resistance has been shown to be crucial for protection against *C.
+difficile* infections (CDI), which are the leading source of hospital-acquired infections in the United States.
+Antibiotics are a major risk factor for acquiring CDI due to their effect on the normal structure of the indigenous gut microbiota.
+We showed that diverse antibiotic perturbations gave rise to altered communities that varied in their susceptibility to *C.
+difficile* colonization.
+Instead of being dependent on one specific population of bacteria, we found that multiple co-existing populations conferred resistance.
+By understanding the relationships between *C.
+difficile* and members of the microbiota it will be possible to better manage this important infection.
+
 ### Introduction
 
-The microbiome, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection [ref].
-This native protection by the microbiome from invasive pathogenic species is termed colonization resistance [ref].
-It is well established that the gut bacterial microbiome is critical in the hosts’ defense against the pathogen Clostridium difficile [ref].
+The microbiota, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection {Vollaard, 1994 \#683}.
+This native protection by the microbiota from invasive pathogenic species is termed colonization resistance.
+It is well established that the gut bacterial microbiota is critical in the hosts’ defense against the pathogen *Clostridium difficile* {Wilson, 1985 \#1045}{Chang, 2008 \#590}{Reeves, 2011 \#118}.
 When this indigenous community is perturbed, this often leads to a loss of resistance.
-This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, chemotherapy, but also exposed to the environmental *C.
-difficile* spores [ref].
-Among hospital-acquired infections in the United States, *C.
-difficile* infections (CDI) are the number one reported and are responsible for 14,000 deaths a year [ref].
-It is not completely understood how different perturbations to the taxonomic composition of the gut microbiota result in a loss of colonization resistance.
+This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, and chemotherapy, but they are also exposed to *C.
+difficile* spores from their environment {Bauer, 2015 \#1047}.
+*C.
+difficile* infections (CDI) are the most reported hospital-acquired infection in the United States and are responsible for 14,000 deaths a year {Lessa, 2015 \#950}.
 
-In understanding colonization resistance, there is a clear need for more knowledge on *C.
-difficile* and its interactions between members of the bacterial community.
-Fecal microbiota transplants, which have been shown to increase *Bacteroidetes* and decrease *Proteobacteria* levels in recipients, have successfully restored colonization resistance in patients and effectively resolved their severe CDI [Seekatz et al.
-2014].
-Precisely how this occurs is not fully understood, but it emphasizes the strength of the community as a whole in its protective ability.
-Individual probiotics have not worked with complete success [dendukuri 2005].
-In fact, several studies using probiotic mixtures of several bacteria find these members often work better together than by themselves [lawley, Buffie CG et al 2015, stein, hickson 2007].
-Thus, studying *C.
-difficile* in the context of the community it is exposed to is essential for building a complete picture of the variables at play in its ability to colonize.
-One way of capturing these variables is through statistical modeling.
-Using models of the microbiome, we can not only identify bacteria with positive and negative roles in *C.
-difficile* colonization but also predict levels of colonization based on the structure of the microbiota that *C.
-difficile* encountered.
+It is not completely understood how different perturbations to the gut microbiota result in a loss of colonization resistance to *C.
+difficile*.
+There is a clear need to better understand the ecology of *C.
+difficile* and its interactions with members of the microbiota.
+In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to *C.
+difficile* colonization.
+It was previously shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011 \#118}{Theriot, 2011 \#79}, tigecycline {Bassis, 2014 \#945}, clindamycin {Buffie, 2012 \#977}, or clindamycin in combination with a five antibiotic cocktail {Chang, 2008 \#590} had decreased colonization resistance.
+These studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a bloom of *Lactobacilliaceae* and *Enterobacteriaceae* are responsible for the loss of colonization resistance.
+These results are largely supported by human association studies {Vincent, 2013 \#37}{Schubert, 2014 \#1048}.
+We observed significant differences between the gut microbiota of hospitalized individuals with and without *C.
+difficile* and non-hospitalized controls {Schubert, 2014 \#1048}.
+In addition, fecal microbiota transplants increase *Bacteroidetes* and decrease *Proteobacteria* levels in recipients, resulting in a successful restoration of colonization resistance in patients {Seekatz, 2014 \#1051}.
+Precisely how this occurs is not fully understood, but it emphasizes the importance of the gut microbiota in colonization resistance against *C.
+difficile*.
 
-Although the applications for microbiome modeling are diverse, given the microbiome's relationship to many aspects of human health and disease, there are currently few studies that attempt to model the microbiome to predict disease or estimate health [Pamer, psoriases, schubert].
-Microbiome modeling has been successfully applied to the prediction of psoriasis diagnosis, in which a microbiota signature defining subjects with and without psoriasis was discovered [Statnikov et al 2013].
-Similarly, incorporating microbiota data into epidemiological based models that classify *C.
-difficile* diagnosis significantly enhances predictive power [Schubert et al.
-2013].
-In a time-series model of microbiome dynamics, the microbiota response to *C.
-difficile* alone, clindamycin alone, or clindamycin followed by *C.
-difficile* challenge was predicted (spearman correlation of 0.62) [Stein et al.
-2013].
-Lastly, the microbiota response to *Citrobacter rodentium* infection over time was modeled to capture time-dependent signatures of the microbiota [Belzer et al 2014].
+Because the gut microbiota is a complex community we need tools that can enable us to dissect the interactions within the community and with *C.
+difficile*.
+One approach is the use of models to identify associations between members of the microbiota and *C.
+difficile*.
+Models have been used to predict *C.
+difficile* {Stein, 2013 \#988}{Schubert, 2014 \#1048} and *Citrobacter* infection {Belzer, 2014 \#1053}, colon cancer {Zackular, 2013 \#472}, and psoriasis {Statnikov, 2013 \#477} based on the composition of the gut microbiota.
+By using models to explain the relationship between members of the gut microbiota, we hope to identify the subset of the normal murine microbiota that are responsible for colonization resistance.
 
-The purpose of this investigation was to expand our current knowledge of the effects of various perturbations on the microbiota's colonization resistance abilities against *C.
-difficile* and identify important members for this function.
-Through the administration of different antibiotic types, doses, and varying recovery time, we altered the murine gut microbiota and observed these structural changes using 16S rRNA sequencing.
-The resultant gut communities were then challenged with *C.
-difficile* spores.
-Using this data, we built a random forest model to predict *C.
-difficile* colonization levels based on the microbiota structure.
-Through this analysis, we have identified groups of related species with antagonistic relationships with *C.
-difficile* colonization.
-Modeling identified bacteria that did not have significant correlations with *C.
-difficile* but were important in prediction accuracy.
-These findings show we can successfully apply modeling techniques to accurately measure the colonization resistance ability of a given microbiome.
-
-### Materials and Methods
-
-***Animal care.*** We used 5-8 week old C57Bl/6 mice for all our experiments.
-These mice were reared under SPF conditions within the animal facility at the University of Michigan.
-All animal-related protocols and experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
-
-***Antibiotic administration.*** Mice were administered one of seven different antibiotics, including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
-The route of administration depended on the specific antibiotic.
-Cefoperazone (either 0.5 mg/ml, 0.3mg/ml, or 0.1 mg/ml), vancomycin (0.625 mg/ml, 0.3 mg/ml, or 0.1 mg/ml), streptomycin (5 mg/ml, 0.5 mg/ml, or 0.1 mg/ml), metronidazole (0.5 mg/ml), and ampicillin (0.5 mg/ml) were all administered in the mouse drinking water for 5 days.
-Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via IP injection.
-Mice that did not receive antibiotics were used as negative controls for these experiments.
-
-***C.
-difficile preparation and challenge*** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
-difficile* challenge.
-*C.
-difficile* strain 630Δerm spores were used in all experiments [cite dena lyras].
-Spores were prepared from a single large batch whose concentration was determined within the week prior to each *C.
-difficile* challenge [cite paul?].
-On the day of challenge 10^3 *C.
-difficile* spores were administered to mice via oral gavage.
-Immediately following this challenge, the remaining *C.
-difficile* inoculum was diluted in a series and plated to confirm the correct dosage.
-
-***Sample collection and plating*** Fecal samples were freshly collected for each mouse on the day of *C.
-difficile* challenge (day 0).
-On the day after the challenge (day 1), another fecal sample was weighed and diluted under anaerobic conditions with anaerobic PBS.
-The number of colony forming units (CFU) were counted following 24 hrs growth on TCCFA plates at 37°C under anaerobic conditions [cite plates].
-
-***DNA extraction and sequencing*** Total bacterial DNA was extracted from each  day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
-Sequencing preparation and sequencing was carried out as previously described (Kozich JJ et al.
-2013).
-Briefly, the V4 region of the 16S rRNA gene was amplified, purified, and pooled in equimolar concentrations.
-These libraries were then mixed with Illumina-prepared PhiX control libraries.
-
-```{r error}
-error_table <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.error.summary", header=T)
-
-error_nochim <- error_table[error_table$numparents==1,]
-
-error_rate <- 100 * sum(error_nochim$weight * error_nochim$error)/sum(error_nochim$weight)
-```
-
-***Sequence curation*** These sequences were curated using mothur as previously described (Kozich JJ et al and Schloss PD et al).
-Sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
-Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 9) requiring an 80% bootstrap confidence score.
-In parallel to the fecal samples, we also sequeced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
-Analysis of the mock community data indicated that the error rate following our cuation procedure was `r round(error_rate, 2)`%.
-In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples [Schloss, reducing effects].
-All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession XXXXXX.
-
-***Statistical analysis and modeling*** All analyses were conducted using R version 3.1.2.
-OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (43 OTUs).
-Spearman rank correlation analysis was performed with a Benjamini and Hochberg adjustment between OTU counts or diversity levels and *C.
-difficile* CFU/g feces.
-Comparison of bacterial levels among titration groups or delayed groups of the same antibiotic was performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests with a Benjamini and Hochberg adjustment.
-Comparison of log (base 10) transformed *C.
-difficile* CFU/g feces between experimental groups was calculated using ANOVA and subsequently pairwise T tests with a Benjamini and Hochberg adjustment.
-Random forest regression models were constructed using the randomForest R package using 5,000 trees [ref].
-The regression was performed using the log10 transformation of the number of CFU/g fecal material as the dependent variable and the 43 OTUs as predictor variables.
-Complete analysis scripts are available at the online repository for this study (https://github.com/SchlossLab/Schubert_AbxD01_mBio_2015).
+The purpose of this investigation was to expand our current knowledge of the effects of various perturbations on colonization resistance against *C.
+difficile*.
+Through the administration of different antibiotic classes, doses, recovery times we altered the murine gut microbiota and challenged the communities with C.
+difficile spores and observed differences in colonization resistance.
+We then used 16S rRNA gene sequencing to identify structural changes within the microbiota that would be predictive of colonization resistance.
+Using these data, we built a random forest model to predict *C.
+difficile* colonization levels.
+Through this analysis, we have identified groups of related bacteria that are associated with *C.
+difficile* colonization resistance.
+This model revealed that the interactions that gave rise to colonization resistance were non-linear and context dependent.
+These findings show we can successfully apply modeling techniques to accurately measure the colonization resistance ability of a given microbiota.
 
@@
 sig <- p.adjust(pvalue, method="BH")
+
+control_comparison <- comparison[grepl("control", comparison)]
+control_sig <- sig[grepl("control", comparison)]
+
+other_comparison <- comparison[!grepl("control", comparison)]
+other_sig <- sig[!grepl("control", comparison)]
+
 ```
 
-***Levels of colonization resistance are associated with the initial structure of the gut microbiota.*** In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to *C.
-difficile* colonization.
-It was previously shown that C57Bl/6 mice treated with cefoperazone [], clindamycin [], or clindamycin in combination with a five antibiotic cocktail [] had decreased colonization resistance.
-To further characterize structural shifts in the microbiota associated with this loss of protective function, we chose a panel of seven representatives from six classes of antibiotics to differentially alter the microbiota and assess their resistance to *C.
+**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to* C.
+difficile*.** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C.
 difficile* colonization (**Table 1**).
@@
 We observed reproducibly high levels of *C.
-difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin.
+difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figure 1**).
 We observed variable levels of *C.
-difficile* colonization in mice treated with amipicilin.
+difficile* colonization in mice treated with ampicillin.
 Only one of six mice receiving vancomycin was colonized with *C.
 difficile* and none of the mice that received ciprofloxacin were colonized.
-We next sequenced the 16S rRNA genes from the fecal communities sampled at the time of *C.
-difficile* challenge to characterize the populations within untreated mice that provided for colonization resistance.
-The antibiotic treatments resulted in distinct and reproducible microbiota structures (Figure 1; AMOVA, P<`r round(max(sig[grepl("control", comparison)]), 2)`).
-In addition, all of the other comparisons were significantly different from each other (AMOVA, P<`r round(max(sig[!grepl("control", comparison)]), 2)`).
+We sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C.
+difficile* challenge to identify populations within the microbiota that conferred colonization resistance.
+All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, , P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (**Figures 1 and S1**; AMOVA, , P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`).
+Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`).
+The community structures of mice receiving beta-lactams (i.e.
+cefoperazone and ampicillin) were not significantly different from each other (AMOVA, P=`r round(max(other_sig[other_comparison=="amp-cef"]), 2)`).
 These results indicate that perturbing the gut microbiota with antibiotics resulted in non-overlapping community structures that yielded significant variation in susceptibility to colonization when challenged with *C.
@@
 
-```{r top_dose_correlation}
-corr_data <- read.table(file="data/process/correlation_analysis.tsv", header=T)
-families <- gsub(".*;", "", corr_data$taxonomy)
-ave_corr_family_agg <- aggregate(corr_data$sig_corrs, by=list(families), function(x){ c(mean(x), length(x))})
-ave_corr_family <- round(ave_corr_family_agg$x[,1], 2)
-names(ave_corr_family) <- ave_corr_family_agg$Group.1
-
-N_family <- round(ave_corr_family_agg$x[,2], 2)
-names(N_family) <- ave_corr_family_agg$Group.1
-```
-
-***Correlation analysis reveals potentially protective bacteria.*** To identify bacterial taxa that could be associated with resistance or susceptibility to *C.
-difficile*, we measured the correlation between the relative abundance of each OTU that had an average abundance greater than 0.1% on the day of inoculation with the level of *C.
-difficile* colonization 24 hours later (Figure 2).
-OTUs associated with providing resistance against *C.
-difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
-Among various bacterial families, three were consistently associated with *C.
-difficile* resistance: *Porphyromonadaceae* (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs), *Lachnospiraceae* (ρ<sub>average</sub>=`r ave_corr_family["Lachnospiraceae"]`, N=`r N_family["Lachnospiraceae"]` OTUs), and *Ruminococcaceae* (ρ<sub>average</sub>=`r ave_corr_family["Ruminococcaceae"]`, N=`r N_family["Ruminococcaceae"]` OTUs).
-Two OTUs from the *Proteobacteria* had a significant association with *C.
-difficile* colonization.
-These included OTUs associated with the genera *Pseudomonas* (ρ=`r round(corr_data[families=="Pseudomonadaceae","sig_corrs"], 2)`) and *Escherichia* (ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
-Overall, these results suggest that several large groups of related OTUs, particularly within the *Firmicutes* and *Bacteroidetes*, may participate in colonization resistance against *C.
-difficile*, while members of the *Proteobacteria* phylum may be involved in susceptibility.
+**Reduced perturbations result in altered levels of colonization.** Based on the *C.
+difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C.
+difficile*.
+In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin (**Figures 2 and S2**).
+These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e.
+cefoperazone), Gram-negative (i.e.
+streptomycin), and Gram-positive (i.e.
+vancomycin) bacteria (**Table 1**).
+As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<0.02; **Figure 2**).
+Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 5) and an OTU associated with the genus *Pseudomonas* (OTU 65; Figure 2).
+The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis.
+Reducing the dose of streptomycin significantly reduced the colonization levels (P\<0.01; Figure 2).
+Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Alistipes* (OTU 11), and *Bacteroidales* (OTU 17).
+In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
+We anticipated that lower doses might result in a community structure that would result in colonization.
+In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=0.96).
+Seven OTUs were differentially represented across the three doses of vancomycin.
+Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
+Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
+Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
+These results suggest that the context of the microbiota is important in determining the overall resistance to *C.
+difficile*.
+For example, the relationship between the *Bacteroides* (OTU 1) and *C.
+difficile* colonization is positive in streptomycin-treated mice and it is negative in cefoperazone-treated mice.
+In addition, cefoperazone and streptomycin-treated mice have high levels of *C.
+difficile* although the former have significantly higher levels of *Escherichia* (OTU 4), which are absent in the streptomycin-treated mice.
+Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance.
+In light of such results, resistance is likely a product of the overall composition of the community.
 
@@
 cef.test <- pairwise.t.test(log(cef$CFU+1), factor(cef$dose), method="BH")
-
 ```
 
-***Reduced perturbation of the microbiota increases colonization resistance.*** Based on the *C.
-difficile* colonization levels in our seven antibiotic treatments, we hypothesized that reducing the dose of antibiotics that the mice received would result in a reduced perturbation to the microbiota and a reduced sensitivity to *C.
+**Reduced perturbations result in altered levels of colonization.** Based on the *C.
+difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C.
 difficile*.
-In addition to the previous treatments, we treated mice with cefoperazone (0.3 and 0.1 mg/mL), streptomycin (0.5 and 0.1 mg/mL), and vancomycin (0.3 and 0.1 mg/mL).
+In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin (Figure S2).
 These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e.
 cefoperazone), Gram-negative (i.e.
-streptomycin) and Gram-positive (i.e.
+streptomycin), and Gram-positive (i.e.
 vancomycin) bacteria.
-Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
+As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2).
+Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 4) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 9) and an OTU associated with the genus *Pseudomonas* (OTU 65; **Figure 2**).
+The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis.
+Reducing the dose of streptomycin significantly reduced the colonization levels (P\<`r round(max(strep.test$p.value, na.rm=T), 2)`; **Figure 2**).
+Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Alistipes* (OTU 11), and *Bacteroidales* (OTU 17).
+In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
 We anticipated that lower doses might result in a community structure that would result in colonization.
-As expected, colonization levels decreased significantly in all mice receiving reduced doses of cefoperazone (P<`r round(max(cef.test$p.value, na.rm=T), 2)`).
-Reducing the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 4) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 9) and an OTU associated with the genus *Pseudomonas* (OTU 65; Figure 3).
-The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis.
-Reducing the dose of streptomycin significantly reduced the colonization levels (P<`r round(max(strep.test$p.value, na.rm=T), 2)`; Figure 3).
-These changes matched the correlations that we observed across all of the antibiotic treatments.
-Reducing the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Allistipes* (OTU 11), and *Bacteroidales* (OTU 17).
-Interestingly, the OTUs that classified as members of the *Porphyromonadaceae* family did not all follow the same dose response.
-Of the 8 OTUs that changed with dose of streptomycin, 5 followed the response predicted from the correlation-based analysis (OTUs 5, 9, 10, 11, and 13), 1 contradicted the expected response (OTU 3), and 2 had a mixed response (OTUs 2 and 17).
-Although the 0.625 mg/mL dose of vancomycin did not result in colonization, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`).
-Seven OTUs were differentially represented across the three vancomycin doses.
+In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`)).
+Seven OTUs were differentially represented across the three doses of vancomycin.
 Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
 Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
-Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
-Together, these results suggest that although individual populations were occasionally predictive of colonization levels, they were not sufficient to consistently predict colonization resistance.
+Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
+These results suggest that the context of the microbiota is important in determining the overall resistance to *C.
+difficile*.
+For example, the relationship between the *Bacteroides* (OTU 1) and *C.
+difficile* colonization is positive in streptomycin-treated mice and it is negative in cefoperazone-treated mice.
+In addition, cefoperazone and streptomycin-treated mice have high levels of *C.
+difficile* although the former have significantly higher levels of *Escherichia* (OTU 4), which are absent in the streptomycin-treated mice.
+Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance.
 In light of such results, resistance is likely a product of the overall composition of the community.
@@
 
-***Allowing recovery of specific bacterial populations restores colonization resistance.***
-In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hrs before challenging them with *C.
+**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hours before challenging them with *C.
 difficile*.
 Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C.
-difficile* is restored [refs].
-To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover an additional five days.
-Among the metronidazole-treated mice, those that recovered longer had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)` reduction in colonization (P<0.001; Figure 4).
-In addition, of the `r nrow(metro_delay)` mice that experienced the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C.
+difficile* is restored {Reeves, 2011 \#118} {Theriot, 2014 \#956}.
+To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (Figure S3).
+Among the metronidazole-treated mice, those with extended recovery had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**).
+In addition, of the `r nrow(metro_delay)` mice given the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C.
 difficile* 24 hours after challenge.
-We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 4).
-Most notable among these was an OTU associated with the the family *Porphyromonadaceae* (OTU 3) and the genus *Escherichia* (OTU 4).
-The relative abundance of the former increased with the delay and the relative abundance of the latter decreased.
-Based on the correlation analysis, the trajectories of the relative abundances for these OTUs matched the observed colonization levels.
-Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in colonization (P=`r round(amp_test$p.value, 2)`).
-Again, we observed a similar increase and decrease in relative abundance for the *Porphyromonadaceae* (OTU 3) and *Escherichia* (OTU 4) OTUs.
-<Need a conclusion sentence here>
+We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 3).
+Most notable among these was a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3).
+The relative abundance of this *Barnesiella* OTU increased with the delay, and the relative abundance of this *Escherichia* OTU decreased.
+Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C.
+difficile* colonization (P=`r round(amp_test$p.value, 2)`).
+As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3).
+However untreated, fully resistant mice harbor significantly lower levels of *Barnesiella* OTU 2.
+Rather, untreated mice have high levels of various *Porphyromonadaceae* OTUs (**Figure 1**).
+These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.
 
-```{r random_forest}
-data <- read.table(file="data/process/random_forest.data")
-n_full_features <- data[1,1]
-rsq_full_features <- data[1,2]
+```{r top_dose_correlation}
+corr_data <- read.table(file="data/process/correlation_analysis.tsv", header=T)
+families <- gsub(".*;", "", corr_data$taxonomy)
+ave_corr_family_agg <- aggregate(corr_data$sig_corrs, by=list(families), function(x){ c(mean(x), length(x))})
+ave_corr_family <- round(ave_corr_family_agg$x[,1], 2)
+names(ave_corr_family) <- ave_corr_family_agg$Group.1
 
-n_part_features <- data[2,1]
-rsq_part_features <- data[2,2]
+N_family <- round(ave_corr_family_agg$x[,2], 2)
+names(N_family) <- ave_corr_family_agg$Group.1
 ```
 
-***The composition of the microbiota on day 0 is predictive of *C.
-difficile* colonization levels on day 1.*** The three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C.
+**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C.
+difficile*, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C.
+difficile* colonization 24 hours later (**Figure 4**).
+OTUs associated with providing resistance against *C.
+difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
+Among various bacterial families the *Porphyromonadaceae* (ρ<sub>average<\sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) and *Ruminococcaceae* (ρ<sub>average<\sub>=`r ave_corr_family["Ruminococcaceae"]`, N=`r N_family["Ruminococcaceae"]` OTUs) were consistently associated with low levels of *C.
+difficile* colonization.
+Two OTUs from the *Proteobacteria* had a significant association with *C.
+difficile* colonization.
+These included OTUs associated with the genera (ρ=`r round(corr_data[families=="Pseudomonadaceae","sig_corrs"], 2)`) and *Escherichia* (ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
+These associations have been suggested previously {Reeves, 2011 \#118;Schubert, 2014 \#1048;Bassis, 2014 \#945;Buffie, 2012 \#977}.
+In addition, by performing an OTU-based analysis we observed several bacterial families that were represented by OTUs that were associated with low and high levels of *C.
+difficile* colonization.
+For example, the Lachnospiraceae have been associated with protection against C.
+difficile and although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
+difficile* and `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU that was associated with high levels of *C.
+difficile*.
+In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` *Lactobacillus* OTUs from the *Lactobacilliaceae* where `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` were associated with low levels of *C.
+difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels.
+The broad taxonomic representation of OTUs associated with low levels of *C.
+difficile* suggests that a diverse community is required to prevent the colonization *C.
+difficile*.
+
+**The composition of the disturbed gut microbiota is predictive of C.
+difficile colonization levels.** The three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C.
 difficile* colonization, but in other contexts they were not.
-For example, the OTU affiliated with *Escherichia* (OTU 4) was associated with high levels of colonization in the ampicillin, metronidazole, and cefoperazone-treated mice; however, it was not detected in the streptomycin-treated mice although *C.
-difficile was able to colonize those mice to high levels.
+Correlation-based analyses cannot predict these types of context dependencies because they do not take into account the non-linearity and statistical interactions between populations.
 This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization.
@@
 difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge.
-The model explained `r round(rsq_full_features*100, 1)`% of the variation in the observed *C.
-difficile* colonization levels (Figure 5).
-Noticing a natural jump following the twelfth OTU in the percent increase in the mean squared error when each OTU was removed, we regenerated the model using only the top 12 OTUs.
-This model explained `r round(rsq_part_features*100, 1)`% of the variation in the observed *C.
+The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C.
+difficile* colonization levels (Figures 5 and S4).
+When we only included the top 7 OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C.
 difficile* colonization levels.
-
-Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia*.
+Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia* (**Figure 6**).
 These results further validate the observations from the correlation-based analysis.
-Together these results suggest that colonization resistance is likely conferred by *porphyromonadaceae*, *Alistipes*, and *lachnospiraceae* and a loss in these bacteria, concurrently with a gain of *Escherichia*, can result in increased susceptibility to infection.
-<needs help>Interestingly, the relationship between these OTUs and colonization resistance was not a simple dichotomous or linear relationship.
-For example, among the streptomycin-treated mice which altogether lacked the *Enterobacteriaceae*, *Porphyromonadaceae* (OTU 3) and *Lachnospiraceae* (OTU 39), were the two most important features in accurately predicting subsequent *C.
-difficile* levels</needs help>
+Together these results suggest that colonization resistance is likely conferred by *Porphyromonadaceae* (OTU 15, 10, and 6), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), and *Alistipes* (OTU 12).
+A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), can result in increased susceptibility to infection (Figure 6).
+As we observed in the titration experiments, the relationship between *Akkermansia* (OTU 4) and *C.
+difficile* colonization appears to be context dependent.
+There were varying abundances of *Akkermansia* in mice regardless of the level of *C.
+difficile* colonization.
 
@@
 
-Through our antibiotic perturbation experiments we identified a panel of bacteria with either strong positive or negative relationships with subsequent levels of *C.
+Previous attempts to study the role of the gut microbiota in colonization resistance against C.
+difficile infection have utilized a single perturbation to the community.
+Here, we used 7 antibiotics from 6 classes that were given to mice in varying doses and with varying post-antibiotic recovery periods.
+The result was a combination of 15 different perturbations that allowed us to generate distinct community profiles that displayed varying susceptibility to *C.
 difficile* colonization.
-A large number of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, other *Clostridia*, and *Alistipes* were found to have strong potential resistance abilities against *C.
-difficile*.
-Conversely, several members of the *Proteobacteria* phylum, including *Pseudomonas*, *Streptococcus*, and especially *Escherichia* have strong potential positive influences on *C.
+Our findings demonstrated that colonization resistance is not driven by individual populations, but by a community of organisms.
+Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are protective against *C.
+difficile* {Buffie, 2012 \#977}{Reeves, 2012 \#126}.
+Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for colonization resistance.
+Overall, these results suggest that *C.
+difficile* is a generalist that is capable of exploiting a variety of niches in the gastrointestinal tract.
+
+There is clear need for more efficient therapies for treatment of *C.
+difficile* infections in humans aimed at restoration of the microbiota.
+Current first line treatments of CDI include regimens of either metronidazole or vancomycin, which further perturb the microbiota.
+As such, relapse rates for CDI are typically around 25-30% {Wilcox, 1998 \#599}.
+Interestingly, we observed that treatment with either antibiotic induced susceptibility to *C.
+difficile* in mice.
+This result has implications for understanding the causes of recurrent infections.
+Previous efforts to restore the microbiota and reestablish colonization resistance also support our findings.
+For instance, association of germ-free mice with a *Lachnospiraceae* isolate only reduced the level of *C.
+difficile* colonization by 10 to 100-fold {Reeves, 2012 \#126}.
+Using conventional mice, mixtures of bacteria rather than individual populations have been shown to restore colonization resistance and mediate clearance of *C.
+difficile* {Lawley, 2012 \#36}{Buffie, 2015 \#981}.
+Fecal transplants, which represent a diverse collection of bacterial populations, have been highly effective in treating humans with recurrent *C.
+difficile* {Kassam, 2013 \#1003}{Seekatz, 2014 \#1051}{Weingarden, 2015 \#1054}.
+By generating a diverse collection of communities that were challenged with C.
+difficile, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance.
+These would include members of the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes*.
+[ It would be nice to comment on how these overlap with other cocktails ]
+
+Random forest regression models allowed us to describe community resistance as a byproduct of an assemblage of bacterial populations rather than as individual populations.
+A correlation-based analysis was unable to identify populations that had a context dependent or non-linear associations with *C.
 difficile* colonization.
-These results were supported by their significant correlation values across several antibiotic regimens.
-By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had increased resistance to *C.
-difficile* colonization.
-Using this collective microbiota data spanning a wide range of colonization resistance levels, we moved beyond correlation to prediction of *C.
-difficile* colonization based on the composition of the microbiota.
-Through modeling we identified bacteria within the community which contributed greatest to accurately predicting *C.
-difficile* levels.
-This is the first comprehensive analysis both measuring and predicting colonization resistance levels based on the microbiota structure.
+Although the murine and human microbiota do not fully overlap, our previous analysis of humans infected with *C.
+difficile* supports the populations that we associated with colonization {Schubert, 2014 \#1048}.
+For instance, *Escherichia* was previously associated with infected individuals and *Lachnospiraceae*, *Ruminococcaceae*, and *Alistipes* were absent from infected individuals.
+The results of the current study suggest that it should be possible to model a patient’s risk of developing a *C.
+difficile* infection based on their gut microbiota composition at admission.
 
-Correlation analysis identified major families with significant negative or positive relationships with *C.
-difficile*.
-Many of the correlaions we observed were not surprising given other studies looking at the relationship between *C.
-difficile* and the microbiome.
-For instance, monocolonization of germfree mice with either a *Lachnospiraceae* species or *E.
-coli* were tested for their resistance to *C.
-difficile* colonization [Reeves].
-Pre-colonization with *Lachnospiraceae* was able to decrease the levels of *C.
-difficile*, while *E.
-coli* pre-colonization and completely germfree mice had similar *C.
-difficile* levels.
-Association studies in humans with and without CDI in hospitals have also highlighted several bacteria found in this study, including *Lachnospiraceae*, *Rumicoccaceae*, *Clostridia*, and *Alistipes* as decreased and *E.
-coli* and *Streptococcus* as increased in subjects with CDI [Schubert, Antharam].
-While we observe overlaps in potentially protective bacteria between mice and humans at broad taxonomy levels, there are notable differences at finer taxonomic levels.
-Two different *Bacteroidales* species, a *Porphyromonadaceae* and a *Bacteroides*, have the strongest negative correlations with *C.
-difficile* in mice and humans, respectively [Schubert].
-*Porphyromonadaceae* is more commonly found in mice than in humans, indicating that there are host specific bacteria in colonization resistance.
+### Materials and Methods
 
-Modeling of the microbiome allows us to characterize *C.
-difficile* colonization in terms of a community, which is necessary to capture these complex relationships among its members.
-Correlation analysis alone was insufficient in identifying all informative bacteria.
-Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C.
-difficile* yet improved model accuracy.
-*Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*).
-While these bacteria may not have important direct effects on *C.
-difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C.
-difficile*.
+**Animal care.** We used 5-8 week old C57Bl/6 mice for all of our experiments.
+These mice were reared under SPF conditions within the animal facility at the University of Michigan.
+All animal-related protocols and experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan and carried out in accordance with the approved guidelines.
 
-Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance.
-No single bacteria was able to consistently explain the subsequent level of *C.
-difficile* observed.
-For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C.
-difficile*.
-The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C.
-difficile* (*Figure 5C*).
-However, a small group of microbiotas with greater than 1% OTU3 abundance were not fully protected by the presence of this OTU alone and ranged in *C.
-difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces.
-These included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas.
-Similarly, having low amounts of one "protective" bacteria such as *Lachnospriaceae* OTU39 does not automatically render the host susceptible to colonization.
-For instance streptomycin treated (0.1 mg/ml) mice did not harbor *Lachnospiraceae* OTU39, yet had undetectable levels of *C.
-difficile*.
-Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with protective capacities.
+**Antibiotic administration.** Mice were administered one of seven different antibiotics including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
+The route of administration depended on the specific antibiotic.
+Cefoperazone (0.5, 0.3, or 0.1 mg/ml), vancomycin (0.625, 0.3, or 0.1 mg/ml), streptomycin (5, 0.5, or 0.1 mg/ml), metronidazole (0.5 mg/ml), and ampicillin (0.5 mg/ml) were all administered in the mouse drinking water for 5 days.
+Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via IP injection.
+Mice that did not receive antibiotics were used as negative controls for these experiments.
 
-Furthermore, the dynamics of *C.
-difficile* colonization does not appear to be influenced solely by the loss of protective species.
-Both *Escherichia* OTU11 and *Streptococcus* OTU78 were positively correlated with *C.
-difficile* colonization and important in the random forest model.
-Notably, *Escherichia* OTU11 was among the strongest positively correlated OTUs with *C.
-difficile* across the majority of antibiotic treatments.
-Though *Escherichia* may aid in colonization of *C.
-difficile*, our data suggests it is not essential.
-*Escherichia* OTU11 was near absent in streptomycin treated (5 mg/ml) mice, but *C.
-difficile* colonized to high levels (*Figure 1*).
-Together These observations lead to a model of colonization by *C.
-difficile* in which antibiotic perturbation leads to a significant loss of protective species.
-This in turn leads to a loss of immune homeostasis [Smith et al.
-2013 Science], an increase in available resources [Theriot 2014, Ferreyra et al 2014, Ng KM et al  2013, Koenigsknecht et al 2015] and eliminates direct or indirect inhibitory effects of bacterial products made by these protective bacteria (e.g.
-bacteriocins or short chain fatty acids) [Buffie CG et al 2015, Theriot 2014].
-While this may be suitable enough for *C.
-difficile* to colonize, other opportunistic species such as *Escherichia coli* and other *Proteobacteria* may also take advantage of these microbiota perturbations.
-The bloom in these bacteria may further change the gut milieu in favor of *C.
-difficile* colonization.
-Increased growth by these bacteria can trigger inflammation, which has been shown to be advantageous for *Salmonella tyhpumurium* to further disrupt and suppress the indigenous microbiota [Stecher et al 2007 Plos Biol].
-Increased *Proteobacteria* abundances were also found among mice with increased mucus layer penetrability [Jakobsson et al.
-2015 EMBO reports].
+**C.
+difficile preparation and challenge.** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
+difficile* challenge.
+*C.
+difficile* strain 630Δerm spores were used in all experiments.
+Spores were prepared from a single large batch whose concentration was determined within the week prior to each *C.
+difficile* challenge {Sorg, 2009 \#1055}.
+Spores were stored long term at 4 °Celsius.
+On the day of challenge 10^3^ *C.
+difficile* spores were administered to mice via oral gavage.
+Immediately following this challenge, the remaining *C.
+difficile* inoculum was diluted in a series and plated to confirm the correct dosage.
 
-We have shown across several different types of perturbations that we can accurately predict the resistance levels of these communities to *C.
-difficile*.
-This has great applications for predicting patient risk of CDI in the hospital.
-Furthermore, this analysis can be used to inform the design of personalized medicine, whether through tailoring probiotic or antibiotic regimens.
-Future studies are needed to examine of the human-specific bacteria in colonization resistance and also to examine the changes in models necessary to account for physiological differences between *C.
-difficile* strains.
-<End sentence is weak.>
+**Sample collection and plating.** Fecal samples were freshly collected for each mouse on the day of *C.
+difficile* challenge.
+On the day after the challenge another fecal sample was weighed and diluted under anaerobic conditions with anaerobic PBS.
+The number of colony forming units (CFU) were counted following 24 hours growth on TCCFA plates at 37°C under anaerobic conditions {Buggy, 1983 \#1059}.
 
-Table 1. Description of Antibiotics used in this study
+**DNA extraction and sequencing.** Total bacterial DNA was extracted from each day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
+We generated amplicons of the V4 region within the 16S rRNA gene and sequenced the fragments using an Illumina MiSeq as previously described {Kozich, 2013 \#40}.
+
+```{r error}
+error_table <- read.table(file="data/process/abxD0.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.error.summary", header=T)
+
+error_nochim <- error_table[error_table$numparents==1,]
+
+error_rate <- 100 * sum(error_nochim$weight * error_nochim$error)/sum(error_nochim$weight)
+```
+
+**Sequence curation.** These sequences were curated using mothur as previously described {Kozich, 2013 \#40; Schloss, 2009 \#812}.
+Briefly, sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
+Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 10) requiring an 80% bootstrap confidence score.
+In parallel to the fecal samples, we also sequenced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
+Analysis of the mock community data indicated that the error rate following our curation procedure was `r round(error_rate, 2)`%.
+In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples {Schloss, 2011 \#163}.
+All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession XXXXXX.
+
+**Statistical analysis and modeling.** All analyses were conducted using R version 3.1.2.
+OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (39 OTUs).
+Comparison of bacterial levels among titration groups or delayed groups of the same antibiotic was performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
+Comparison of log (base 10) transformed *C.
+difficile* CFU/g feces between experimental groups was calculated using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
+Spearman rank correlation analysis was performed between OTU counts and *C.
+difficile* CFU/g feces.
+All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05.
+Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#1061} .
+The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 39 OTUs as predictor variables.
+Complete analysis scripts are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015).
+
+### Acknowledgements
+
+This work was supported by several grants from the National Institutes
+for Health R01GM099514, R01HG005975, U19AI090871, and P30DK034933. The
+funding agencies had no role in study design, data collection and
+analysis, decision to publish, or preparation of the manuscript.
+
+**Table 1. Description of Antibiotics used in this study**
 
@@
 
-**Figure Legends**
+### Figure Legends
 
-**Figure 1.
-Antibiotic treatments result in significant alterations to the structure of the microbiome and variation in colonization resistance.** The median percent relative abundances of the most abundant phyla found in each experimental group on the day of *C.
-difficile* challenge.
-Each bargraph depicts a different antibiotic perturbation.
-The median level *C.
-difficile* colonization found 24 hours post microbiota sampling is indicated in the top right for each treatment.
-Error bars indicate the interquartile range.
+**Figure 1. Antibiotic treatments result in significant alterations to
+the structure of the microbiota and variation in colonization
+resistance.** Bars indicate the median percent relative abundance of
+those selected OTUs from all treatment groups on the day of *C.
+difficile* challenge. Stars along the x-axis indicate those OTUs that
+were significantly different from untreated mice for that antibiotic
+treatment. The error bars indicate the interquartile range. The median
+level *C. difficile* colonization found 24 hours post microbiota
+sampling and the number of animals in the treatment group is indicated
+in the top right for each treatment with the interquartile range in
+parentheses. The concentration next to the name of the antibiotic
+indicates the dose of the antibiotic that was given to the animals.
 
-**Figure 2.
-Broader taxonomic groups share similar relationships with *C.
-difficile* colonization.** The relative abundance of OTUs found on the day that mice were challenged with *C.
-difficile* spores were correlated with the level of C.
-difficile CFU/g feces 24 hours later.
-Only OTUs with an average abundance of at least 0.1% across all original treatments were considered in the correlation analysis.
-All correlations shown in this graph are significant after controlling for multiple comparisons using the Benjimani-Hochberg correction (P<0.05).
-OTUs are grouped by the taxonomic family that its sequences classified as or the next deeper classification that provided a classification.
-The letters in the parentheses correspond to the phylum that the taxa belongs to.
-B: Bacteroidetes, F: Firmicutes, P: Proteobacteria, V: Verrucomicrobia, A: Actinobacteria, T: Tenericutes.
+**Figure 2. Titration of antibiotic perturbations results in altered
+community structures and *C. difficile* colonization resistance.** Bars
+indicate the median percent relative abundance of those selected OTUs
+from all treatment groups on the day of *C. difficile* challenge. Stars
+along the x-axis indicate those OTUs that varied significantly across
+doses of the same antibiotic. The error bars indicate the interquartile
+range. The median level *C. difficile* colonization found 24 hours post
+microbiota sampling is plotted on the right for each treatment with
+error bars indicating the interquartile range. The number of animals
+used in each treatment group is indicated in the legend, which depicts
+the doses of each antibiotic that were used.
 
-**Figure 3.
-The average percent relative abundances for family level phylotypes found on the day of *C.
-difficile* challenge in each titration group.** Each row depicts a different antibiotic that was titrated.
-High, medium, and low antibiotic doses are represented by black, grey, and white colored bars respectively.
-The specific dose amount is shown for each antibiotic.
-Error bars show the standard error of the mean.
+**Figure 3. Increasing the recovery time following antibiotic
+perturbation restores colonization resistance.** Bars indicate the
+median percent relative abundance of those selected OTUs from all
+treatment groups on the day of *C. difficile* challenge. Stars along the
+x-axis indicate those OTUs that varied significantly between those mice
+that were allowed 1 or 6 days of recovery. The error bars indicate the
+interquartile range. The median level *C. difficile* colonization found
+24 hours post microbiota sampling is plotted on the right for each
+recovery period and antibiotic with error bars indicating the
+interquartile range. The number of mice used in each treatment group is
+indicated above the *C. difficile* colonization data. The dose of each
+antibiotic is indicated next to the name of the antibiotic.
 
-**Figure 4.
-Changes in the average relative abundance for family level phylotypes durin the period of recovery prior to challenge with *C.
-difficile*.** Each row depicts a different antibiotic that was allowed to recover.
-Black bars represent untreated animals.
-Grey bars represent animals given 1 day of recovery off antibiotics as in Figure 1.
-Error bars indicate the standard error of the mean.
+**Figure 4. Diverse taxonomic groups are associated with low levels of
+*C. difficile* colonization.** Spearman correlation coefficients were
+calculated using the relative abundance of OTUs found on the day that
+mice were challenged with *C. difficile* spores and the amount of *C.
+difficile* observed 24 hours later. Only significant correlations are
+presented. OTUs are grouped by the taxonomic family and the letters in
+the parentheses correspond to the phylum that the taxa belong to. B:
+Bacteroidetes, F: Firmicutes, P: Proteobacteria, A: Actinobacteria, T:
+Tenericutes.
 
-**Figure 5.
-Random forest model to predict *C.
-difficile* colonization levels based on the microbiota.** (A) The fit of the random forest model by experimental group.
-The dashed line represents the y=x line.
-(B) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model.
-(C) The percent relative abundance of the top 8 OTUs from (B) on day 0 are plotted against the log<sub>10</sub> *C.
-difficile* CFU/g feces.
+**Figure 5. Random forest regression model predicts *C. difficile*
+colonization levels based on the structure of the microbiota.** The
+overall model explained XX% of the variation in the data. Each pane
+shows antibiotic treatment groups in color and the other points as gray
+circles.
 
-**Supplemental Figure 2.
-Changes between untreated and recovery days.** A) The average thetaYC distance between pretreatment (day -11) and recovery days (days -5 through -1) is represented by circles connected by solid lines.
-Dashed lines with the same symbol indicate individual mice from the same cage.
-Antibiotic treatment is indicated by color.
-B) The average level of the inverse simpson diversity index is plotted by antibiotic treatment across time.
-Error bars show standard error of the mean.
-Mice were challenged with *C.
-difficile* on day 0.
+**Figure 6. Relationship between OTU relative abundance and *C.
+difficile* colonization levels indicates non-linearity and
+context-dependency.** The 9 OTUs that resulted in the greatest change in
+percent mean squared error when removed from the random forest
+regression model are shown in each pane and together explain XX.X% of
+the variation in the data. The Spearman correlation value between that
+OTUs abundance and C. difficile levels are shown for each pane. The
+color and symbols represent the same antibiotic dose and recovery period
+as in Figure 5.
+
+**Figure S1. Effect of antibiotic perturbations on phylum-level
+representation of communities on day of *C. difficile* challenge.** Bars
+depict the median relative abundance across mice within the treatment
+group and error bars indicate the interquartile range.
+
+**Figure S2. Effect of titrated antibiotic treatments on phylum-level
+representation of communities on day of *C. difficile* challenge.** Bars
+depict the median relative abundance across mice within the treatment
+group and error bars indicate the interquartile range.
+
+**Figure S3. Effect of recovery period following antibiotic treatments
+on phylum-level representation of communities on day of *C. difficile*
+challenge.** Bars depict the median relative abundance across mice
+within the treatment group and error bars indicate the interquartile
+range.
+
+**Figure S4. The change in percent mean squared error when each OTU was
+removed from the random forest regression model.**

### 2015-04-09 code/build_figure1.R
Make everything a PDF

@@ -13,7 +13,7 @@
 #   * data/process/abxD1.counts
 #
 # Output...
-#   * results/figures/figure1.tiff
+#   * results/figures/figure1.pdf
 #
 ################################################################################
 
@@ -146,7 +146,7 @@ single_drug_bars <- function(drug, drug_sig_otus, drug_label){
 
 
 
-tiff(file="results/figures/figure1.tiff", width=4.5, height=10.0, units="in", res=300)
+pdf(file="results/figures/figure1.pdf", width=4.5, height=10.0)
     par(cex=1.2)
 
     layout(matrix(c(

### 2015-04-09 code/build_figure2.R

@@ -11,7 +11,7 @@
 #   * data/process/abxD1.counts
 #
 # Output...
-#   * results/figures/figure2.tiff
+#   * results/figures/figure2.pdf
 #
 ################################################################################
 
@@ -111,7 +111,7 @@ otu <- gsub("Otu0*", "", names(taxonomy))
 
 label <- paste0(taxonomy, " (OTU ", otu, ")")
 
-tiff(file="results/figures/figure2.tiff", width=6.875, height=5.75, units="in", res=300)
+pdf(file="results/figures/figure2.pdf", width=6.875, height=5.75)
     par(cex=1.2)
 
     layout(matrix(c(1,5,2,6,3,7,4,8), nrow=4, byrow=T), width=c(1,0.25), height=c(1,1,1,1.2))

### 2015-04-09 code/build_figure3.R

@@ -11,7 +11,7 @@
 #   * data/process/abxD1.counts
 #
 # Output...
-#   * results/figures/figure3.tiff
+#   * results/figures/figure3.pdf
 #
 ################################################################################
 
@@ -95,7 +95,7 @@ otu <- gsub("Otu0*", "", names(taxonomy))
 
 label <- paste0(taxonomy, " (OTU ", otu, ")")
 
-tiff(file="results/figures/figure3.tiff", width=6.875, height=4.5, units="in", res=300)
+pdf(file="results/figures/figure3.pdf", width=6.875, height=4.5)
     par(cex=1.2)
 
     layout(matrix(c(1,4,2,5,3,6), nrow=3, byrow=T), width=c(1,0.25), height=c(1,1,1.1))

### 2015-04-09 code/build_figure4.R

@@ -16,7 +16,7 @@
 #   * data/process/abxD1.counts
 #
 # Output...
-#   * results/figures/figure4.tiff
+#   * results/figures/figure4.pdf
 #
 ################################################################################
 
@@ -105,7 +105,7 @@ phylum_levels <- gsub("Bacteria;\\d", "", phylum_levels)
 tax_labels <- paste(family_levels, phylum_levels)
 
 
-tiff(file="results/figures/figure4.tiff", width=6.5, height=4.0, units="in", res=300)
+pdf(file="results/figures/figure4.pdf", width=6.5, height=4.0)
 
     set.seed("6201976")
 

### 2015-04-09 code/build_figure5_6_S4.R

@@ -11,8 +11,8 @@
 #   * data/process/abxD1.counts
 #
 # Output...
-#   * results/figures/figure5.tiff
-#   * results/figures/figure6.tiff
+#   * results/figures/figure5.pdf
+#   * results/figures/figure6.pdf
 #
 ################################################################################
 
@@ -105,7 +105,7 @@ write(c(n_features, rf_partial$rsq[n_trees]), file="data/process/random_forest.d
 
 
 # supplemental figure 4: full feature importance plot
-tiff(file="results/figures/figureS4.tiff", width=3.5, height=5.0, units="in", res=300)
+pdf(file="results/figures/figureS4.pdf", width=3.5, height=5.0)
 
     par(mar=c(3,8,0.5,0.5))
     plot(NA, yaxt="n", xlab="", ylab="", xlim=c(min(importance_sorted), max(importance_sorted)),
@@ -178,7 +178,7 @@ make_colored_plot <- function(drug){
 
 
 
-tiff(file="results/figures/figure5.tiff", width=5.0, height=6.0, units="in", res=300)
+pdf(file="results/figures/figure5.pdf", width=5.0, height=6.0)
 
     design <- matrix(1:8, nrow=4, byrow=T)
     design <- cbind(c(rep(9,4)), design)
@@ -267,7 +267,7 @@ dev.off()
 n_features <- 9
 
 # let's build Figure 6 (w/ color & pch)
-tiff(file="results/figures/figure6.tiff", width=6.875, height=7.5, units="in", res=300)
+pdf(file="results/figures/figure6.pdf", width=6.875, height=7.5)
 
     #want to jitter the relative abundance for those mice that had no Cdiff
     #colonization

### 2015-04-09 code/build_figureS1.R

@@ -13,7 +13,7 @@
 #   * data/process/abxD1.counts
 #
 # Output...
-#   * results/figures/figure1.tiff
+#   * results/figures/figure1.pdf
 #
 ################################################################################
 
@@ -89,7 +89,7 @@ fulldose_phylum_barplot <- function(drug, label){
     pos
 }
 
-tiff(height=9, width=3.75, file="results/figures/figureS1.tiff", units="in", res=300)
+pdf(height=9, width=3.75, file="results/figures/figureS1.pdf")
 
     z <- layout(
         matrix( c(  1,

### 2015-04-09 code/build_figureS2.R

@@ -86,7 +86,7 @@ titration_phylum_barplot <- function(drug, label){
     z
 }
 
-tiff(height=5, width=3.75, file="results/figures/figureS2.tiff", units="in", res=300)
+pdf(height=5, width=3.75, file="results/figures/figureS2.pdf")
 
     z <- layout(
         matrix( c(  1,

### 2015-04-09 code/build_figureS3.R

@@ -93,7 +93,7 @@ delay_phylum_barplot <- function(drug, label){
     z
 }
 
-tiff(height=3.25, width=3.75, file="results/figures/figureS3.tiff", units="in", res=300)
+pdf(height=3.25, width=3.75, file="results/figures/figureS3.pdf")
 
     z <- layout(
         matrix( c(  1,

### 2015-04-09 Schubert_abxD01_mBio_2015.Rmd
Fix table formatting

@@
 
-Antibiotic | Administration | Class | Mechanism | Target
+| Antibiotic | Administration | Class | Mechanism | Target |
 :---------:|:--------------:|:-----:|:---------:|:---------:

### 2015-04-09 Schubert_abxD01_mBio_2015.Rmd
Fix table formatting

--- 2015-04-09 Schubert_abxD01_mBio_2015.Rmd
@@
 
-Antibiotic | Administration | Class | Mechanism | Target
-:---------:|:--------------:|:-----:|:---------:|:---------:
+| Antibiotic | Administration | Class | Mechanism | Target |
+|:---------:|:--------------:|:-----:|:---------:|:---------:|
 Ciprofloxacin | Oral gavage, one time | Fluoroquinolone | Inhibits DNA gyrase | Gram +/-

### 2015-04-10 scratch/fun_with_plotmath.R
Make a 12 panel figure

@@ -1,0 +1,31 @@
+my_names <- c("Name1", "Name2", "Name3", "Name4")
+my_numbers <- c(2,4,5,9)
+my_data <- c(24, 47, 55, 90)
+
+
+par(mar=c(4,10,1,1))
+plot(x=my_data, y=1:4, yaxt="n", ylab="", xlim=c(0,100))
+)
+
+# axis(2, label=paste0(my_names, " (Number ", my_numbers, ")"), at=1:4, las=1)
+
+# no... just says "expresssion(...)"
+# axis(2, label=paste0(expression(italic(my_names)), " (Number ", my_numbers, ")"), at=1:4, las=1)
+
+# no... just italicizes "expresssion(...)"
+#axis(2, label=parse(text=paste0(expression(italic(my_names)), "~(Number~", my_numbers, ")")), at=1:4, las=1)
+
+
+# no... expression makes vector into a single value
+#axis(2, label=expression(paste0(italic(my_names), " (Number ", my_numbers, ")")), at=1:4, las=1)
+
+my_text <- paste0("italic(", my_names, ")~(Number~", my_numbers, ")")
+axis(2, label=parse(text=my_text), at=1:4, las=1)
+
+
+m <- "2.3~x~10^6"
+l <- "1.2~x~10^6"
+u <- "4.5~x~10^6"
+
+my_string <- paste0(m, "~(", l, "~-~", u, ")")
+text(x=50, y=2.5, parse(text=my_string))

### 2015-04-10 Schubert_abxD01_mBio_2015.Rmd
Incorporate Schloss edits

@@
 difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner.
-Using relative abundance data, we built machine learning regression models to predict the levels *C.
+Using relative abundance data, we built machine learning regression models to predict the levels of *C.
 difficile* that were found 24 hours after challenging the perturbed communities.
@@
 difficile*.
-Rather, resistance can be achieved with multiple diverse assemblages.
+Rather, resistance and sensitivity can be achieved with multiple diverse assemblages.
 
@@
 
-The gastrointestinal tract harbors a complex community of bacteria, known as the microbiota, which plays an integral role in resisting the invasion by gut pathogens.
+The gastrointestinal tract harbors a complex community of bacteria, known as the microbiota, which plays an integral role preventing its colonization by gut pathogens.
 This resistance has been shown to be crucial for protection against *C.
@@
 Antibiotics are a major risk factor for acquiring CDI due to their effect on the normal structure of the indigenous gut microbiota.
-We showed that diverse antibiotic perturbations gave rise to altered communities that varied in their susceptibility to *C.
+We found that diverse antibiotic perturbations gave rise to altered communities that varied in their susceptibility to *C.
 difficile* colonization.
@@
 This native protection by the microbiota from invasive pathogenic species is termed colonization resistance.
-It is well established that the gut bacterial microbiota is critical in the hosts’ defense against the pathogen *Clostridium difficile* {Wilson, 1985 \#1045}{Chang, 2008 \#590}{Reeves, 2011 \#118}.
+It is well established that the gut bacterial microbiota is critical in the host's defense against the pathogen *Clostridium difficile* {Wilson, 1985 \#1045; Chang, 2008 \#590;Reeves, 2011 \#118}.
 When this indigenous community is perturbed, this often leads to a loss of resistance.
@@
 difficile* and its interactions with members of the microbiota.
-In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to *C.
+In mouse models of CDI, the perturbed, untreated murine microbiota is completely resistant to *C.
 difficile* colonization.
-It was previously shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011 \#118}{Theriot, 2011 \#79}, tigecycline {Bassis, 2014 \#945}, clindamycin {Buffie, 2012 \#977}, or clindamycin in combination with a five antibiotic cocktail {Chang, 2008 \#590} had decreased colonization resistance.
+It was previously shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011 \#118; Theriot, 2011 \#79}, tigecycline {Bassis, 2014 \#945}, clindamycin {Buffie, 2012 \#977}, or clindamycin in combination with a five antibiotic cocktail {Chang, 2008 \#590} had decreased colonization resistance.
 These studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a bloom of *Lactobacilliaceae* and *Enterobacteriaceae* are responsible for the loss of colonization resistance.
-These results are largely supported by human association studies {Vincent, 2013 \#37}{Schubert, 2014 \#1048}.
+These results are largely supported by human association studies {Vincent, 2013 \#37; Schubert, 2014 \#1048}.
 We observed significant differences between the gut microbiota of hospitalized individuals with and without *C.
@@
 Models have been used to predict *C.
-difficile* {Stein, 2013 \#988}{Schubert, 2014 \#1048} and *Citrobacter* infection {Belzer, 2014 \#1053}, colon cancer {Zackular, 2013 \#472}, and psoriasis {Statnikov, 2013 \#477} based on the composition of the gut microbiota.
-By using models to explain the relationship between members of the gut microbiota, we hope to identify the subset of the normal murine microbiota that are responsible for colonization resistance.
+difficile* {Stein, 2013 \#988; Schubert, 2014 \#1048} and *Citrobacter* infection {Belzer, 2014 \#1053}, colon cancer {Zackular, 2013 \#472}, and psoriasis {Statnikov, 2013 \#477} based on the composition of the gut microbiota.
+By using models to explain the relationship between members of the gut microbiota, we similarly sought to identify the subset of the normal murine microbiota that are responsible for colonization resistance.
 
@@
 difficile*.
-Through the administration of different antibiotic classes, doses, recovery times we altered the murine gut microbiota and challenged the communities with C.
-difficile spores and observed differences in colonization resistance.
+Through the administration of different antibiotic classes, doses, and recovery times we altered the murine gut microbiota and challenged the communities with *C.
+difficile* spores and to quantify differences in colonization resistance.
 We then used 16S rRNA gene sequencing to identify structural changes within the microbiota that would be predictive of colonization resistance.
@@
 
-**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to* C.
-difficile*.** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C.
+**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to** ***C.
+difficile.*** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C.
 difficile* colonization (**Table 1**).
-On the day after the mice were challenged with *C.
+On the day after mice were challenged with *C.
 difficile* we enumerated the density of *C.
@@
 We observed reproducibly high levels of *C.
-difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figure 1**).
+difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**).
 We observed variable levels of *C.
@@
 difficile* challenge to identify populations within the microbiota that conferred colonization resistance.
-All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, , P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (**Figures 1 and S1**; AMOVA, , P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`).
+All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, , P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`).
 Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`).
@@
 difficile*.
-
-**Reduced perturbations result in altered levels of colonization.** Based on the *C.
-difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C.
-difficile*.
-In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin (**Figures 2 and S2**).
-These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e.
-cefoperazone), Gram-negative (i.e.
-streptomycin), and Gram-positive (i.e.
-vancomycin) bacteria (**Table 1**).
-As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<0.02; **Figure 2**).
-Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 5) and an OTU associated with the genus *Pseudomonas* (OTU 65; Figure 2).
-The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis.
-Reducing the dose of streptomycin significantly reduced the colonization levels (P\<0.01; Figure 2).
-Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Alistipes* (OTU 11), and *Bacteroidales* (OTU 17).
-In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
-We anticipated that lower doses might result in a community structure that would result in colonization.
-In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=0.96).
-Seven OTUs were differentially represented across the three doses of vancomycin.
-Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
-Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
-Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
-These results suggest that the context of the microbiota is important in determining the overall resistance to *C.
-difficile*.
-For example, the relationship between the *Bacteroides* (OTU 1) and *C.
-difficile* colonization is positive in streptomycin-treated mice and it is negative in cefoperazone-treated mice.
-In addition, cefoperazone and streptomycin-treated mice have high levels of *C.
-difficile* although the former have significantly higher levels of *Escherichia* (OTU 4), which are absent in the streptomycin-treated mice.
-Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance.
-In light of such results, resistance is likely a product of the overall composition of the community.
 
@@
 As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2).
-Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 4) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 9) and an OTU associated with the genus *Pseudomonas* (OTU 65; **Figure 2**).
-The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis.
+Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs.
+We also observed increases in the relative abundances of OTUs associated with the family *Porphyromonadaceae* (OTU 5, 10, 11, 13, and 21; **Figure 2**).
 Reducing the dose of streptomycin significantly reduced the colonization levels (P\<`r round(max(strep.test$p.value, na.rm=T), 2)`; **Figure 2**).
-Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Alistipes* (OTU 11), and *Bacteroidales* (OTU 17).
+Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 5, 6, 10, and 11), *Alistipes* (OTU 12), and *Bacteroidales* (OTU 17).
 In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
@@
 Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
-Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
+Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 5, and 6) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
 Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
@@
 For example, the relationship between the *Bacteroides* (OTU 1) and *C.
-difficile* colonization is positive in streptomycin-treated mice and it is negative in cefoperazone-treated mice.
-In addition, cefoperazone and streptomycin-treated mice have high levels of *C.
-difficile* although the former have significantly higher levels of *Escherichia* (OTU 4), which are absent in the streptomycin-treated mice.
+difficile* colonization was positive in streptomycin-treated mice and it was negative in cefoperazone-treated mice.
+In addition, cefoperazone and streptomycin-treated mice had high levels of *C.
+difficile* although the former had significantly higher levels of *Escherichia* (OTU 3), which were absent in the streptomycin-treated mice.
 Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance.
@@
 
-**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hours before challenging them with *C.
+**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we allowed the gut microbiota to recover for 24 hours before challenging them with *C.
 difficile*.
 Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C.
-difficile* is restored {Reeves, 2011 \#118} {Theriot, 2014 \#956}.
-To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (Figure S3).
-Among the metronidazole-treated mice, those with extended recovery had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**).
-In addition, of the `r nrow(metro_delay)` mice given the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C.
+difficile* is restored {Reeves, 2011 \#118; Theriot, 2014 \#956}.
+To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**).
+Among the metronidazole-treated mice, those with extended recovery had a $`r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`$-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**).
+In addition, `r sum(metro_delay$CFU==0)` of the `r nrow(metro_delay)` mice given the longer recovery period had no detectable *C.
 difficile* 24 hours after challenge.
-We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 3).
+We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**).
 Most notable among these was a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3).
@@
 As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3).
-However untreated, fully resistant mice harbor significantly lower levels of *Barnesiella* OTU 2.
-Rather, untreated mice have high levels of various *Porphyromonadaceae* OTUs (**Figure 1**).
+However untreated, fully resistant mice harbored significantly lower levels of *Barnesiella* (OTU 2).
+Rather, untreated mice had high levels of various *Porphyromonadaceae* OTUs (**Figure 1**).
 These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.
@@
 difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
-Among various bacterial families the *Porphyromonadaceae* (ρ<sub>average<\sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) and *Ruminococcaceae* (ρ<sub>average<\sub>=`r ave_corr_family["Ruminococcaceae"]`, N=`r N_family["Ruminococcaceae"]` OTUs) were consistently associated with low levels of *C.
+The *Porphyromonadaceae* (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C.
 difficile* colonization.
-Two OTUs from the *Proteobacteria* had a significant association with *C.
+Among the three OTUs from the *Proteobacteria* had a significant positive association with *C.
+difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
+By performing an OTU-based analysis we were able to observe intra-family and genus differences in association with *C.
 difficile* colonization.
-These included OTUs associated with the genera (ρ=`r round(corr_data[families=="Pseudomonadaceae","sig_corrs"], 2)`) and *Escherichia* (ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
-These associations have been suggested previously {Reeves, 2011 \#118;Schubert, 2014 \#1048;Bassis, 2014 \#945;Buffie, 2012 \#977}.
-In addition, by performing an OTU-based analysis we observed several bacterial families that were represented by OTUs that were associated with low and high levels of *C.
-difficile* colonization.
-For example, the Lachnospiraceae have been associated with protection against C.
-difficile and although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
+For example, the Lachnospiraceae have been associated with protection against *C.
+difficile* and although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
 difficile* and `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU that was associated with high levels of *C.
@@
 The broad taxonomic representation of OTUs associated with low levels of *C.
-difficile* suggests that a diverse community is required to prevent the colonization *C.
+difficile* again suggests that a diverse community is required to prevent the colonization *C.
 difficile*.
 
-**The composition of the disturbed gut microbiota is predictive of C.
-difficile colonization levels.** The three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C.
-difficile* colonization, but in other contexts they were not.
+**The composition of the disturbed gut microbiota is predictive of** ***C.
+difficile*** **colonization levels.** These three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C.
+difficile* colonization, but in other contexts those OTUs had the opposite or not association.
+This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization.
 Correlation-based analyses cannot predict these types of context dependencies because they do not take into account the non-linearity and statistical interactions between populations.
-This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization.
 Therefore, we used a regression-based random forest machine learning algorithm to predict the level of *C.
-difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge.
+difficile* colonization observed in the three sets of experiments using the composition of the microbiota at the time of challenge as predictor variables.
 The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C.
-difficile* colonization levels (Figures 5 and S4).
-When we only included the top 7 OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C.
+difficile* colonization levels (Figures 5).
+When we only included the top `r n_part_features` OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C.
 difficile* colonization levels.
-Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia* (**Figure 6**).
-These results further validate the observations from the correlation-based analysis.
-Together these results suggest that colonization resistance is likely conferred by *Porphyromonadaceae* (OTU 15, 10, and 6), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), and *Alistipes* (OTU 12).
-A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), can result in increased susceptibility to infection (Figure 6).
+The OTUs that were ranked as being the most important in defining the random forest model, further validated the observations from the correlation-based analysis (**Figures S4**).
+According to the random forest model, colonization resistance was associated with members of the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**).
+A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), was associated with an increased susceptibility to infection (**Figure 6**).
 As we observed in the titration experiments, the relationship between *Akkermansia* (OTU 4) and *C.
-difficile* colonization appears to be context dependent.
+difficile* indicated that wide variation in the relative abundance of *Akkermansia* was associated with varying levels of *C.
+difficile*.
 There were varying abundances of *Akkermansia* in mice regardless of the level of *C.
 difficile* colonization.
+Finally, as indicated by the number of OTUs with relative abundances below the limit of detection, those mice could harbor varying levels of *C.
+difficle*.
+These observations bolster the hypothesis that colonization is context dependent.
 
@@
 
-Previous attempts to study the role of the gut microbiota in colonization resistance against C.
-difficile infection have utilized a single perturbation to the community.
+Previous attempts to study the role of the gut microbiota in colonization resistance against *C.
+difficile* infection have utilized a single perturbation to the community.
 Here, we used 7 antibiotics from 6 classes that were given to mice in varying doses and with varying post-antibiotic recovery periods.
-The result was a combination of 15 different perturbations that allowed us to generate distinct community profiles that displayed varying susceptibility to *C.
+The result was a combination of 15 different perturbations and the non-pertrubed microbiota, which allowed us to generate distinct community profiles that displayed varying susceptibility to *C.
 difficile* colonization.
-Our findings demonstrated that colonization resistance is not driven by individual populations, but by a community of organisms.
+Our findings demonstrated that colonization resistance is not driven by individual populations, but by a consortium of organisms.
 Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are protective against *C.
-difficile* {Buffie, 2012 \#977}{Reeves, 2012 \#126}.
+difficile* {Buffie, 2012 \#977; Reeves, 2012 \#126}.
 Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for colonization resistance.
@@
 Using conventional mice, mixtures of bacteria rather than individual populations have been shown to restore colonization resistance and mediate clearance of *C.
-difficile* {Lawley, 2012 \#36}{Buffie, 2015 \#981}.
+difficile* {Lawley, 2012 \#36; Buffie, 2015 \#981}.
 Fecal transplants, which represent a diverse collection of bacterial populations, have been highly effective in treating humans with recurrent *C.
-difficile* {Kassam, 2013 \#1003}{Seekatz, 2014 \#1051}{Weingarden, 2015 \#1054}.
-By generating a diverse collection of communities that were challenged with C.
-difficile, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance.
+difficile* {Kassam, 2013 \#1003; Seekatz, 2014 \#1051; Weingarden, 2015 \#1054}.
+By generating a diverse collection of communities that were challenged with **C.
+difficile**, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance.
 These would include members of the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes*.
-[ It would be nice to comment on how these overlap with other cocktails ]
+Several of these populatins have been examined for their potential as a probiotic for preventing *C.
+difficile* infection.
+A 6-species cocktail, including isolates of *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, *Coriobacteriaceae*, *Staphylococcus*, and *Enterococcus*, successfully resolved CDI in mice {Lawley, 2012 #36}.
+In humans, *Lactobacillus*-based probiotics have been co-administered with antibiotics to deter the onset of antibiotic-associated diarrhea (AAD) and *C.
+difficile* infection {Pattani, 2013 #1004}.
+A more diverse probiotic contained 33 bacterial species included *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaeae*, *Eubacteriaceae*, and *Lactobacillus* isolates, successfully restored colonization resistance in recurrent *C.
+difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 #2}.
+Given this evidence, we feel confident that an effective probiotic community could be designed based on our findings to recover colonization resistance against **C.
+difficile**.
 
@@
 For instance, *Escherichia* was previously associated with infected individuals and *Lachnospiraceae*, *Ruminococcaceae*, and *Alistipes* were absent from infected individuals.
-The results of the current study suggest that it should be possible to model a patient’s risk of developing a *C.
+The overlap between the results from the current study and past human studies and the power of random forest models suggest that it should be possible to model a patient’s risk of developing a *C.
 difficile* infection based on their gut microbiota composition at admission.
@@
 
-**Antibiotic administration.** Mice were administered one of seven different antibiotics including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin.
+**Antibiotic administration.** Mice were administered one of seven different antibiotics including cefoperazone, vancomycin, metronidazole, streptomycin, ciprofloxacin, ampicillin, and clindamycin (**Table 1**).
 The route of administration depended on the specific antibiotic.
@@
 
-**C.
-difficile preparation and challenge.** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
+***C.
+difficile*** **preparation and challenge.** All antibiotic-treated mice were given 24 hours to recover with untreated drinking water prior to *C.
 difficile* challenge.
@@
 difficile* challenge {Sorg, 2009 \#1055}.
-Spores were stored long term at 4 °Celsius.
+Spores were stored long term at 4°C.
 On the day of challenge 10^3^ *C.
@@
 
-**DNA extraction and sequencing.** Total bacterial DNA was extracted from each day 0 stool sample using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
+**DNA extraction and sequencing.** Total bacterial DNA was extracted from each stool sample collected prior to challenge using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
 We generated amplicons of the V4 region within the 16S rRNA gene and sequenced the fragments using an Illumina MiSeq as previously described {Kozich, 2013 \#40}.
@@
 
-**Sequence curation.** These sequences were curated using mothur as previously described {Kozich, 2013 \#40; Schloss, 2009 \#812}.
+**Sequence curation.** These sequences were curated using mothur (v.1.35) as previously described {Kozich, 2013 \#40; Schloss, 2009 \#812}.
 Briefly, sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
@@
 Analysis of the mock community data indicated that the error rate following our curation procedure was `r round(error_rate, 2)`%.
-In order to avoid biases due to uneven sampling, samples were normalized to 1,625 sequences per samples {Schloss, 2011 \#163}.
 All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession XXXXXX.
 
-**Statistical analysis and modeling.** All analyses were conducted using R version 3.1.2.
-OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (39 OTUs).
-Comparison of bacterial levels among titration groups or delayed groups of the same antibiotic was performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
+**Statistical analysis and modeling.** Complete scripts for regenerating our analysis adn this paper are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015).
+Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#1039}.
+To avoid biases due to uneven sampling, the dissimiarlity matrices were calculated by rarefying the samples to 1,625 sequences per sample.
+We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 #788}.
+OTU-based analyses were performed using R (v.3.1.2).
+After subsampling the OTU frequency data to 1,625 sequences per sample, OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (N=`r n_full_features` OTUs).
+All OTU-by-OTU comparisons were performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
 Comparison of log (base 10) transformed *C.
@@
 difficile* CFU/g feces.
-All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05.
-Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#1061} .
-The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 39 OTUs as predictor variables.
-Complete analysis scripts are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015).
+All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#1041}.
+Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#1061}.
+The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the `r n_full_features` OTUs as predictor variables.
 
@@
 
-**Figure 1. Antibiotic treatments result in significant alterations to
-the structure of the microbiota and variation in colonization
-resistance.** Bars indicate the median percent relative abundance of
-those selected OTUs from all treatment groups on the day of *C.
-difficile* challenge. Stars along the x-axis indicate those OTUs that
-were significantly different from untreated mice for that antibiotic
-treatment. The error bars indicate the interquartile range. The median
-level *C. difficile* colonization found 24 hours post microbiota
-sampling and the number of animals in the treatment group is indicated
-in the top right for each treatment with the interquartile range in
-parentheses. The concentration next to the name of the antibiotic
-indicates the dose of the antibiotic that was given to the animals.
+**Figure 1.
+Antibiotic treatments result in significant alterations to the structure of the microbiota and variation in colonization resistance.** Bars indicate the median percent relative abundance of those selected OTUs from all treatment groups on the day of *C.
+difficile* challenge.
+Stars along the x-axis indicate those OTUs that were significantly different from untreated mice for that antibiotic treatment.
+The error bars indicate the interquartile range.
+The median level *C.
+difficile* colonization found 24 hours post microbiota sampling and the number of animals in the treatment group is indicated in the top right for each treatment with the interquartile range in parentheses.
+The concentration next to the name of the antibiotic indicates the dose of the antibiotic that was given to the animals.
 
-**Figure 2. Titration of antibiotic perturbations results in altered
-community structures and *C. difficile* colonization resistance.** Bars
-indicate the median percent relative abundance of those selected OTUs
-from all treatment groups on the day of *C. difficile* challenge. Stars
-along the x-axis indicate those OTUs that varied significantly across
-doses of the same antibiotic. The error bars indicate the interquartile
-range. The median level *C. difficile* colonization found 24 hours post
-microbiota sampling is plotted on the right for each treatment with
-error bars indicating the interquartile range. The number of animals
-used in each treatment group is indicated in the legend, which depicts
-the doses of each antibiotic that were used.
+**Figure 2.
+Titration of antibiotic perturbations results in altered community structures and** ***C.
+difficile*** **colonization resistance.** Bars indicate the median percent relative abundance of those selected OTUs from all treatment groups on the day of *C.
+difficile* challenge.
+Stars along the x-axis indicate those OTUs that varied significantly across doses of the same antibiotic.
+The error bars indicate the interquartile range.
+The median level *C.
+difficile* colonization found 24 hours post microbiota sampling is plotted on the right for each treatment with error bars indicating the interquartile range.
+The number of animals used in each treatment group is indicated in the legend, which depicts the doses of each antibiotic that were used.
 
-**Figure 3. Increasing the recovery time following antibiotic
-perturbation restores colonization resistance.** Bars indicate the
-median percent relative abundance of those selected OTUs from all
-treatment groups on the day of *C. difficile* challenge. Stars along the
-x-axis indicate those OTUs that varied significantly between those mice
-that were allowed 1 or 6 days of recovery. The error bars indicate the
-interquartile range. The median level *C. difficile* colonization found
-24 hours post microbiota sampling is plotted on the right for each
-recovery period and antibiotic with error bars indicating the
-interquartile range. The number of mice used in each treatment group is
-indicated above the *C. difficile* colonization data. The dose of each
-antibiotic is indicated next to the name of the antibiotic.
+**Figure 3.
+Increasing the recovery time following antibiotic perturbation restores colonization resistance.** Bars indicate the median percent relative abundance of those selected OTUs from all treatment groups on the day of *C.
+difficile* challenge.
+Stars along the x-axis indicate those OTUs that varied significantly between those mice that were allowed 1 or 6 days of recovery.
+The error bars indicate the interquartile range.
+The median level *C.
+difficile* colonization found 24 hours post microbiota sampling is plotted on the right for each recovery period and antibiotic with error bars indicating the interquartile range.
+The number of mice used in each treatment group is indicated above the *C.
+difficile* colonization data.
+The dose of each antibiotic is indicated next to the name of the antibiotic.
 
-**Figure 4. Diverse taxonomic groups are associated with low levels of
-*C. difficile* colonization.** Spearman correlation coefficients were
-calculated using the relative abundance of OTUs found on the day that
-mice were challenged with *C. difficile* spores and the amount of *C.
-difficile* observed 24 hours later. Only significant correlations are
-presented. OTUs are grouped by the taxonomic family and the letters in
-the parentheses correspond to the phylum that the taxa belong to. B:
-Bacteroidetes, F: Firmicutes, P: Proteobacteria, A: Actinobacteria, T:
-Tenericutes.
+**Figure 4.
+Diverse taxonomic groups are associated with low levels of** ***C.
+difficile*** **colonization.** Spearman correlation coefficients were calculated using the relative abundance of OTUs found on the day that mice were challenged with *C.
+difficile* spores and the amount of *C.
+difficile* observed 24 hours later.
+Only significant correlations are presented.
+OTUs are grouped by the taxonomic family and the letters in the parentheses correspond to the phylum that the taxa belong to.
+B: Bacteroidetes, F: Firmicutes, P: Proteobacteria, A: Actinobacteria, T: Tenericutes.
 
-**Figure 5. Random forest regression model predicts *C. difficile*
-colonization levels based on the structure of the microbiota.** The
-overall model explained XX% of the variation in the data. Each pane
-shows antibiotic treatment groups in color and the other points as gray
-circles.
+**Figure 5.
+Random forest regression model predicts** ***C.
+difficile*** **colonization levels based on the structure of the microbiota.** The overall model explained `r round(rsq_full_features, 1)`% of the variation in the data.
+Each pane shows antibiotic treatment groups in color and the other points as gray circles.
 
-**Figure 6. Relationship between OTU relative abundance and *C.
-difficile* colonization levels indicates non-linearity and
-context-dependency.** The 9 OTUs that resulted in the greatest change in
-percent mean squared error when removed from the random forest
-regression model are shown in each pane and together explain XX.X% of
-the variation in the data. The Spearman correlation value between that
-OTUs abundance and C. difficile levels are shown for each pane. The
-color and symbols represent the same antibiotic dose and recovery period
-as in Figure 5.
+**Figure 6.
+Relationship between OTU relative abundance and** ***C.
+difficile*** **colonization levels indicates non-linearity and context-dependency.** The `r n_part_features` OTUs that resulted in the greatest change in percent mean squared error when removed from the random forest regression model are shown in each pane and together explain `r round(rsq_part_features, 1)`% of the variation in the data.
+The Spearman correlation value between that OTUs abundance and **C.
+difficile** levels are shown for each pane.
+The color and symbols represent the same antibiotic dose and recovery period as in Figure 5.
 
-**Figure S1. Effect of antibiotic perturbations on phylum-level
-representation of communities on day of *C. difficile* challenge.** Bars
-depict the median relative abundance across mice within the treatment
-group and error bars indicate the interquartile range.
+**Figure S1.
+Effect of antibiotic perturbations on phylum-level representation of communities on day of** ***C.
+difficile*** **challenge.** Bars depict the median relative abundance across mice within the treatment group and error bars indicate the interquartile range.
 
-**Figure S2. Effect of titrated antibiotic treatments on phylum-level
-representation of communities on day of *C. difficile* challenge.** Bars
-depict the median relative abundance across mice within the treatment
-group and error bars indicate the interquartile range.
+**Figure S2.
+Effect of titrated antibiotic treatments on phylum-level representation of communities on day of** ***C.
+difficile*** **challenge.** Bars depict the median relative abundance across mice within the treatment group and error bars indicate the interquartile range.
 
-**Figure S3. Effect of recovery period following antibiotic treatments
-on phylum-level representation of communities on day of *C. difficile*
-challenge.** Bars depict the median relative abundance across mice
-within the treatment group and error bars indicate the interquartile
-range.
+**Figure S3.
+Effect of recovery period following antibiotic treatments on phylum-level representation of communities on day of** ***C.
+difficile*** **challenge.** Bars depict the median relative abundance across mice within the treatment group and error bars indicate the interquartile range.
 
-**Figure S4. The change in percent mean squared error when each OTU was
-removed from the random forest regression model.**
+**Figure S4. The change in percent mean squared error when each OTU was removed from the random forest regression model.**

### 2015-04-10 Schubert_abxD01_mBio_2015.Rmd
Format document

@@
 difficile* challenge to identify populations within the microbiota that conferred colonization resistance.
-All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, , P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`).
+All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`).
 Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`).
@@
 difficile* {Kassam, 2013 \#1003; Seekatz, 2014 \#1051; Weingarden, 2015 \#1054}.
-By generating a diverse collection of communities that were challenged with **C.
-difficile**, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance.
+By generating a diverse collection of communities that were challenged with *C.
+difficile*, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance.
 These would include members of the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes*.
@@
 difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 #2}.
-Given this evidence, we feel confident that an effective probiotic community could be designed based on our findings to recover colonization resistance against **C.
-difficile**.
+Given this evidence, we feel confident that an effective probiotic community could be designed based on our findings to recover colonization resistance against *C.
+difficile*.
 
@@
 |:---------:|:--------------:|:-----:|:---------:|:---------:|
-Ciprofloxacin | Oral gavage, one time | Fluoroquinolone | Inhibits DNA gyrase | Gram +/-
-Clindamycin | IP injection, one time | Lincosamide | Inhibits protein synthesis | Anaerobes
-Vancomycin | *Ad libitum* in drinking water, five days | Glycopeptide | Inhibits peptidoglycan synthesis | Gram +
-Streptomycin | *Ad libitum* in drinking water, five days | Aminoglycoside | Inhibits protein synthesis | Gram +/-
-Cefoperazone | *Ad libitum* in drinking water, five days | β-lactam: Cephalosporin | Inhibits peptidoglycan synthesis | Gram +/-
-Ampicillin | *Ad libitum* in drinking water, five days | β-lactam: Penicillin | Inhibits peptidoglycan synthesis | Gram +/-
-Metronidazole | *Ad libitum* in drinking water, five days | Nitromidazole | Destabilizes bacterial DNA | Anaerobes
+| Ciprofloxacin | Oral gavage, one time | Fluoroquinolone | Inhibits DNA gyrase | Gram +/- |
+| Clindamycin | IP injection, one time | Lincosamide | Inhibits protein synthesis | Anaerobes |
+| Vancomycin | *Ad libitum* in drinking water, five days | Glycopeptide | Inhibits peptidoglycan synthesis | Gram + |
+| Streptomycin | *Ad libitum* in drinking water, five days | Aminoglycoside | Inhibits protein synthesis | Gram +/- |
+| Cefoperazone | *Ad libitum* in drinking water, five days | β-lactam: Cephalosporin | Inhibits peptidoglycan synthesis | Gram +/- |
+| Ampicillin | *Ad libitum* in drinking water, five days | β-lactam: Penicillin | Inhibits peptidoglycan synthesis | Gram +/- |
+| Metronidazole | *Ad libitum* in drinking water, five days | Nitromidazole | Destabilizes bacterial DNA | Anaerobes |
 
@@
 difficile*** **colonization levels indicates non-linearity and context-dependency.** The `r n_part_features` OTUs that resulted in the greatest change in percent mean squared error when removed from the random forest regression model are shown in each pane and together explain `r round(rsq_part_features, 1)`% of the variation in the data.
-The Spearman correlation value between that OTUs abundance and **C.
-difficile** levels are shown for each pane.
+The Spearman correlation value between that OTUs abundance and *C.
+difficile* levels are shown for each pane.
 The color and symbols represent the same antibiotic dose and recovery period as in Figure 5.

### 2015-04-21 Schubert_abxD01_mBio_2015.Rmd
Incorporate edits from Vince and others

@@
 difficile* infection is commonly associated with antibiotic use, the precise alterations to the microbiota associated with this loss in function are unknown.
-We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which we then challenged with *C.
+We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which were then challenged with *C.
 difficile* spores.
@@
 difficile*.
-Rather, resistance and sensitivity can be achieved with multiple diverse assemblages.
+Rather, multiple diverse assemblages act in concert to mediate colonization resistance.
 
@@
 
-The microbiota, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection {Vollaard, 1994 \#683}.
-This native protection by the microbiota from invasive pathogenic species is termed colonization resistance.
-It is well established that the gut bacterial microbiota is critical in the host's defense against the pathogen *Clostridium difficile* {Wilson, 1985 \#1045; Chang, 2008 \#590;Reeves, 2011 \#118}.
-When this indigenous community is perturbed, this often leads to a loss of resistance.
-This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, and chemotherapy, but they are also exposed to *C.
-difficile* spores from their environment {Bauer, 2015 \#1047}.
-*C.
-difficile* infections (CDI) are the most reported hospital-acquired infection in the United States and are responsible for 14,000 deaths a year {Lessa, 2015 \#950}.
+The microbiota, or the diverse community of microorganisms living in and
+on the body, has an integral role in deterring pathogen colonization and
+infection {Vollaard, 1994 \#1}. This native protection by the microbiota
+from invasive pathogenic species is termed colonization resistance. It
+is well established that the gut bacterial microbiota is critical in the
+host's defense against the pathogen *Clostridium difficile* {Chang, 2008
+\#3;Reeves, 2011 \#4;Wilson, 1985 \#2}. Perturbations to this indigenous
+community often lead to a loss of resistance. This is especially
+important in many hospital settings where patients are not only exposed
+to various types and degrees of perturbations, such as antibiotics, diet
+changes, and chemotherapy, but they are also exposed to *C. difficile*
+spores from their environment {Bauer, 2015 \#5}. *C. difficile*
+infections (CDI) are the most reported hospital-acquired infection in
+the United States and are responsible for 14,000 deaths a year {Lessa,
+2015 \#6}.
 
-It is not completely understood how different perturbations to the gut microbiota result in a loss of colonization resistance to *C.
-difficile*.
-There is a clear need to better understand the ecology of *C.
-difficile* and its interactions with members of the microbiota.
-In mouse models of CDI, the perturbed, untreated murine microbiota is completely resistant to *C.
-difficile* colonization.
-It was previously shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011 \#118; Theriot, 2011 \#79}, tigecycline {Bassis, 2014 \#945}, clindamycin {Buffie, 2012 \#977}, or clindamycin in combination with a five antibiotic cocktail {Chang, 2008 \#590} had decreased colonization resistance.
-These studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a bloom of *Lactobacilliaceae* and *Enterobacteriaceae* are responsible for the loss of colonization resistance.
-These results are largely supported by human association studies {Vincent, 2013 \#37; Schubert, 2014 \#1048}.
-We observed significant differences between the gut microbiota of hospitalized individuals with and without *C.
-difficile* and non-hospitalized controls {Schubert, 2014 \#1048}.
-In addition, fecal microbiota transplants increase *Bacteroidetes* and decrease *Proteobacteria* levels in recipients, resulting in a successful restoration of colonization resistance in patients {Seekatz, 2014 \#1051}.
-Precisely how this occurs is not fully understood, but it emphasizes the importance of the gut microbiota in colonization resistance against *C.
+It is not completely understood how different perturbations to the gut
+microbiota result in a loss of colonization resistance to *C.
+difficile*. There is a clear need to better understand the ecology of
+*C. difficile* and its interactions with members of the microbiota. In
+mouse models of CDI, the unperturbed, untreated murine microbiome is
+completely resistant to *C. difficile* colonization. It was previously
+shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011
+\#4;Theriot, 2011 \#7}, tigecycline {Bassis, 2014 \#8}, clindamycin
+{Buffie, 2012 \#9}, or clindamycin in combination with a five antibiotic
+cocktail {Chang, 2008 \#3} had decreased colonization resistance. These
+studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a
+bloom of *Lactobacillaceae* and *Enterobacteriaceae* are responsible for
+the loss of colonization resistance. These results are largely supported
+by human association studies {Schubert, 2014 \#11;Vincent, 2013 \#10}.
+We observed significant differences between the gut microbiota of
+hospitalized individuals with and without *C. difficile* and
+non-hospitalized controls {Schubert, 2014 \#11}. In addition, fecal
+microbiota transplants increase *Bacteroidetes* and decrease
+*Proteobacteria* levels in recipients, resulting in a successful
+restoration of colonization resistance in patients {Seekatz, 2014 \#12}.
+Precisely how this occurs is not fully understood, but it emphasizes the
+importance of the gut microbiota in colonization resistance against *C.
 difficile*.
 
-Because the gut microbiota is a complex community we need tools that can enable us to dissect the interactions within the community and with *C.
-difficile*.
-One approach is the use of models to identify associations between members of the microbiota and *C.
-difficile*.
-Models have been used to predict *C.
-difficile* {Stein, 2013 \#988; Schubert, 2014 \#1048} and *Citrobacter* infection {Belzer, 2014 \#1053}, colon cancer {Zackular, 2013 \#472}, and psoriasis {Statnikov, 2013 \#477} based on the composition of the gut microbiota.
-By using models to explain the relationship between members of the gut microbiota, we similarly sought to identify the subset of the normal murine microbiota that are responsible for colonization resistance.
+Because the gut microbiota is a complex community we need tools that
+enable us to dissect the interactions within the community and with *C.
+difficile*. One approach is the use of mathematical models to identify
+associations between members of the microbiota and *C. difficile*.
+Mathematical models have been used to predict *C. difficile* {Schubert,
+2014 \#11;Stein, 2013 \#13} and *Citrobacter* infection {Belzer, 2014
+\#14}, colon cancer {Zackular, 2013 \#15}, and psoriasis {Statnikov,
+2013 \#16} based on the composition of the gut microbiota. We similarly
+sought to identify the subset of the normal murine microbiota that are
+responsible for colonization resistance by using mathematical models to
+explain the relationship between members of the gut microbiota.
 
-The purpose of this investigation was to expand our current knowledge of the effects of various perturbations on colonization resistance against *C.
-difficile*.
-Through the administration of different antibiotic classes, doses, and recovery times we altered the murine gut microbiota and challenged the communities with *C.
-difficile* spores and to quantify differences in colonization resistance.
-We then used 16S rRNA gene sequencing to identify structural changes within the microbiota that would be predictive of colonization resistance.
-Using these data, we built a random forest model to predict *C.
-difficile* colonization levels.
-Through this analysis, we have identified groups of related bacteria that are associated with *C.
-difficile* colonization resistance.
-This model revealed that the interactions that gave rise to colonization resistance were non-linear and context dependent.
-These findings show we can successfully apply modeling techniques to accurately measure the colonization resistance ability of a given microbiota.
+The purpose of this investigation was to expand our current knowledge of
+the effects of various perturbations on colonization resistance against
+*C. difficile*. Through the administration of different antibiotic
+classes, doses, and recovery times we altered the murine gut microbiota
+and challenged the communities with *C. difficile* spores to quantify
+differences in colonization resistance. We then used 16S rRNA gene
+sequencing to identify structural changes within the microbiota that
+would be predictive of colonization resistance. Using these data, we
+built a random forest model to predict *C. difficile* colonization
+levels. Through this analysis, we have identified groups of related
+bacteria that are associated with *C. difficile* colonization
+resistance. This model revealed that the interactions giving rise to
+colonization resistance were non-linear and context dependent. These
+findings show we can successfully apply modeling techniques to
+accurately measure the colonization resistance ability of a given
+microbiota.
 
@@
 difficile* colonization (**Table 1**).
-On the day after mice were challenged with *C.
-difficile* we enumerated the density of *C.
+Following the cessation of antibiotics, each treatment group was given one day of recovery prior to challenge with *C.
+difficile* spores.
+One day post challenge we enumerated the density of *C.
 difficile* in the animals' feces.
@@
 difficile* colonization in mice treated with ampicillin.
+None of the mice that received ciprofloxacin were colonized.
+In addition to administering ciprofloxacin by oral gavage, we provided ciprofloxacin by intraperitoneal injection (10 mg/mL).
+For both approaches we provided one or two days of recovery.
+Regardless of the method, the resulting communities were resistant to *C.
+difficile* colonization.
 Only one of six mice receiving vancomycin was colonized with *C.
-difficile* and none of the mice that received ciprofloxacin were colonized.
-We sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C.
+difficile.* We suspected that this was due to residual vancomycin repressing *C.
+difficile* growth.
+In fact, two days post *C.
+difficile* challenge, *C.
+difficile* bloomed in these mice to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces.
+Furthermore, given two days of post-vancomycin recovery, there was no delay in *C.
+difficile* colonization to high levels, and on day one post challenge we observed a median of 3.0x10^7^ (interquartile range 2.6x10^7^–3.6x10^7^, N=4) CFU/g feces.
+These results suggest that although vancomycin is not absorbed by the gut tissue, the absence of *C.
+difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment.
+Overall, the various antibiotic perturbations provided varying levels of colonization by *C.
+difficile,* which suggested that the resulting communities varied in their composition.
+
+To test this hypothesis, we sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C.
 difficile* challenge to identify populations within the microbiota that conferred colonization resistance.
@@
 
-**Reduced perturbations result in altered levels of colonization.** Based on the *C.
-difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C.
-difficile*.
+**Reduced perturbations result in altered levels of colonization.** Based on the C.
+difficile colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota.
+Consequently we expected a greater maintenance of resistance against *C.
+difficile* colonization in these titrated treatment groups.
 In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (Figure S2).
@@
 vancomycin) bacteria.
-As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2).
+As expected in all mice receiving titrated doses of cefoperazone, *C.
+difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2).
 Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs.
@@
 Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 5, 6, 10, and 11), *Alistipes* (OTU 12), and *Bacteroidales* (OTU 17).
-In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse.
+In addition to its anti-Gram-positive activity, vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of *C.
+difficile* colonization in only one mouse.
 We anticipated that lower doses might result in a community structure that would result in colonization.
-In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`)).
+In fact, the 0.3 and 0.1 mg/mL doses of vancomycin resulted in similarly high levels of *C.
+difficile* colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`)).
 Seven OTUs were differentially represented across the three doses of vancomycin.
-Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
+Surprisingly, even though the colonization levels of *C.
+difficile* did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose.
 Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 5, and 6) and one was affiliated with a member of the genus *Bacteroides* (OTU 1).
 Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level.
-These results suggest that the context of the microbiota is important in determining the overall resistance to *C.
+These results suggest that the context in which specific members of the microbiota are found is important in determining the overall resistance to *C.
 difficile*.
@@
 difficile* colonization.
-Among the three OTUs from the *Proteobacteria* had a significant positive association with *C.
+Among the three *Proteobacteria* OTUS with a significant positive association with *C.
 difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
@@
 For example, the Lachnospiraceae have been associated with protection against *C.
-difficile* and although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
-difficile* and `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU that was associated with high levels of *C.
 difficile*.
-In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` *Lactobacillus* OTUs from the *Lactobacilliaceae* where `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` were associated with low levels of *C.
+Although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
+difficile* colonization, `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU was associated with high levels of *C.
+difficile*.
+In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus* OTUs (family *Lactobacilliaceae*), `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` of which were associated with low levels of *C.
 difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels.
 The broad taxonomic representation of OTUs associated with low levels of *C.
-difficile* again suggests that a diverse community is required to prevent the colonization *C.
+difficile* again suggests that a functionally diverse community is required to prevent the colonization *C.
 difficile*.
@@
 difficile*** **colonization levels.** These three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C.
-difficile* colonization, but in other contexts those OTUs had the opposite or not association.
+difficile* colonization, but in other contexts those OTUs had the opposite or no association.
 This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization.
@@
 The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C.
-difficile* colonization levels (Figures 5).
+difficile* colonization levels (**Figures 5**).
 When we only included the top `r n_part_features` OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C.
@@
 difficle*.
-These observations bolster the hypothesis that colonization is context dependent.
+These observations bolster the hypothesis that colonization resistance is context dependent.
 
@@
 difficile* infection have utilized a single perturbation to the community.
-Here, we used 7 antibiotics from 6 classes that were given to mice in varying doses and with varying post-antibiotic recovery periods.
-The result was a combination of 15 different perturbations and the non-pertrubed microbiota, which allowed us to generate distinct community profiles that displayed varying susceptibility to *C.
+Here, we used seven antibiotics from six classes that were given to mice in varying doses and with varying post-antibiotic recovery periods.
+The result was a combination of 15 different perturbations and the non-perturbed microbiota, which allowed us to generate distinct community profiles that displayed varying susceptibility to *C.
 difficile* colonization.
@@
 Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are protective against *C.
-difficile* {Buffie, 2012 \#977; Reeves, 2012 \#126}.
+difficile* {Buffie, 2012 \#9;Reeves, 2012 \#18}.
 Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for colonization resistance.
-Overall, these results suggest that *C.
-difficile* is a generalist that is capable of exploiting a variety of niches in the gastrointestinal tract.
+Thus colonization resistance can be described as an emergent property of the microbiome, in which individual bacterial populations integrated in a community contribute to the overall resistance to *C.
+difficile* {Novikoff, 1945 \#34}.
 
@@
 Current first line treatments of CDI include regimens of either metronidazole or vancomycin, which further perturb the microbiota.
-As such, relapse rates for CDI are typically around 25-30% {Wilcox, 1998 \#599}.
+As such, relapse rates for CDI are typically around 25-30% {Wilcox, 1998 \#19}.
 Interestingly, we observed that treatment with either antibiotic induced susceptibility to *C.
@@
 For instance, association of germ-free mice with a *Lachnospiraceae* isolate only reduced the level of *C.
-difficile* colonization by 10 to 100-fold {Reeves, 2012 \#126}.
+difficile* colonization by 10 to 100-fold {Reeves, 2012 \#18}.
 Using conventional mice, mixtures of bacteria rather than individual populations have been shown to restore colonization resistance and mediate clearance of *C.
-difficile* {Lawley, 2012 \#36; Buffie, 2015 \#981}.
+difficile* {Buffie, 2015 \#21;Lawley, 2012 \#20}.
 Fecal transplants, which represent a diverse collection of bacterial populations, have been highly effective in treating humans with recurrent *C.
-difficile* {Kassam, 2013 \#1003; Seekatz, 2014 \#1051; Weingarden, 2015 \#1054}.
+difficile* {Kassam, 2013 \#22;Seekatz, 2014 \#12;Weingarden, 2015 \#23}.
 By generating a diverse collection of communities that were challenged with *C.
@@
 These would include members of the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes*.
-Several of these populatins have been examined for their potential as a probiotic for preventing *C.
+Several of these populations have been examined for their potential as a probiotic for preventing *C.
 difficile* infection.
-A 6-species cocktail, including isolates of *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, *Coriobacteriaceae*, *Staphylococcus*, and *Enterococcus*, successfully resolved CDI in mice {Lawley, 2012 #36}.
+A 6-species cocktail, including isolates of *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, *Coriobacteriaceae*, *Staphylococcus*, and *Enterococcus*, successfully resolved CDI in mice {Lawley, 2012 \#20}.
 In humans, *Lactobacillus*-based probiotics have been co-administered with antibiotics to deter the onset of antibiotic-associated diarrhea (AAD) and *C.
-difficile* infection {Pattani, 2013 #1004}.
-A more diverse probiotic contained 33 bacterial species included *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaeae*, *Eubacteriaceae*, and *Lactobacillus* isolates, successfully restored colonization resistance in recurrent *C.
-difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 #2}.
+difficile* infection {Lawley, 2012 \#20}.
+A more diverse probiotic, which contained 33 bacterial species including *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, *Eubacteriaceae*, and *Lactobacillus* isolates, successfully restored colonization resistance in recurrent *C.
+difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 \#25}.
 Given this evidence, we feel confident that an effective probiotic community could be designed based on our findings to recover colonization resistance against *C.
@@
 Although the murine and human microbiota do not fully overlap, our previous analysis of humans infected with *C.
-difficile* supports the populations that we associated with colonization {Schubert, 2014 \#1048}.
+difficile* supports the populations that we associated with colonization {Schubert, 2014 \#11}.
 For instance, *Escherichia* was previously associated with infected individuals and *Lachnospiraceae*, *Ruminococcaceae*, and *Alistipes* were absent from infected individuals.
-The overlap between the results from the current study and past human studies and the power of random forest models suggest that it should be possible to model a patient’s risk of developing a *C.
+The overlap between the results from the current study and past human studies along with the power of random forest models suggest that it should be possible to model a patient’s risk of developing a *C.
 difficile* infection based on their gut microbiota composition at admission.
+Overall these findings demonstrate the significance of the microbiota as an interconnected bacterial community in assessing resistance to pathogen colonization.
 
@@
 Spores were prepared from a single large batch whose concentration was determined within the week prior to each *C.
-difficile* challenge {Sorg, 2009 \#1055}.
+difficile* challenge {Sorg, 2009 \#26}.
 Spores were stored long term at 4°C.
@@
 On the day after the challenge another fecal sample was weighed and diluted under anaerobic conditions with anaerobic PBS.
-The number of colony forming units (CFU) were counted following 24 hours growth on TCCFA plates at 37°C under anaerobic conditions {Buggy, 1983 \#1059}.
+The number of colony forming units (CFU) were counted following 24 hours growth on TCCFA plates at 37°C under anaerobic conditions {Buggy, 1983 \#27}.
 
 **DNA extraction and sequencing.** Total bacterial DNA was extracted from each stool sample collected prior to challenge using the MOBIO PowerSoil®-htp 96 Well Soil DNA Isolation Kit.
-We generated amplicons of the V4 region within the 16S rRNA gene and sequenced the fragments using an Illumina MiSeq as previously described {Kozich, 2013 \#40}.
+We generated amplicons of the V4 region within the 16S rRNA gene and sequenced the fragments using an Illumina MiSeq as previously described {Kozich, 2013 \#28}.
 
@@
 
-**Sequence curation.** These sequences were curated using mothur (v.1.35) as previously described {Kozich, 2013 \#40; Schloss, 2009 \#812}.
+**Sequence curation.** These sequences were curated using mothur (v.1.35) as previously described {Kozich, 2013 \#28;Schloss, 2009 \#29}.
 Briefly, sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
-Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 10) requiring an 80% bootstrap confidence score.
+Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 10) requiring an 80% bootstrap confidence score {Wang, 2007 \#1809}.
 In parallel to the fecal samples, we also sequenced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
 Analysis of the mock community data indicated that the error rate following our curation procedure was `r round(error_rate, 2)`%.
-All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession XXXXXX.
+All 16S rRNA gene sequence data and metadata are available through the Sequence Read Archive under accession SRP057386.
 
-**Statistical analysis and modeling.** Complete scripts for regenerating our analysis adn this paper are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015).
-Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#1039}.
-To avoid biases due to uneven sampling, the dissimiarlity matrices were calculated by rarefying the samples to 1,625 sequences per sample.
-We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 #788}.
+**Statistical analysis and modeling.** Complete scripts for regenerating our analysis and this paper are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015).
+Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#1061}.
+To avoid biases due to uneven sampling, the dissimilarity matrices were calculated by rarefying the samples to 1,625 sequences per sample.
+We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 \#804}.
 OTU-based analyses were performed using R (v.3.1.2).
-After subsampling the OTU frequency data to 1,625 sequences per sample, OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (N=`r n_full_features` OTUs).
-All OTU-by-OTU comparisons were performed using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
+After subsampling the OTU frequency data to 1,625 sequences per sample, OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (N=38 OTUs).
+All OTU-by-OTU comparisons were performed using the Kruskal-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
 Comparison of log (base 10) transformed *C.
-difficile* CFU/g feces between experimental groups was calculated using the Kruskall-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
+difficile* CFU/g feces between experimental groups was calculated using the Kruskal-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests.
 Spearman rank correlation analysis was performed between OTU counts and *C.
 difficile* CFU/g feces.
-All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#1041}.
-Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#1061}.
-The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the `r n_full_features` OTUs as predictor variables.
+All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#604}.
+Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}.
+The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 38 OTUs as predictor variables.
 
@@
 
-This work was supported by several grants from the National Institutes
-for Health R01GM099514, R01HG005975, U19AI090871, and P30DK034933. The
-funding agencies had no role in study design, data collection and
-analysis, decision to publish, or preparation of the manuscript.
+We thank Dr.
+Vincent Young for providing a critical review of an earlier version of this manuscript.
+This work was supported by several grants from the National Institutes for Health R01GM099514, R01HG005975, U19AI090871, and P30DK034933.
+The funding agencies had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 

### 2015-04-21 Schubert_abxD01_mBio_2015.Rmd
Minor edits

@@
 difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner.
-Using relative abundance data, we built machine learning regression models to predict the levels of *C.
+Using relative abundance data, we built a machine learning regression model to predict the levels of *C.
 difficile* that were found 24 hours after challenging the perturbed communities.
@@
 would be predictive of colonization resistance. Using these data, we
-built a random forest model to predict *C. difficile* colonization
+built a random forest regression model to predict *C. difficile* colonization
 levels. Through this analysis, we have identified groups of related
@@
 colonization resistance were non-linear and context dependent. These
-findings show we can successfully apply modeling techniques to
-accurately measure the colonization resistance ability of a given
-microbiota.
+findings show we can successfully apply modeling techniques to accurately
+predict the colonization resistance of a given microbiota..
 
@@
 
-**Reduced perturbations result in altered levels of colonization.** Based on the C.
-difficile colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota.
+**Reduced perturbations result in altered levels of colonization.** Based on the *C.
+difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota.
 Consequently we expected a greater maintenance of resistance against *C.
@@
 In fact, the 0.3 and 0.1 mg/mL doses of vancomycin resulted in similarly high levels of *C.
-difficile* colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`)).
+difficile* colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`).
 Seven OTUs were differentially represented across the three doses of vancomycin.
@@
 **Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C.
-difficile*, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C.
+difficile* across the three sets of experiments, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C.
 difficile* colonization 24 hours later (**Figure 4**).
@@
 difficile* colonization.
-Among the three *Proteobacteria* OTUS with a significant positive association with *C.
+Among the three *Proteobacteria* OTUs with a significant positive association with *C.
 difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
@@
 difficile* colonization.
-For example, the Lachnospiraceae have been associated with protection against *C.
+For example, the *Lachnospiraceae* have been associated with protection against *C.
 difficile*.
@@
 difficile* colonization levels.
-The OTUs that were ranked as being the most important in defining the random forest model, further validated the observations from the correlation-based analysis (**Figures S4**).
+The OTUs that were ranked as being the most important in defining the random forest model further validated the observations from the correlation-based analysis (**Figures S4**).
 According to the random forest model, colonization resistance was associated with members of the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**).
@@
 Finally, as indicated by the number of OTUs with relative abundances below the limit of detection, those mice could harbor varying levels of *C.
-difficle*.
+difficile*.
 These observations bolster the hypothesis that colonization resistance is context dependent.
@@
 difficile* colonization.
-Our findings demonstrated that colonization resistance is not driven by individual populations, but by a consortium of organisms.
-Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are protective against *C.
+Our findings demonstrated that colonization resistance was not driven by individual populations, but by a consortium of organisms.
+Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are partially protective against *C.
 difficile* {Buffie, 2012 \#9;Reeves, 2012 \#18}.
-Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for colonization resistance.
+Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for complete colonization resistance.
 Thus colonization resistance can be described as an emergent property of the microbiome, in which individual bacterial populations integrated in a community contribute to the overall resistance to *C.
@@
 Cefoperazone (0.5, 0.3, or 0.1 mg/ml), vancomycin (0.625, 0.3, or 0.1 mg/ml), streptomycin (5, 0.5, or 0.1 mg/ml), metronidazole (0.5 mg/ml), and ampicillin (0.5 mg/ml) were all administered in the mouse drinking water for 5 days.
-Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via IP injection.
+Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via intraperitoneal injection.
 Mice that did not receive antibiotics were used as negative controls for these experiments.
@@
 | Ciprofloxacin | Oral gavage, one time | Fluoroquinolone | Inhibits DNA gyrase | Gram +/- |
-| Clindamycin | IP injection, one time | Lincosamide | Inhibits protein synthesis | Anaerobes |
+| Clindamycin | Intraperitoneal injection, one time | Lincosamide | Inhibits protein synthesis | Anaerobes |
 | Vancomycin | *Ad libitum* in drinking water, five days | Glycopeptide | Inhibits peptidoglycan synthesis | Gram + |

### 2015-04-22 Schubert_abxD01_mBio_2015.Rmd
Incorporate Schubert minor edits

@@
 by human association studies {Schubert, 2014 \#11;Vincent, 2013 \#10}.
-We observed significant differences between the gut microbiota of
+We previously observed significant differences between the gut microbiota of
 hospitalized individuals with and without *C. difficile* and
@@
 difficile* colonization in these titrated treatment groups.
-In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (Figure S2).
+In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (*Figure S2*).
 These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e.
@@
 As expected in all mice receiving titrated doses of cefoperazone, *C.
-difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2).
+difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; *Figure 2*).
 Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs.
@@
 difficile*.
-In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus* OTUs (family *Lactobacilliaceae*), `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` of which were associated with low levels of *C.
+In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus* OTUs (family *Lactobacillaceae*), `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` of which were associated with low levels of *C.
 difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels.
@@
 Random forest regression models allowed us to describe community resistance as a byproduct of an assemblage of bacterial populations rather than as individual populations.
-A correlation-based analysis was unable to identify populations that had a context dependent or non-linear associations with *C.
+A correlation-based analysis was unable to identify populations that had a context dependent or non-linear association with *C.
 difficile* colonization.

### 2015-06-07 code/build_figure5_6_S4.R
Reorder plots and add binary model generation

@@ -79,7 +79,7 @@ n_trees <- 1e4
 set.seed("6201976")
 rf_full <- randomForest(logCFU ~ ., data=abund_good, importance=TRUE, ntree=n_trees)
 
-write(c(ncol(abund_good), rf_full$rsq[n_trees]), file="data/process/random_forest.data")
+write(paste("full", ncol(abund_good), rf_full$rsq[n_trees], sep="\t"), file="data/process/random_forest.data")
 
 # sort the features by their effect on % increase in the standard error
 importance_sorted <- sort(importance(rf_full)[,1], decreasing=T)
@@ -92,13 +92,17 @@ n_features <- 12
 importance_subset <- importance_sorted[1:n_features]
 
 
-
-
 # let's get Rsq for the subset
 set.seed("6201976")
 rf_partial <- randomForest(logCFU ~ ., data=abund_good[,names(importance_sorted)[1:n_features]], importance=TRUE, ntree=n_trees)
-write(c(n_features, rf_partial$rsq[n_trees]), file="data/process/random_forest.data", append=T)
+write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)
 
+
+
+colonized <- rep("yes", length(logCFU))
+colonized[logCFU<2] <- "no"
+binary_model <- randomForest(x=abund_good, y=as.factor(colonized), importance=TRUE, ntree=n_trees)
+write(paste("binary", ncol(abund_good), binary_model$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
 
 # supplemental figure 4: full feature importance plot
 tiff(file="results/figures/figureS4.tiff", width=4, height=5.0, unit="in", res=300)
@@ -111,7 +115,10 @@ tiff(file="results/figures/figureS4.tiff", width=4, height=5.0, unit="in", res=3
     points(x=rev(importance_sorted), y=1:length(importance_sorted), pch=19, cex=0.8)
     axis(1, at=seq(0,100,25), label=c("0", "", "50", "", "100"), cex=0.8)
     box()
[…]

### 2015-06-07 Schubert_abxD01_mBio_2015.Rmd
PDS incorporated AMS and reviewers comments

@@
 data <- read.table(file="data/process/random_forest.data")
-n_full_features <- data[1,1]
-rsq_full_features <- data[1,2] * 100
+n_full_features <- data[1,2]
+rsq_full_features <- data[1,3] * 100
 
-n_part_features <- data[2,1]
-rsq_part_features <- data[2,2] * 100
+n_part_features <- data[2,2]
+rsq_part_features <- data[2,3] * 100
+
+binary_error <- data[3,3] * 100
 ```
@@
 
-Perturbations to the gut microbiota result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*.
+Perturbations to the gut microbiota can result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*.
 Although *C.
@@
 non-hospitalized controls {Schubert, 2014 \#11}. In addition, fecal
-microbiota transplants increase *Bacteroidetes* and decrease
-*Proteobacteria* levels in recipients, resulting in a successful
+microbiota transplants have been shown to increase the relative abundance of *Bacteroidetes* and decrease the relative abundance of *Proteobacteria* and result in a successful
 restoration of colonization resistance in patients {Seekatz, 2014 \#12}.
-Precisely how this occurs is not fully understood, but it emphasizes the
-importance of the gut microbiota in colonization resistance against *C.
+The mechanisms involved in restoring colonization resistance is not fully understood, but it emphasizes the importance of the gut microbiota in protecting against *C.
 difficile*.
@@
 difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**).
-We observed variable levels of *C.
+We observed the most variance in levels of *C.
 difficile* colonization in mice treated with ampicillin.
@@
 In addition to administering ciprofloxacin by oral gavage, we provided ciprofloxacin by intraperitoneal injection (10 mg/mL).
-For both approaches we provided one or two days of recovery.
+For both approaches we provided one or two days of recovery from the antibiotic treatment.
 Regardless of the method, the resulting communities were resistant to *C.
@@
 difficile* challenge, *C.
-difficile* bloomed in these mice to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces.
+difficile* bloomed in this treatment group to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces.
 Furthermore, given two days of post-vancomycin recovery, there was no delay in *C.
@@
 All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`).
+The similarity in the structure of the microbiota in ciprofloxacin-treated and untreated mice suggests that a higher dose of ciprofloxacin may have been necessary to significantly perturb the microbiota to allow *C.
+difficile* to overcome colonization resistance.
 Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`).
@@
 difficile*.
-For example, the relationship between the *Bacteroides* (OTU 1) and *C.
+For example, the relationship between the *Bacteroides*-affiliated OTU (OTU 1) and *C.
 difficile* colonization was positive in streptomycin-treated mice and it was negative in cefoperazone-treated mice.
 In addition, cefoperazone and streptomycin-treated mice had high levels of *C.
-difficile* although the former had significantly higher levels of *Escherichia* (OTU 3), which were absent in the streptomycin-treated mice.
+difficile* although the former had significantly higher levels of an *Escherichia*-affiliated OTU (OTU 3), which were absent in the streptomycin-treated mice.
 Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance.
@@
 We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**).
-Most notable among these was a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3).
-The relative abundance of this *Barnesiella* OTU increased with the delay, and the relative abundance of this *Escherichia* OTU decreased.
+Most notable among these were two OTUs that affiliated with a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3).
+The relative abundance of this *Barnesiella*-affiliated OTU increased with the delay, and the relative abundance of this *Escherichia*-affiliated OTU decreased.
 Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C.
 difficile* colonization (P=`r round(amp_test$p.value, 2)`).
-As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3).
-However untreated, fully resistant mice harbored significantly lower levels of *Barnesiella* (OTU 2).
-Rather, untreated mice had high levels of various *Porphyromonadaceae* OTUs (**Figure 1**).
+As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3)-affiliated OTUs.
+However untreated, fully resistant mice harbored significantly lower levels of the *Barnesiella*-affiliated OTU (OTU 2).
+Rather, untreated mice had high levels of various *Porphyromonadaceae*-affiliated OTUs (**Figure 1**).
 These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.
@@
 difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
-The *Porphyromonadaceae* (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C.
+The *Porphyromonadaceae*-affiliated OTUs (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C.
 difficile* colonization.
-Among the three *Proteobacteria* OTUs with a significant positive association with *C.
+Among the three *Proteobacteria*-affiliated OTUs with a significant positive association with *C.
 difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`).
@@
 difficile*.
-Although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
+Although within the *Lachnospiraceae* familiy we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
 difficile* colonization, `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU was associated with high levels of *C.
 difficile*.
-In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus* OTUs (family *Lactobacillaceae*), `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` of which were associated with low levels of *C.
+In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*, `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)`) of which were associated with low levels of *C.
 difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels.
 The broad taxonomic representation of OTUs associated with low levels of *C.
-difficile* again suggests that a functionally diverse community is required to prevent the colonization *C.
-difficile*.
+difficile* again suggests that a diverse community may be advantageous in preventing *C.
+difficile* colonization.
 
@@
 The OTUs that were ranked as being the most important in defining the random forest model further validated the observations from the correlation-based analysis (**Figures S4**).
-According to the random forest model, colonization resistance was associated with members of the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**).
-A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), was associated with an increased susceptibility to infection (**Figure 6**).
-As we observed in the titration experiments, the relationship between *Akkermansia* (OTU 4) and *C.
+According to the random forest model, colonization resistance was associated with OTUs that affiliated with the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**).
+A loss in these populations, concurrently with a gain in OTUs affiliated with the *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), was associated with an increased susceptibility to infection (**Figure 6**).
+As we observed in the titration experiments, the relationship between an *Akkermansia*-affiliated OTU (OTU 4) and *C.
 difficile* indicated that wide variation in the relative abundance of *Akkermansia* was associated with varying levels of *C.
 difficile*.
-There were varying abundances of *Akkermansia* in mice regardless of the level of *C.
+There were varying abundances of the *Akkermansia*-affiliated OTU in mice regardless of the level of *C.
 difficile* colonization.
@@
 difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 \#25}.
-Given this evidence, we feel confident that an effective probiotic community could be designed based on our findings to recover colonization resistance against *C.
+Given this evidence, we feel confident that an effective probiotic mixture could be designed based on our findings to recover colonization resistance against *C.
 difficile*.
+Moreover, this line of study will be useful towards the development of personalized treatments based on an individual's specific gut microbiota, which may be a more efficient strategy for preventing and treating CDI.
+Further examination of the bacterial populations identified in this study is necessary to identify causal relationships and assess the specific mechanisms of colonization resistance.
+Such research will further advance the development of protocols to prevent and treat CDI.
 
@@
 difficile* infection based on their gut microbiota composition at admission.
+As a demonstration of this, we generated a random forest model to predict the binary outcome of whether a mouse would become colonized, regardless of *C.
+difficile* abundance.
+Using the same OTUs, we observed an error rate of `r round(binary_error, digits=1)`%.
+This suggests that such an approach would be valuable and could perhaps be improved by incorporating other clinical data {Schubert, 2014 \#11}.
 Overall these findings demonstrate the significance of the microbiota as an interconnected bacterial community in assessing resistance to pathogen colonization.
@@
 
-**Animal care.** We used 5-8 week old C57Bl/6 mice for all of our experiments.
+**Animal care.** We used 5-8 week old C57Bl/6 mice obtained from a single breeding colony maintained at the University of Michigan for all of our experiments.
 These mice were reared under SPF conditions within the animal facility at the University of Michigan.
@@
 Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via intraperitoneal injection.
-Mice that did not receive antibiotics were used as negative controls for these experiments.
+Mice that had not received antibiotics were used as negative controls for these experiments.
+*C.
+difficile* is unable to colonize mice that are not perturbed by antibiotics.
 
@@
 Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}.
+To construct each tree, two-thirds of the samples are used to train the model and the other one-third is used to test the model.
 The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 38 OTUs as predictor variables.
@@
 |:---------:|:--------------:|:-----:|:---------:|:---------:|
+| Ampicillin | *Ad libitum* in drinking water, five days | β-lactam: Penicillin | Inhibits peptidoglycan synthesis | Gram +/- |
+| Cefoperazone | *Ad libitum* in drinking water, five days | β-lactam: Cephalosporin | Inhibits peptidoglycan synthesis | Gram +/- |
 | Ciprofloxacin | Oral gavage, one time | Fluoroquinolone | Inhibits DNA gyrase | Gram +/- |
 | Clindamycin | Intraperitoneal injection, one time | Lincosamide | Inhibits protein synthesis | Anaerobes |
+| Metronidazole | *Ad libitum* in drinking water, five days | Nitromidazole | Destabilizes bacterial DNA | Anaerobes |
+| Streptomycin | *Ad libitum* in drinking water, five days | Aminoglycoside | Inhibits protein synthesis | Gram +/- |
 | Vancomycin | *Ad libitum* in drinking water, five days | Glycopeptide | Inhibits peptidoglycan synthesis | Gram + |
-| Streptomycin | *Ad libitum* in drinking water, five days | Aminoglycoside | Inhibits protein synthesis | Gram +/- |
-| Cefoperazone | *Ad libitum* in drinking water, five days | β-lactam: Cephalosporin | Inhibits peptidoglycan synthesis | Gram +/- |
-| Ampicillin | *Ad libitum* in drinking water, five days | β-lactam: Penicillin | Inhibits peptidoglycan synthesis | Gram +/- |
-| Metronidazole | *Ad libitum* in drinking water, five days | Nitromidazole | Destabilizes bacterial DNA | Anaerobes |
 
@@
 difficile* challenge.
-Stars along the x-axis indicate those OTUs that were significantly different from untreated mice for that antibiotic treatment.
+Stars along the x-axis indicate those OTUs that were significantly different from untreated mice for that antibiotic treatment after correcting for multiple comparisons.
 The error bars indicate the interquartile range.
 The median level *C.
-difficile* colonization found 24 hours post microbiota sampling and the number of animals in the treatment group is indicated in the top right for each treatment with the interquartile range in parentheses.
-The concentration next to the name of the antibiotic indicates the dose of the antibiotic that was given to the animals.
+difficile* colonization found 24 hours post microbiota sampling is plotted on the right for each treatment with error bars indicating the interquartile range.
+The dose of antibiotic and the number of animals used in each treatment group is indicated for antibiotic treatment group.
+The treatment groups are sorted according to level of *C.
+difficile* colonization.
 
@@
 difficile* challenge.
-Stars along the x-axis indicate those OTUs that varied significantly across doses of the same antibiotic.
+Stars along the x-axis indicate those OTUs that varied significantly across doses of the same antibiotic after correting for multiple comparisons.
 The error bars indicate the interquartile range.
@@
 difficile* challenge.
-Stars along the x-axis indicate those OTUs that varied significantly between those mice that were allowed 1 or 6 days of recovery.
+Stars along the x-axis indicate those OTUs that varied significantly between those mice that were allowed 1 or 6 days of recovery after correting for multiple comparisons.
 The error bars indicate the interquartile range.
@@
 difficile* observed 24 hours later.
-Only significant correlations are presented.
+Only significant correlations are presented after correting for multiple comparisons.
 OTUs are grouped by the taxonomic family and the letters in the parentheses correspond to the phylum that the taxa belong to.
-B: Bacteroidetes, F: Firmicutes, P: Proteobacteria, A: Actinobacteria, T: Tenericutes.
+B: *Bacteroidetes*, F: *Firmicutes*, P: *Proteobacteria*, A: *Actinobacteria*, T: *Tenericutes*.
 
@@
 Each pane shows antibiotic treatment groups in color and the other points as gray circles.
+The panels are sorted by the level of C.
+difficile colonization when mice were treated with the highest dose.
 
@@
 The Spearman correlation value between that OTUs abundance and *C.
-difficile* levels are shown for each pane.
+difficile* levels are shown for each pane when the corrected P-value was significant.
 The color and symbols represent the same antibiotic dose and recovery period as in Figure 5.
@@
 
-**Figure S4. The change in percent mean squared error when each OTU was removed from the random forest regression model.**
+**Figure S4.
+The change in percent mean squared error when each OTU was removed from the random forest regression model.** The top 12 OTUs with the highest percent increase in mean square error are denoted in red text and are depicted in Figure 6.

### 2015-06-10 Schubert_abxD01_mBio_2015.Rmd
Incorporate Alyx's edits

@@
 restoration of colonization resistance in patients {Seekatz, 2014 \#12}.
-The mechanisms involved in restoring colonization resistance is not fully understood, but it emphasizes the importance of the gut microbiota in protecting against *C.
-difficile*.
+The mechanisms involved in restoring colonization resistance are not fully understood, but this treatment emphasizes the importance of the overall gut microbiota in protecting against *C. difficile*.
 
@@
 difficile* colonization to high levels, and on day one post challenge we observed a median of 3.0x10^7^ (interquartile range 2.6x10^7^–3.6x10^7^, N=4) CFU/g feces.
-These results suggest that although vancomycin is not absorbed by the gut tissue, the absence of *C.
+These results suggest that although the gut tissue does not absorb vancomycin, the absence of *C.
 difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment.
-Overall, the various antibiotic perturbations provided varying levels of colonization by *C.
-difficile,* which suggested that the resulting communities varied in their composition.
+Overall, these various antibiotic perturbations provided varying levels of colonization by *C.
+difficile,* which led us to hypothesize that the resulting communities varied in their composition.
 
@@
 difficile* colonization in these titrated treatment groups.
-In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (*Figure S2*).
+In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (**Figure S2**).
 These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e.
@@
 As expected in all mice receiving titrated doses of cefoperazone, *C.
-difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; *Figure 2*).
+difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; **Figure 2**).
 Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs.
@@
 names(N_family) <- ave_corr_family_agg$Group.1
+
+number_to_text <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten")
 ```
@@
 difficile*.
-Although within the *Lachnospiraceae* familiy we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C.
-difficile* colonization, `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU was associated with high levels of *C.
+Although within the *Lachnospiraceae* family we observed `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)]` OTUs that were associated with low levels of *C.
+difficile* colonization, `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)]` OTU was associated with high levels of *C.
 difficile*.
-In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*, `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)`) of which were associated with low levels of *C.
-difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels.
+In addition we observed `r number_to_text[nrow(corr_data[families=="Lactobacillaceae",])]` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*, `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)]`) of which were associated with low levels of *C.
+difficile* and `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)]` was associated with high levels.
 The broad taxonomic representation of OTUs associated with low levels of *C.
@@
 The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C.
-difficile* colonization levels (**Figures 5**).
+difficile* colonization levels (**Figure 5**).
 When we only included the top `r n_part_features` OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C.
 difficile* colonization levels.
-The OTUs that were ranked as being the most important in defining the random forest model further validated the observations from the correlation-based analysis (**Figures S4**).
+The OTUs that were ranked as being the most important in defining the random forest model further validated the observations from the correlation-based analysis (**Figure S4**).
 According to the random forest model, colonization resistance was associated with OTUs that affiliated with the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**).
@@
 difficile* {Kassam, 2013 \#22;Seekatz, 2014 \#12;Weingarden, 2015 \#23}.
-By generating a diverse collection of communities that were challenged with *C.
+By generating a varied collection of communities that were challenged with *C.
 difficile*, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance.
@@
 Further examination of the bacterial populations identified in this study is necessary to identify causal relationships and assess the specific mechanisms of colonization resistance.
+Additional investigations into the importance of taxonomic diversity of the microbiota in colonization resistance may highlight its underlying connection to functional redundancy, which is hypothesized to contribute to overall functional stability of the community, known as the "insurance hypothesis" {Yachi, 1999 #38}.
 Such research will further advance the development of protocols to prevent and treat CDI.
@@
 difficile* abundance.
-Using the same OTUs, we observed an error rate of `r round(binary_error, digits=1)`%.
+With this model, which used the same OTUs as the regression model, we observed an error rate of `r round(binary_error, digits=1)`%.
 This suggests that such an approach would be valuable and could perhaps be improved by incorporating other clinical data {Schubert, 2014 \#11}.
@@
 Ciprofloxacin (10 mg/kg) was administered via oral gavage and clindamycin (10 mg/kg) was administered via intraperitoneal injection.
-Mice that had not received antibiotics were used as negative controls for these experiments.
-*C.
+Mice that had not received antibiotics were used as negative controls for these experiments because *C.
 difficile* is unable to colonize mice that are not perturbed by antibiotics.
+The "No antibiotics" group in Figures 1 and S1 collectively refer to the microbiota of these untreated animals prior to their challenge with C.
+difficile, as well as the microbiota of animals prior to treatment with their respective antibiotic (i.e.
+the baseline for antibiotic treated mice).
 
@@
 Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}.
-To construct each tree, two-thirds of the samples are used to train the model and the other one-third is used to test the model.
+To construct each tree, two-thirds of the samples were randomly selected were used to train the model and the other one-third of the samples were sampled with replacement to test the model.
 The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 38 OTUs as predictor variables.
@@
 difficile* challenge.
-Stars along the x-axis indicate those OTUs that varied significantly across doses of the same antibiotic after correting for multiple comparisons.
+Stars along the x-axis indicate those OTUs that varied significantly across doses of the same antibiotic after correcting for multiple comparisons.
 The error bars indicate the interquartile range.
@@
 difficile* challenge.
-Stars along the x-axis indicate those OTUs that varied significantly between those mice that were allowed 1 or 6 days of recovery after correting for multiple comparisons.
+Stars along the x-axis indicate those OTUs that varied significantly between those mice that were allowed 1 or 6 days of recovery after correcting for multiple comparisons.
 The error bars indicate the interquartile range.
@@
 difficile* observed 24 hours later.
-Only significant correlations are presented after correting for multiple comparisons.
+Only significant correlations are presented after correcting for multiple comparisons.
 OTUs are grouped by the taxonomic family and the letters in the parentheses correspond to the phylum that the taxa belong to.
@@
 Each pane shows antibiotic treatment groups in color and the other points as gray circles.
-The panels are sorted by the level of C.
-difficile colonization when mice were treated with the highest dose.
+The panes are sorted by the level of *C.
+difficile* colonization when mice were treated with the highest dose of their respective antibiotic.
 

### 2015-06-10 Schubert_abxD01_mBio_2015.Rmd
Fine tuning

@@
 findings show we can successfully apply modeling techniques to accurately
-predict the colonization resistance of a given microbiota..
+predict the colonization resistance of a given microbiota.
 
-### Results
+# Results
 
@@
 Overall, these various antibiotic perturbations provided varying levels of colonization by *C.
-difficile,* which led us to hypothesize that the resulting communities varied in their composition.
+difficile*, which led us to hypothesize that the resulting communities varied in their composition.
 
@@
 Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C.
-difficile* is restored {Reeves, 2011 \#118; Theriot, 2014 \#956}.
+difficile* is restored {Reeves, 2011 \#4; Theriot, 2014 \#17}.
 To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**).
@@
 difficile*.
-In addition we observed `r number_to_text[nrow(corr_data[families=="Lactobacillaceae",])]` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*, `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)]`) of which were associated with low levels of *C.
+In addition we observed `r number_to_text[nrow(corr_data[families=="Lactobacillaceae",])]` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*), `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)]` of which were associated with low levels of *C.
 difficile* and `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)]` was associated with high levels.
@@
 difficile* is unable to colonize mice that are not perturbed by antibiotics.
-The "No antibiotics" group in Figures 1 and S1 collectively refer to the microbiota of these untreated animals prior to their challenge with C.
-difficile, as well as the microbiota of animals prior to treatment with their respective antibiotic (i.e.
+The "No antibiotics" group in Figures 1 and S1 collectively refer to the microbiota of these untreated animals prior to their challenge with *C.
+difficile*, as well as the microbiota of animals prior to treatment with their respective antibiotic (i.e.
 the baseline for antibiotic treated mice).
@@
 Briefly, sequences were binned into operational taxonomic units (OTUs) using a 3% dissimilarity cutoff.
-Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 10) requiring an 80% bootstrap confidence score {Wang, 2007 \#1809}.
+Taxonomic assignments were determined by using a naïve Bayesian classifier with the Ribosomal Database Project (RDP) training set (version 10) requiring an 80% bootstrap confidence score {Wang, 2007 \#35}.
 In parallel to the fecal samples, we also sequenced a mock community where we knew the true sequence of the 16S rRNA gene sequences.
@@
 **Statistical analysis and modeling.** Complete scripts for regenerating our analysis and this paper are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015).
-Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#1061}.
+Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#36}.
 To avoid biases due to uneven sampling, the dissimilarity matrices were calculated by rarefying the samples to 1,625 sequences per sample.
-We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 \#804}.
+We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 \#37}.
 OTU-based analyses were performed using R (v.3.1.2).
@@
 difficile* CFU/g feces.
-All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#604}.
+All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#39}.
 Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}.

### 2015-06-11 Schubert_abxD01_mBio_2015.Rmd
Small edits

@@
 difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment.
-Overall, these various antibiotic perturbations provided varying levels of colonization by *C.
+Overall, these antibiotic perturbations provided different levels of colonization by *C.
 difficile*, which led us to hypothesize that the resulting communities varied in their composition.
@@
 Most notable among these were two OTUs that affiliated with a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3).
-The relative abundance of this *Barnesiella*-affiliated OTU increased with the delay, and the relative abundance of this *Escherichia*-affiliated OTU decreased.
+The relative abundance of this *Barnesiella*-affiliated OTU increased with the extendend recovery, and the relative abundance of this *Escherichia*-affiliated OTU decreased.
 Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C.
@@
 OTUs associated with providing resistance against *C.
-difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
+difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`).
 The *Porphyromonadaceae*-affiliated OTUs (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C.
@@
 For example, the *Lachnospiraceae* have been associated with protection against *C.
-difficile*.
+difficile* {Reeves, 2012 #18}.
 Although within the *Lachnospiraceae* family we observed `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)]` OTUs that were associated with low levels of *C.
@@
 Here, we used seven antibiotics from six classes that were given to mice in varying doses and with varying post-antibiotic recovery periods.
-The result was a combination of 15 different perturbations and the non-perturbed microbiota, which allowed us to generate distinct community profiles that displayed varying susceptibility to *C.
+The result was a combination of 15 different perturbations and the non-perturbed microbiota, which allowed us to generate distinct community profiles that displayed a range of susceptibilities to *C.
 difficile* colonization.
@@
 Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}.
-To construct each tree, two-thirds of the samples were randomly selected were used to train the model and the other one-third of the samples were sampled with replacement to test the model.
+To construct each tree, two-thirds of the samples were randomly selected to train the model and one-third of the samples were selected to test the model.
 The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 38 OTUs as predictor variables.
